---
title: "First-cohort metabolic scores by skin progression"
author: "Alex Matei"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
source("data_preprocessing/common/config.R")
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  dev = "png",
  dpi = 300
)
```

```{r packages-and-paths}
suppressPackageStartupMessages({
  library(SpatialExperiment)
  library(SummarizedExperiment)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(ggplot2)
  library(ggpubr)
  library(rstatix)
  library(svglite)
})

data_path <- paste0(
  paste0(project_path(""), "/"),
  "IMC_metabolic_analysis_AEM/spe_list.RData"
)

report_path <- knitr::current_input(dir = TRUE)
report_dir <- dirname(normalizePath(report_path, winslash = "/", mustWork = TRUE))
plot_dir <- file.path(
  report_dir,
  "Met_expr_progression_scores_numeric_pvalues_plots"
)
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)
```

# Analysis definition

This focused report reproduces the four first-cohort score plots used in the
figure. Epithelial cells are excluded as in the original analysis. Each point
is one donor. Glycolysis and TCA/OXPHOS are donor medians and use the original
unadjusted Dunn comparisons (`p.adjust.method = "none"`). HIF1A and NOX4 are
donor means from the `z-rescaled` protein assay and use the original exact,
unadjusted Wilcoxon rank-sum comparisons. Numerical raw p-values are shown
instead of significance symbols.

```{r prepare-data}
input_env <- new.env(parent = emptyenv())
load(data_path, envir = input_env)
spe <- input_env$spe_list$spe
rm(input_env)

cell_data <- as.data.frame(colData(spe), check.names = FALSE)
keep <- as.character(cell_data$new_celllabels) != "E-cad+"

group_levels <- c("Healthy", "Progressive SSc", "Stable SSc")
group_palette <- c(
  "Healthy" = "#B9DCE8",
  "Progressive SSc" = "#C03830",
  "Stable SSc" = "#F6BF93"
)

cell_scores <- tibble(
  group = recode(
    as.character(cell_data$Skin.progression.at.biopsy[keep]),
    healthy = "Healthy",
    progressive = "Progressive SSc",
    stable = "Stable SSc"
  ),
  donor = as.character(cell_data$patient_id[keep]),
  Glycolysis = as.numeric(cell_data$Glycolysis_z_c[keep]),
  `TCA/OXPHOS` = as.numeric(cell_data$TCA_z_c[keep]),
  `Hypoxia signaling` = as.numeric(
    assay(spe, "z-rescaled")["HIF1a", keep]
  ),
  `ROS signaling` = as.numeric(
    assay(spe, "z-rescaled")["NOX4", keep]
  )
) %>%
  filter(!is.na(group), !is.na(donor)) %>%
  mutate(group = factor(group, levels = group_levels))

donor_scores <- cell_scores %>%
  group_by(group, donor) %>%
  summarise(
    Glycolysis = median(Glycolysis, na.rm = TRUE),
    `TCA/OXPHOS` = median(`TCA/OXPHOS`, na.rm = TRUE),
    `Hypoxia signaling` = mean(`Hypoxia signaling`, na.rm = TRUE),
    `ROS signaling` = mean(`ROS signaling`, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(Glycolysis, `TCA/OXPHOS`, `Hypoxia signaling`, `ROS signaling`),
    names_to = "score",
    values_to = "value"
  ) %>%
  mutate(
    score = factor(
      score,
      levels = c(
        "Glycolysis",
        "TCA/OXPHOS",
        "Hypoxia signaling",
        "ROS signaling"
      )
    )
  )

format_numeric_p <- function(p) {
  case_when(
    p < 0.001 ~ sprintf("p = %.2e", p),
    p < 0.01 ~ sprintf("p = %.5f", p),
    p < 0.1 ~ sprintf("p = %.4f", p),
    TRUE ~ sprintf("p = %.3f", p)
  )
}

overall_tests <- donor_scores %>%
  group_by(score) %>%
  kruskal_test(value ~ group) %>%
  ungroup() %>%
  mutate(p_label = format_numeric_p(p))

dunn_tests <- donor_scores %>%
  filter(score %in% c("Glycolysis", "TCA/OXPHOS")) %>%
  group_by(score) %>%
  dunn_test(value ~ group, p.adjust.method = "none") %>%
  ungroup() %>%
  mutate(method = "Dunn; unadjusted")

pairwise_group_levels <- combn(group_levels, 2, simplify = FALSE)

wilcox_tests <- bind_rows(lapply(
  c("Hypoxia signaling", "ROS signaling"),
  function(score_name) {
    bind_rows(lapply(pairwise_group_levels, function(comparison) {
      comparison_data <- donor_scores %>%
        filter(
          as.character(score) == score_name,
          as.character(group) %in% comparison
        )

      x <- comparison_data$value[
        as.character(comparison_data$group) == comparison[[1]]
      ]
      y <- comparison_data$value[
        as.character(comparison_data$group) == comparison[[2]]
      ]
      test_result <- stats::wilcox.test(x, y, exact = TRUE)

      tibble(
        score = score_name,
        group1 = comparison[[1]],
        group2 = comparison[[2]],
        n1 = length(x),
        n2 = length(y),
        statistic = unname(test_result$statistic),
        p = test_result$p.value,
        method = "Exact Wilcoxon rank-sum; unadjusted"
      )
    }))
  }
))

pairwise_tests <- bind_rows(dunn_tests, wilcox_tests) %>%
  mutate(
    score = factor(
      score,
      levels = c(
        "Glycolysis",
        "TCA/OXPHOS",
        "Hypoxia signaling",
        "ROS signaling"
      )
    )
  ) %>%
  group_by(score) %>%
  mutate(
    score_max = max(donor_scores$value[donor_scores$score == first(score)]),
    score_min = min(donor_scores$value[donor_scores$score == first(score)]),
    score_span = pmax(score_max - score_min, 0.25),
    comparison_level = if_else(
      group1 == "Healthy" & group2 == "Stable SSc",
      2,
      1
    ),
    y.position = score_max + score_span * (
      0.13 + 0.19 * (comparison_level - 1)
    ),
    p_label = format_numeric_p(p)
  ) %>%
  ungroup()

write.csv(
  overall_tests,
  file.path(plot_dir, "overall_kruskal_wallis_pvalues.csv"),
  row.names = FALSE
)
write.csv(
  dunn_tests,
  file.path(plot_dir, "pairwise_dunn_unadjusted_pvalues.csv"),
  row.names = FALSE
)
write.csv(
  pairwise_tests,
  file.path(plot_dir, "pairwise_unadjusted_pvalues.csv"),
  row.names = FALSE
)
```

# Numerical p-values

```{r statistical-tables}
overall_tests %>%
  select(score, n, statistic, df, p, p_label)

pairwise_tests %>%
  select(score, group1, group2, n1, n2, method, statistic, p, p_label)
```

# Figure panels

```{r build-plots}
score_titles <- c(
  "Glycolysis" = "Glycolysis score",
  "TCA/OXPHOS" = "TCA/OXPHOS score",
  "Hypoxia signaling" = "Hypoxia signaling",
  "ROS signaling" = "ROS signaling"
)

make_score_plot <- function(score_name) {
  plot_data <- donor_scores %>% filter(score == score_name)
  plot_stats <- pairwise_tests %>% filter(score == score_name)

  ggplot(plot_data, aes(x = group, y = value, fill = group)) +
    geom_boxplot(
      width = 0.58,
      outlier.shape = NA,
      color = "#4D4D4D",
      linewidth = 0.55
    ) +
    geom_point(
      position = position_jitter(width = 0.07, height = 0, seed = 2026),
      size = 2.3,
      shape = 16,
      color = "black"
    ) +
    stat_pvalue_manual(
      plot_stats,
      label = "p_label",
      xmin = "group1",
      xmax = "group2",
      y.position = "y.position",
      hide.ns = FALSE,
      tip.length = 0.012,
      bracket.size = 0.45,
      size = 3.5
    ) +
    scale_fill_manual(values = group_palette, drop = FALSE) +
    scale_y_continuous(expand = expansion(mult = c(0.08, 0.08))) +
    coord_cartesian(clip = "off") +
    labs(
      title = unname(score_titles[score_name]),
      x = NULL,
      y = "score (z-rescaled)"
    ) +
    theme_classic(base_size = 12) +
    theme(
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),
      plot.title = element_text(
        face = "bold",
        size = 13,
        hjust = 0.5,
        margin = margin(b = 8)
      ),
      axis.title.y = element_text(face = "bold", size = 11),
      axis.text.x = element_text(
        angle = 38,
        hjust = 1,
        vjust = 1,
        size = 10,
        color = "#333333"
      ),
      axis.text.y = element_text(size = 10, color = "#333333"),
      axis.line = element_line(linewidth = 0.5, color = "#333333"),
      axis.ticks = element_line(linewidth = 0.45, color = "#333333"),
      legend.position = "none",
      plot.margin = margin(12, 18, 12, 12)
    )
}

plot_names <- levels(donor_scores$score)
score_plots <- setNames(lapply(plot_names, make_score_plot), plot_names)

combined_score_plots <- lapply(
  score_plots,
  function(plot) {
    plot +
      scale_x_discrete(
        labels = c(
          "Healthy" = "Healthy",
          "Progressive SSc" = "Progressive\nSSc",
          "Stable SSc" = "Stable\nSSc"
        )
      ) +
      theme(
        axis.text.x = element_text(
          angle = 25,
          hjust = 1,
          vjust = 1,
          size = 9.5,
          color = "#333333"
        ),
        plot.margin = margin(12, 28, 14, 28)
      )
  }
)

combined_plot <- ggarrange(
  plotlist = combined_score_plots,
  ncol = 2,
  nrow = 2,
  align = "hv"
)

save_plot <- function(plot, stem, width, height) {
  ggsave(
    file.path(plot_dir, paste0(stem, ".svg")),
    plot,
    width = width,
    height = height,
    units = "in",
    device = svglite::svglite,
    bg = "white"
  )
  ggsave(
    file.path(plot_dir, paste0(stem, ".pdf")),
    plot,
    width = width,
    height = height,
    units = "in",
    device = cairo_pdf,
    bg = "white"
  )
  ggsave(
    file.path(plot_dir, paste0(stem, ".png")),
    plot,
    width = width,
    height = height,
    units = "in",
    dpi = 450,
    bg = "white"
  )
}

invisible(mapply(
  FUN = function(plot, stem) save_plot(plot, stem, 4.7, 4.4),
  plot = score_plots,
  stem = c(
    "glycolysis_score_numeric_pvalues",
    "tca_oxphos_score_numeric_pvalues",
    "hypoxia_signaling_numeric_pvalues",
    "ros_signaling_numeric_pvalues"
  ),
  SIMPLIFY = FALSE
))

save_plot(
  combined_plot,
  "first_cohort_four_scores_numeric_pvalues",
  width = 10.8,
  height = 8.8
)
```

```{r combined-plot, echo=FALSE, fig.width=10.8, fig.height=8.8}
combined_plot
```

## Individual panels {.tabset .tabset-pill}

### Glycolysis

```{r glycolysis-panel, echo=FALSE, fig.width=4.7, fig.height=4.4}
score_plots[["Glycolysis"]]
```

### TCA/OXPHOS

```{r tca-panel, echo=FALSE, fig.width=4.7, fig.height=4.4}
score_plots[["TCA/OXPHOS"]]
```

### Hypoxia signaling

```{r hypoxia-panel, echo=FALSE, fig.width=4.7, fig.height=4.4}
score_plots[["Hypoxia signaling"]]
```

### ROS signaling

```{r ros-panel, echo=FALSE, fig.width=4.7, fig.height=4.4}
score_plots[["ROS signaling"]]
```

# Session information

```{r session-info}
sessionInfo()
```
