---
title: "Xenium cell-filtering reproducibility"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

<style>
.main-container {
  max-width: 1380px !important;
}
table {
  font-size: 13px;
}
th, td {
  vertical-align: top !important;
  overflow-wrap: anywhere;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
library(dplyr)
library(knitr)

thresholds <- read.csv(
  "outputs/xenium_qc_thresholds_and_retention.csv",
  check.names = FALSE
)
validation <- read.csv(
  "outputs/xenium_qc_validation_checks.csv",
  check.names = FALSE
)
formula_sensitivity <- read.csv(
  "outputs/excluded_validation_negative_control_formula_sensitivity.csv",
  check.names = FALSE
)
excluded_validation_area <- read.csv(
  "outputs/excluded_validation_area_threshold_audit.csv",
  check.names = FALSE
)
blinding <- read.csv(
  "outputs/xenium_qc_blinding_audit.csv",
  check.names = FALSE
)
sources <- read.csv(
  "outputs/source_inventory_md5.csv",
  check.names = FALSE
)

six <- thresholds |> filter(pipeline == "Original six-donor QC")
excluded_validation <- thresholds |> filter(sample == "ExcludedValidation")
six_before <- sum(six$cells_before_qc)
six_retained <- sum(six$cells_retained)
seven_before <- sum(thresholds$cells_before_qc)
seven_retained <- sum(thresholds$cells_retained)
```

# Technical summary

The executed 2025-07-23 QC report contains the exact original six-donor
filter. Cells were retained when they had **more than 20 gene-expression
transcripts**, **more than 10 detected genes**, a **negative-control
proportion below 0.001**, and a cell area within an **inclusive,
sample-specific interval**. This retained
`r format(six_retained, big.mark = ",")` of
`r format(six_before, big.mark = ",")` cells
(`r sprintf("%.2f", 100 * six_retained / six_before)`%).

For ExcludedValidation, an inclusive **5-400 um^2^** area predicate can be stated
explicitly without excluding any supplied cell: its observed area range
is `r sprintf("%.3f-%.3f", excluded_validation_area$observed_area_min_um2, excluded_validation_area$observed_area_max_um2)`
um^2^, and all
`r format(excluded_validation_area$cells_evaluated, big.mark = ",")` cells pass these
bounds. The current rule retained
`r format(excluded_validation$cells_retained, big.mark = ",")` of
`r format(excluded_validation$cells_before_qc, big.mark = ",")` cells after the other
QC criteria. ExcludedValidation still does **not** use an identical negative-control
formula: its numerator and denominator differ from those used for the
original six donors. The seven-donor Methods statement below therefore
documents that remaining exception.

# Exact thresholds and retained cells

```{r threshold-table}
thresholds |>
  transmute(
    Donor = sample,
    Pipeline = pipeline,
    `Transcripts/cell` = transcript_threshold,
    `Detected genes` = detected_gene_threshold,
    `Negative-control proportion` = negative_control_threshold,
    `Cell area` = area_filter,
    `Before QC` = cells_before_qc,
    Retained = cells_retained,
    Excluded = cells_excluded,
    `Retained (%)` = sprintf("%.2f", percent_retained)
  ) |>
  kable(
    caption = paste0(
      "Exact QC predicates and cell counts. ",
      "Area bounds are inclusive."
    ),
    align = c("l", "l", "c", "c", "c", "l", "r", "r", "r", "r")
  )
```

# ExcludedValidation area-threshold audit

The **5-400 um^2^** bounds are round outward limits around the observed
ExcludedValidation range. They make the area criterion explicit and reproducible
while deliberately leaving cell membership unchanged; they should be
described as non-excluding data-bracketing bounds, not as an independently
optimized exclusion rule.

```{r excluded_validation-area-audit}
excluded_validation_area |>
  transmute(
    Donor = sample,
    `Observed minimum (um^2)` = sprintf("%.3f", observed_area_min_um2),
    `Observed maximum (um^2)` = sprintf("%.3f", observed_area_max_um2),
    `Applied inclusive bounds (um^2)` = paste0(
      applied_area_min_um2_inclusive,
      "-",
      applied_area_max_um2_inclusive
    ),
    `Cells evaluated` = cells_evaluated,
    `Cells passing area` = cells_passing_area,
    `Cells excluded by area` = cells_excluded_by_area
  ) |>
  kable(
    caption = paste0(
      "ExcludedValidation area audit. The explicit inclusive criterion excludes ",
      "no supplied cells."
    ),
    align = c("l", "r", "r", "c", "r", "r", "r")
  )
```

# Metric definitions

For the original six donors, the negative-control proportion was:

\[
  \frac{
    C_{\mathrm{control\ probe}} +
    C_{\mathrm{genomic\ control}} +
    C_{\mathrm{control\ codeword}} +
    C_{\mathrm{unassigned\ codeword}}
  }{
    T_{\mathrm{gene\ expression}} +
    C_{\mathrm{control\ probe}} +
    C_{\mathrm{genomic\ control}} +
    C_{\mathrm{control\ codeword}} +
    C_{\mathrm{unassigned\ codeword}}
  }.
\]

This is the executed report's expression
`1 - transcript_counts / (transcript_counts + the four control-count
terms)`. Deprecated-codeword counts were not included.

For ExcludedValidation, the stored proportion was instead:

\[
  \frac{
    C_{\mathrm{control\ probe}} +
    C_{\mathrm{control\ codeword}} +
    C_{\mathrm{unassigned\ codeword}}
  }{
    C_{\mathrm{total}}
  }.
\]

# Manuscript-ready Methods

## Verified original six-donor wording

Cells were retained when they contained more than 20 gene-expression
transcripts, more than 10 detected genes, a negative-control count
proportion below 0.001, and a cell area within the following
sample-specific inclusive bounds: Validation5, 15-270 um^2^; Validation6, 15-250
um^2^; Validation1, 19-220 um^2^; Validation2, 13-300 um^2^; Validation3, 13-280
um^2^; and Validation4, 14-300 um^2^. The negative-control proportion was
calculated as the sum of control-probe, genomic-control,
control-codeword, and unassigned-codeword counts divided by the sum of
these counts and gene-expression transcript counts. These criteria
retained `r format(six_retained, big.mark = ",")` of
`r format(six_before, big.mark = ",")` segmented cells: Validation5,
`r format(six$cells_retained[six$sample == "Validation5"], big.mark = ",")`;
Validation6,
`r format(six$cells_retained[six$sample == "Validation6"], big.mark = ",")`;
Validation1,
`r format(six$cells_retained[six$sample == "Validation1"], big.mark = ",")`;
Validation2,
`r format(six$cells_retained[six$sample == "Validation2"], big.mark = ",")`;
Validation3,
`r format(six$cells_retained[six$sample == "Validation3"], big.mark = ",")`;
and Validation4,
`r format(six$cells_retained[six$sample == "Validation4"], big.mark = ",")`.
Thresholds were
selected from sample-wise QC distributions before ResolVI integration,
Leiden clustering, and cell-phenotype assignment. Sample identity was
visible during QC; however, downstream cell-phenotype labels and
progressive/stable status were not present in the QC object and were
not used for threshold selection. ResolVI used sample ID as the batch
variable. Expression was normalized with Seurat `NormalizeData()`
using a scale factor of 1,000, and cell phenotypes were assigned by
Leiden clustering on the ResolVI embeddings.

## Exact wording for the currently joined seven-donor object

Cells from the original six donors were filtered using the criteria and
sample-specific area bounds stated above. ExcludedValidation was processed
subsequently using more than 20 gene-expression transcripts, more than
10 detected genes, a negative-control proportion below 0.001, and an
inclusive cell-area range of 5-400 um^2^. The observed ExcludedValidation cell-area
range was
`r sprintf("%.3f-%.3f", excluded_validation_area$observed_area_min_um2, excluded_validation_area$observed_area_max_um2)`
um^2^, so the
area criterion retained all
`r format(excluded_validation_area$cells_evaluated, big.mark = ",")` supplied ExcludedValidation
cells and did not alter the final retained set. For ExcludedValidation, the
negative-control proportion was calculated as the sum of
negative-control-probe, negative-control-codeword, and
unassigned-codeword counts divided by total counts. The combined
filters retained `r format(seven_retained, big.mark = ",")` of
`r format(seven_before, big.mark = ",")` supplied cells across seven
donors, including `r format(excluded_validation$cells_retained, big.mark = ",")` of
`r format(excluded_validation$cells_before_qc, big.mark = ",")` ExcludedValidation cells.

This wording is reproducible. The area predicate is now explicit for all
seven donors, but the ExcludedValidation bounds are non-excluding and its
negative-control formula remains different. It should not be replaced by
a claim that all seven donors used an identical QC pipeline.

# Blinding audit

```{r blinding-table}
blinding |>
  kable(
    caption = "What can be established from the archived QC object and report.",
    align = c("l", "l", "l")
  )
```

The defensible statement is therefore: **sample identity was visible,
but thresholding preceded cell phenotyping and did not use downstream
cell-phenotype labels or progressive/stable status**. A claim of formal
blinding to clinical phenotype is not supported by the archived report.

# Validation and pipeline caveat

```{r validation-table}
validation |>
  mutate(Result = ifelse(passed, "PASS", "FAIL")) |>
  select(Check = check, Result) |>
  kable(
    caption = "Object-level checks against the recovered predicates.",
    align = c("l", "c")
  )
```

The current ExcludedValidation rule retains
`r formula_sensitivity$retained_current_excluded_validation_formula` cells. Applying
the original six-donor negative-control formula to ExcludedValidation together with
the inclusive 5-400 um^2^ area predicate would retain
`r formula_sensitivity$retained_with_original_six_donor_negative_formula_and_5_400_area_filter`
cells. The difference is
`r formula_sensitivity$retained_only_by_current_excluded_validation_formula` cells.
This is a material pipeline difference to resolve before describing all
seven donors as identically filtered.

# Source provenance

```{r source-table}
sources |>
  transmute(
    Source = source,
    Path = path,
    `Modified` = modified,
    MD5 = md5
  ) |>
  kable(
    caption = "Archived sources used for this audit.",
    align = c("l", "l", "l", "l")
  )
```
