---
title: "IMC Protein and Xenium scMetabolism Scores"
author: "Codex"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
source("data_preprocessing/common/config.R")
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.retina = 2,
  dev.args = list(bg = "white")
)
```

```{r libraries}
suppressPackageStartupMessages({
  library(SummarizedExperiment)
  library(SingleCellExperiment)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(knitr)
})
```

```{r data_and_helpers}
project_dir <- project_path("Revision/Nan annotation/Data/RMD files")
imc_file <- project_path("Revision/Nan annotation/Data/IMC spe objects/New IMC/spe2025-12-17 umap.RData")
output_dir <- file.path(project_dir, "metabolism_score_comparison_outputs")
scmetab_output_dir <- file.path(project_dir, "scmetabolism_outputs")
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

if (!exists("spe")) {
  load(imc_file)
}

score_assay <- "zrescaled"
if (!score_assay %in% SummarizedExperiment::assayNames(spe)) {
  stop("The IMC spe object does not contain assay 'zrescaled'.")
}

glycolysis <- c("GLUT1", "HK1", "PFKL1_PFKM", "PKM2", "LDHA", "GAPDH")
tca_oxphos <- c("CS", "OGDH", "ATP5A", "SDHA")

pathway_markers <- list(
  "Glycolysis" = glycolysis,
  "TCA / OXPHOS" = tca_oxphos
)
pathway_levels <- names(pathway_markers)
pathway_labels <- c(
  "Glycolysis" = "Glycolysis",
  "TCA / OXPHOS" = "TCA /\nOXPHOS"
)

group_colors <- c(
  "Other_Fib" = "#4E79A7",
  "Met_hi_Fib" = "#E15759",
  "Met_int_EC" = "#4E79A7",
  "Met_hi_EC" = "#E15759"
)
group_colors_dark <- c(
  "Other_Fib" = "#2F5F8F",
  "Met_hi_Fib" = "#B23B3F",
  "Met_int_EC" = "#2F5F8F",
  "Met_hi_EC" = "#B23B3F"
)

marker_status <- dplyr::bind_rows(lapply(names(pathway_markers), function(pathway_name) {
  data.frame(
    pathway = pathway_name,
    marker = pathway_markers[[pathway_name]],
    available = pathway_markers[[pathway_name]] %in% rownames(spe),
    stringsAsFactors = FALSE
  )
}))

missing_pathways <- marker_status %>%
  dplyr::group_by(pathway) %>%
  dplyr::summarise(n_available = sum(available), .groups = "drop") %>%
  dplyr::filter(n_available == 0)

if (nrow(missing_pathways) > 0) {
  stop("No IMC markers were found for: ", paste(missing_pathways$pathway, collapse = ", "))
}

if (!"Rphenograph_500" %in% colnames(SummarizedExperiment::colData(spe))) {
  stop("The IMC spe object must contain colData(spe)$Rphenograph_500 for fibroblast labels.")
}

if (!"Rphenograph_200EC" %in% colnames(SummarizedExperiment::colData(spe))) {
  stop("The IMC spe object must contain colData(spe)$Rphenograph_200EC for EC labels.")
}

spe$metfiblabel <- dplyr::case_when(
  as.character(spe$Rphenograph_500) %in% c("1", "2") ~ "Met_hi_Fib",
  as.character(spe$Rphenograph_500) == "3" ~ "Other_Fib",
  TRUE ~ NA_character_
)

spe$metEClabel <- dplyr::case_when(
  as.character(spe$Rphenograph_200EC) %in% c("1", "2") ~ "Met_hi_EC",
  as.character(spe$Rphenograph_200EC) == "3" ~ "Met_int_EC",
  TRUE ~ NA_character_
)

make_imc_protein_score <- function(markers) {
  markers <- intersect(markers, rownames(spe))
  colMeans(as.matrix(SummarizedExperiment::assay(spe, score_assay)[markers, , drop = FALSE]), na.rm = TRUE)
}

score_wide_all <- as.data.frame(
  vapply(pathway_markers, make_imc_protein_score, numeric(ncol(spe))),
  check.names = FALSE
)
score_wide_all$cell <- colnames(spe)
score_wide_all$sample_id <- if ("sample_id" %in% colnames(SummarizedExperiment::colData(spe))) {
  as.character(spe$sample_id)
} else {
  NA_character_
}
score_wide_all$metfiblabel <- as.character(spe$metfiblabel)
score_wide_all$metEClabel <- as.character(spe$metEClabel)

analysis_configs <- list(
  fibroblasts = list(
    label = "Fibroblast",
    section_label = "Fibroblasts",
    group_col = "metfiblabel",
    group_levels = c("Other_Fib", "Met_hi_Fib"),
    file_prefix = "combined_fibroblast",
    scmetab_cell_file = "scmetabolism_metfiblabel_energy_cell_scores.csv",
    scmetab_joined_file = "scmetabolism_imc_xenium_cell_joined.csv"
  ),
  endothelial = list(
    label = "EC",
    section_label = "Endothelial Cells",
    group_col = "metEClabel",
    group_levels = c("Met_int_EC", "Met_hi_EC"),
    file_prefix = "combined_ec",
    scmetab_cell_file = "scmetabolism_ec_metEClabel_energy_cell_scores.csv",
    scmetab_joined_file = "scmetabolism_ec_imc_xenium_cell_joined.csv"
  )
)

rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  if (!all(is.finite(rng)) || rng[1] == rng[2]) {
    return(rep(0.5, length(x)))
  }
  (x - rng[1]) / (rng[2] - rng[1])
}

format_p <- function(p_value) {
  dplyr::case_when(
    is.na(p_value) ~ "p = NA",
    p_value < 0.001 ~ "p < 0.001",
    TRUE ~ paste0("p = ", formatC(p_value, format = "f", digits = 3))
  )
}

save_plot_set <- function(plot, basename, width, height) {
  png_file <- file.path(output_dir, paste0(basename, ".png"))
  pdf_file <- file.path(output_dir, paste0(basename, ".pdf"))
  svg_file <- file.path(output_dir, paste0(basename, ".svg"))

  ggsave(png_file, plot, width = width, height = height, dpi = 300, bg = "white")

  grDevices::pdf(pdf_file, width = width, height = height, useDingbats = FALSE, bg = "white")
  print(plot)
  grDevices::dev.off()

  grDevices::svg(svg_file, width = width, height = height, bg = "white")
  print(plot)
  grDevices::dev.off()

  invisible(c(png_file, pdf_file, svg_file))
}

summarise_score_long <- function(score_long) {
  score_long %>%
    dplyr::group_by(pathway, group) %>%
    dplyr::summarise(
      n_cells = dplyr::n(),
      n_samples = dplyr::n_distinct(sample_id[!is.na(sample_id) & sample_id != ""]),
      mean_score = mean(score, na.rm = TRUE),
      median_score = median(score, na.rm = TRUE),
      q25 = stats::quantile(score, 0.25, na.rm = TRUE),
      q75 = stats::quantile(score, 0.75, na.rm = TRUE),
      sd_score = stats::sd(score, na.rm = TRUE),
      se_score = sd_score / sqrt(n_cells),
      ci95 = 1.96 * se_score,
      .groups = "drop"
    )
}

summarise_sample_scores <- function(score_long) {
  score_long %>%
    dplyr::filter(!is.na(sample_id), sample_id != "") %>%
    dplyr::group_by(sample_id, pathway, group) %>%
    dplyr::summarise(
      n_cells = dplyr::n(),
      mean_score = mean(score, na.rm = TRUE),
      median_score = median(score, na.rm = TRUE),
      .groups = "drop"
    )
}

make_imc_score_long <- function(config) {
  score_wide_all %>%
    dplyr::mutate(group = as.character(.data[[config$group_col]])) %>%
    dplyr::filter(!is.na(group), group %in% config$group_levels) %>%
    tidyr::pivot_longer(
      cols = dplyr::all_of(pathway_levels),
      names_to = "pathway",
      values_to = "score"
    ) %>%
    dplyr::mutate(
      group = factor(group, levels = config$group_levels),
      pathway = factor(pathway, levels = pathway_levels),
      source = "IMC protein"
    ) %>%
    dplyr::select(cell, sample_id, group, pathway, score, source)
}

load_scmetab_score_long <- function(config) {
  score_file <- file.path(scmetab_output_dir, config$scmetab_cell_file)
  if (!file.exists(score_file)) {
    warning("Missing scMetabolism score file: ", score_file)
    return(data.frame())
  }

  cell_scores <- read.csv(score_file, check.names = FALSE, stringsAsFactors = FALSE)
  required_cols <- c("cell", config$group_col, "pathway", "score")
  missing_cols <- setdiff(required_cols, colnames(cell_scores))
  if (length(missing_cols) > 0) {
    stop("Missing columns in ", score_file, ": ", paste(missing_cols, collapse = ", "))
  }

  score_long <- cell_scores %>%
    dplyr::transmute(
      cell = as.character(cell),
      group = as.character(.data[[config$group_col]]),
      pathway = dplyr::case_when(
        pathway == "Glycolysis / Gluconeogenesis" ~ "Glycolysis",
        pathway == "Citrate cycle (TCA cycle)" ~ "TCA / OXPHOS",
        TRUE ~ as.character(pathway)
      ),
      score = as.numeric(score),
      source = "Xenium scMetabolism"
    ) %>%
    dplyr::filter(
      !is.na(group),
      group %in% config$group_levels,
      pathway %in% pathway_levels
    )

  joined_file <- file.path(scmetab_output_dir, config$scmetab_joined_file)
  if (file.exists(joined_file)) {
    joined <- read.csv(joined_file, check.names = FALSE, stringsAsFactors = FALSE)
    if ("cell" %in% colnames(joined)) {
      if ("sample_id" %in% colnames(joined)) {
        joined$sample_id_lookup <- as.character(joined$sample_id)
      } else if ("sample_id_xenium" %in% colnames(joined)) {
        joined$sample_id_lookup <- as.character(joined$sample_id_xenium)
      } else {
        joined$sample_id_lookup <- NA_character_
      }

      sample_lookup <- joined %>%
        dplyr::transmute(
          cell = as.character(cell),
          sample_id = sample_id_lookup
        ) %>%
        dplyr::distinct(cell, .keep_all = TRUE)

      score_long <- score_long %>%
        dplyr::left_join(sample_lookup, by = "cell")
    } else {
      score_long$sample_id <- NA_character_
    }
  } else {
    score_long$sample_id <- NA_character_
  }

  score_long %>%
    dplyr::mutate(
      group = factor(group, levels = config$group_levels),
      pathway = factor(pathway, levels = pathway_levels)
    ) %>%
    dplyr::select(cell, sample_id, group, pathway, score, source)
}

make_energy_dotplot <- function(summary_df, config, title, legend_title, scale_by_column = FALSE) {
  plot_df <- summary_df %>%
    dplyr::mutate(
      pathway = factor(as.character(pathway), levels = rev(pathway_levels)),
      group = factor(as.character(group), levels = config$group_levels)
    )

  if (isTRUE(scale_by_column)) {
    plot_df <- plot_df %>%
      dplyr::group_by(pathway) %>%
      dplyr::mutate(
        fill_score = rescale01(mean_score),
        size_score = fill_score
      ) %>%
      dplyr::ungroup()
    legend_title <- "Column-scaled mean"
  } else {
    plot_df <- plot_df %>%
      dplyr::mutate(
        fill_score = mean_score,
        size_score = rescale01(mean_score)
      )
  }

  ggplot(plot_df, aes(x = group, y = pathway)) +
    geom_point(
      aes(fill = fill_score, size = size_score),
      shape = 21,
      color = "grey35",
      stroke = 0.35
    ) +
    scale_y_discrete(labels = pathway_labels) +
    scale_fill_gradient(low = "white", high = "#D7301F", name = legend_title) +
    scale_size(range = c(6.5, 12.5), guide = "none") +
    labs(title = title, x = NULL, y = NULL) +
    theme_bw(base_size = 14) +
    theme(
      axis.text.x = element_text(size = 13, face = "bold"),
      axis.text.y = element_text(size = 13, lineheight = 0.95),
      legend.position = "right",
      legend.title = element_text(size = 12.5),
      legend.text = element_text(size = 11),
      panel.grid.major = element_line(color = "grey88"),
      panel.grid.minor = element_blank(),
      plot.title = element_text(size = 14.5, hjust = 0.5, margin = margin(b = 6)),
      plot.margin = margin(5.5, 6, 5.5, 5.5)
    )
}

make_violin_plot <- function(score_long, title, y_label) {
  ggplot(score_long, aes(x = group, y = score, fill = group)) +
    geom_violin(width = 0.82, alpha = 0.8, color = "grey30", trim = TRUE) +
    geom_boxplot(width = 0.16, outlier.shape = NA, fill = "white", color = "grey15") +
    facet_wrap(~ pathway, nrow = 1, scales = "free_y") +
    scale_fill_manual(values = group_colors) +
    labs(title = title, x = NULL, y = y_label, fill = NULL) +
    theme_bw(base_size = 13) +
    theme(
      axis.text.x = element_text(size = 12, face = "bold", angle = 20, hjust = 1),
      strip.text = element_text(size = 12, face = "bold"),
      legend.position = "none",
      plot.title = element_text(hjust = 0.5)
    )
}

make_density_plot <- function(score_long, title, x_label) {
  ggplot(score_long, aes(x = score, fill = group, color = group)) +
    geom_density(alpha = 0.32, linewidth = 0.8, adjust = 1.1) +
    facet_wrap(~ pathway, nrow = 1, scales = "free") +
    scale_fill_manual(values = group_colors) +
    scale_color_manual(values = group_colors_dark) +
    labs(title = title, x = x_label, y = "Density", fill = NULL, color = NULL) +
    theme_bw(base_size = 13) +
    theme(
      strip.text = element_text(size = 12, face = "bold"),
      legend.position = "top",
      plot.title = element_text(hjust = 0.5)
    )
}

make_sample_plot <- function(sample_summary, title, y_label) {
  if (nrow(sample_summary) == 0) {
    return(
      ggplot() +
        annotate("text", x = 0, y = 0, label = "No sample IDs available", size = 4) +
        labs(title = title) +
        theme_void()
    )
  }

  ggplot(sample_summary, aes(x = group, y = mean_score, color = group)) +
    geom_point(
      position = position_jitter(width = 0.08, height = 0),
      size = 2.8,
      alpha = 0.85
    ) +
    stat_summary(
      fun = mean,
      geom = "crossbar",
      width = 0.42,
      color = "grey20",
      linewidth = 0.35
    ) +
    facet_wrap(~ pathway, nrow = 1, scales = "free_y") +
    scale_color_manual(values = group_colors) +
    labs(
      title = title,
      subtitle = "Each dot is one sample/group mean",
      x = NULL,
      y = y_label,
      color = NULL
    ) +
    theme_bw(base_size = 13) +
    theme(
      axis.text.x = element_text(size = 12, face = "bold", angle = 20, hjust = 1),
      strip.text = element_text(size = 12, face = "bold"),
      legend.position = "none",
      plot.title = element_text(hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5)
    )
}

cor_summary <- function(df) {
  df <- df %>%
    dplyr::filter(
      is.finite(imc_protein_score),
      is.finite(scmetabolism_score)
    )

  if (
    nrow(df) < 3 ||
    length(unique(df$imc_protein_score)) < 2 ||
    length(unique(df$scmetabolism_score)) < 2
  ) {
    return(data.frame(
      n = nrow(df),
      spearman_rho = NA_real_,
      spearman_p = NA_real_,
      pearson_r = NA_real_,
      pearson_p = NA_real_
    ))
  }

  spearman_test <- suppressWarnings(stats::cor.test(
    df$imc_protein_score,
    df$scmetabolism_score,
    method = "spearman",
    exact = FALSE
  ))
  pearson_test <- suppressWarnings(stats::cor.test(
    df$imc_protein_score,
    df$scmetabolism_score,
    method = "pearson"
  ))

  data.frame(
    n = nrow(df),
    spearman_rho = unname(spearman_test$estimate),
    spearman_p = spearman_test$p.value,
    pearson_r = unname(pearson_test$estimate),
    pearson_p = pearson_test$p.value
  )
}

make_score_bundle <- function(score_long, config, source_id, source_label, y_label, raw_legend_title) {
  score_summary <- summarise_score_long(score_long)
  sample_summary <- summarise_sample_scores(score_long)

  p_dot_raw <- make_energy_dotplot(
    score_summary,
    config,
    paste0(config$label, " ", source_label, " scores"),
    raw_legend_title,
    scale_by_column = FALSE
  )
  p_dot_scaled <- make_energy_dotplot(
    score_summary,
    config,
    paste0(config$label, " ", source_label, " scores - scaled"),
    "Column-scaled mean",
    scale_by_column = TRUE
  )
  p_violin <- make_violin_plot(
    score_long,
    paste0(config$label, " ", source_label, " score distributions"),
    y_label
  )
  p_density <- make_density_plot(
    score_long,
    paste0(config$label, " ", source_label, " score densities"),
    y_label
  )
  p_sample <- make_sample_plot(
    sample_summary,
    paste0(config$label, " ", source_label, " sample means"),
    paste0("Sample mean ", y_label)
  )

  basename <- paste0(config$file_prefix, "_", source_id)
  save_plot_set(p_dot_raw, paste0(basename, "_dotplot_raw_reversed_axes"), width = 6.8, height = 3.4)
  save_plot_set(p_dot_scaled, paste0(basename, "_dotplot_scaled_by_column_reversed_axes"), width = 6.8, height = 3.4)
  save_plot_set(p_violin, paste0(basename, "_violin"), width = 7.2, height = 4.4)
  save_plot_set(p_density, paste0(basename, "_density"), width = 7.2, height = 4.4)
  save_plot_set(p_sample, paste0(basename, "_sample_means"), width = 7.2, height = 4.4)

  write.csv(score_long, file.path(output_dir, paste0(basename, "_cell_scores.csv")), row.names = FALSE)
  write.csv(score_summary, file.path(output_dir, paste0(basename, "_summary.csv")), row.names = FALSE)
  write.csv(sample_summary, file.path(output_dir, paste0(basename, "_sample_summary.csv")), row.names = FALSE)

  list(
    score_long = score_long,
    summary = score_summary,
    sample_summary = sample_summary,
    plots = list(
      dot_raw = p_dot_raw,
      dot_scaled = p_dot_scaled,
      violin = p_violin,
      density = p_density,
      sample = p_sample
    )
  )
}

make_correlation_bundle <- function(config) {
  joined_file <- file.path(scmetab_output_dir, config$scmetab_joined_file)
  if (!file.exists(joined_file)) {
    warning("Missing scMetabolism joined file: ", joined_file)
    return(list(joined = data.frame(), long = data.frame(), stats = data.frame(), plots = list()))
  }

  scmetab_cell <- read.csv(joined_file, check.names = FALSE, stringsAsFactors = FALSE)
  required_cols <- c("KEY", config$group_col, "xenium_glycolysis_score", "xenium_tca_score")
  missing_cols <- setdiff(required_cols, colnames(scmetab_cell))
  if (length(missing_cols) > 0) {
    stop("Missing columns in ", joined_file, ": ", paste(missing_cols, collapse = ", "))
  }

  scmetab_cell$xenium_cell <- if ("cell" %in% colnames(scmetab_cell)) {
    as.character(scmetab_cell$cell)
  } else {
    NA_character_
  }
  scmetab_cell$scmetab_sample_id <- if ("sample_id" %in% colnames(scmetab_cell)) {
    as.character(scmetab_cell$sample_id)
  } else if ("sample_id_xenium" %in% colnames(scmetab_cell)) {
    as.character(scmetab_cell$sample_id_xenium)
  } else {
    NA_character_
  }

  scmetab_keep <- scmetab_cell %>%
    dplyr::transmute(
      KEY = as.character(KEY),
      xenium_cell,
      scmetab_sample_id,
      scmetab_group = as.character(.data[[config$group_col]]),
      scmetabolism_glycolysis_score = xenium_glycolysis_score,
      scmetabolism_tca_score = xenium_tca_score
    )

  imc_protein_keep <- score_wide_all %>%
    dplyr::mutate(
      KEY = as.character(cell),
      group = as.character(.data[[config$group_col]])
    ) %>%
    dplyr::select(
      KEY,
      imc_cell = cell,
      imc_sample_id = sample_id,
      group,
      imc_protein_glycolysis_score = Glycolysis,
      imc_protein_tca_oxphos_score = `TCA / OXPHOS`
    )

  joined <- imc_protein_keep %>%
    dplyr::inner_join(scmetab_keep, by = "KEY") %>%
    dplyr::filter(
      group %in% config$group_levels,
      scmetab_group %in% config$group_levels,
      group == scmetab_group
    ) %>%
    dplyr::mutate(group = factor(group, levels = config$group_levels))

  cor_long <- dplyr::bind_rows(
    joined %>%
      dplyr::transmute(
        KEY,
        imc_cell,
        xenium_cell,
        imc_sample_id,
        scmetab_sample_id,
        group,
        comparison = "Glycolysis",
        imc_protein_score = imc_protein_glycolysis_score,
        scmetabolism_score = scmetabolism_glycolysis_score
      ),
    joined %>%
      dplyr::transmute(
        KEY,
        imc_cell,
        xenium_cell,
        imc_sample_id,
        scmetab_sample_id,
        group,
        comparison = "TCA / OXPHOS",
        imc_protein_score = imc_protein_tca_oxphos_score,
        scmetabolism_score = scmetabolism_tca_score
      )
  ) %>%
    dplyr::filter(
      is.finite(imc_protein_score),
      is.finite(scmetabolism_score)
    ) %>%
    dplyr::mutate(
      group = factor(as.character(group), levels = config$group_levels),
      comparison = factor(comparison, levels = pathway_levels)
    )

  cor_stats_by_group <- cor_long %>%
    dplyr::group_by(comparison, group) %>%
    dplyr::group_modify(~ cor_summary(.x)) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(level = "group")

  cor_stats_all <- cor_long %>%
    dplyr::group_by(comparison) %>%
    dplyr::group_modify(~ cor_summary(.x)) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(group = "All matched cells", level = "all")

  cor_stats <- dplyr::bind_rows(cor_stats_by_group, cor_stats_all) %>%
    dplyr::select(level, comparison, group, dplyr::everything())

  cor_plot_stats <- cor_stats %>%
    dplyr::filter(level == "group") %>%
    dplyr::mutate(
      comparison = factor(as.character(comparison), levels = pathway_levels),
      group = factor(as.character(group), levels = config$group_levels),
      label = paste0(
        "Spearman rho = ",
        formatC(spearman_rho, format = "f", digits = 2),
        "\n",
        format_p(spearman_p),
        "\nn = ",
        n
      )
    )

  if (nrow(cor_long) > 0) {
    p_points <- ggplot(cor_long, aes(x = imc_protein_score, y = scmetabolism_score, color = group)) +
      geom_point(alpha = 0.45, size = 1.25) +
      geom_smooth(method = "lm", formula = y ~ x, se = FALSE, linewidth = 0.65) +
      geom_label(
        data = cor_plot_stats,
        aes(x = -Inf, y = Inf, label = label),
        inherit.aes = FALSE,
        hjust = -0.04,
        vjust = 1.08,
        size = 3.1,
        label.size = 0.2,
        fill = "white",
        alpha = 0.9
      ) +
      facet_grid(group ~ comparison, scales = "free") +
      scale_color_manual(values = group_colors) +
      labs(
        title = paste0(config$label, " IMC protein score vs Xenium scMetabolism"),
        x = "IMC protein score",
        y = "Xenium scMetabolism AUCell score",
        color = NULL
      ) +
      theme_bw(base_size = 12) +
      theme(
        legend.position = "none",
        strip.text = element_text(size = 11.5, face = "bold"),
        plot.title = element_text(hjust = 0.5)
      )

    p_bins <- ggplot(cor_long, aes(x = imc_protein_score, y = scmetabolism_score)) +
      geom_bin2d(bins = 32) +
      geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "grey15", linewidth = 0.65) +
      geom_label(
        data = cor_plot_stats,
        aes(x = -Inf, y = Inf, label = label),
        inherit.aes = FALSE,
        hjust = -0.04,
        vjust = 1.08,
        size = 3.1,
        label.size = 0.2,
        fill = "white",
        alpha = 0.9
      ) +
      facet_grid(group ~ comparison, scales = "free") +
      scale_fill_gradient(low = "white", high = "#D7301F", name = "Matched cells") +
      labs(
        title = paste0(config$label, " binned IMC protein vs scMetabolism correlation"),
        x = "IMC protein score",
        y = "Xenium scMetabolism AUCell score"
      ) +
      theme_bw(base_size = 12) +
      theme(
        strip.text = element_text(size = 11.5, face = "bold"),
        plot.title = element_text(hjust = 0.5)
      )
  } else {
    p_points <- ggplot() +
      annotate("text", x = 0, y = 0, label = "No matched cells available", size = 4) +
      labs(title = paste0(config$label, " IMC protein score vs Xenium scMetabolism")) +
      theme_void()
    p_bins <- ggplot() +
      annotate("text", x = 0, y = 0, label = "No matched cells available", size = 4) +
      labs(title = paste0(config$label, " binned IMC protein vs scMetabolism correlation")) +
      theme_void()
  }

  basename <- paste0(config$file_prefix, "_imc_protein_vs_scmetabolism")
  save_plot_set(p_points, paste0(basename, "_correlation_points"), width = 8.2, height = 5.2)
  save_plot_set(p_bins, paste0(basename, "_correlation_bins"), width = 8.2, height = 5.2)

  write.csv(joined, file.path(output_dir, paste0(basename, "_cell_joined.csv")), row.names = FALSE)
  write.csv(cor_long, file.path(output_dir, paste0(basename, "_correlation_long.csv")), row.names = FALSE)
  write.csv(cor_stats, file.path(output_dir, paste0(basename, "_correlation_stats.csv")), row.names = FALSE)

  list(
    joined = joined,
    long = cor_long,
    stats = cor_stats,
    plots = list(points = p_points, bins = p_bins)
  )
}

build_analysis_result <- function(config) {
  imc_long <- make_imc_score_long(config)
  scmetab_long <- load_scmetab_score_long(config)

  list(
    imc = make_score_bundle(
      imc_long,
      config,
      source_id = "imc_protein",
      source_label = "IMC protein",
      y_label = "Mean zrescaled IMC protein score",
      raw_legend_title = "Mean IMC protein score"
    ),
    scmetabolism = make_score_bundle(
      scmetab_long,
      config,
      source_id = "scmetabolism",
      source_label = "Xenium scMetabolism",
      y_label = "scMetabolism AUCell score",
      raw_legend_title = "Mean AUCell score"
    ),
    correlation = make_correlation_bundle(config)
  )
}

analysis_results <- lapply(analysis_configs, build_analysis_result)

write.csv(marker_status, file.path(output_dir, "imc_protein_marker_status.csv"), row.names = FALSE)
```

# Marker Sets

The IMC protein scores are means of the available `zrescaled` IMC markers in
each pathway. The scMetabolism scores are the existing Xenium AUCell scores for
Glycolysis / Gluconeogenesis and Citrate cycle (TCA cycle), shown with the same
plot styles as the IMC protein scores.

```{r marker_status}
kable(marker_status, digits = 3)
```

# Fibroblasts

## Fibroblast IMC Protein Scores

```{r fibroblast_imc_summary}
kable(analysis_results$fibroblasts$imc$summary, digits = 4)
```

```{r fibroblast_imc_dotplots, fig.width=6.8, fig.height=3.4}
analysis_results$fibroblasts$imc$plots$dot_raw
analysis_results$fibroblasts$imc$plots$dot_scaled
```

```{r fibroblast_imc_distributions, fig.width=7.2, fig.height=4.4}
analysis_results$fibroblasts$imc$plots$violin
analysis_results$fibroblasts$imc$plots$density
```

```{r fibroblast_imc_samples, fig.width=7.2, fig.height=4.4}
analysis_results$fibroblasts$imc$plots$sample
```

## Fibroblast Xenium scMetabolism Scores

```{r fibroblast_scmetabolism_summary}
kable(analysis_results$fibroblasts$scmetabolism$summary, digits = 4)
```

```{r fibroblast_scmetabolism_dotplots, fig.width=6.8, fig.height=3.4}
analysis_results$fibroblasts$scmetabolism$plots$dot_raw
analysis_results$fibroblasts$scmetabolism$plots$dot_scaled
```

```{r fibroblast_scmetabolism_distributions, fig.width=7.2, fig.height=4.4}
analysis_results$fibroblasts$scmetabolism$plots$violin
analysis_results$fibroblasts$scmetabolism$plots$density
```

```{r fibroblast_scmetabolism_samples, fig.width=7.2, fig.height=4.4}
analysis_results$fibroblasts$scmetabolism$plots$sample
```

## Fibroblast IMC Protein vs scMetabolism Correlation

```{r fibroblast_correlation_stats}
kable(analysis_results$fibroblasts$correlation$stats, digits = 4)
```

```{r fibroblast_correlation_plots, fig.width=8.2, fig.height=5.2}
analysis_results$fibroblasts$correlation$plots$points
analysis_results$fibroblasts$correlation$plots$bins
```

# Endothelial Cells

## EC IMC Protein Scores

```{r ec_imc_summary}
kable(analysis_results$endothelial$imc$summary, digits = 4)
```

```{r ec_imc_dotplots, fig.width=6.8, fig.height=3.4}
analysis_results$endothelial$imc$plots$dot_raw
analysis_results$endothelial$imc$plots$dot_scaled
```

```{r ec_imc_distributions, fig.width=7.2, fig.height=4.4}
analysis_results$endothelial$imc$plots$violin
analysis_results$endothelial$imc$plots$density
```

```{r ec_imc_samples, fig.width=7.2, fig.height=4.4}
analysis_results$endothelial$imc$plots$sample
```

## EC Xenium scMetabolism Scores

```{r ec_scmetabolism_summary}
kable(analysis_results$endothelial$scmetabolism$summary, digits = 4)
```

```{r ec_scmetabolism_dotplots, fig.width=6.8, fig.height=3.4}
analysis_results$endothelial$scmetabolism$plots$dot_raw
analysis_results$endothelial$scmetabolism$plots$dot_scaled
```

```{r ec_scmetabolism_distributions, fig.width=7.2, fig.height=4.4}
analysis_results$endothelial$scmetabolism$plots$violin
analysis_results$endothelial$scmetabolism$plots$density
```

```{r ec_scmetabolism_samples, fig.width=7.2, fig.height=4.4}
analysis_results$endothelial$scmetabolism$plots$sample
```

## EC IMC Protein vs scMetabolism Correlation

```{r ec_correlation_stats}
kable(analysis_results$endothelial$correlation$stats, digits = 4)
```

```{r ec_correlation_plots, fig.width=8.2, fig.height=5.2}
analysis_results$endothelial$correlation$plots$points
analysis_results$endothelial$correlation$plots$bins
```
