---
title: "IMC heatmap sankey plot (new)"
author: "Veda"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    wrap: 72
    code_folding: hide
---

```{r setup, include=FALSE}
source("data_preprocessing/common/config.R")
knitr::opts_chunk$set(echo = TRUE)
```

```{r library set up}
library(SingleCellExperiment)
library(SpatialExperiment)
library(DelayedMatrixStats)
library(dittoSeq)
library(ggsankey)
library(ggplot2)
library(dplyr)
library(viridis)
library(scuttle)
library(SpatialFeatureExperiment)
# Assume `sfe` and `spe` objects are already loaded in your environment

```

# Heatmaps metabolic marker expression 

```{r new IMC Rphenograph_500 heatmap}
load(project_path("Revision/Nan annotation/Data/IMC spe objects/New IMC/spe2025-12-17 umap.RData"))
# sfe<-readRDS(project_path("Revision/Nan annotation/Data/Xenium sfe objects/Xenium sfe object/XeniumSfe_FibAnno_251104104458.rds"))
sfe<-readRDS(project_path("Revision/Nan annotation/Data/Xenium sfe objects/Xenium sfe object/sfe_09032026.rds"))

z<-list("glycolysis"=c(2,30,32,7,15,26),"TCA_cycle"=c(29,24,31,17),"Amino_Acid"=c(8),"Fatty_acid_synthesis"=c(36),
        "Fatty_acid_oxidation"=(34),"Pentose_Pathway"=12,"anabolic_signaling"=c(9),
        "catabolic_signaling"=c(21),"Oxidative_stress"=c(37),
        "hypoxia"=38,"Mitochondrial_biogenesis"=c(22,13,10))

spe$metfiblabel <- case_when(
  spe$Rphenograph_500 %in% c("1", "2") ~ "Met_hi_Fib",
  spe$Rphenograph_500 == "3" ~ "Other_Fib"
)

markers_ordered <- names(spe)[do.call(c, z)]

rowData(spe)$use_channel_met_markers<-grepl("GLUT1|PKM2|CD98|p_mTOR|pNRF2|G6PD|mtTFA|LDHA|SDHA|pAMPK|PGC1a|OGDH|GAPDH|CS|HK1|ATP5A|PFKL1_PFKM|CPT1a|ACAC|NOX4|HIF1a", rownames(spe)) ## for new spe

celltype_mean <- aggregateAcrossCells(as(spe[rowData(spe)$use_channel_met_markers,], "SingleCellExperiment"),  
                                              ids = spe$"metfiblabel", 
                                              statistics = "mean", 
                                              use.assay.type = "zrescaled")

spe_test <- celltype_mean
spe_test$metfiblabel <- as.character(spe_test$metfiblabel)
dittoHeatmap(
  spe_test,
  assay = "zrescaled",
  genes = markers_ordered,
  cluster_cols = T,
  cluster_rows = F,
  #scale = "zscaled",
  heatmap.colors = c(viridis(50), rep("#FDE725FF", 1.5)),
  annot.by = c("metfiblabel", 'ncells')
)

dittoHeatmap(
  spe_test,
  assay = "zrescaled",
  genes = markers_ordered,
  cluster_cols = TRUE,
  cluster_rows = FALSE,
  heatmap.colors = viridis(100),
  annot.by = c("metfiblabel", "ncells"),
  scale = "none",
  breaks = seq(-1, 0.5, length.out = 101)
)


```

```{r data}
library(SpatialExperiment)
library(scater)
library(tidyverse)

spe2 <- spe

spe1 <- readRDS(project_path("Revision/Nan annotation/Data/IMC spe objects/spe_old.rds"))
#spe2 <- readRDS(project_path("external/spe_final_with_all_gates.rds"))
spe1$dataset <- 'orig_imc'
spe2$dataset <- 'postxenium_imc'


meta <- dplyr::full_join(
  dplyr::select(as.data.frame(spe1@colData), ObjectNumber, sample_id, cell_labels, celltype, dataset) %>%
    mutate(gates = dplyr::case_when(
      .$cell_labels == 'CD31+' ~ 'Endothelial',
      .$cell_labels == 'E-cad+' ~ 'Epithelial',
      .$cell_labels == 'CD45+' ~ 'Immune',
      .default = .$cell_labels)) %>%
    mutate(gates = ifelse(startsWith(.$celltype, 'CD45_CD68'), 'Macrophage', .$gates)),
  dplyr::select(as.data.frame(spe2@colData), ObjectNumber, sample_id, gates, dataset, metfiblabel) 
)
ch_used <- rowData(spe1)$use_channel_met_markers

expr <- cbind(assay(spe1, 'z-rescaled_new'), assay(spe2, 'zrescaled'))
counts <- cbind(assay(spe1, 'counts'), assay(spe2, 'counts'))
asinh <- cbind(assay(spe1, 'asinh'), assay(spe2, 'asinh'))

merged <- SingleCellExperiment(
  list(counts = counts, asinh = asinh, expr = expr),
  colData = meta,
  rowData = rowData(spe1)
)
set.seed(171025)
merged <- scater::runUMAP(merged, subset_row = ch_used, exprs_values = 'expr', name = 'umap',
                  BPPARAM = BiocParallel::MulticoreParam())

library(dittoSeq)

# dittoDimPlot(merged, var = 'dataset', reduction.use = 'umap')
# dittoDimPlot(merged, var = 'gates', reduction.use = 'umap')

```

# Seurat anchor-based label transfer - Stromal

```{r stromal-seu, fig.width=12, fig.height=4}
library(Seurat)
gate <- 'Stromal'

obj <- merged[,merged$dataset == 'orig_imc' & merged$gates == gate]

assay1 <- CreateAssay5Object(counts = assay(obj, 'counts'), data = assay(obj, 'asinh'))

ref <- CreateSeuratObject(
  assay1,
  meta.data = as.data.frame(obj@colData)
)

ref@assays$RNA$scale.data <- assay(obj, 'expr') %>% `rownames<-`(rownames(ref))
VariableFeatures(ref) <- rownames(ref)[ch_used]
# ref <- ScaleData(ref)
ref <- RunPCA(ref)
ElbowPlot(ref)

ref <- RunUMAP(ref,  dims = 1:20, return.model = TRUE)
  

DimPlot(ref, group.by = 'celltype', alpha = .3)

obj <- merged[,merged$dataset == 'postxenium_imc' & merged$gates == gate]

assay(obj, "z_scaled_new") <- t(apply(assay(obj, "asinh"), 1, function(x) (x - mean(x))/sd(x)))

assay2 <- CreateAssay5Object(counts = assay(obj, 'counts'), data = assay(obj, 'asinh'))

to_dgC <- function(x) {
  if (inherits(x, "dgCMatrix")) return(x)
  x <- as(x, "CsparseMatrix")
  as(x, "dgCMatrix")
}

if (exists("cur_sfe")) {
  assay(cur_sfe, "counts") <- to_dgC(assay(cur_sfe, "counts"))
  assay(cur_sfe, "logcounts") <- to_dgC(assay(cur_sfe, "logcounts"))

  seurat_obj <- as.Seurat(
    cur_sfe,
    counts = "counts",
    data   = "logcounts"
  )
}

query <- CreateSeuratObject(
  assay2,
  meta.data = as.data.frame(obj@colData)
)

VariableFeatures(query) <- rownames(query)[ch_used]
query@assays$RNA$scale.data <- assay(obj, 'z_scaled_new') %>% `rownames<-`(rownames(ref))
query <- RunPCA(query)
ElbowPlot(query)

anchors <- FindTransferAnchors(reference = ref, query = query, dims = 1:20,
    reference.reduction  = "pca")
query <- MapQuery(anchorset = anchors, reference = ref, query = query, reference.dims = 1:20,
    refdata = list(celltype = "celltype"), reference.reduction = "pca", reduction.model = "umap")
p1 <- DimPlot(ref, group.by = 'celltype', alpha = .3)
p2 <- DimPlot(query, group.by = 'predicted.celltype', alpha = .3)
p3 <- DimPlot(query, group.by = 'metfiblabel', alpha = .3)
print(p1|p2|p3)


merged$predicted.celltype <- NA
pred <- setNames(query$predicted.celltype, colnames(query))
lab <- case_when(is.na(merged$predicted.celltype) ~ pred[colnames(merged)],
                 .default = merged$predicted.celltype)
merged$predicted.celltype <- lab

table(merged$predicted.celltype)
table(query$predicted.celltype,query$metfiblabel)

```


```{r IMC Rphenograph_500 transfer}
#load(project_path("external/spe2025-12-17 umap.RData"))

head(colnames(spe))
head(colData(sfe)$KEY)

unique(colData(spe)$metfiblabel)

# Create a lookup dataframe
df_map <- data.frame(
  KEY = colnames(spe),
  metfiblabel = spe$metfiblabel,
  stringsAsFactors = FALSE
)

# Align based on shared KEYs
idx <- match(colData(sfe)$KEY, df_map$KEY)

# Transfer the predicted.celltype column
colData(sfe)$metfiblabel <- df_map$metfiblabel[idx]
# Check a few transferred entries
head(colData(sfe)$metfiblabel)

# Count how many matches were successful
sum(!is.na(colData(sfe)$metfiblabel))
sum(is.na(colData(sfe)$metfiblabel))



```

# Sankey plots - mapping metabolic IMC to Xenium clusters

```{r new IMC Rphenograph_500 sankey}

if ("lv3_anno"%in%colnames(colData(sfe))){
  sfe$lv3_anno <- sfe$lv3_anno
} else {
  sfe$lv3_anno <- sfe$lv2_anno_new
}

df2 <- as.data.frame(S4Vectors::as.data.frame(colData(sfe)))[, c("lv3_anno", "metfiblabel")]
df2 <- df2[!is.na(df2$lv3_anno) & !is.na(df2$metfiblabel) & df2$lv3_anno != "" & grepl("_Fib$", df2$lv3_anno),]

sankey3 <- ggsankey::make_long(df2, lv3_anno, metfiblabel)

ggplot(sankey3,
       aes(x = x, next_x = next_x,
           node = node, next_node = next_node,
           fill = node, label = node)) +
  geom_sankey(flow.alpha = 0.7, node.color = "grey30") +
  geom_sankey_label(size = 3, color = "black") +
  theme_minimal() +
  labs(title = "Sankey 3 – lv3_anno → Met State (met_hi / met_others only)",
       x = NULL, y = NULL) +
  theme(axis.text.x = element_text(size = 10, face = "bold"))

matching <- as.data.frame(S4Vectors::as.data.frame(colData(sfe)))[, c("lv3_anno", "metfiblabel")]
matching <- df2[!is.na(df2$lv3_anno) & !is.na(df2$metfiblabel) & df2$lv3_anno != "", ]

# Summarize and calculate frequency
matching.summary <- matching %>%
  group_by(across(all_of(colnames(matching)))) %>%
  summarize(count = n(), .groups = "drop") %>%
  group_by(across(all_of(colnames(matching)[2]))) %>%
  mutate(tot_cells = sum(count), freq = count / tot_cells) %>%
  ungroup()

# Identify rare matches (freq < 0.1 or count < 20)
rare.matches <- matching.summary %>%
  filter(freq < 0.02 | count < 5) %>%
  select(all_of(colnames(matching[1:2])))

# Remove rare matches from the original data
matching.filtered <- matching %>%
  anti_join(rare.matches, by = colnames(matching)[1:2])

# Create sankey data with filtered matches
matching.sankey <- make_long(matching.filtered, !!sym(colnames(matching)[1]), !!sym(colnames(matching)[2]))

# Plot the simplified Sankey
ggplot(
  matching.sankey,
  aes(
    x = x,
    next_x = next_x,
    node = node,
    next_node = next_node,
    fill = factor(node),
    label = node
  )
) +
  geom_sankey(flow.alpha = 0.3, node.color = "gray30") +
  geom_sankey_label(size = 3,
                    color = "white",
                    fill = "gray40") +
  theme_sankey(base_size = 18) +
  labs(x = NULL) +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5))


```

```{r sankey figure, fig.width=6, fig.height=4}
ggplot(
  matching.sankey,
  aes(
    x = x,
    next_x = next_x,
    node = node,
    next_node = next_node,
    fill = factor(node),
    label = node
  )
) +
  geom_sankey(
    flow.alpha = 0.3,
    node.color = "gray30",
    width = 0.18
  ) +

  # left-side labels: move left of nodes
  geom_sankey_text(
    data = subset(matching.sankey, x == "lv3_anno"),
    size = 5.5,
    color = "black",
    hjust = 1,
    position = position_nudge(x = -0.12)
  ) +

  # right-side labels: move right of nodes
  geom_sankey_text(
    data = subset(matching.sankey, x == "metfiblabel"),
    size = 5.5,
    color = "black",
    hjust = 0,
    position = position_nudge(x = 0.12)
  ) +

  theme_sankey(base_size = 20) +
  labs(x = NULL) +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, size = 20),
    plot.margin = margin(10, 35, 10, 35)
  ) +
  coord_cartesian(clip = "off")
```


# Dot plots - gene expression across metabolic clusters

```{r dotplot 2}

# genes_use_subset <- c(
#   "CTHRC1","POSTN","FN1","COL1A1","COL1A2","COL8A1","SFRP4","SFRP2","FAP","LRRC15",
#   "CCL19","CXCL12","APOE","PI16","LGR5","COL18A1","WIF1","ACTA2"
# )

genes_use_subset <- unique(c(
  # your original set
  "CTHRC1","POSTN","FN1","COL1A1","COL1A2","COL8A1","SFRP4","SFRP2","FAP","LRRC15",
  "CCL19","CXCL12","APOE","PI16","LGR5","COL18A1","WIF1","ACTA2",

  # pan-fibro / “universal” fibroblasts (often used to define less-activated states)
  "DCN","LUM","DPT","CD34","PDGFRA",
  "COL15A1","COL14A1","C7","CFD","CLU","APOD","FBLN1","FBLN2","TNXB",

  # inflammatory fibroblasts (iCAF-like chemokine/cytokine programs)
  "IL6","IL11","CXCL1","CXCL8","CCL2","CCL20","IL1R1",

  # antigen-presenting fibroblasts (apCAFs; MHC-II module)
  "CD74","HLA-DRA","HLA-DRB1","HLA-DPA1","HLA-DPB1",

  # profibrotic / myofibroblast-like programs (ECM production, contractility, crosslinking)
  "CTGF","TGFBI","SERPINE1",
  "TAGLN","MYL9","CALD1","MYH11",
  "ITGA11","COL3A1","COL5A1","COL5A2","COL6A1","COL6A2","COL12A1","COL11A1",
  "SPARC","BGN","THBS1","TNC",
  "LOX","LOXL2",
  "MMP2","TIMP1",

  # matrix-remodeling CAF-like programs (often separate from pure myofibroblast markers)
  "MMP1","MMP3","MMP9"
))

genes_use_subset_selected <- unique(c(
  # your original set
  "ACTA2","COL8A1","FN1","SFRP4","CCL19","COL18A1","WIF1",
  
  # inflammatory fibroblasts (iCAF-like chemokine/cytokine programs)
  "IL6","IL11","CCL20","IL1R1",
  
  # profibrotic / myofibroblast-like programs (ECM production, contractility, crosslinking)
  "TNC",
  "LOX","LOXL2",
  "MMP2","TIMP1",

  # core fibroblast identity / stromal ECM
  "DCN", "LUM", "PDGFRA", "DPT",

  # universal / less-activated fibroblast state
  "PI16", "CD34",

  # common fibroblast-enriched ECM marker from cross-organ analyses
  "FBLN2"
))

# Optional: if you want to explicitly flag pericyte-/mural-like clusters to exclude or separate
genes_pericyte_like <- c("RGS5","PDGFRB","CSPG4","MCAM","ABCC9","KCNJ8","NOTCH3")

stromal_types <- c("Other_Fib", "Met_hi_Fib")

cur_sfe <- sfe[, sfe$metfiblabel %in% stromal_types]

library(Seurat)

assay(cur_sfe, "counts") <- to_dgC(assay(cur_sfe, "counts"))
assay(cur_sfe, "logcounts") <- to_dgC(assay(cur_sfe, "logcounts"))

seurat_obj <- as.Seurat(
  cur_sfe,
  counts = "counts",
  data   = "logcounts"
)
unique(Idents(seurat_obj))
Idents(seurat_obj) <- "metfiblabel"

# ✅ enforce stromal_1 ... stromal_8 order
Idents(seurat_obj) <- as.factor(Idents(seurat_obj))#, levels = stromal_types)

## --- Prepare data: only stromal_2, stromal_4, stromal_8 ---

keep_stromal <- c("Other_Fib", "Met_hi_Fib")
#keep_stromal <- unique(seurat_obj$predicted.celltype)

# Idents(seurat_obj) <- factor(
#   seurat_obj$predicted.celltype,
#   levels = metfiblabel
# )
# 
 obj_sub <- subset(seurat_obj, idents = keep_stromal)


## --- Extract DotPlot data and manually row-scale gene expression ---

ddata1 <- DotPlot(obj_sub, features = genes_use_subset)$data
ddata2 <- DotPlot(obj_sub, features = genes_use_subset_selected)$data

library(dplyr)

# Row-wise scaling (per gene)
ddata1 <- ddata1 %>%
  group_by(features.plot) %>%
  mutate(avg.exp.scaled = base::scale(avg.exp)[,1]) %>%
  ungroup()

ddata2 <- ddata2 %>%
  group_by(features.plot) %>%
  mutate(avg.exp.scaled = base::scale(avg.exp)[,1]) %>%
  ungroup()

# Ensure x-axis = stromal clusters in correct order
ddata1$id <- factor(ddata1$id, levels = keep_stromal)
ddata2$id <- factor(ddata2$id, levels = keep_stromal)


## --- Final plot: x = clusters, y = genes ---

p1 <- ggplot(ddata1, aes(x = id, y = features.plot)) +
  geom_point(aes(size = pct.exp, color = avg.exp.scaled)) +
  scale_color_gradientn(colors = rev(RColorBrewer::brewer.pal(7, "RdBu"))) +
  scale_size(range = c(1, 8)) +
  labs(x = NULL, y = NULL) +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8)
  )

p2 <- ggplot(ddata2, aes(x = id, y = features.plot)) +
  geom_point(aes(size = pct.exp, color = avg.exp.scaled)) +
  scale_color_gradientn(colors = rev(RColorBrewer::brewer.pal(7, "RdBu"))) +
  scale_size(range = c(1, 8)) +
  labs(x = NULL, y = NULL) +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8)
  )

p1
p2


```

```{r dotplot figure, fig.width=6, fig.height=6}
p2 <- p2 +
  scale_y_discrete(limits = rev) +
  theme(
    axis.text.x  = element_text(size = 18, angle = 45, hjust = 1),
    axis.text.y  = element_text(size = 18),
    legend.title = element_text(size = 18),
    legend.text  = element_text(size = 16)
  )
p2
```


# ECM scores 1 {.tabset .tabset-pill}

```{r ecm1, results='asis'}

#seurat <- seurat_obj

matrisome <- read.csv(project_path("Revision/Nan annotation/Data/Hs_Matrisome_Masterlist_Naba.csv"))
ecm_gene <- matrisome %>% filter(`Matrisome.Division` == "Core matrisome")
ecm_gene <- dplyr::select(ecm_gene, `Matrisome.Category`, `Gene.Symbol`)

## correct gene names in the data set
`%notin%` <- Negate(`%in%`)
all(ecm_gene$`Gene.Symbol` %in% rownames(seurat_obj))

ecm_gene$`Gene.Symbol`[which(ecm_gene$`Gene.Symbol` %notin% rownames(seurat_obj))]

ecm_gene <- rbind(ecm_gene,
                  data.frame(`Matrisome.Category` = rep('CoreMatrisome', nrow(ecm_gene)), 
                             `Gene.Symbol` = ecm_gene$`Gene.Symbol`))

ecm_list <- split(ecm_gene$`Gene.Symbol`, ecm_gene$`Matrisome.Category`)
names(ecm_list)[names(ecm_list) == "ECM Glycoproteins"] <- "Glycoproteins"
ecm_list[["ECM"]] <- pull(ecm_gene, `Gene.Symbol`)

library(decoupleR)

mat <- seurat_obj@assays$originalexp$data
scores <- run_aucell(mat, ecm_gene, minsize = 0, .source = `Matrisome.Category`, .target = `Gene.Symbol`, nproc = 1)

seurat_obj[['Naba_matrisome_aucell']] <- scores %>%
  tidyr::pivot_wider(id_cols = 'source', names_from = 'condition', values_from = 'score') %>%
  dplyr::mutate(source = paste0(source, ' score')) %>%
  tibble::column_to_rownames(var = 'source') %>%
  Seurat::CreateAssayObject(.)

#seurat <- AddModuleScore(seurat, ecm_list, name = 'ECM_score', ctrl = 10)


Seurat::DefaultAssay(object = seurat_obj) <- "Naba_matrisome_aucell"

for (i in rownames(seurat_obj)) {
  cat('\n\n##', i, '\n\n')
  # p <- FeaturePlot(
  #   seurat_obj,
  #   i,
  #   reduction = 'umap_fib',
  #   cols = c('#dcdcdc90', 'darkred'),
  #   order = T,
  #   alpha = .3
  # )
  p2 <- VlnPlot(seurat_obj, features = i, group.by = 'metfiblabel', pt.size = .00001, alpha = .01, sort = T) + NoLegend()
 # p3 <- VlnPlot(seurat_obj, features = i, group.by = 'metfiblabel', split.by = 'Indication', pt.size = .00001, alpha = .01, sort = T)
  if (exists("p")) print(p)
  print(p2)
 # print(p3)
  cat('\n\n')
}

#   for (j in seq_along(ecm_list)) {
#     colnames(seurat@meta.data)[which(colnames(seurat@meta.data) == paste0('ECM_score', j))] <- paste0(names(ecm_list)[j], '_score')
#   }
# p <- VlnPlot(seurat, grep('_score', colnames(seurat@meta.data), value = TRUE), 
#                group.by = 'metfiblabel', combine = FALSE, pt.size = .2, alpha = .2)
#   plot_list <- lapply(p, function(plot) {
#     plot + theme(legend.position = "none")
# })
# for (j in seq_along(plot_list)) {
#     cat("\n\n### ", names(ecm_list)[j], '\n')
#     print(plot_list[[j]])
#   }
```

# Dot plots gene expression across Xenium clusters separated by metabolism

```{r dotplot 3}

# genes_use_subset <- c(
#   "CTHRC1","POSTN","FN1","COL1A1","COL1A2","COL8A1","SFRP4","SFRP2","FAP","LRRC15",
#   "CCL19","CXCL12","APOE","PI16","LGR5","COL18A1","WIF1", "ACTA2"
# )

stromal_types <- c("Other_Fib", "Met_hi_Fib")
cur_sfe <- sfe[, sfe$metfiblabel %in% stromal_types]

cur_sfe$lv3_by_met <- paste(cur_sfe$lv3_anno, cur_sfe$metfiblabel, sep = "_")
cur_sfe$lv3_by_met <- factor(cur_sfe$lv3_by_met)

table(cur_sfe$lv3_by_met)


library(Seurat)

assay(cur_sfe, "counts") <- to_dgC(assay(cur_sfe, "counts"))
assay(cur_sfe, "logcounts") <- to_dgC(assay(cur_sfe, "logcounts"))

seurat_obj <- as.Seurat(
  cur_sfe,
  counts = "counts",
  data   = "logcounts"
)
unique(Idents(seurat_obj))
Idents(seurat_obj) <- "lv3_by_met"

# ✅ enforce stromal_1 ... stromal_8 order
Idents(seurat_obj) <- as.factor(Idents(seurat_obj))#, levels = stromal_types)

## --- Prepare data: only stromal_2, stromal_4, stromal_8 ---

keep_stromal <- c(
  "papillary_Fib_Other_Fib",
  "papillary_Fib_Met_hi_Fib",
  "PI16_Fib_Other_Fib",
  "PI16_Fib_Met_hi_Fib"
  # "CXCL12_Fib_Met_int_Fib",
  # "CXCL12_Fib_Met_hi_Fib",
  # "COL8A1_Fib_Met_int_Fib",
  # "COL8A1_Fib_Met_hi_Fib",
  # "CCL19_Fib_Met_int_Fib",
  # "CCL19_Fib_Met_hi_Fib",
  # "SPARC_Fib_Met_int_Fib",
  # "SPARC_Fib_Met_hi_Fib",
  # "COCH_Fib_Met_int_Fib",
  # "COCH_Fib_Met_hi_Fib",
  # "NGFR_Fib_Met_int_Fib",
  # "NGFR_Fib_Met_hi_Fib"
)
#keep_stromal <- unique(seurat_obj$predicted.celltype)

Idents(seurat_obj) <- factor(
  seurat_obj$lv3_by_met,
  levels = keep_stromal
)

 obj_sub <- subset(seurat_obj, idents = keep_stromal)


## --- Extract DotPlot data and manually row-scale gene expression ---

ddata <- DotPlot(obj_sub, features = genes_use_subset_selected)$data

library(dplyr)

# Row-wise scaling (per gene)
ddata <- ddata %>%
  group_by(features.plot) %>%
  mutate(avg.exp.scaled = base::scale(avg.exp)[,1]) %>%
  ungroup()

# Ensure x-axis = stromal clusters in correct order
ddata$id <- factor(ddata$id, levels = keep_stromal)


## --- Final plot: x = clusters, y = genes ---

p <- ggplot(ddata, aes(x = id, y = features.plot)) +
  geom_point(aes(size = pct.exp, color = avg.exp.scaled)) +
  scale_color_gradientn(colors = rev(RColorBrewer::brewer.pal(7, "RdBu"))) +
  scale_size(range = c(1, 8)) +
  labs(x = NULL, y = NULL) +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8)
  )

p


```

# ECM scores 2 {.tabset .tabset-pill}

```{r ecm2, results='asis'}

matrisome <- read.csv(project_path("Revision/Nan annotation/Data/Hs_Matrisome_Masterlist_Naba.csv"))
ecm_gene <- matrisome %>% filter(`Matrisome.Division` == "Core matrisome")
ecm_gene <- dplyr::select(ecm_gene, `Matrisome.Category`, `Gene.Symbol`)

## correct gene names in the data set
`%notin%` <- Negate(`%in%`)
all(ecm_gene$`Gene.Symbol` %in% rownames(seurat_obj))

ecm_gene$`Gene.Symbol`[which(ecm_gene$`Gene.Symbol` %notin% rownames(seurat_obj))]

ecm_gene <- rbind(ecm_gene,
                  data.frame(`Matrisome.Category` = rep('CoreMatrisome', nrow(ecm_gene)), 
                             `Gene.Symbol` = ecm_gene$`Gene.Symbol`))

library(decoupleR)

keep_stromal <- c(
  "papillary_Fib_Other_Fib",
  "papillary_Fib_Met_hi_Fib",
  "PI16_Fib_Other_Fib",
  "PI16_Fib_Met_hi_Fib",
  "CXCL12_Fib_Other_Fib",
  "CXCL12_Fib_Met_hi_Fib",
  "COL8A1_Fib_Other_Fib",
  "COL8A1_Fib_Met_hi_Fib",
  "CCL19_Fib_Other_Fib",
  "CCL19_Fib_Met_hi_Fib",
  "COMP_Fib_Other_Fib",
  "COMP_Fib_Met_hi_Fib",
  "COCH_Fib_Other_Fib",
  "COCH_Fib_Met_hi_Fib",
  "NGFR_Fib_Other_Fib",
  "NGFR_Fib_Met_hi_Fib"
)
#keep_stromal <- unique(seurat_obj$predicted.celltype)

Idents(seurat_obj) <- factor(
  seurat_obj$lv3_by_met,
  levels = keep_stromal
)

 obj_sub <- subset(seurat_obj, idents = keep_stromal)

mat <- obj_sub@assays$originalexp$data
scores <- run_aucell(mat, ecm_gene, minsize = 0, .source = `Matrisome.Category`, .target = `Gene.Symbol`, nproc = 1)

obj_sub[['Naba_matrisome_aucell']] <- scores %>%
  tidyr::pivot_wider(id_cols = 'source', names_from = 'condition', values_from = 'score') %>%
  dplyr::mutate(source = paste0(source, ' score')) %>%
  tibble::column_to_rownames(var = 'source') %>%
  Seurat::CreateAssayObject(.)


Seurat::DefaultAssay(object = obj_sub) <- "Naba_matrisome_aucell"

obj_sub$lv3_by_met <- factor(obj_sub$lv3_by_met, levels=keep_stromal)

for (i in rownames(obj_sub)) {
  cat('\n\n##', i, '\n\n')
  # p <- FeaturePlot(
  #   obj_sub,
  #   i,
  #   reduction = 'umap_fib',
  #   cols = c('#dcdcdc90', 'darkred'),
  #   order = T,
  #   alpha = .3
  # )
  p2 <- VlnPlot(obj_sub, features = i, group.by = "lv3_by_met", pt.size = .00001, alpha = .01, sort = F) + NoLegend()
 # p3 <- VlnPlot(seurat_obj, features = i, group.by = 'metfiblabel', split.by = 'Indication', pt.size = .00001, alpha = .01, sort = T)
  if (exists("p")) print(p)
  print(p2)
 # print(p3)
  cat('\n\n')
}
```

# scMetabolism on Xenium fibroblast clusters {.tabset .tabset-pill}

```{r scmetabolism_setup, message=FALSE, warning=FALSE}
if (!requireNamespace("scMetabolism", quietly = TRUE)) {
  stop(
    "The scMetabolism package is required for this section. Install it with ",
  )
}

library(scMetabolism)
library(SpatialFeatureExperiment)
library(SummarizedExperiment)
library(Seurat)
library(Matrix)
library(ggplot2)
library(dplyr)
library(tibble)
library(tidyr)
library(RColorBrewer)

if (!exists("to_dgC")) {
  to_dgC <- function(x) {
    if (inherits(x, "dgCMatrix")) return(x)
    x <- as(x, "CsparseMatrix")
    as(x, "dgCMatrix")
  }
}
```

```{r scmetabolism_prepare_sfe}
if (!exists("spe")) {
  load(project_path("Revision/Nan annotation/Data/IMC spe objects/New IMC/spe2025-12-17 umap.RData"))
}

if (!exists("sfe")) {
  sfe <- readRDS(project_path("Revision/Nan annotation/Data/Xenium sfe objects/Xenium sfe object/sfe_09032026.rds"))
}

spe$metfiblabel <- dplyr::case_when(
  spe$Rphenograph_500 %in% c("1", "2") ~ "Met_hi_Fib",
  spe$Rphenograph_500 == "3" ~ "Other_Fib"
)

if (!"KEY" %in% colnames(SummarizedExperiment::colData(sfe))) {
  stop("The Xenium sfe object must contain colData(sfe)$KEY to transfer metfiblabel from the IMC spe.")
}

df_map <- data.frame(
  KEY = colnames(spe),
  metfiblabel = spe$metfiblabel,
  stringsAsFactors = FALSE
)

idx <- match(SummarizedExperiment::colData(sfe)$KEY, df_map$KEY)
sfe$metfiblabel <- df_map$metfiblabel[idx]

if (!"lv3_anno" %in% colnames(SummarizedExperiment::colData(sfe))) {
  if ("lv2_anno_new" %in% colnames(SummarizedExperiment::colData(sfe))) {
    sfe$lv3_anno <- sfe$lv2_anno_new
  } else {
    stop("The Xenium sfe object must contain lv3_anno or lv2_anno_new.")
  }
}

stromal_types <- c("Other_Fib", "Met_hi_Fib")
keep_scmetab <- !is.na(sfe$lv3_anno) &
  sfe$lv3_anno != "" &
  grepl("_Fib$", sfe$lv3_anno) &
  !is.na(sfe$metfiblabel) &
  sfe$metfiblabel %in% stromal_types

sfe_scmetab <- sfe[, keep_scmetab]

lv3_preferred_order <- c(
  "papillary_Fib",
  "PI16_Fib",
  "CXCL12_Fib",
  "COL8A1_Fib",
  "CCL19_Fib",
  "COMP_Fib",
  "COCH_Fib",
  "NGFR_Fib"
)

lv3_levels <- unique(as.character(sfe_scmetab$lv3_anno))
lv3_levels <- c(
  intersect(lv3_preferred_order, lv3_levels),
  setdiff(sort(lv3_levels), lv3_preferred_order)
)

sfe_scmetab$lv3_anno <- factor(as.character(sfe_scmetab$lv3_anno), levels = lv3_levels)
sfe_scmetab$metfiblabel <- factor(as.character(sfe_scmetab$metfiblabel), levels = stromal_types)
sfe_scmetab$lv3_by_met <- paste(sfe_scmetab$lv3_anno, sfe_scmetab$metfiblabel, sep = "_")

lv3_by_met_levels <- as.vector(t(outer(lv3_levels, stromal_types, paste, sep = "_")))
lv3_by_met_levels <- intersect(lv3_by_met_levels, unique(sfe_scmetab$lv3_by_met))
sfe_scmetab$lv3_by_met <- factor(sfe_scmetab$lv3_by_met, levels = lv3_by_met_levels)

table(sfe_scmetab$lv3_anno, sfe_scmetab$metfiblabel)
```

```{r scmetabolism_run, message=FALSE, warning=FALSE}
counts_mat <- to_dgC(SummarizedExperiment::assay(sfe_scmetab, "counts"))
scmetab_meta <- as.data.frame(SummarizedExperiment::colData(sfe_scmetab))
rownames(scmetab_meta) <- colnames(sfe_scmetab)

old_assay_version <- getOption("Seurat.object.assay.version")
options(Seurat.object.assay.version = "v3")

xenium_scmetab <- Seurat::CreateSeuratObject(
  counts = counts_mat,
  meta.data = scmetab_meta,
  assay = "RNA",
  project = "Xenium_scMetabolism"
)

if (is.null(old_assay_version)) {
  options(Seurat.object.assay.version = NULL)
} else {
  options(Seurat.object.assay.version = old_assay_version)
}

xenium_scmetab <- Seurat::NormalizeData(xenium_scmetab, verbose = FALSE)
xenium_scmetab$lv3_anno <- factor(as.character(xenium_scmetab$lv3_anno), levels = lv3_levels)
xenium_scmetab$metfiblabel <- factor(as.character(xenium_scmetab$metfiblabel), levels = stromal_types)
xenium_scmetab$lv3_by_met <- factor(as.character(xenium_scmetab$lv3_by_met), levels = lv3_by_met_levels)
Seurat::Idents(xenium_scmetab) <- "lv3_anno"

detected_cores <- parallel::detectCores(logical = FALSE)
scmetab_ncores <- ifelse(is.na(detected_cores), 1, min(2, detected_cores))

xenium_scmetab <- scMetabolism::sc.metabolism.Seurat(
  obj = xenium_scmetab,
  method = "AUCell",
  imputation = FALSE,
  ncores = scmetab_ncores,
  metabolism.type = "KEGG"
)

scmetab_scores <- xenium_scmetab@assays$METABOLISM$score
scmetab_pathway_variance <- sort(
  apply(as.matrix(scmetab_scores), 1, stats::var, na.rm = TRUE),
  decreasing = TRUE
)

scmetab_preferred_pathways <- c(
  "Glycolysis / Gluconeogenesis",
  "Citrate cycle (TCA cycle)",
  "Oxidative phosphorylation",
  "Pentose phosphate pathway",
  "Pyruvate metabolism",
  "Fatty acid biosynthesis",
  "Fatty acid degradation",
  "Glutathione metabolism",
  "Purine metabolism",
  "Pyrimidine metabolism"
)

scmetab_dot_pathways <- unique(c(
  intersect(scmetab_preferred_pathways, rownames(scmetab_scores)),
  names(head(scmetab_pathway_variance, 20))
))
scmetab_dot_pathways <- scmetab_dot_pathways[seq_len(min(30, length(scmetab_dot_pathways)))]

scmetab_heatmap_pathways <- names(head(scmetab_pathway_variance, 30))
```

```{r scmetabolism_helpers}
summarize_scmetabolism <- function(score_matrix, meta_data, group_var, pathways = rownames(score_matrix)) {
  score_df <- as.data.frame(t(as.matrix(score_matrix)), check.names = FALSE)
  score_cells <- colnames(score_matrix)
  if (is.null(score_cells) || !all(score_cells %in% rownames(meta_data))) {
    score_cells <- rownames(meta_data)[seq_len(ncol(score_matrix))]
  }
  score_df$cell <- score_cells
  score_df$group <- as.character(meta_data[score_df$cell, group_var, drop = TRUE])

  tidyr::pivot_longer(
    score_df,
    cols = dplyr::all_of(pathways),
    names_to = "pathway",
    values_to = "score"
  ) %>%
    dplyr::filter(!is.na(group), group != "") %>%
    dplyr::group_by(group, pathway) %>%
    dplyr::summarise(
      mean_score = mean(score, na.rm = TRUE),
      median_score = median(score, na.rm = TRUE),
      n_cells = dplyr::n(),
      .groups = "drop"
    )
}

plot_scmetabolism_heatmap <- function(summary_df, group_levels, pathway_levels, plot_title) {
  summary_df %>%
    dplyr::filter(pathway %in% pathway_levels) %>%
    dplyr::group_by(pathway) %>%
    dplyr::mutate(
      mean_score_z = if (dplyr::n_distinct(mean_score, na.rm = TRUE) > 1) {
        as.numeric(base::scale(mean_score))
      } else {
        0
      }
    ) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(
      group = factor(group, levels = group_levels),
      pathway = factor(pathway, levels = rev(pathway_levels))
    ) %>%
    ggplot(aes(x = group, y = pathway, fill = mean_score_z)) +
    geom_tile(color = "white", linewidth = 0.2) +
    scale_fill_gradientn(
      colors = rev(RColorBrewer::brewer.pal(7, "RdBu")),
      name = "Row z-score",
      na.value = "grey90"
    ) +
    labs(title = plot_title, x = NULL, y = NULL) +
    theme_bw() +
    theme(
      axis.text.x = element_text(size = 9, angle = 45, hjust = 1),
      axis.text.y = element_text(size = 8),
      plot.title = element_text(hjust = 0.5)
    )
}

make_scmetabolism_cell_df <- function(score_matrix, meta_data, pathways) {
  score_matrix <- as.matrix(score_matrix)
  score_cells <- colnames(score_matrix)
  if (is.null(score_cells) || !all(score_cells %in% rownames(meta_data))) {
    score_cells <- rownames(meta_data)[seq_len(ncol(score_matrix))]
  }

  score_df <- as.data.frame(t(score_matrix[pathways, , drop = FALSE]), check.names = FALSE)
  score_df$cell <- score_cells
  score_df$lv3_anno <- meta_data[score_df$cell, "lv3_anno", drop = TRUE]
  score_df$metfiblabel <- meta_data[score_df$cell, "metfiblabel", drop = TRUE]
  score_df$lv3_by_met <- meta_data[score_df$cell, "lv3_by_met", drop = TRUE]

  score_df %>%
    tidyr::pivot_longer(
      cols = dplyr::all_of(pathways),
      names_to = "pathway",
      values_to = "score"
    ) %>%
    dplyr::mutate(
      lv3_anno = factor(as.character(lv3_anno), levels = lv3_levels),
      metfiblabel = factor(as.character(metfiblabel), levels = stromal_types),
      lv3_by_met = factor(as.character(lv3_by_met), levels = lv3_by_met_levels),
      pathway = factor(pathway, levels = pathways)
    )
}
```

## Xenium lv3_anno

```{r scmetabolism_lv3_plots, fig.width=12, fig.height=8}
p_scmetab_dot_lv3 <- scMetabolism::DotPlot.metabolism(
  obj = xenium_scmetab,
  pathway = scmetab_dot_pathways,
  phenotype = "lv3_anno",
  norm = "y"
) +
  theme(
    axis.text.x = element_text(size = 9, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8)
  )

p_scmetab_dot_lv3
```

```{r scmetabolism_lv3_heatmap, fig.width=8, fig.height=8}
scmetab_lv3_summary <- summarize_scmetabolism(
  score_matrix = scmetab_scores,
  meta_data = xenium_scmetab@meta.data,
  group_var = "lv3_anno",
  pathways = scmetab_heatmap_pathways
)

p_scmetab_heatmap_lv3 <- plot_scmetabolism_heatmap(
  summary_df = scmetab_lv3_summary,
  group_levels = lv3_levels,
  pathway_levels = scmetab_heatmap_pathways,
  plot_title = "scMetabolism KEGG activity across Xenium lv3_anno fibroblast clusters"
)

p_scmetab_heatmap_lv3
```

## Xenium lv3_anno split by IMC metfiblabel

```{r scmetabolism_lv3_by_met_plots, fig.width=14, fig.height=8}
p_scmetab_dot_lv3_by_met <- scMetabolism::DotPlot.metabolism(
  obj = xenium_scmetab,
  pathway = scmetab_dot_pathways,
  phenotype = "lv3_by_met",
  norm = "y"
) +
  theme(
    axis.text.x = element_text(size = 9, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8)
  )

p_scmetab_dot_lv3_by_met
```

```{r scmetabolism_lv3_by_met_heatmap, fig.width=11, fig.height=8}
scmetab_lv3_by_met_summary <- summarize_scmetabolism(
  score_matrix = scmetab_scores,
  meta_data = xenium_scmetab@meta.data,
  group_var = "lv3_by_met",
  pathways = scmetab_heatmap_pathways
)

p_scmetab_heatmap_lv3_by_met <- plot_scmetabolism_heatmap(
  summary_df = scmetab_lv3_by_met_summary,
  group_levels = lv3_by_met_levels,
  pathway_levels = scmetab_heatmap_pathways,
  plot_title = "scMetabolism KEGG activity across Xenium lv3_anno clusters split by IMC metfiblabel"
)

p_scmetab_heatmap_lv3_by_met
```

```{r scmetabolism_met_hi_delta_heatmap, fig.width=8, fig.height=8}
scmetab_met_delta <- scmetab_lv3_by_met_summary %>%
  dplyr::mutate(
    lv3_anno = sub("_(Other_Fib|Met_hi_Fib)$", "", group),
    metfiblabel = dplyr::case_when(
      grepl("_Other_Fib$", group) ~ "Other_Fib",
      grepl("_Met_hi_Fib$", group) ~ "Met_hi_Fib",
      TRUE ~ NA_character_
    )
  ) %>%
  dplyr::filter(!is.na(metfiblabel)) %>%
  dplyr::select(lv3_anno, metfiblabel, pathway, mean_score) %>%
  tidyr::pivot_wider(names_from = metfiblabel, values_from = mean_score) %>%
  dplyr::filter(!is.na(Met_hi_Fib), !is.na(Other_Fib)) %>%
  dplyr::mutate(delta_met_hi_vs_other = Met_hi_Fib - Other_Fib)

p_scmetab_met_hi_delta <- scmetab_met_delta %>%
  dplyr::filter(pathway %in% scmetab_heatmap_pathways) %>%
  dplyr::mutate(
    lv3_anno = factor(lv3_anno, levels = lv3_levels),
    pathway = factor(pathway, levels = rev(scmetab_heatmap_pathways))
  ) %>%
  ggplot(aes(x = lv3_anno, y = pathway, fill = delta_met_hi_vs_other)) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0,
    name = "Met_hi - Other"
  ) +
  labs(
    title = "Within-lv3 scMetabolism difference: Met_hi_Fib minus Other_Fib",
    x = NULL,
    y = NULL
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 9, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8),
    plot.title = element_text(hjust = 0.5)
  )

p_scmetab_met_hi_delta
```

## Glycolysis and TCA/OXPHOS focus

```{r scmetabolism_energy_pathways}
scmetab_energy_requested <- c(
  "Glycolysis / Gluconeogenesis",
  "Citrate cycle (TCA cycle)",
  "Oxidative phosphorylation"
)

scmetab_energy_pathways <- intersect(scmetab_energy_requested, rownames(scmetab_scores))
scmetab_energy_missing <- setdiff(scmetab_energy_requested, scmetab_energy_pathways)

if (length(scmetab_energy_pathways) == 0) {
  stop("None of the requested glycolysis/TCA/OXPHOS pathways are present in scMetabolism scores.")
}

scmetab_energy_availability <- data.frame(
  pathway = scmetab_energy_requested,
  plotted = scmetab_energy_requested %in% scmetab_energy_pathways,
  note = ifelse(
    scmetab_energy_requested %in% scmetab_energy_pathways,
    "available in scMetabolism KEGG scores",
    "not plotted; excluded by scMetabolism because this Xenium panel has insufficient pathway gene coverage"
  )
)

scmetab_energy_caption <- if (length(scmetab_energy_missing) > 0) {
  paste("Not plotted:", paste(scmetab_energy_missing, collapse = ", "),
        "was excluded by scMetabolism because this Xenium panel has insufficient pathway gene coverage.")
} else {
  NULL
}

scmetab_energy_scores <- make_scmetabolism_cell_df(
  score_matrix = scmetab_scores,
  meta_data = xenium_scmetab@meta.data,
  pathways = scmetab_energy_pathways
)

scmetab_energy_by_met_summary <- summarize_scmetabolism(
  score_matrix = scmetab_scores,
  meta_data = xenium_scmetab@meta.data,
  group_var = "lv3_by_met",
  pathways = scmetab_energy_pathways
)

scmetab_energy_delta <- scmetab_energy_by_met_summary %>%
  dplyr::mutate(
    lv3_anno = sub("_(Other_Fib|Met_hi_Fib)$", "", group),
    metfiblabel = dplyr::case_when(
      grepl("_Other_Fib$", group) ~ "Other_Fib",
      grepl("_Met_hi_Fib$", group) ~ "Met_hi_Fib",
      TRUE ~ NA_character_
    )
  ) %>%
  dplyr::filter(!is.na(metfiblabel)) %>%
  dplyr::select(lv3_anno, metfiblabel, pathway, mean_score) %>%
  tidyr::pivot_wider(names_from = metfiblabel, values_from = mean_score) %>%
  dplyr::filter(!is.na(Met_hi_Fib), !is.na(Other_Fib)) %>%
  dplyr::mutate(delta_met_hi_vs_other = Met_hi_Fib - Other_Fib)

scmetab_output_dir <- file.path(getwd(), "scmetabolism_outputs")
dir.create(scmetab_output_dir, showWarnings = FALSE, recursive = TRUE)

write.csv(
  scmetab_energy_availability,
  file.path(scmetab_output_dir, "scmetabolism_energy_pathway_availability.csv"),
  row.names = FALSE
)
write.csv(
  scmetab_energy_scores,
  file.path(scmetab_output_dir, "scmetabolism_energy_cell_scores.csv"),
  row.names = FALSE
)
write.csv(
  scmetab_energy_delta,
  file.path(scmetab_output_dir, "scmetabolism_energy_met_hi_delta.csv"),
  row.names = FALSE
)

scmetab_energy_availability
```

```{r scmetabolism_energy_lv3_violin, fig.width=11, fig.height=5}
p_scmetab_energy_lv3_violin <- ggplot(
  scmetab_energy_scores,
  aes(x = lv3_anno, y = score)
) +
  geom_violin(fill = "#B8D8D8", color = "grey35", scale = "width", trim = TRUE) +
  geom_boxplot(width = 0.15, outlier.shape = NA, fill = "white", color = "grey25") +
  facet_wrap(~ pathway, ncol = 1, scales = "free_y") +
  labs(
    title = "Focused scMetabolism scores across Xenium fibroblast lv3_anno clusters",
    subtitle = "Glycolysis and TCA/OXPHOS focus",
    caption = scmetab_energy_caption,
    x = NULL,
    y = "scMetabolism AUCell score"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 9, angle = 45, hjust = 1),
    strip.text = element_text(size = 11, face = "bold"),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_energy_lv3_violin.png"),
  p_scmetab_energy_lv3_violin,
  width = 11,
  height = 5,
  dpi = 300
)

p_scmetab_energy_lv3_violin
```

```{r scmetabolism_energy_lv3_by_met_violin, fig.width=12, fig.height=6}
p_scmetab_energy_lv3_by_met_violin <- ggplot(
  scmetab_energy_scores,
  aes(x = lv3_anno, y = score, fill = metfiblabel)
) +
  geom_violin(
    position = position_dodge(width = 0.8),
    color = "grey35",
    scale = "width",
    trim = TRUE
  ) +
  geom_boxplot(
    position = position_dodge(width = 0.8),
    width = 0.14,
    outlier.shape = NA,
    color = "grey25"
  ) +
  scale_fill_manual(values = c("Other_Fib" = "#4E79A7", "Met_hi_Fib" = "#E15759")) +
  facet_wrap(~ pathway, ncol = 1, scales = "free_y") +
  labs(
    title = "Focused scMetabolism scores split by IMC metfiblabel",
    subtitle = "Glycolysis and TCA/OXPHOS focus within Xenium lv3_anno clusters",
    caption = scmetab_energy_caption,
    x = NULL,
    y = "scMetabolism AUCell score",
    fill = NULL
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 9, angle = 45, hjust = 1),
    strip.text = element_text(size = 11, face = "bold"),
    legend.position = "top",
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_energy_lv3_by_met_violin.png"),
  p_scmetab_energy_lv3_by_met_violin,
  width = 12,
  height = 6,
  dpi = 300
)

p_scmetab_energy_lv3_by_met_violin
```

```{r scmetabolism_energy_delta_bar, fig.width=10, fig.height=5}
p_scmetab_energy_delta_bar <- scmetab_energy_delta %>%
  dplyr::mutate(
    lv3_anno = factor(lv3_anno, levels = lv3_levels),
    pathway = factor(pathway, levels = scmetab_energy_pathways)
  ) %>%
  ggplot(aes(x = lv3_anno, y = delta_met_hi_vs_other, fill = delta_met_hi_vs_other > 0)) +
  geom_hline(yintercept = 0, linewidth = 0.35, color = "grey35") +
  geom_col(width = 0.72) +
  scale_fill_manual(values = c("TRUE" = "#B2182B", "FALSE" = "#2166AC"), guide = "none") +
  facet_wrap(~ pathway, ncol = 1, scales = "free_y") +
  labs(
    title = "Focused scMetabolism difference by Xenium cluster",
    subtitle = "Met_hi_Fib minus Other_Fib",
    caption = scmetab_energy_caption,
    x = NULL,
    y = "Mean score difference"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 9, angle = 45, hjust = 1),
    strip.text = element_text(size = 11, face = "bold"),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_energy_met_hi_delta_bar.png"),
  p_scmetab_energy_delta_bar,
  width = 10,
  height = 5,
  dpi = 300
)

p_scmetab_energy_delta_bar
```

## Glycolysis and TCA by IMC metfiblabel only

```{r scmetabolism_metfiblabel_energy_data}
scmetab_metfiblabel_pathways <- intersect(
  c("Glycolysis / Gluconeogenesis", "Citrate cycle (TCA cycle)"),
  rownames(scmetab_scores)
)

if (length(scmetab_metfiblabel_pathways) == 0) {
  stop("Glycolysis and TCA pathways are not present in the scMetabolism scores.")
}

scmetab_metfiblabel_scores <- make_scmetabolism_cell_df(
  score_matrix = scmetab_scores,
  meta_data = xenium_scmetab@meta.data,
  pathways = scmetab_metfiblabel_pathways
) %>%
  dplyr::filter(!is.na(metfiblabel)) %>%
  dplyr::mutate(
    metfiblabel = factor(as.character(metfiblabel), levels = stromal_types),
    pathway = factor(as.character(pathway), levels = scmetab_metfiblabel_pathways)
  ) %>%
  dplyr::select(cell, metfiblabel, pathway, score)

scmetab_metfiblabel_summary <- scmetab_metfiblabel_scores %>%
  dplyr::group_by(pathway, metfiblabel) %>%
  dplyr::summarise(
    n_cells = dplyr::n(),
    mean_score = mean(score, na.rm = TRUE),
    median_score = median(score, na.rm = TRUE),
    q25 = stats::quantile(score, 0.25, na.rm = TRUE),
    q75 = stats::quantile(score, 0.75, na.rm = TRUE),
    sd_score = stats::sd(score, na.rm = TRUE),
    se_score = sd_score / sqrt(n_cells),
    ci95 = 1.96 * se_score,
    .groups = "drop"
  )

scmetab_metfiblabel_delta <- scmetab_metfiblabel_summary %>%
  dplyr::select(pathway, metfiblabel, mean_score, median_score) %>%
  tidyr::pivot_wider(names_from = metfiblabel, values_from = c(mean_score, median_score)) %>%
  dplyr::mutate(
    mean_delta_met_hi_vs_other = mean_score_Met_hi_Fib - mean_score_Other_Fib,
    median_delta_met_hi_vs_other = median_score_Met_hi_Fib - median_score_Other_Fib
  )

scmetab_output_dir <- file.path(getwd(), "scmetabolism_outputs")
dir.create(scmetab_output_dir, showWarnings = FALSE, recursive = TRUE)

write.csv(
  scmetab_metfiblabel_scores,
  file.path(scmetab_output_dir, "scmetabolism_metfiblabel_energy_cell_scores.csv"),
  row.names = FALSE
)
write.csv(
  scmetab_metfiblabel_summary,
  file.path(scmetab_output_dir, "scmetabolism_metfiblabel_energy_summary.csv"),
  row.names = FALSE
)
write.csv(
  scmetab_metfiblabel_delta,
  file.path(scmetab_output_dir, "scmetabolism_metfiblabel_energy_delta.csv"),
  row.names = FALSE
)

scmetab_metfiblabel_summary
```

```{r scmetabolism_metfiblabel_energy_violin_jitter, fig.width=8, fig.height=5}
p_scmetab_metfiblabel_violin_jitter <- ggplot(
  scmetab_metfiblabel_scores,
  aes(x = metfiblabel, y = score, fill = metfiblabel)
) +
  geom_violin(width = 0.82, alpha = 0.78, color = "grey30", trim = TRUE) +
  geom_jitter(
    aes(color = metfiblabel),
    width = 0.08,
    size = 0.7,
    alpha = 0.28,
    show.legend = FALSE
  ) +
  geom_boxplot(width = 0.16, outlier.shape = NA, fill = "white", color = "grey15") +
  scale_fill_manual(values = c("Other_Fib" = "#4E79A7", "Met_hi_Fib" = "#E15759")) +
  scale_color_manual(values = c("Other_Fib" = "#2F5F8F", "Met_hi_Fib" = "#B23B3F")) +
  facet_wrap(~ pathway, nrow = 1, scales = "free_y") +
  labs(
    title = "Glycolysis and TCA scMetabolism scores by IMC metfiblabel",
    subtitle = "Violin, boxplot, and cell-level jitter",
    x = NULL,
    y = "scMetabolism AUCell score",
    fill = NULL
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 11, angle = 25, hjust = 1),
    strip.text = element_text(size = 11, face = "bold"),
    legend.position = "top",
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_metfiblabel_energy_violin_jitter.png"),
  p_scmetab_metfiblabel_violin_jitter,
  width = 8,
  height = 5,
  dpi = 300
)

p_scmetab_metfiblabel_violin_jitter
```

```{r scmetabolism_metfiblabel_energy_density, fig.width=8, fig.height=4.8}
p_scmetab_metfiblabel_density <- ggplot(
  scmetab_metfiblabel_scores,
  aes(x = score, fill = metfiblabel, color = metfiblabel)
) +
  geom_density(alpha = 0.36, linewidth = 0.8, adjust = 1.1) +
  scale_fill_manual(values = c("Other_Fib" = "#4E79A7", "Met_hi_Fib" = "#E15759")) +
  scale_color_manual(values = c("Other_Fib" = "#2F5F8F", "Met_hi_Fib" = "#B23B3F")) +
  facet_wrap(~ pathway, nrow = 1, scales = "free") +
  labs(
    title = "Glycolysis and TCA score distributions",
    subtitle = "Density view by IMC metfiblabel",
    x = "scMetabolism AUCell score",
    y = "Density",
    fill = NULL,
    color = NULL
  ) +
  theme_bw() +
  theme(
    strip.text = element_text(size = 11, face = "bold"),
    legend.position = "top",
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_metfiblabel_energy_density.png"),
  p_scmetab_metfiblabel_density,
  width = 8,
  height = 4.8,
  dpi = 300
)

p_scmetab_metfiblabel_density
```

```{r scmetabolism_metfiblabel_energy_dumbbell, fig.width=8, fig.height=3.8}
scmetab_metfiblabel_dumbbell <- scmetab_metfiblabel_summary %>%
  dplyr::select(pathway, metfiblabel, mean_score) %>%
  tidyr::pivot_wider(names_from = metfiblabel, values_from = mean_score) %>%
  dplyr::mutate(pathway = factor(pathway, levels = rev(scmetab_metfiblabel_pathways)))

p_scmetab_metfiblabel_dumbbell <- ggplot(scmetab_metfiblabel_dumbbell) +
  geom_segment(
    aes(x = Other_Fib, xend = Met_hi_Fib, y = pathway, yend = pathway),
    color = "grey55",
    linewidth = 1.2
  ) +
  geom_point(aes(x = Other_Fib, y = pathway), color = "#4E79A7", size = 4) +
  geom_point(aes(x = Met_hi_Fib, y = pathway), color = "#E15759", size = 4) +
  labs(
    title = "Mean Glycolysis and TCA scores by IMC metfiblabel",
    subtitle = "Dumbbell view highlights direction and size of the group difference",
    x = "Mean scMetabolism AUCell score",
    y = NULL
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_text(size = 11, face = "bold"),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_metfiblabel_energy_dumbbell.png"),
  p_scmetab_metfiblabel_dumbbell,
  width = 8,
  height = 3.8,
  dpi = 300
)

p_scmetab_metfiblabel_dumbbell
```

```{r scmetabolism_metfiblabel_energy_mean_ci, fig.width=6.6, fig.height=3.8}
scmetab_metfiblabel_dot_labels <- c(
  "Glycolysis / Gluconeogenesis" = "Glycolysis /\nGluconeogenesis",
  "Citrate cycle (TCA cycle)" = "Citrate cycle\n(TCA cycle)"
)

make_energy_dotplot <- function(
  summary_df,
  group_col,
  group_levels,
  pathway_levels,
  pathway_labels,
  plot_title,
  scale_by_column = FALSE
) {
  plot_df <- summary_df %>%
    dplyr::mutate(
      pathway = factor(as.character(pathway), levels = pathway_levels),
      group_label = factor(as.character(.data[[group_col]]), levels = rev(group_levels))
    )

  if (isTRUE(scale_by_column)) {
    plot_df <- plot_df %>%
      dplyr::group_by(pathway) %>%
      dplyr::mutate(
        dot_min = min(mean_score, na.rm = TRUE),
        dot_max = max(mean_score, na.rm = TRUE),
        dot_score = dplyr::if_else(
          dot_max > dot_min,
          (mean_score - dot_min) / (dot_max - dot_min),
          0.5
        )
      ) %>%
      dplyr::ungroup() %>%
      dplyr::select(-dot_min, -dot_max)
    legend_title <- "Column-scaled mean"
  } else {
    plot_df <- plot_df %>% dplyr::mutate(dot_score = mean_score)
    legend_title <- "Mean AUCell score"
  }

  ggplot(plot_df, aes(x = pathway, y = group_label)) +
    geom_point(
      aes(fill = dot_score, size = dot_score),
      shape = 21,
      color = "grey35",
      stroke = 0.35
    ) +
    scale_x_discrete(labels = pathway_labels) +
    scale_fill_gradient(low = "white", high = "#D7301F", name = legend_title) +
    scale_size(range = c(8, 16), guide = "none") +
    labs(
      title = plot_title,
      x = NULL,
      y = NULL
    ) +
    theme_bw() +
    theme(
      axis.text.x = element_text(size = 10),
      axis.text.y = element_text(size = 11, face = "bold"),
      legend.position = "right",
      panel.grid.major = element_line(color = "grey88"),
      panel.grid.minor = element_blank(),
      plot.title = element_text(hjust = 0.5)
    )
}

p_scmetab_metfiblabel_dotplot_raw <- make_energy_dotplot(
  summary_df = scmetab_metfiblabel_summary,
  group_col = "metfiblabel",
  group_levels = stromal_types,
  pathway_levels = scmetab_metfiblabel_pathways,
  pathway_labels = scmetab_metfiblabel_dot_labels,
  plot_title = "Fibroblast mean Glycolysis and TCA scores",
  scale_by_column = FALSE
)

p_scmetab_metfiblabel_dotplot_scaled <- make_energy_dotplot(
  summary_df = scmetab_metfiblabel_summary,
  group_col = "metfiblabel",
  group_levels = stromal_types,
  pathway_levels = scmetab_metfiblabel_pathways,
  pathway_labels = scmetab_metfiblabel_dot_labels,
  plot_title = "Fibroblast mean Glycolysis and TCA scores - scaled by column",
  scale_by_column = TRUE
)

p_scmetab_metfiblabel_mean_ci <- p_scmetab_metfiblabel_dotplot_raw

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_metfiblabel_energy_mean_ci.png"),
  p_scmetab_metfiblabel_dotplot_raw,
  width = 7.5,
  height = 4.2,
  dpi = 300
)

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_metfiblabel_energy_dotplot.png"),
  p_scmetab_metfiblabel_dotplot_raw,
  width = 6.6,
  height = 3.8,
  dpi = 300
)

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_metfiblabel_energy_dotplot_raw.png"),
  p_scmetab_metfiblabel_dotplot_raw,
  width = 6.6,
  height = 3.8,
  dpi = 300
)

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_metfiblabel_energy_dotplot_scaled_by_column.png"),
  p_scmetab_metfiblabel_dotplot_scaled,
  width = 6.6,
  height = 3.8,
  dpi = 300
)

p_scmetab_metfiblabel_dotplot_raw
p_scmetab_metfiblabel_dotplot_scaled
```

```{r scmetabolism_metfiblabel_energy_heatmap, fig.width=5.5, fig.height=3.8}
p_scmetab_metfiblabel_heatmap <- scmetab_metfiblabel_summary %>%
  dplyr::group_by(pathway) %>%
  dplyr::mutate(mean_score_z = as.numeric(base::scale(mean_score))) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(
    pathway = factor(pathway, levels = rev(scmetab_metfiblabel_pathways)),
    metfiblabel = factor(as.character(metfiblabel), levels = stromal_types)
  ) %>%
  ggplot(aes(x = metfiblabel, y = pathway, fill = mean_score_z)) +
  geom_tile(color = "white", linewidth = 0.4) +
  geom_text(aes(label = sprintf("%.3f", mean_score)), size = 4) +
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0,
    name = "Row z-score"
  ) +
  labs(
    title = "Mean Glycolysis and TCA scores",
    subtitle = "Numbers show raw mean AUCell scores",
    x = NULL,
    y = NULL
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 11),
    axis.text.y = element_text(size = 10, face = "bold"),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_metfiblabel_energy_heatmap.png"),
  p_scmetab_metfiblabel_heatmap,
  width = 5.5,
  height = 3.8,
  dpi = 300
)

p_scmetab_metfiblabel_heatmap
```

## Glycolysis and TCA by IMC metfiblabel - single-cell distribution views

```{r scmetabolism_metfiblabel_energy_ecdf, fig.width=8, fig.height=4.8}
p_scmetab_metfiblabel_ecdf <- ggplot(
  scmetab_metfiblabel_scores,
  aes(x = score, color = metfiblabel)
) +
  stat_ecdf(linewidth = 1.15) +
  scale_color_manual(values = c("Other_Fib" = "#4E79A7", "Met_hi_Fib" = "#E15759")) +
  facet_wrap(~ pathway, nrow = 1, scales = "free_x") +
  labs(
    title = "Single-cell score distributions by IMC metfiblabel",
    subtitle = "ECDF view; right-shifted curves indicate higher scores",
    x = "scMetabolism AUCell score",
    y = "Cumulative fraction of cells",
    color = NULL
  ) +
  theme_bw() +
  theme(
    strip.text = element_text(size = 11, face = "bold"),
    legend.position = "top",
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_metfiblabel_energy_ecdf.png"),
  p_scmetab_metfiblabel_ecdf,
  width = 8,
  height = 4.8,
  dpi = 300
)

p_scmetab_metfiblabel_ecdf
```

```{r scmetabolism_metfiblabel_energy_quantile_shift, fig.width=8, fig.height=4.8}
scmetab_metfiblabel_quantile_shift <- scmetab_metfiblabel_scores %>%
  dplyr::group_by(pathway, metfiblabel) %>%
  dplyr::summarise(
    quantile = list(seq(0.05, 0.95, by = 0.05)),
    score_quantile = list(stats::quantile(score, probs = seq(0.05, 0.95, by = 0.05), na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  tidyr::unnest(c(quantile, score_quantile)) %>%
  tidyr::pivot_wider(names_from = metfiblabel, values_from = score_quantile) %>%
  dplyr::mutate(delta_met_hi_vs_other = Met_hi_Fib - Other_Fib)

write.csv(
  scmetab_metfiblabel_quantile_shift,
  file.path(scmetab_output_dir, "scmetabolism_metfiblabel_energy_quantile_shift.csv"),
  row.names = FALSE
)

p_scmetab_metfiblabel_quantile_shift <- ggplot(
  scmetab_metfiblabel_quantile_shift,
  aes(x = quantile, y = delta_met_hi_vs_other)
) +
  geom_hline(yintercept = 0, color = "grey35", linewidth = 0.4) +
  geom_area(aes(fill = delta_met_hi_vs_other > 0), alpha = 0.45) +
  geom_line(linewidth = 1, color = "grey20") +
  geom_point(size = 1.7, color = "grey20") +
  scale_fill_manual(values = c("TRUE" = "#B2182B", "FALSE" = "#2166AC"), guide = "none") +
  facet_wrap(~ pathway, nrow = 1, scales = "free_y") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Single-cell quantile shift: Met_hi_Fib minus Other_Fib",
    subtitle = "Shows where in the distribution the group difference is largest",
    x = "Distribution quantile",
    y = "Delta AUCell score"
  ) +
  theme_bw() +
  theme(
    strip.text = element_text(size = 11, face = "bold"),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_metfiblabel_energy_quantile_shift.png"),
  p_scmetab_metfiblabel_quantile_shift,
  width = 8,
  height = 4.8,
  dpi = 300
)

p_scmetab_metfiblabel_quantile_shift
```

```{r scmetabolism_metfiblabel_energy_threshold_enrichment, fig.width=8, fig.height=4.8}
scmetab_metfiblabel_threshold_enrichment <- scmetab_metfiblabel_scores %>%
  dplyr::group_by(pathway) %>%
  dplyr::summarise(
    threshold = list(stats::quantile(score, probs = seq(0.1, 0.9, by = 0.05), na.rm = TRUE)),
    threshold_quantile = list(seq(0.1, 0.9, by = 0.05)),
    .groups = "drop"
  ) %>%
  tidyr::unnest(c(threshold, threshold_quantile)) %>%
  dplyr::left_join(scmetab_metfiblabel_scores, by = "pathway") %>%
  dplyr::group_by(pathway, threshold_quantile, threshold, metfiblabel) %>%
  dplyr::summarise(frac_above = mean(score >= threshold, na.rm = TRUE), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = metfiblabel, values_from = frac_above) %>%
  dplyr::mutate(delta_frac_met_hi_vs_other = Met_hi_Fib - Other_Fib)

write.csv(
  scmetab_metfiblabel_threshold_enrichment,
  file.path(scmetab_output_dir, "scmetabolism_metfiblabel_energy_threshold_enrichment.csv"),
  row.names = FALSE
)

p_scmetab_metfiblabel_threshold_enrichment <- ggplot(
  scmetab_metfiblabel_threshold_enrichment,
  aes(x = threshold_quantile, y = delta_frac_met_hi_vs_other)
) +
  geom_hline(yintercept = 0, color = "grey35", linewidth = 0.4) +
  geom_col(aes(fill = delta_frac_met_hi_vs_other > 0), width = 0.035) +
  scale_fill_manual(values = c("TRUE" = "#B2182B", "FALSE" = "#2166AC"), guide = "none") +
  facet_wrap(~ pathway, nrow = 1, scales = "free_y") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "High-score enrichment by threshold",
    subtitle = "Met_hi_Fib fraction above threshold minus Other_Fib fraction",
    x = "Pooled score threshold quantile",
    y = "Delta fraction above threshold"
  ) +
  theme_bw() +
  theme(
    strip.text = element_text(size = 11, face = "bold"),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_metfiblabel_energy_threshold_enrichment.png"),
  p_scmetab_metfiblabel_threshold_enrichment,
  width = 8,
  height = 4.8,
  dpi = 300
)

p_scmetab_metfiblabel_threshold_enrichment
```

## IMC vs Xenium Glycolysis and TCA scores

```{r scmetabolism_imc_xenium_correlation_data}
score_cols <- c("glycolysis_score", "tca_score", "hif1a_score", "nox4_score")

format_p <- function(p) {
  dplyr::case_when(
    is.na(p) ~ "NA",
    p < 0.001 ~ "<0.001",
    TRUE ~ sprintf("%.3f", p)
  )
}

make_imc_score <- function(markers, assay_name = "zrescaled") {
  markers <- intersect(markers, rownames(spe))
  if (length(markers) == 0) {
    return(rep(NA_real_, ncol(spe)))
  }
  colMeans(as.matrix(SummarizedExperiment::assay(spe, assay_name)[markers, , drop = FALSE]), na.rm = TRUE)
}

if (!all(score_cols %in% colnames(SummarizedExperiment::colData(spe)))) {
  spe$glycolysis_score <- make_imc_score(c("GLUT1", "HK1", "PFKL1_PFKM", "PKM2", "LDHA", "GAPDH"))
  spe$tca_score <- make_imc_score(c("CS", "OGDH", "ATP5A", "SDHA"))
  spe$hif1a_score <- make_imc_score("HIF1a")
  spe$nox4_score <- make_imc_score("NOX4")
}

df_sample <- as.data.frame(SummarizedExperiment::colData(spe)[, c("sample_id", "metfiblabel", score_cols)])

df_sample <- df_sample[df_sample$metfiblabel %in% c("Met_hi_Fib", "Met_int_Fib", "Other_Fib"), ]

df_sample <- df_sample %>%
  dplyr::group_by(sample_id, metfiblabel) %>%
  dplyr::summarise(dplyr::across(dplyr::all_of(score_cols), mean, na.rm = TRUE), .groups = "drop")

df_cell <- as.data.frame(SummarizedExperiment::colData(spe)[, c("sample_id", "metfiblabel", score_cols)]) %>%
  dplyr::mutate(
    KEY = colnames(spe),
    sample_id = as.character(sample_id),
    metfiblabel = as.character(metfiblabel),
    .before = 1
  ) %>%
  dplyr::filter(
    metfiblabel %in% c("Met_hi_Fib", "Met_int_Fib", "Other_Fib"),
    !is.na(KEY),
    KEY != ""
  ) %>%
  dplyr::distinct(KEY, .keep_all = TRUE)

xenium_sample_candidates <- intersect(
  c("SampleID", "sample_id", "SampleId", "Patient", "ROI"),
  colnames(xenium_scmetab@meta.data)
)

xenium_sample_overlap <- vapply(
  xenium_sample_candidates,
  function(sample_col) {
    cur_values <- as.character(xenium_scmetab@meta.data[[sample_col]])
    cur_values <- cur_values[!is.na(cur_values) & cur_values != "" & cur_values != "not found"]
    length(intersect(unique(df_sample$sample_id), unique(cur_values)))
  },
  numeric(1)
)

xenium_correlation_sample_col <- names(which.max(xenium_sample_overlap))

if (length(xenium_correlation_sample_col) == 0 || is.na(xenium_correlation_sample_col)) {
  xenium_correlation_sample_col <- "sample_id"
}

xenium_cell_key_col <- if ("KEY" %in% colnames(xenium_scmetab@meta.data)) "KEY" else "cell"

xenium_correlation_pathways <- intersect(
  c("Glycolysis / Gluconeogenesis", "Citrate cycle (TCA cycle)", "Oxidative phosphorylation"),
  rownames(scmetab_scores)
)

xenium_correlation_cells <- make_scmetabolism_cell_df(
  score_matrix = scmetab_scores,
  meta_data = xenium_scmetab@meta.data,
  pathways = xenium_correlation_pathways
) %>%
  dplyr::mutate(
    sample_id = as.character(xenium_scmetab@meta.data[cell, xenium_correlation_sample_col, drop = TRUE]),
    metfiblabel = as.character(metfiblabel),
    xenium_cell_key = if (xenium_cell_key_col %in% colnames(xenium_scmetab@meta.data)) {
      as.character(xenium_scmetab@meta.data[cell, xenium_cell_key_col, drop = TRUE])
    } else {
      as.character(cell)
    },
    xenium_score_name = dplyr::case_when(
      pathway == "Glycolysis / Gluconeogenesis" ~ "xenium_glycolysis_score",
      pathway == "Citrate cycle (TCA cycle)" ~ "xenium_tca_score",
      pathway == "Oxidative phosphorylation" ~ "xenium_oxphos_score",
      TRUE ~ NA_character_
    )
  ) %>%
  dplyr::filter(
    !is.na(xenium_score_name),
    !is.na(sample_id),
    sample_id != "",
    sample_id != "not found",
    !is.na(xenium_cell_key),
    xenium_cell_key != "",
    !is.na(metfiblabel)
  )

xenium_sample_scores <- xenium_correlation_cells %>%
  dplyr::group_by(sample_id, metfiblabel, xenium_score_name) %>%
  dplyr::summarise(xenium_score = mean(score, na.rm = TRUE), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = xenium_score_name, values_from = xenium_score)

scmetab_imc_xenium_joined <- df_sample %>%
  dplyr::inner_join(xenium_sample_scores, by = c("sample_id", "metfiblabel"))

scmetab_imc_xenium_cor_long <- dplyr::bind_rows(
  scmetab_imc_xenium_joined %>%
    dplyr::transmute(
      sample_id,
      metfiblabel,
      comparison = "Glycolysis",
      imc_score = glycolysis_score,
      xenium_score = xenium_glycolysis_score
    ),
  scmetab_imc_xenium_joined %>%
    dplyr::transmute(
      sample_id,
      metfiblabel,
      comparison = "TCA",
      imc_score = tca_score,
      xenium_score = xenium_tca_score
    )
) %>%
  dplyr::filter(!is.na(imc_score), !is.na(xenium_score)) %>%
  dplyr::mutate(comparison = factor(comparison, levels = c("Glycolysis", "TCA")))

xenium_cell_scores <- xenium_correlation_cells %>%
  dplyr::select(
    xenium_cell_key,
    cell,
    sample_id,
    metfiblabel,
    xenium_score_name,
    score
  ) %>%
  dplyr::distinct(xenium_cell_key, xenium_score_name, .keep_all = TRUE) %>%
  tidyr::pivot_wider(names_from = xenium_score_name, values_from = score)

scmetab_imc_xenium_cell_joined <- df_cell %>%
  dplyr::inner_join(xenium_cell_scores, by = c("KEY" = "xenium_cell_key"), suffix = c("_imc", "_xenium")) %>%
  dplyr::mutate(
    sample_id = dplyr::coalesce(sample_id_xenium, sample_id_imc),
    metfiblabel = dplyr::coalesce(metfiblabel_xenium, metfiblabel_imc),
    metfiblabel_match = metfiblabel_imc == metfiblabel_xenium
  ) %>%
  dplyr::filter(!is.na(metfiblabel), metfiblabel %in% stromal_types)

cell_correlation_specs <- data.frame(
  comparison = c("Glycolysis", "TCA", "OXPHOS"),
  imc_col = c("glycolysis_score", "tca_score", "tca_score"),
  xenium_col = c("xenium_glycolysis_score", "xenium_tca_score", "xenium_oxphos_score"),
  stringsAsFactors = FALSE
) %>%
  dplyr::filter(imc_col %in% colnames(scmetab_imc_xenium_cell_joined), xenium_col %in% colnames(scmetab_imc_xenium_cell_joined))

if (nrow(cell_correlation_specs) > 0) {
  scmetab_imc_xenium_cell_cor_long <- dplyr::bind_rows(
    lapply(seq_len(nrow(cell_correlation_specs)), function(i) {
      cur_spec <- cell_correlation_specs[i, ]
      scmetab_imc_xenium_cell_joined %>%
        dplyr::transmute(
          KEY,
          cell,
          sample_id,
          metfiblabel,
          metfiblabel_match,
          comparison = cur_spec$comparison,
          imc_score = .data[[cur_spec$imc_col]],
          xenium_score = .data[[cur_spec$xenium_col]]
        )
    })
  ) %>%
    dplyr::filter(!is.na(imc_score), !is.na(xenium_score)) %>%
    dplyr::mutate(
      comparison = factor(comparison, levels = cell_correlation_specs$comparison),
      metfiblabel = factor(as.character(metfiblabel), levels = stromal_types)
    )
} else {
  scmetab_imc_xenium_cell_cor_long <- tibble::tibble(
    KEY = character(),
    cell = character(),
    sample_id = character(),
    metfiblabel = factor(character(), levels = stromal_types),
    metfiblabel_match = logical(),
    comparison = factor(character(), levels = character()),
    imc_score = numeric(),
    xenium_score = numeric()
  )
}

cor_test_one <- function(x, y, method = "spearman") {
  if (length(stats::complete.cases(x, y)) < 3 || sum(stats::complete.cases(x, y)) < 3) {
    return(c(estimate = NA_real_, p.value = NA_real_))
  }
  res <- suppressWarnings(stats::cor.test(x, y, method = method, exact = FALSE))
  c(estimate = unname(res$estimate), p.value = res$p.value)
}

scmetab_imc_xenium_cor_stats <- scmetab_imc_xenium_cor_long %>%
  dplyr::group_by(comparison) %>%
  dplyr::summarise(
    n = dplyr::n(),
    spearman_rho = cor_test_one(imc_score, xenium_score, "spearman")["estimate"],
    spearman_p = cor_test_one(imc_score, xenium_score, "spearman")["p.value"],
    pearson_r = cor_test_one(imc_score, xenium_score, "pearson")["estimate"],
    pearson_p = cor_test_one(imc_score, xenium_score, "pearson")["p.value"],
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    label = paste0(
      "Spearman rho=", sprintf("%.2f", spearman_rho),
      "\np=", format_p(spearman_p),
      "\nn=", n
    )
  )

scmetab_imc_xenium_cell_cor_stats_overall <- scmetab_imc_xenium_cell_cor_long %>%
  dplyr::group_by(comparison) %>%
  dplyr::summarise(
    stratum = "All matched cells",
    n = dplyr::n(),
    n_samples = dplyr::n_distinct(sample_id),
    spearman_rho = cor_test_one(imc_score, xenium_score, "spearman")["estimate"],
    spearman_p = cor_test_one(imc_score, xenium_score, "spearman")["p.value"],
    pearson_r = cor_test_one(imc_score, xenium_score, "pearson")["estimate"],
    pearson_p = cor_test_one(imc_score, xenium_score, "pearson")["p.value"],
    .groups = "drop"
  )

scmetab_imc_xenium_cell_cor_stats_by_label <- scmetab_imc_xenium_cell_cor_long %>%
  dplyr::group_by(comparison, metfiblabel) %>%
  dplyr::summarise(
    stratum = as.character(dplyr::first(metfiblabel)),
    n = dplyr::n(),
    n_samples = dplyr::n_distinct(sample_id),
    spearman_rho = cor_test_one(imc_score, xenium_score, "spearman")["estimate"],
    spearman_p = cor_test_one(imc_score, xenium_score, "spearman")["p.value"],
    pearson_r = cor_test_one(imc_score, xenium_score, "pearson")["estimate"],
    pearson_p = cor_test_one(imc_score, xenium_score, "pearson")["p.value"],
    .groups = "drop"
  )

scmetab_imc_xenium_cell_cor_stats <- dplyr::bind_rows(
  scmetab_imc_xenium_cell_cor_stats_overall,
  scmetab_imc_xenium_cell_cor_stats_by_label
) %>%
  dplyr::mutate(
    label = paste0(
      "rho=", sprintf("%.2f", spearman_rho),
      "\np=", format_p(spearman_p),
      "\nn=", n
    )
  )

scmetab_imc_xenium_availability <- data.frame(
  xenium_pathway = c("Glycolysis / Gluconeogenesis", "Citrate cycle (TCA cycle)", "Oxidative phosphorylation"),
  available_in_xenium_scmetabolism = c(
    "Glycolysis / Gluconeogenesis",
    "Citrate cycle (TCA cycle)",
    "Oxidative phosphorylation"
  ) %in% rownames(scmetab_scores)
)

scmetab_imc_xenium_key_overlap <- data.frame(
  xenium_sample_column = names(xenium_sample_overlap),
  n_matching_imc_sample_ids = as.integer(xenium_sample_overlap),
  selected = names(xenium_sample_overlap) == xenium_correlation_sample_col
)

scmetab_imc_xenium_cell_key_overlap <- data.frame(
  xenium_cell_key_column = xenium_cell_key_col,
  n_imc_cells_with_key = dplyr::n_distinct(df_cell$KEY),
  n_xenium_cells_with_key = dplyr::n_distinct(xenium_cell_scores$xenium_cell_key),
  n_matched_cells = dplyr::n_distinct(scmetab_imc_xenium_cell_joined$KEY),
  n_matched_cell_pathway_pairs = nrow(scmetab_imc_xenium_cell_cor_long)
)

write.csv(df_sample, file.path(scmetab_output_dir, "imc_metfiblabel_sample_scores.csv"), row.names = FALSE)
write.csv(df_cell, file.path(scmetab_output_dir, "imc_metfiblabel_cell_scores.csv"), row.names = FALSE)
write.csv(
  scmetab_imc_xenium_joined,
  file.path(scmetab_output_dir, "scmetabolism_imc_xenium_sample_joined.csv"),
  row.names = FALSE
)
write.csv(
  scmetab_imc_xenium_cell_joined,
  file.path(scmetab_output_dir, "scmetabolism_imc_xenium_cell_joined.csv"),
  row.names = FALSE
)
write.csv(
  scmetab_imc_xenium_cor_long,
  file.path(scmetab_output_dir, "scmetabolism_imc_xenium_correlation_long.csv"),
  row.names = FALSE
)
write.csv(
  scmetab_imc_xenium_cell_cor_long,
  file.path(scmetab_output_dir, "scmetabolism_imc_xenium_cell_correlation_long.csv"),
  row.names = FALSE
)
write.csv(
  scmetab_imc_xenium_cor_stats,
  file.path(scmetab_output_dir, "scmetabolism_imc_xenium_correlation_stats.csv"),
  row.names = FALSE
)
write.csv(
  scmetab_imc_xenium_cell_cor_stats,
  file.path(scmetab_output_dir, "scmetabolism_imc_xenium_cell_correlation_stats.csv"),
  row.names = FALSE
)
write.csv(
  scmetab_imc_xenium_availability,
  file.path(scmetab_output_dir, "scmetabolism_imc_xenium_pathway_availability.csv"),
  row.names = FALSE
)
write.csv(
  scmetab_imc_xenium_key_overlap,
  file.path(scmetab_output_dir, "scmetabolism_imc_xenium_sample_key_overlap.csv"),
  row.names = FALSE
)
write.csv(
  scmetab_imc_xenium_cell_key_overlap,
  file.path(scmetab_output_dir, "scmetabolism_imc_xenium_cell_key_overlap.csv"),
  row.names = FALSE
)

dplyr::bind_rows(
  scmetab_imc_xenium_cor_stats %>% dplyr::mutate(level = "sample"),
  scmetab_imc_xenium_cell_cor_stats_overall %>% dplyr::mutate(level = "single_cell")
)
```

```{r scmetabolism_imc_xenium_correlation_plot, fig.width=8.5, fig.height=4.8}
if (nrow(scmetab_imc_xenium_cor_long) > 0) {
  p_scmetab_imc_xenium_correlation <- ggplot(
    scmetab_imc_xenium_cor_long,
    aes(x = imc_score, y = xenium_score, color = metfiblabel)
  ) +
    geom_point(size = 3, alpha = 0.9) +
    geom_smooth(method = "lm", se = FALSE, color = "grey35", linewidth = 0.7) +
    geom_text(
      data = scmetab_imc_xenium_cor_stats,
      aes(x = -Inf, y = Inf, label = label),
      inherit.aes = FALSE,
      hjust = -0.05,
      vjust = 1.08,
      size = 3.3
    ) +
    scale_color_manual(values = c("Other_Fib" = "#4E79A7", "Met_hi_Fib" = "#E15759", "Met_int_Fib" = "#59A14F")) +
    facet_wrap(~ comparison, nrow = 1, scales = "free") +
    labs(
      title = "IMC marker-score vs Xenium scMetabolism score",
      subtitle = paste0("Aggregated by IMC sample_id and Xenium ", xenium_correlation_sample_col, " + metfiblabel"),
      x = "IMC mean score",
      y = "Xenium mean scMetabolism AUCell score",
      color = NULL
    ) +
    theme_bw() +
    theme(
      strip.text = element_text(size = 11, face = "bold"),
      legend.position = "top",
      plot.title = element_text(hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5)
    )
} else {
  p_scmetab_imc_xenium_correlation <- ggplot() +
    annotate(
      "text",
      x = 0,
      y = 0,
      label = paste0(
        "No overlapping IMC/Xenium sample_id + metfiblabel groups found.\n",
        "Selected Xenium sample column: ", xenium_correlation_sample_col
      ),
      size = 5
    ) +
    theme_void() +
    labs(title = "IMC marker-score vs Xenium scMetabolism score")
}

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_imc_xenium_correlation.png"),
  p_scmetab_imc_xenium_correlation,
  width = 8.5,
  height = 4.8,
  dpi = 300
)

p_scmetab_imc_xenium_correlation
```

```{r scmetabolism_imc_xenium_cell_correlation_density_plot, fig.width=8.5, fig.height=6.2}
cell_correlation_plot_stats <- scmetab_imc_xenium_cell_cor_stats %>%
  dplyr::filter(!is.na(metfiblabel)) %>%
  dplyr::mutate(
    metfiblabel = factor(as.character(metfiblabel), levels = stromal_types),
    comparison = factor(as.character(comparison), levels = levels(scmetab_imc_xenium_cell_cor_long$comparison))
  )

if (nrow(scmetab_imc_xenium_cell_cor_long) > 0) {
  p_scmetab_imc_xenium_cell_density <- ggplot(
    scmetab_imc_xenium_cell_cor_long,
    aes(x = imc_score, y = xenium_score)
  ) +
    geom_bin2d(bins = 45) +
    geom_smooth(method = "lm", se = FALSE, color = "grey15", linewidth = 0.55) +
    geom_text(
      data = cell_correlation_plot_stats,
      aes(x = -Inf, y = Inf, label = label),
      inherit.aes = FALSE,
      hjust = -0.05,
      vjust = 1.08,
      size = 3.1
    ) +
    scale_fill_gradient(low = "grey94", high = "#2C7FB8", trans = "sqrt", name = "Cells") +
    facet_grid(metfiblabel ~ comparison, scales = "free") +
    labs(
      title = "Single-cell matched IMC marker score vs Xenium scMetabolism score",
      subtitle = paste0("Joined by ", xenium_cell_key_col, "; binned cells, statistics from all matched cells"),
      x = "IMC single-cell marker score",
      y = "Xenium single-cell scMetabolism AUCell score"
    ) +
    theme_bw() +
    theme(
      strip.text = element_text(size = 10, face = "bold"),
      legend.position = "right",
      plot.title = element_text(hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5)
    )
} else {
  p_scmetab_imc_xenium_cell_density <- ggplot() +
    annotate(
      "text",
      x = 0,
      y = 0,
      label = paste0(
        "No overlapping IMC/Xenium cells found.\n",
        "Selected Xenium cell key column: ", xenium_cell_key_col
      ),
      size = 5
    ) +
    theme_void() +
    labs(title = "Single-cell matched IMC marker score vs Xenium scMetabolism score")
}

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_imc_xenium_cell_correlation_density.png"),
  p_scmetab_imc_xenium_cell_density,
  width = 8.5,
  height = 6.2,
  dpi = 300
)

p_scmetab_imc_xenium_cell_density
```

```{r scmetabolism_imc_xenium_cell_correlation_point_plot, fig.width=8.5, fig.height=4.8}
cell_point_limit <- 40000

if (nrow(scmetab_imc_xenium_cell_cor_long) > 0) {
  set.seed(1)
  cell_point_data <- scmetab_imc_xenium_cell_cor_long %>%
    dplyr::group_by(comparison, metfiblabel) %>%
    dplyr::group_modify(function(.x, .y) {
      if (nrow(.x) > cell_point_limit) {
        dplyr::slice_sample(.x, n = cell_point_limit)
      } else {
        .x
      }
    }) %>%
    dplyr::ungroup()

  cell_point_subtitle <- if (nrow(cell_point_data) < nrow(scmetab_imc_xenium_cell_cor_long)) {
    paste0("Up to ", cell_point_limit, " cells per facet plotted; statistics use all matched cells")
  } else {
    "All matched cells plotted and used for statistics"
  }

  p_scmetab_imc_xenium_cell_points <- ggplot(
    cell_point_data,
    aes(x = imc_score, y = xenium_score, color = metfiblabel)
  ) +
    geom_point(alpha = 0.12, size = 0.35, stroke = 0) +
    geom_smooth(
      data = scmetab_imc_xenium_cell_cor_long,
      method = "lm",
      se = FALSE,
      linewidth = 0.65,
      alpha = 0.8
    ) +
    geom_text(
      data = scmetab_imc_xenium_cell_cor_stats_overall %>%
        dplyr::mutate(
          label = paste0(
            "rho=", sprintf("%.2f", spearman_rho),
            "\np=", format_p(spearman_p),
            "\nn=", n
          ),
          comparison = factor(as.character(comparison), levels = levels(scmetab_imc_xenium_cell_cor_long$comparison))
        ),
      aes(x = -Inf, y = Inf, label = label),
      inherit.aes = FALSE,
      hjust = -0.05,
      vjust = 1.08,
      size = 3.2
    ) +
    scale_color_manual(values = c("Other_Fib" = "#4E79A7", "Met_hi_Fib" = "#E15759", "Met_int_Fib" = "#59A14F")) +
    facet_wrap(~ comparison, nrow = 1, scales = "free") +
    labs(
      title = "Single-cell IMC-Xenium correlation point cloud",
      subtitle = cell_point_subtitle,
      x = "IMC single-cell marker score",
      y = "Xenium single-cell scMetabolism AUCell score",
      color = NULL
    ) +
    theme_bw() +
    theme(
      strip.text = element_text(size = 11, face = "bold"),
      legend.position = "top",
      plot.title = element_text(hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5)
    )
} else {
  p_scmetab_imc_xenium_cell_points <- ggplot() +
    annotate(
      "text",
      x = 0,
      y = 0,
      label = paste0(
        "No overlapping IMC/Xenium cells found.\n",
        "Selected Xenium cell key column: ", xenium_cell_key_col
      ),
      size = 5
    ) +
    theme_void() +
    labs(title = "Single-cell IMC-Xenium correlation point cloud")
}

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_imc_xenium_cell_correlation_points.png"),
  p_scmetab_imc_xenium_cell_points,
  width = 8.5,
  height = 4.8,
  dpi = 300
)

p_scmetab_imc_xenium_cell_points
```

## Glycolysis and TCA by IMC metfiblabel - sample-level paired views (not run)

```{r scmetabolism_metfiblabel_energy_sample_stats, eval=FALSE}
scmetab_sample_col <- intersect(
  c("sample_id", "SampleID", "SampleId", "Patient", "ROI"),
  colnames(xenium_scmetab@meta.data)
)[1]

if (is.na(scmetab_sample_col)) {
  stop("No sample-level column was found in xenium_scmetab@meta.data.")
}

scmetab_score_cells <- colnames(scmetab_scores)
if (is.null(scmetab_score_cells) || !all(scmetab_score_cells %in% rownames(xenium_scmetab@meta.data))) {
  scmetab_score_cells <- rownames(xenium_scmetab@meta.data)[seq_len(ncol(scmetab_scores))]
}

scmetab_metfiblabel_sample_cell_scores <- as.data.frame(
  t(as.matrix(scmetab_scores[scmetab_metfiblabel_pathways, , drop = FALSE])),
  check.names = FALSE
) %>%
  dplyr::mutate(
    cell = scmetab_score_cells,
    sample_id = as.character(xenium_scmetab@meta.data[cell, scmetab_sample_col, drop = TRUE]),
    metfiblabel = factor(
      as.character(xenium_scmetab@meta.data[cell, "metfiblabel", drop = TRUE]),
      levels = stromal_types
    )
  ) %>%
  tidyr::pivot_longer(
    cols = dplyr::all_of(scmetab_metfiblabel_pathways),
    names_to = "pathway",
    values_to = "score"
  ) %>%
  dplyr::filter(!is.na(sample_id), sample_id != "", !is.na(metfiblabel)) %>%
  dplyr::mutate(pathway = factor(pathway, levels = scmetab_metfiblabel_pathways))

scmetab_metfiblabel_sample_summary <- scmetab_metfiblabel_sample_cell_scores %>%
  dplyr::group_by(sample_id, pathway, metfiblabel) %>%
  dplyr::summarise(
    n_cells = dplyr::n(),
    mean_score = mean(score, na.rm = TRUE),
    median_score = median(score, na.rm = TRUE),
    .groups = "drop"
  )

scmetab_metfiblabel_sample_paired <- scmetab_metfiblabel_sample_summary %>%
  dplyr::select(sample_id, pathway, metfiblabel, mean_score, median_score, n_cells) %>%
  tidyr::pivot_wider(
    names_from = metfiblabel,
    values_from = c(mean_score, median_score, n_cells)
  ) %>%
  dplyr::filter(!is.na(mean_score_Other_Fib), !is.na(mean_score_Met_hi_Fib)) %>%
  dplyr::mutate(
    mean_delta_met_hi_vs_other = mean_score_Met_hi_Fib - mean_score_Other_Fib,
    median_delta_met_hi_vs_other = median_score_Met_hi_Fib - median_score_Other_Fib,
    pathway = factor(as.character(pathway), levels = scmetab_metfiblabel_pathways)
  )

format_p <- function(p) {
  dplyr::case_when(
    is.na(p) ~ "NA",
    p < 0.001 ~ "<0.001",
    TRUE ~ sprintf("%.3f", p)
  )
}

paired_wilcox_p <- function(df) {
  if (nrow(df) < 2) return(NA_real_)
  stats::wilcox.test(
    df$mean_score_Met_hi_Fib,
    df$mean_score_Other_Fib,
    paired = TRUE,
    exact = FALSE
  )$p.value
}

single_cell_wilcox_p <- function(df) {
  if (dplyr::n_distinct(df$metfiblabel) < 2) return(NA_real_)
  stats::wilcox.test(score ~ metfiblabel, data = df, exact = FALSE)$p.value
}

scmetab_metfiblabel_sample_stats <- scmetab_metfiblabel_sample_paired %>%
  dplyr::group_by(pathway) %>%
  dplyr::summarise(
    n_paired_samples = dplyr::n(),
    paired_wilcox_p = paired_wilcox_p(dplyr::cur_data()),
    mean_delta_met_hi_vs_other = mean(mean_delta_met_hi_vs_other, na.rm = TRUE),
    median_delta_met_hi_vs_other = median(mean_delta_met_hi_vs_other, na.rm = TRUE),
    .groups = "drop"
  )

scmetab_metfiblabel_single_cell_stats <- scmetab_metfiblabel_sample_cell_scores %>%
  dplyr::group_by(pathway) %>%
  dplyr::summarise(
    n_other_cells = sum(metfiblabel == "Other_Fib"),
    n_met_hi_cells = sum(metfiblabel == "Met_hi_Fib"),
    single_cell_wilcox_p = single_cell_wilcox_p(dplyr::cur_data()),
    .groups = "drop"
  )

scmetab_metfiblabel_test_summary <- scmetab_metfiblabel_sample_stats %>%
  dplyr::left_join(scmetab_metfiblabel_single_cell_stats, by = "pathway") %>%
  dplyr::mutate(
    paired_wilcox_label = paste0("paired Wilcoxon p=", format_p(paired_wilcox_p)),
    single_cell_wilcox_label = paste0("single-cell Wilcoxon p=", format_p(single_cell_wilcox_p))
  )

scmetab_metfiblabel_paired_long <- scmetab_metfiblabel_sample_paired %>%
  dplyr::select(sample_id, pathway, Other_Fib = mean_score_Other_Fib, Met_hi_Fib = mean_score_Met_hi_Fib) %>%
  tidyr::pivot_longer(
    cols = c("Other_Fib", "Met_hi_Fib"),
    names_to = "metfiblabel",
    values_to = "mean_score"
  ) %>%
  dplyr::mutate(metfiblabel = factor(metfiblabel, levels = stromal_types))

write.csv(
  scmetab_metfiblabel_sample_cell_scores,
  file.path(scmetab_output_dir, "scmetabolism_metfiblabel_energy_sample_cell_scores.csv"),
  row.names = FALSE
)
write.csv(
  scmetab_metfiblabel_sample_summary,
  file.path(scmetab_output_dir, "scmetabolism_metfiblabel_energy_sample_summary.csv"),
  row.names = FALSE
)
write.csv(
  scmetab_metfiblabel_sample_paired,
  file.path(scmetab_output_dir, "scmetabolism_metfiblabel_energy_sample_paired.csv"),
  row.names = FALSE
)
write.csv(
  scmetab_metfiblabel_test_summary,
  file.path(scmetab_output_dir, "scmetabolism_metfiblabel_energy_tests.csv"),
  row.names = FALSE
)

scmetab_metfiblabel_test_summary
```

```{r scmetabolism_metfiblabel_energy_sample_paired_lines, fig.width=7.5, fig.height=5, eval=FALSE}
p_scmetab_metfiblabel_sample_paired_lines <- ggplot(
  scmetab_metfiblabel_paired_long,
  aes(x = metfiblabel, y = mean_score, group = sample_id)
) +
  geom_line(color = "grey55", alpha = 0.65, linewidth = 0.6) +
  geom_point(aes(color = metfiblabel), size = 2.6, alpha = 0.9) +
  geom_text(
    data = scmetab_metfiblabel_test_summary,
    aes(x = 1.5, y = Inf, label = paired_wilcox_label),
    inherit.aes = FALSE,
    vjust = 1.4,
    size = 3.3
  ) +
  scale_color_manual(values = c("Other_Fib" = "#4E79A7", "Met_hi_Fib" = "#E15759")) +
  facet_wrap(~ pathway, nrow = 1, scales = "free_y") +
  labs(
    title = "Sample-level paired Glycolysis and TCA scores",
    subtitle = paste0("One dot per sample per metfiblabel; sample column: ", scmetab_sample_col),
    x = NULL,
    y = "Sample mean scMetabolism AUCell score",
    color = NULL
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 11, angle = 20, hjust = 1),
    strip.text = element_text(size = 11, face = "bold"),
    legend.position = "top",
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_metfiblabel_energy_sample_paired_lines.png"),
  p_scmetab_metfiblabel_sample_paired_lines,
  width = 7.5,
  height = 5,
  dpi = 300
)

p_scmetab_metfiblabel_sample_paired_lines
```

```{r scmetabolism_metfiblabel_energy_sample_delta, fig.width=7.5, fig.height=4.6, eval=FALSE}
p_scmetab_metfiblabel_sample_delta <- ggplot(
  scmetab_metfiblabel_sample_paired,
  aes(x = pathway, y = mean_delta_met_hi_vs_other)
) +
  geom_hline(yintercept = 0, color = "grey35", linewidth = 0.4) +
  geom_point(
    aes(color = mean_delta_met_hi_vs_other > 0),
    position = position_jitter(width = 0.08, height = 0),
    size = 3,
    alpha = 0.9
  ) +
  stat_summary(
    fun = median,
    geom = "crossbar",
    width = 0.42,
    color = "black",
    linewidth = 0.35
  ) +
  geom_text(
    data = scmetab_metfiblabel_test_summary,
    aes(x = pathway, y = Inf, label = paired_wilcox_label),
    inherit.aes = FALSE,
    vjust = 1.4,
    size = 3.3
  ) +
  scale_color_manual(values = c("TRUE" = "#B2182B", "FALSE" = "#2166AC"), guide = "none") +
  labs(
    title = "Sample-level paired differences",
    subtitle = "Each dot is one sample: Met_hi_Fib mean minus Other_Fib mean",
    x = NULL,
    y = "Delta scMetabolism AUCell score"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(size = 10, angle = 20, hjust = 1),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_metfiblabel_energy_sample_delta.png"),
  p_scmetab_metfiblabel_sample_delta,
  width = 7.5,
  height = 4.6,
  dpi = 300
)

p_scmetab_metfiblabel_sample_delta
```

```{r scmetabolism_metfiblabel_energy_sample_dumbbell, fig.width=7.5, fig.height=4.5, eval=FALSE}
p_scmetab_metfiblabel_sample_dumbbell <- ggplot(scmetab_metfiblabel_sample_paired) +
  geom_segment(
    aes(
      x = mean_score_Other_Fib,
      xend = mean_score_Met_hi_Fib,
      y = sample_id,
      yend = sample_id
    ),
    color = "grey65",
    linewidth = 0.55,
    alpha = 0.75
  ) +
  geom_point(aes(x = mean_score_Other_Fib, y = sample_id), color = "#4E79A7", size = 2.2) +
  geom_point(aes(x = mean_score_Met_hi_Fib, y = sample_id), color = "#E15759", size = 2.2) +
  geom_text(
    data = scmetab_metfiblabel_test_summary,
    aes(x = Inf, y = Inf, label = paired_wilcox_label),
    inherit.aes = FALSE,
    hjust = 1.05,
    vjust = 1.3,
    size = 3.2
  ) +
  facet_wrap(~ pathway, nrow = 1, scales = "free_x") +
  labs(
    title = "Sample-level dumbbell view",
    subtitle = "One connected pair per sample",
    x = "Sample mean scMetabolism AUCell score",
    y = "Sample"
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_text(size = 8),
    strip.text = element_text(size = 11, face = "bold"),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_metfiblabel_energy_sample_dumbbell.png"),
  p_scmetab_metfiblabel_sample_dumbbell,
  width = 7.5,
  height = 4.5,
  dpi = 300
)

p_scmetab_metfiblabel_sample_dumbbell
```

# scMetabolism on Xenium endothelial clusters {.tabset .tabset-pill}

```{r scmetabolism_ec_prepare_sfe}
if (!exists("spe")) {
  load(project_path("Revision/Nan annotation/Data/IMC spe objects/New IMC/spe2025-12-17 umap.RData"))
}

if (!exists("sfe")) {
  sfe <- readRDS(project_path("Revision/Nan annotation/Data/Xenium sfe objects/Xenium sfe object/sfe_09032026.rds"))
}

if (!"Rphenograph_200EC" %in% colnames(SummarizedExperiment::colData(spe))) {
  stop("The IMC spe object must contain colData(spe)$Rphenograph_200EC for the EC metabolic labels.")
}

if (!"KEY" %in% colnames(SummarizedExperiment::colData(sfe))) {
  stop("The Xenium sfe object must contain colData(sfe)$KEY to transfer metEClabel from the IMC spe.")
}

if (!"lv2_anno_new" %in% colnames(SummarizedExperiment::colData(sfe))) {
  stop("The Xenium sfe object must contain lv2_anno_new for the EC cluster selection.")
}

spe$metEClabel <- dplyr::case_when(
  spe$Rphenograph_200EC %in% c("1", "2") ~ "Met_hi_EC",
  spe$Rphenograph_200EC == "3" ~ "Met_int_EC",
  TRUE ~ NA_character_
)

ec_map <- data.frame(
  KEY = colnames(spe),
  metEClabel = spe$metEClabel,
  stringsAsFactors = FALSE
)

ec_idx <- match(SummarizedExperiment::colData(sfe)$KEY, ec_map$KEY)
sfe$metEClabel <- ec_map$metEClabel[ec_idx]

ec_met_levels <- c("Met_int_EC", "Met_hi_EC")
ec_met_levels_plot <- c("Met_hi_EC", "Met_int_EC")
ec_lv2_preferred_order <- c(
  "ACKR1_EC",
  "ACTA2_EC",
  "cycling_EC",
  "EPC",
  "HEY1_EC",
  "lymphatic_EC"
)

keep_ec_scmetab <- !is.na(sfe$lv2_anno_new) &
  sfe$lv2_anno_new %in% ec_lv2_preferred_order &
  !is.na(sfe$metEClabel) &
  sfe$metEClabel %in% ec_met_levels

sfe_ec_scmetab <- sfe[, keep_ec_scmetab]

ec_lv2_levels <- c(
  intersect(ec_lv2_preferred_order, unique(as.character(sfe_ec_scmetab$lv2_anno_new))),
  setdiff(sort(unique(as.character(sfe_ec_scmetab$lv2_anno_new))), ec_lv2_preferred_order)
)

sfe_ec_scmetab$lv2_anno_new <- factor(as.character(sfe_ec_scmetab$lv2_anno_new), levels = ec_lv2_levels)
sfe_ec_scmetab$metEClabel <- factor(as.character(sfe_ec_scmetab$metEClabel), levels = ec_met_levels)
sfe_ec_scmetab$lv2_by_met <- paste(sfe_ec_scmetab$lv2_anno_new, sfe_ec_scmetab$metEClabel, sep = "_")

ec_lv2_by_met_levels <- unlist(lapply(ec_lv2_levels, function(cur_lv2) {
  paste(cur_lv2, ec_met_levels_plot, sep = "_")
}), use.names = FALSE)
ec_lv2_by_met_levels <- intersect(ec_lv2_by_met_levels, unique(as.character(sfe_ec_scmetab$lv2_by_met)))
sfe_ec_scmetab$lv2_by_met <- factor(sfe_ec_scmetab$lv2_by_met, levels = ec_lv2_by_met_levels)

table(sfe_ec_scmetab$lv2_anno_new, sfe_ec_scmetab$metEClabel)
```

```{r scmetabolism_ec_run, message=FALSE, warning=FALSE}
ec_counts_mat <- to_dgC(SummarizedExperiment::assay(sfe_ec_scmetab, "counts"))
ec_scmetab_meta <- as.data.frame(SummarizedExperiment::colData(sfe_ec_scmetab))
rownames(ec_scmetab_meta) <- colnames(sfe_ec_scmetab)

old_assay_version_ec <- getOption("Seurat.object.assay.version")
options(Seurat.object.assay.version = "v3")

ec_xenium_scmetab <- Seurat::CreateSeuratObject(
  counts = ec_counts_mat,
  meta.data = ec_scmetab_meta,
  assay = "RNA",
  project = "Xenium_EC_scMetabolism"
)

if (is.null(old_assay_version_ec)) {
  options(Seurat.object.assay.version = NULL)
} else {
  options(Seurat.object.assay.version = old_assay_version_ec)
}

ec_xenium_scmetab <- Seurat::NormalizeData(ec_xenium_scmetab, verbose = FALSE)
ec_xenium_scmetab$lv2_anno_new <- factor(as.character(ec_xenium_scmetab$lv2_anno_new), levels = ec_lv2_levels)
ec_xenium_scmetab$metEClabel <- factor(as.character(ec_xenium_scmetab$metEClabel), levels = ec_met_levels)
ec_xenium_scmetab$lv2_by_met <- factor(as.character(ec_xenium_scmetab$lv2_by_met), levels = ec_lv2_by_met_levels)
Seurat::Idents(ec_xenium_scmetab) <- "lv2_anno_new"

ec_detected_cores <- parallel::detectCores(logical = FALSE)
ec_scmetab_ncores <- ifelse(is.na(ec_detected_cores), 1, min(2, ec_detected_cores))

ec_xenium_scmetab <- scMetabolism::sc.metabolism.Seurat(
  obj = ec_xenium_scmetab,
  method = "AUCell",
  imputation = FALSE,
  ncores = ec_scmetab_ncores,
  metabolism.type = "KEGG"
)

ec_scmetab_scores <- ec_xenium_scmetab@assays$METABOLISM$score
ec_scmetab_pathway_variance <- sort(
  apply(as.matrix(ec_scmetab_scores), 1, stats::var, na.rm = TRUE),
  decreasing = TRUE
)

ec_scmetab_preferred_pathways <- c(
  "Glycolysis / Gluconeogenesis",
  "Citrate cycle (TCA cycle)",
  "Oxidative phosphorylation",
  "Pentose phosphate pathway",
  "Pyruvate metabolism",
  "Fatty acid biosynthesis",
  "Fatty acid degradation",
  "Glutathione metabolism",
  "Purine metabolism",
  "Pyrimidine metabolism"
)

ec_scmetab_dot_pathways <- unique(c(
  intersect(ec_scmetab_preferred_pathways, rownames(ec_scmetab_scores)),
  names(head(ec_scmetab_pathway_variance, 20))
))
ec_scmetab_dot_pathways <- ec_scmetab_dot_pathways[seq_len(min(30, length(ec_scmetab_dot_pathways)))]
ec_scmetab_heatmap_pathways <- names(head(ec_scmetab_pathway_variance, 30))

scmetab_output_dir <- file.path(getwd(), "scmetabolism_outputs")
dir.create(scmetab_output_dir, showWarnings = FALSE, recursive = TRUE)
```

```{r scmetabolism_ec_helpers}
make_scmetabolism_ec_cell_df <- function(score_matrix, meta_data, pathways) {
  score_matrix <- as.matrix(score_matrix)
  score_cells <- colnames(score_matrix)
  if (is.null(score_cells) || !all(score_cells %in% rownames(meta_data))) {
    score_cells <- rownames(meta_data)[seq_len(ncol(score_matrix))]
  }

  score_df <- as.data.frame(t(score_matrix[pathways, , drop = FALSE]), check.names = FALSE)
  score_df$cell <- score_cells
  score_df$lv2_anno_new <- meta_data[score_df$cell, "lv2_anno_new", drop = TRUE]
  score_df$metEClabel <- meta_data[score_df$cell, "metEClabel", drop = TRUE]
  score_df$lv2_by_met <- meta_data[score_df$cell, "lv2_by_met", drop = TRUE]

  score_df %>%
    tidyr::pivot_longer(
      cols = dplyr::all_of(pathways),
      names_to = "pathway",
      values_to = "score"
    ) %>%
    dplyr::mutate(
      lv2_anno_new = factor(as.character(lv2_anno_new), levels = ec_lv2_levels),
      metEClabel = factor(as.character(metEClabel), levels = ec_met_levels),
      lv2_by_met = factor(as.character(lv2_by_met), levels = ec_lv2_by_met_levels),
      pathway = factor(pathway, levels = pathways)
    )
}

summarize_ec_scmetabolism <- function(score_matrix, meta_data, group_var, pathways = rownames(score_matrix)) {
  score_df <- as.data.frame(t(as.matrix(score_matrix)), check.names = FALSE)
  score_cells <- colnames(score_matrix)
  if (is.null(score_cells) || !all(score_cells %in% rownames(meta_data))) {
    score_cells <- rownames(meta_data)[seq_len(ncol(score_matrix))]
  }
  score_df$cell <- score_cells
  score_df$group <- as.character(meta_data[score_df$cell, group_var, drop = TRUE])

  tidyr::pivot_longer(
    score_df,
    cols = dplyr::all_of(pathways),
    names_to = "pathway",
    values_to = "score"
  ) %>%
    dplyr::filter(!is.na(group), group != "") %>%
    dplyr::group_by(group, pathway) %>%
    dplyr::summarise(
      mean_score = mean(score, na.rm = TRUE),
      median_score = median(score, na.rm = TRUE),
      n_cells = dplyr::n(),
      .groups = "drop"
    )
}
```

## EC lv2_anno_new

```{r scmetabolism_ec_lv2_plots, fig.width=12, fig.height=8}
p_scmetab_ec_dot_lv2 <- scMetabolism::DotPlot.metabolism(
  obj = ec_xenium_scmetab,
  pathway = ec_scmetab_dot_pathways,
  phenotype = "lv2_anno_new",
  norm = "y"
) +
  theme(
    axis.text.x = element_text(size = 9, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8)
  )

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_ec_lv2_dotplot.png"),
  p_scmetab_ec_dot_lv2,
  width = 12,
  height = 8,
  dpi = 300
)

p_scmetab_ec_dot_lv2
```

```{r scmetabolism_ec_lv2_heatmap, fig.width=8.5, fig.height=7.5}
ec_scmetab_summary_lv2 <- summarize_ec_scmetabolism(
  ec_scmetab_scores,
  ec_xenium_scmetab@meta.data,
  "lv2_anno_new",
  ec_scmetab_heatmap_pathways
)

p_scmetab_ec_heatmap_lv2 <- plot_scmetabolism_heatmap(
  ec_scmetab_summary_lv2,
  ec_lv2_levels,
  ec_scmetab_heatmap_pathways,
  "Xenium EC scMetabolism by lv2_anno_new"
)

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_ec_lv2_heatmap.png"),
  p_scmetab_ec_heatmap_lv2,
  width = 8.5,
  height = 7.5,
  dpi = 300
)

p_scmetab_ec_heatmap_lv2
```

## EC lv2_by_met

```{r scmetabolism_ec_lv2_by_met_plots, fig.width=14, fig.height=8}
Seurat::Idents(ec_xenium_scmetab) <- "lv2_by_met"

p_scmetab_ec_dot_lv2_by_met <- scMetabolism::DotPlot.metabolism(
  obj = ec_xenium_scmetab,
  pathway = ec_scmetab_dot_pathways,
  phenotype = "lv2_by_met",
  norm = "y"
) +
  theme(
    axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8)
  )

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_ec_lv2_by_met_dotplot.png"),
  p_scmetab_ec_dot_lv2_by_met,
  width = 14,
  height = 8,
  dpi = 300
)

p_scmetab_ec_dot_lv2_by_met
```

```{r scmetabolism_ec_lv2_by_met_heatmap, fig.width=11, fig.height=7.5}
ec_scmetab_summary_lv2_by_met <- summarize_ec_scmetabolism(
  ec_scmetab_scores,
  ec_xenium_scmetab@meta.data,
  "lv2_by_met",
  ec_scmetab_heatmap_pathways
)

p_scmetab_ec_heatmap_lv2_by_met <- plot_scmetabolism_heatmap(
  ec_scmetab_summary_lv2_by_met,
  ec_lv2_by_met_levels,
  ec_scmetab_heatmap_pathways,
  "Xenium EC scMetabolism by lv2_by_met"
)

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_ec_lv2_by_met_heatmap.png"),
  p_scmetab_ec_heatmap_lv2_by_met,
  width = 11,
  height = 7.5,
  dpi = 300
)

p_scmetab_ec_heatmap_lv2_by_met
```

## EC Glycolysis and TCA by metEClabel

```{r scmetabolism_ec_energy_metlabel_data}
ec_energy_pathways <- intersect(
  c("Glycolysis / Gluconeogenesis", "Citrate cycle (TCA cycle)", "Oxidative phosphorylation"),
  rownames(ec_scmetab_scores)
)

if (length(ec_energy_pathways) == 0) {
  stop("Glycolysis, TCA, and OXPHOS pathways are not present in the EC scMetabolism scores.")
}

ec_energy_display_labels <- c(
  "Glycolysis / Gluconeogenesis" = "Glycolysis / Gluconeogenesis",
  "Citrate cycle (TCA cycle)" = "TCA cycle",
  "Oxidative phosphorylation" = "OXPHOS"
)

ec_scmetab_metlabel_scores <- make_scmetabolism_ec_cell_df(
  score_matrix = ec_scmetab_scores,
  meta_data = ec_xenium_scmetab@meta.data,
  pathways = ec_energy_pathways
) %>%
  dplyr::filter(!is.na(metEClabel)) %>%
  dplyr::mutate(
    metEClabel = factor(as.character(metEClabel), levels = ec_met_levels),
    pathway = factor(as.character(pathway), levels = ec_energy_pathways)
  ) %>%
  dplyr::select(cell, metEClabel, pathway, score)

ec_scmetab_metlabel_summary <- ec_scmetab_metlabel_scores %>%
  dplyr::group_by(pathway, metEClabel) %>%
  dplyr::summarise(
    n_cells = dplyr::n(),
    mean_score = mean(score, na.rm = TRUE),
    median_score = median(score, na.rm = TRUE),
    q25 = stats::quantile(score, 0.25, na.rm = TRUE),
    q75 = stats::quantile(score, 0.75, na.rm = TRUE),
    sd_score = stats::sd(score, na.rm = TRUE),
    se_score = sd_score / sqrt(n_cells),
    ci95 = 1.96 * se_score,
    .groups = "drop"
  )

ec_scmetab_metlabel_delta <- ec_scmetab_metlabel_summary %>%
  dplyr::filter(pathway %in% c("Glycolysis / Gluconeogenesis", "Citrate cycle (TCA cycle)")) %>%
  dplyr::select(pathway, metEClabel, mean_score, median_score) %>%
  tidyr::pivot_wider(names_from = metEClabel, values_from = c(mean_score, median_score)) %>%
  dplyr::mutate(
    mean_delta_met_hi_vs_int = mean_score_Met_hi_EC - mean_score_Met_int_EC,
    median_delta_met_hi_vs_int = median_score_Met_hi_EC - median_score_Met_int_EC
  )

ec_pathway_availability <- data.frame(
  xenium_pathway = c("Glycolysis / Gluconeogenesis", "Citrate cycle (TCA cycle)", "Oxidative phosphorylation"),
  available_in_ec_xenium_scmetabolism = c(
    "Glycolysis / Gluconeogenesis",
    "Citrate cycle (TCA cycle)",
    "Oxidative phosphorylation"
  ) %in% rownames(ec_scmetab_scores)
)

write.csv(
  ec_scmetab_metlabel_scores,
  file.path(scmetab_output_dir, "scmetabolism_ec_metEClabel_energy_cell_scores.csv"),
  row.names = FALSE
)
write.csv(
  ec_scmetab_metlabel_summary,
  file.path(scmetab_output_dir, "scmetabolism_ec_metEClabel_energy_summary.csv"),
  row.names = FALSE
)
write.csv(
  ec_scmetab_metlabel_delta,
  file.path(scmetab_output_dir, "scmetabolism_ec_metEClabel_energy_delta.csv"),
  row.names = FALSE
)
write.csv(
  ec_pathway_availability,
  file.path(scmetab_output_dir, "scmetabolism_ec_pathway_availability.csv"),
  row.names = FALSE
)

ec_scmetab_metlabel_summary
```

```{r scmetabolism_ec_energy_dotplot, fig.width=7.2, fig.height=3.8}
ec_dot_labels <- c(
  "Glycolysis / Gluconeogenesis" = "Glycolysis /\nGluconeogenesis",
  "Citrate cycle (TCA cycle)" = "Citrate cycle\n(TCA cycle)",
  "Oxidative phosphorylation" = "OXPHOS"
)

p_scmetab_ec_metlabel_dotplot_raw <- make_energy_dotplot(
  summary_df = ec_scmetab_metlabel_summary,
  group_col = "metEClabel",
  group_levels = ec_met_levels,
  pathway_levels = ec_energy_pathways,
  pathway_labels = ec_dot_labels,
  plot_title = "EC mean Glycolysis and TCA scores",
  scale_by_column = FALSE
)

p_scmetab_ec_metlabel_dotplot_scaled <- make_energy_dotplot(
  summary_df = ec_scmetab_metlabel_summary,
  group_col = "metEClabel",
  group_levels = ec_met_levels,
  pathway_levels = ec_energy_pathways,
  pathway_labels = ec_dot_labels,
  plot_title = "EC mean Glycolysis and TCA scores - scaled by column",
  scale_by_column = TRUE
)

p_scmetab_ec_metlabel_dotplot <- p_scmetab_ec_metlabel_dotplot_raw

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_ec_metEClabel_energy_dotplot.png"),
  p_scmetab_ec_metlabel_dotplot_raw,
  width = 7.2,
  height = 3.8,
  dpi = 300
)

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_ec_metEClabel_energy_dotplot_raw.png"),
  p_scmetab_ec_metlabel_dotplot_raw,
  width = 7.2,
  height = 3.8,
  dpi = 300
)

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_ec_metEClabel_energy_dotplot_scaled_by_column.png"),
  p_scmetab_ec_metlabel_dotplot_scaled,
  width = 7.2,
  height = 3.8,
  dpi = 300
)

p_scmetab_ec_metlabel_dotplot_raw
p_scmetab_ec_metlabel_dotplot_scaled
```

```{r scmetabolism_ec_energy_distribution_plots, fig.width=8.2, fig.height=4.8}
p_scmetab_ec_metlabel_density <- ggplot(
  ec_scmetab_metlabel_scores,
  aes(x = score, fill = metEClabel, color = metEClabel)
) +
  geom_density(alpha = 0.36, linewidth = 0.8, adjust = 1.1) +
  scale_fill_manual(values = c("Met_int_EC" = "#4E79A7", "Met_hi_EC" = "#E15759")) +
  scale_color_manual(values = c("Met_int_EC" = "#2F5F8F", "Met_hi_EC" = "#B23B3F")) +
  facet_wrap(~ pathway, nrow = 1, scales = "free", labeller = labeller(pathway = ec_energy_display_labels)) +
  labs(
    title = "EC Glycolysis and TCA score distributions",
    x = "scMetabolism AUCell score",
    y = "Density",
    fill = NULL,
    color = NULL
  ) +
  theme_bw() +
  theme(
    strip.text = element_text(size = 11, face = "bold"),
    legend.position = "top",
    plot.title = element_text(hjust = 0.5)
  )

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_ec_metEClabel_energy_density.png"),
  p_scmetab_ec_metlabel_density,
  width = 8.2,
  height = 4.8,
  dpi = 300
)

p_scmetab_ec_metlabel_density
```

```{r scmetabolism_ec_energy_dumbbell, fig.width=8, fig.height=3.8}
ec_scmetab_metlabel_dumbbell <- ec_scmetab_metlabel_summary %>%
  dplyr::filter(pathway %in% c("Glycolysis / Gluconeogenesis", "Citrate cycle (TCA cycle)")) %>%
  dplyr::select(pathway, metEClabel, mean_score) %>%
  tidyr::pivot_wider(names_from = metEClabel, values_from = mean_score) %>%
  dplyr::mutate(pathway = factor(pathway, levels = rev(c("Glycolysis / Gluconeogenesis", "Citrate cycle (TCA cycle)"))))

p_scmetab_ec_metlabel_dumbbell <- ggplot(ec_scmetab_metlabel_dumbbell) +
  geom_segment(
    aes(x = Met_int_EC, xend = Met_hi_EC, y = pathway, yend = pathway),
    color = "grey55",
    linewidth = 1.2
  ) +
  geom_point(aes(x = Met_int_EC, y = pathway), color = "#4E79A7", size = 4) +
  geom_point(aes(x = Met_hi_EC, y = pathway), color = "#E15759", size = 4) +
  scale_y_discrete(labels = ec_energy_display_labels) +
  labs(
    title = "EC mean Glycolysis and TCA scores by metEClabel",
    subtitle = "Dumbbell view highlights Met_hi_EC versus Met_int_EC",
    x = "Mean scMetabolism AUCell score",
    y = NULL
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_text(size = 11, face = "bold"),
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_ec_metEClabel_energy_dumbbell.png"),
  p_scmetab_ec_metlabel_dumbbell,
  width = 8,
  height = 3.8,
  dpi = 300
)

p_scmetab_ec_metlabel_dumbbell
```

## EC IMC vs Xenium Glycolysis and TCA scores

```{r scmetabolism_ec_imc_xenium_correlation_data}
ec_score_cols <- c("glycolysis_score", "tca_score", "hif1a_score", "nox4_score")

ec_format_p <- function(p) {
  dplyr::case_when(
    is.na(p) ~ "NA",
    p < 0.001 ~ "<0.001",
    TRUE ~ sprintf("%.3f", p)
  )
}

ec_make_imc_score <- function(markers, assay_name = "zrescaled") {
  markers <- intersect(markers, rownames(spe))
  if (length(markers) == 0) {
    return(rep(NA_real_, ncol(spe)))
  }
  colMeans(as.matrix(SummarizedExperiment::assay(spe, assay_name)[markers, , drop = FALSE]), na.rm = TRUE)
}

if (!all(ec_score_cols %in% colnames(SummarizedExperiment::colData(spe)))) {
  spe$glycolysis_score <- ec_make_imc_score(c("GLUT1", "HK1", "PFKL1_PFKM", "PKM2", "LDHA", "GAPDH"))
  spe$tca_score <- ec_make_imc_score(c("CS", "OGDH", "ATP5A", "SDHA"))
  spe$hif1a_score <- ec_make_imc_score("HIF1a")
  spe$nox4_score <- ec_make_imc_score("NOX4")
}

ec_df_sample <- as.data.frame(SummarizedExperiment::colData(spe)[, c("sample_id", "metEClabel", ec_score_cols)]) %>%
  dplyr::filter(metEClabel %in% ec_met_levels) %>%
  dplyr::group_by(sample_id, metEClabel) %>%
  dplyr::summarise(dplyr::across(dplyr::all_of(ec_score_cols), mean, na.rm = TRUE), .groups = "drop")

ec_df_cell <- as.data.frame(SummarizedExperiment::colData(spe)[, c("sample_id", "metEClabel", ec_score_cols)]) %>%
  dplyr::mutate(
    KEY = colnames(spe),
    sample_id = as.character(sample_id),
    metEClabel = as.character(metEClabel),
    .before = 1
  ) %>%
  dplyr::filter(
    metEClabel %in% ec_met_levels,
    !is.na(KEY),
    KEY != ""
  ) %>%
  dplyr::distinct(KEY, .keep_all = TRUE)

ec_xenium_sample_candidates <- intersect(
  c("SampleID", "sample_id", "SampleId", "Patient", "ROI"),
  colnames(ec_xenium_scmetab@meta.data)
)

ec_xenium_sample_overlap <- vapply(
  ec_xenium_sample_candidates,
  function(sample_col) {
    cur_values <- as.character(ec_xenium_scmetab@meta.data[[sample_col]])
    cur_values <- cur_values[!is.na(cur_values) & cur_values != "" & cur_values != "not found"]
    length(intersect(unique(ec_df_sample$sample_id), unique(cur_values)))
  },
  numeric(1)
)

ec_xenium_correlation_sample_col <- names(which.max(ec_xenium_sample_overlap))

if (length(ec_xenium_correlation_sample_col) == 0 || is.na(ec_xenium_correlation_sample_col)) {
  ec_xenium_correlation_sample_col <- "sample_id"
}

ec_xenium_cell_key_col <- if ("KEY" %in% colnames(ec_xenium_scmetab@meta.data)) "KEY" else "cell"

ec_correlation_pathways <- intersect(
  c("Glycolysis / Gluconeogenesis", "Citrate cycle (TCA cycle)"),
  rownames(ec_scmetab_scores)
)

ec_xenium_correlation_cells <- make_scmetabolism_ec_cell_df(
  score_matrix = ec_scmetab_scores,
  meta_data = ec_xenium_scmetab@meta.data,
  pathways = ec_correlation_pathways
) %>%
  dplyr::mutate(
    sample_id = as.character(ec_xenium_scmetab@meta.data[cell, ec_xenium_correlation_sample_col, drop = TRUE]),
    metEClabel = as.character(metEClabel),
    xenium_cell_key = if (ec_xenium_cell_key_col %in% colnames(ec_xenium_scmetab@meta.data)) {
      as.character(ec_xenium_scmetab@meta.data[cell, ec_xenium_cell_key_col, drop = TRUE])
    } else {
      as.character(cell)
    },
    xenium_score_name = dplyr::case_when(
      pathway == "Glycolysis / Gluconeogenesis" ~ "xenium_glycolysis_score",
      pathway == "Citrate cycle (TCA cycle)" ~ "xenium_tca_score",
      TRUE ~ NA_character_
    )
  ) %>%
  dplyr::filter(
    !is.na(xenium_score_name),
    !is.na(sample_id),
    sample_id != "",
    sample_id != "not found",
    !is.na(xenium_cell_key),
    xenium_cell_key != "",
    !is.na(metEClabel)
  )

ec_xenium_sample_scores <- ec_xenium_correlation_cells %>%
  dplyr::group_by(sample_id, metEClabel, xenium_score_name) %>%
  dplyr::summarise(xenium_score = mean(score, na.rm = TRUE), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = xenium_score_name, values_from = xenium_score)

ec_scmetab_imc_xenium_joined <- ec_df_sample %>%
  dplyr::inner_join(ec_xenium_sample_scores, by = c("sample_id", "metEClabel"))

ec_scmetab_imc_xenium_cor_long <- dplyr::bind_rows(
  ec_scmetab_imc_xenium_joined %>%
    dplyr::transmute(
      sample_id,
      metEClabel,
      comparison = "Glycolysis",
      imc_score = glycolysis_score,
      xenium_score = xenium_glycolysis_score
    ),
  ec_scmetab_imc_xenium_joined %>%
    dplyr::transmute(
      sample_id,
      metEClabel,
      comparison = "TCA",
      imc_score = tca_score,
      xenium_score = xenium_tca_score
    )
) %>%
  dplyr::filter(!is.na(imc_score), !is.na(xenium_score)) %>%
  dplyr::mutate(
    comparison = factor(comparison, levels = c("Glycolysis", "TCA")),
    metEClabel = factor(as.character(metEClabel), levels = ec_met_levels)
  )

ec_xenium_cell_scores <- ec_xenium_correlation_cells %>%
  dplyr::select(
    xenium_cell_key,
    cell,
    sample_id,
    metEClabel,
    xenium_score_name,
    score
  ) %>%
  dplyr::distinct(xenium_cell_key, xenium_score_name, .keep_all = TRUE) %>%
  tidyr::pivot_wider(names_from = xenium_score_name, values_from = score)

ec_scmetab_imc_xenium_cell_joined <- ec_df_cell %>%
  dplyr::inner_join(ec_xenium_cell_scores, by = c("KEY" = "xenium_cell_key"), suffix = c("_imc", "_xenium")) %>%
  dplyr::mutate(
    sample_id = dplyr::coalesce(sample_id_xenium, sample_id_imc),
    metEClabel = dplyr::coalesce(metEClabel_xenium, metEClabel_imc),
    metEClabel_match = metEClabel_imc == metEClabel_xenium
  ) %>%
  dplyr::filter(!is.na(metEClabel), metEClabel %in% ec_met_levels)

ec_cell_correlation_specs <- data.frame(
  comparison = c("Glycolysis", "TCA"),
  imc_col = c("glycolysis_score", "tca_score"),
  xenium_col = c("xenium_glycolysis_score", "xenium_tca_score"),
  stringsAsFactors = FALSE
) %>%
  dplyr::filter(imc_col %in% colnames(ec_scmetab_imc_xenium_cell_joined), xenium_col %in% colnames(ec_scmetab_imc_xenium_cell_joined))

if (nrow(ec_cell_correlation_specs) > 0) {
  ec_scmetab_imc_xenium_cell_cor_long <- dplyr::bind_rows(
    lapply(seq_len(nrow(ec_cell_correlation_specs)), function(i) {
      cur_spec <- ec_cell_correlation_specs[i, ]
      ec_scmetab_imc_xenium_cell_joined %>%
        dplyr::transmute(
          KEY,
          cell,
          sample_id,
          metEClabel,
          metEClabel_match,
          comparison = cur_spec$comparison,
          imc_score = .data[[cur_spec$imc_col]],
          xenium_score = .data[[cur_spec$xenium_col]]
        )
    })
  ) %>%
    dplyr::filter(!is.na(imc_score), !is.na(xenium_score)) %>%
    dplyr::mutate(
      comparison = factor(comparison, levels = ec_cell_correlation_specs$comparison),
      metEClabel = factor(as.character(metEClabel), levels = ec_met_levels)
    )
} else {
  ec_scmetab_imc_xenium_cell_cor_long <- tibble::tibble(
    KEY = character(),
    cell = character(),
    sample_id = character(),
    metEClabel = factor(character(), levels = ec_met_levels),
    metEClabel_match = logical(),
    comparison = factor(character(), levels = character()),
    imc_score = numeric(),
    xenium_score = numeric()
  )
}

ec_cor_test_one <- function(x, y, method = "spearman") {
  if (length(stats::complete.cases(x, y)) < 3 || sum(stats::complete.cases(x, y)) < 3) {
    return(c(estimate = NA_real_, p.value = NA_real_))
  }
  res <- suppressWarnings(stats::cor.test(x, y, method = method, exact = FALSE))
  c(estimate = unname(res$estimate), p.value = res$p.value)
}

ec_scmetab_imc_xenium_cor_stats <- ec_scmetab_imc_xenium_cor_long %>%
  dplyr::group_by(comparison) %>%
  dplyr::summarise(
    n = dplyr::n(),
    spearman_rho = ec_cor_test_one(imc_score, xenium_score, "spearman")["estimate"],
    spearman_p = ec_cor_test_one(imc_score, xenium_score, "spearman")["p.value"],
    pearson_r = ec_cor_test_one(imc_score, xenium_score, "pearson")["estimate"],
    pearson_p = ec_cor_test_one(imc_score, xenium_score, "pearson")["p.value"],
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    label = paste0(
      "Spearman rho=", sprintf("%.2f", spearman_rho),
      "\np=", ec_format_p(spearman_p),
      "\nn=", n
    )
  )

ec_scmetab_imc_xenium_cell_cor_stats_overall <- ec_scmetab_imc_xenium_cell_cor_long %>%
  dplyr::group_by(comparison) %>%
  dplyr::summarise(
    stratum = "All matched cells",
    n = dplyr::n(),
    n_samples = dplyr::n_distinct(sample_id),
    spearman_rho = ec_cor_test_one(imc_score, xenium_score, "spearman")["estimate"],
    spearman_p = ec_cor_test_one(imc_score, xenium_score, "spearman")["p.value"],
    pearson_r = ec_cor_test_one(imc_score, xenium_score, "pearson")["estimate"],
    pearson_p = ec_cor_test_one(imc_score, xenium_score, "pearson")["p.value"],
    .groups = "drop"
  )

ec_scmetab_imc_xenium_cell_cor_stats_by_label <- ec_scmetab_imc_xenium_cell_cor_long %>%
  dplyr::group_by(comparison, metEClabel) %>%
  dplyr::summarise(
    stratum = as.character(dplyr::first(metEClabel)),
    n = dplyr::n(),
    n_samples = dplyr::n_distinct(sample_id),
    spearman_rho = ec_cor_test_one(imc_score, xenium_score, "spearman")["estimate"],
    spearman_p = ec_cor_test_one(imc_score, xenium_score, "spearman")["p.value"],
    pearson_r = ec_cor_test_one(imc_score, xenium_score, "pearson")["estimate"],
    pearson_p = ec_cor_test_one(imc_score, xenium_score, "pearson")["p.value"],
    .groups = "drop"
  )

ec_scmetab_imc_xenium_cell_cor_stats <- dplyr::bind_rows(
  ec_scmetab_imc_xenium_cell_cor_stats_overall,
  ec_scmetab_imc_xenium_cell_cor_stats_by_label
) %>%
  dplyr::mutate(
    label = paste0(
      "rho=", sprintf("%.2f", spearman_rho),
      "\np=", ec_format_p(spearman_p),
      "\nn=", n
    )
  )

ec_scmetab_imc_xenium_sample_key_overlap <- data.frame(
  xenium_sample_column = names(ec_xenium_sample_overlap),
  n_matching_imc_sample_ids = as.integer(ec_xenium_sample_overlap),
  selected = names(ec_xenium_sample_overlap) == ec_xenium_correlation_sample_col
)

ec_scmetab_imc_xenium_cell_key_overlap <- data.frame(
  xenium_cell_key_column = ec_xenium_cell_key_col,
  n_imc_cells_with_key = dplyr::n_distinct(ec_df_cell$KEY),
  n_xenium_cells_with_key = dplyr::n_distinct(ec_xenium_cell_scores$xenium_cell_key),
  n_matched_cells = dplyr::n_distinct(ec_scmetab_imc_xenium_cell_joined$KEY),
  n_matched_cell_pathway_pairs = nrow(ec_scmetab_imc_xenium_cell_cor_long)
)

write.csv(ec_df_sample, file.path(scmetab_output_dir, "scmetabolism_ec_imc_metEClabel_sample_scores.csv"), row.names = FALSE)
write.csv(ec_df_cell, file.path(scmetab_output_dir, "scmetabolism_ec_imc_metEClabel_cell_scores.csv"), row.names = FALSE)
write.csv(ec_scmetab_imc_xenium_joined, file.path(scmetab_output_dir, "scmetabolism_ec_imc_xenium_sample_joined.csv"), row.names = FALSE)
write.csv(ec_scmetab_imc_xenium_cell_joined, file.path(scmetab_output_dir, "scmetabolism_ec_imc_xenium_cell_joined.csv"), row.names = FALSE)
write.csv(ec_scmetab_imc_xenium_cor_long, file.path(scmetab_output_dir, "scmetabolism_ec_imc_xenium_correlation_long.csv"), row.names = FALSE)
write.csv(ec_scmetab_imc_xenium_cell_cor_long, file.path(scmetab_output_dir, "scmetabolism_ec_imc_xenium_cell_correlation_long.csv"), row.names = FALSE)
write.csv(ec_scmetab_imc_xenium_cor_stats, file.path(scmetab_output_dir, "scmetabolism_ec_imc_xenium_correlation_stats.csv"), row.names = FALSE)
write.csv(ec_scmetab_imc_xenium_cell_cor_stats, file.path(scmetab_output_dir, "scmetabolism_ec_imc_xenium_cell_correlation_stats.csv"), row.names = FALSE)
write.csv(ec_scmetab_imc_xenium_sample_key_overlap, file.path(scmetab_output_dir, "scmetabolism_ec_imc_xenium_sample_key_overlap.csv"), row.names = FALSE)
write.csv(ec_scmetab_imc_xenium_cell_key_overlap, file.path(scmetab_output_dir, "scmetabolism_ec_imc_xenium_cell_key_overlap.csv"), row.names = FALSE)

dplyr::bind_rows(
  ec_scmetab_imc_xenium_cor_stats %>% dplyr::mutate(level = "sample"),
  ec_scmetab_imc_xenium_cell_cor_stats_overall %>% dplyr::mutate(level = "single_cell")
)
```

```{r scmetabolism_ec_imc_xenium_sample_correlation_plot, fig.width=8.5, fig.height=4.8}
if (nrow(ec_scmetab_imc_xenium_cor_long) > 0) {
  p_scmetab_ec_imc_xenium_correlation <- ggplot(
    ec_scmetab_imc_xenium_cor_long,
    aes(x = imc_score, y = xenium_score, color = metEClabel)
  ) +
    geom_point(size = 3, alpha = 0.9) +
    geom_smooth(method = "lm", se = FALSE, color = "grey35", linewidth = 0.7) +
    geom_text(
      data = ec_scmetab_imc_xenium_cor_stats,
      aes(x = -Inf, y = Inf, label = label),
      inherit.aes = FALSE,
      hjust = -0.05,
      vjust = 1.08,
      size = 3.3
    ) +
    scale_color_manual(values = c("Met_int_EC" = "#4E79A7", "Met_hi_EC" = "#E15759")) +
    facet_wrap(~ comparison, nrow = 1, scales = "free") +
    labs(
      title = "EC IMC marker-score vs Xenium scMetabolism score",
      subtitle = paste0("Aggregated by IMC sample_id and Xenium ", ec_xenium_correlation_sample_col, " + metEClabel"),
      x = "IMC mean score",
      y = "Xenium mean scMetabolism AUCell score",
      color = NULL
    ) +
    theme_bw() +
    theme(
      strip.text = element_text(size = 11, face = "bold"),
      legend.position = "top",
      plot.title = element_text(hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5)
    )
} else {
  p_scmetab_ec_imc_xenium_correlation <- ggplot() +
    annotate(
      "text",
      x = 0,
      y = 0,
      label = paste0(
        "No overlapping EC IMC/Xenium sample_id + metEClabel groups found.\n",
        "Selected Xenium sample column: ", ec_xenium_correlation_sample_col
      ),
      size = 5
    ) +
    theme_void() +
    labs(title = "EC IMC marker-score vs Xenium scMetabolism score")
}

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_ec_imc_xenium_correlation.png"),
  p_scmetab_ec_imc_xenium_correlation,
  width = 8.5,
  height = 4.8,
  dpi = 300
)

p_scmetab_ec_imc_xenium_correlation
```

```{r scmetabolism_ec_imc_xenium_cell_correlation_density_plot, fig.width=8.5, fig.height=6.2}
ec_cell_correlation_plot_stats <- ec_scmetab_imc_xenium_cell_cor_stats %>%
  dplyr::filter(!is.na(metEClabel)) %>%
  dplyr::mutate(
    metEClabel = factor(as.character(metEClabel), levels = ec_met_levels),
    comparison = factor(as.character(comparison), levels = levels(ec_scmetab_imc_xenium_cell_cor_long$comparison))
  )

if (nrow(ec_scmetab_imc_xenium_cell_cor_long) > 0) {
  p_scmetab_ec_imc_xenium_cell_density <- ggplot(
    ec_scmetab_imc_xenium_cell_cor_long,
    aes(x = imc_score, y = xenium_score)
  ) +
    geom_bin2d(bins = 45) +
    geom_smooth(method = "lm", se = FALSE, color = "grey15", linewidth = 0.55) +
    geom_text(
      data = ec_cell_correlation_plot_stats,
      aes(x = -Inf, y = Inf, label = label),
      inherit.aes = FALSE,
      hjust = -0.05,
      vjust = 1.08,
      size = 3.1
    ) +
    scale_fill_gradient(low = "grey94", high = "#B23B3F", trans = "sqrt", name = "Cells") +
    facet_grid(metEClabel ~ comparison, scales = "free") +
    labs(
      title = "EC single-cell matched IMC marker score vs Xenium scMetabolism score",
      subtitle = paste0("Joined by ", ec_xenium_cell_key_col, "; binned cells, statistics from all matched cells"),
      x = "IMC single-cell marker score",
      y = "Xenium single-cell scMetabolism AUCell score"
    ) +
    theme_bw() +
    theme(
      strip.text = element_text(size = 10, face = "bold"),
      legend.position = "right",
      plot.title = element_text(hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5)
    )
} else {
  p_scmetab_ec_imc_xenium_cell_density <- ggplot() +
    annotate(
      "text",
      x = 0,
      y = 0,
      label = paste0(
        "No overlapping EC IMC/Xenium cells found.\n",
        "Selected Xenium cell key column: ", ec_xenium_cell_key_col
      ),
      size = 5
    ) +
    theme_void() +
    labs(title = "EC single-cell matched IMC marker score vs Xenium scMetabolism score")
}

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_ec_imc_xenium_cell_correlation_density.png"),
  p_scmetab_ec_imc_xenium_cell_density,
  width = 8.5,
  height = 6.2,
  dpi = 300
)

p_scmetab_ec_imc_xenium_cell_density
```

```{r scmetabolism_ec_imc_xenium_cell_correlation_point_plot, fig.width=8.5, fig.height=4.8}
ec_cell_point_limit <- 40000

if (nrow(ec_scmetab_imc_xenium_cell_cor_long) > 0) {
  set.seed(1)
  ec_cell_point_data <- ec_scmetab_imc_xenium_cell_cor_long %>%
    dplyr::group_by(comparison, metEClabel) %>%
    dplyr::group_modify(function(.x, .y) {
      if (nrow(.x) > ec_cell_point_limit) {
        dplyr::slice_sample(.x, n = ec_cell_point_limit)
      } else {
        .x
      }
    }) %>%
    dplyr::ungroup()

  ec_cell_point_subtitle <- if (nrow(ec_cell_point_data) < nrow(ec_scmetab_imc_xenium_cell_cor_long)) {
    paste0("Up to ", ec_cell_point_limit, " cells per facet plotted; statistics use all matched cells")
  } else {
    "All matched cells plotted and used for statistics"
  }

  p_scmetab_ec_imc_xenium_cell_points <- ggplot(
    ec_cell_point_data,
    aes(x = imc_score, y = xenium_score, color = metEClabel)
  ) +
    geom_point(alpha = 0.2, size = 0.55, stroke = 0) +
    geom_smooth(
      data = ec_scmetab_imc_xenium_cell_cor_long,
      method = "lm",
      se = FALSE,
      linewidth = 0.65,
      alpha = 0.8
    ) +
    geom_text(
      data = ec_scmetab_imc_xenium_cell_cor_stats_overall %>%
        dplyr::mutate(
          label = paste0(
            "rho=", sprintf("%.2f", spearman_rho),
            "\np=", ec_format_p(spearman_p),
            "\nn=", n
          ),
          comparison = factor(as.character(comparison), levels = levels(ec_scmetab_imc_xenium_cell_cor_long$comparison))
        ),
      aes(x = -Inf, y = Inf, label = label),
      inherit.aes = FALSE,
      hjust = -0.05,
      vjust = 1.08,
      size = 3.2
    ) +
    scale_color_manual(values = c("Met_int_EC" = "#4E79A7", "Met_hi_EC" = "#E15759")) +
    facet_wrap(~ comparison, nrow = 1, scales = "free") +
    labs(
      title = "EC single-cell IMC-Xenium correlation point cloud",
      subtitle = ec_cell_point_subtitle,
      x = "IMC single-cell marker score",
      y = "Xenium single-cell scMetabolism AUCell score",
      color = NULL
    ) +
    theme_bw() +
    theme(
      strip.text = element_text(size = 11, face = "bold"),
      legend.position = "top",
      plot.title = element_text(hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5)
    )
} else {
  p_scmetab_ec_imc_xenium_cell_points <- ggplot() +
    annotate(
      "text",
      x = 0,
      y = 0,
      label = paste0(
        "No overlapping EC IMC/Xenium cells found.\n",
        "Selected Xenium cell key column: ", ec_xenium_cell_key_col
      ),
      size = 5
    ) +
    theme_void() +
    labs(title = "EC single-cell IMC-Xenium correlation point cloud")
}

ggsave(
  file.path(scmetab_output_dir, "scmetabolism_ec_imc_xenium_cell_correlation_points.png"),
  p_scmetab_ec_imc_xenium_cell_points,
  width = 8.5,
  height = 4.8,
  dpi = 300
)

p_scmetab_ec_imc_xenium_cell_points
```
