---
title: "First-cohort metabolic-neighborhood correlations with local mRSS"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    code_folding: hide
date: "`r format(Sys.Date(), '%d %B %Y')`"
---

```{r setup, include=FALSE}
source("data_preprocessing/common/config.R")
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dpi = 180
)

suppressPackageStartupMessages({
  library(SpatialExperiment)
  library(readxl)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(patchwork)
  library(scales)
  library(knitr)
})

spe_path <- project_path("IMC_metabolic_analysis_AEM/Datasets/spe20250127174306.rds")
clinical_path <- project_path("Revision/Clinical data_metabolic_AEM_therapy_localmRss.xlsx")
output_dir <- file.path(getwd(), "first_cohort_local_mRSS_outputs")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

theme_set(theme_bw(base_size = 13))
```

## Analysis definition

This report adapts the original first-cohort neighborhood-correlation analysis to the **local skin score** recorded in the revised clinical table. The original analysis unit and normalization are retained: for every SSc donor, each base-cluster/neighbor combination is expressed as a percentage of all stromal cells from that donor, and the donor-level percentages are correlated with the clinical score using Spearman's rank correlation.

The local score is available for eight of the nine SSc donors. Its observed range is limited to 1 or 2, producing many tied ranks. Raw asymptotic Spearman p-values are used for plot colors to preserve the convention of the original figure. Exact two-sided permutation p-values across all possible allocations of the three score-2 labels among eight donors, and Benjamini-Hochberg-adjusted q-values across the 33 displayed correlations, are reported in the audit table.

```{r helpers}
parse_local_score <- function(x) {
  x <- as.character(x)
  out <- suppressWarnings(as.numeric(x))
  needs_parse <- is.na(out) & grepl("[0-9]", x)
  out[needs_parse] <- suppressWarnings(as.numeric(
    sub(".*?([0-9]+(?:\\.[0-9]+)?).*", "\\1", x[needs_parse], perl = TRUE)
  ))
  out
}

cluster_labels <- c(
  stromal_1 = "TCA/OXPHOS\u02b0\u2071\nPGC-1\u03b1\u02b0\u2071_Fib",
  stromal_2 = "Met\u02b0\u2071_Fib",
  stromal_3 = "TCA/OXPHOS\u2071\u207f\u1d57\nPGC-1\u03b1\u02b0\u2071_Fib",
  stromal_4 = "pmTOR\u02b0\u2071CD98\u02b0\u2071_Fib",
  stromal_5 = "Met\u02e1\u1d52\u02b7_Fib",
  stromal_6 = "Glycolysis\u2071\u207f\u1d57\nCS\u02b0\u2071OGDH\u02e1\u1d52\u02b7_Fib",
  stromal_7 = "Glut\u02b0\u2071_Fib",
  stromal_8 = "Glycolysis\u2071\u207f\u1d57\nPGC-1\u03b1\u02e1\u1d52\u02b7FAO\u02e1\u1d52\u02b7_Fib",
  CD31_1 = "Met\u02e1\u1d52\u02b7_EC",
  CD31_2 = "TCA/OXPHOS\u02b0\u2071\nPGC-1\u03b1\u02b0\u2071_EC",
  CD31_3 = "Met\u02b0\u2071_EC",
  CD31_4 = "pmTOR\u02b0\u2071CD98\u02b0\u2071_EC",
  CD31_5 = "Glycolysis\u02b0\u2071\nTCA/OXPHOS\u2071\u207f\u1d57PPP\u02b0\u2071_EC",
  CD45_CD68_1 = "Met\u02b0\u2071_Mf",
  CD45_CD68_2 = "Met\u02e1\u1d52\u02b7_Mf",
  CD45_CD68_3 = "Glycolysis\u02e1\u1d52\u02b7\nTCA/OXPHOS\u02b0\u2071_Mf",
  CD45_CD68_4 = "Glycolysis\u02b0\u2071\nTCA/OXPHOS\u02b0\u2071_Mf"
)

panel_definitions <- bind_rows(
  tibble(panel = "A", base_cluster = "stromal_2", neighbor_cluster = paste0("stromal_", 1:8),
         panel_title = "Met\u02b0\u2071_Fib \u2192 fibroblast neighborhoods"),
  tibble(panel = "B", base_cluster = "stromal_2", neighbor_cluster = paste0("CD31_", 1:5),
         panel_title = "Met\u02b0\u2071_Fib \u2192 endothelial neighborhoods"),
  tibble(panel = "C", base_cluster = "stromal_4", neighbor_cluster = paste0("stromal_", 1:8),
         panel_title = "pmTOR\u02b0\u2071CD98\u02b0\u2071_Fib \u2192 fibroblast neighborhoods"),
  tibble(panel = "D", base_cluster = "stromal_8", neighbor_cluster = paste0("stromal_", 1:8),
         panel_title = "Glycolysis\u2071\u207f\u1d57 PGC-1\u03b1\u02e1\u1d52\u02b7 FAO\u02e1\u1d52\u02b7_Fib \u2192 fibroblast neighborhoods"),
  tibble(panel = "E", base_cluster = "stromal_8", neighbor_cluster = paste0("CD45_CD68_", 1:4),
         panel_title = "Glycolysis\u2071\u207f\u1d57 PGC-1\u03b1\u02e1\u1d52\u02b7 FAO\u02e1\u1d52\u02b7_Fib \u2192 macrophage neighborhoods")
) %>%
  mutate(
    neighbor_variable = paste0("has_", neighbor_cluster, "_neigh"),
    positive_neighbor_value = paste0("with_", neighbor_cluster, "_neigh"),
    display_label = unname(cluster_labels[neighbor_cluster])
  )

exact_binary_permutation_p <- function(score, frequency, observed_rho) {
  stopifnot(length(score) == length(frequency))
  n_high <- sum(score == max(score))
  allocations <- combn(seq_along(score), n_high)
  permuted <- apply(allocations, 2, function(high_indices) {
    permuted_score <- rep(min(score), length(score))
    permuted_score[high_indices] <- max(score)
    suppressWarnings(cor(permuted_score, frequency, method = "spearman"))
  })
  mean(abs(permuted) >= abs(observed_rho) - sqrt(.Machine$double.eps), na.rm = TRUE)
}
```

## Clinical data

```{r clinical-data}
clinical_raw <- suppressWarnings(read_xlsx(clinical_path, sheet = "Summary"))

clinical <- clinical_raw %>%
  transmute(
    patient_id = `Label in metabolic panel`,
    subtype = `Subtype (dc/lc)`,
    current_mRSS = suppressWarnings(as.numeric(`current mRSS`)),
    local_mRSS = parse_local_score(`Local skin score`)
  ) %>%
  filter(subtype != "healthy") %>%
  mutate(included_local_mRSS = !is.na(local_mRSS))

write.csv(clinical, file.path(output_dir, "local_mRSS_donor_inclusion.csv"), row.names = FALSE)

kable(
  clinical,
  digits = 2,
  caption = "SSc donors and local-mRSS availability"
)
```

## Donor-level cluster frequencies

```{r calculate-frequencies}
spe <- readRDS(spe_path)
cell_data <- as.data.frame(colData(spe)) %>%
  transmute(
    patient_id = patient_id,
    indication = indication,
    celltype = new_celllabels,
    metabolic_cluster = pheno_clusters_new2,
    cell_index = seq_len(n())
  )

neighbor_data <- as.data.frame(spe$hasNeigh_df)
stopifnot(nrow(neighbor_data) == nrow(cell_data))
needed_neighbor_variables <- unique(panel_definitions$neighbor_variable)
stopifnot(all(needed_neighbor_variables %in% names(neighbor_data)))

analysis_cells <- bind_cols(
  cell_data,
  neighbor_data[, needed_neighbor_variables, drop = FALSE]
) %>%
  filter(indication == "SSc", celltype == "Stromal")

stromal_denominators <- analysis_cells %>%
  count(patient_id, name = "stromal_denominator")

frequency_long <- lapply(seq_len(nrow(panel_definitions)), function(i) {
  definition <- panel_definitions[i, ]
  analysis_cells %>%
    mutate(
      is_target = metabolic_cluster == definition$base_cluster &
        .data[[definition$neighbor_variable]] == definition$positive_neighbor_value
    ) %>%
    group_by(patient_id) %>%
    summarise(target_cells = sum(is_target, na.rm = TRUE), .groups = "drop") %>%
    right_join(stromal_denominators, by = "patient_id") %>%
    mutate(
      target_cells = replace_na(target_cells, 0L),
      frequency_percent = 100 * target_cells / stromal_denominator,
      panel = definition$panel,
      base_cluster = definition$base_cluster,
      neighbor_cluster = definition$neighbor_cluster,
      display_label = definition$display_label,
      panel_title = definition$panel_title
    )
}) %>%
  bind_rows() %>%
  left_join(clinical, by = "patient_id")

write.csv(
  frequency_long,
  file.path(output_dir, "local_mRSS_donor_level_neighborhood_frequencies.csv"),
  row.names = FALSE
)

frequency_summary <- frequency_long %>%
  filter(included_local_mRSS) %>%
  group_by(panel, base_cluster, neighbor_cluster, display_label, panel_title) %>%
  group_modify(~{
    test <- suppressWarnings(cor.test(
      .x$local_mRSS,
      .x$frequency_percent,
      method = "spearman",
      exact = FALSE
    ))
    observed_rho <- unname(test$estimate)
    tibble(
      n_donors = nrow(.x),
      rho = observed_rho,
      p_spearman_raw = test$p.value,
      p_exact_permutation = exact_binary_permutation_p(
        .x$local_mRSS,
        .x$frequency_percent,
        observed_rho
      ),
      minimum_frequency_percent = min(.x$frequency_percent),
      maximum_frequency_percent = max(.x$frequency_percent)
    )
  }) %>%
  ungroup() %>%
  mutate(
    q_BH_33_tests = p.adjust(p_spearman_raw, method = "BH"),
    direction = case_when(
      rho > 0 ~ "positive",
      rho < 0 ~ "negative",
      TRUE ~ "zero"
    )
  ) %>%
  arrange(panel, desc(rho))

write.csv(
  frequency_summary,
  file.path(output_dir, "local_mRSS_correlation_results_33_tests.csv"),
  row.names = FALSE
)
```

## Correlation plots

```{r plot-functions}
plot_local_mrss_panel <- function(panel_id, show_legends = TRUE) {
  plot_data <- frequency_summary %>%
    filter(panel == panel_id) %>%
    arrange(rho) %>%
    mutate(display_label = factor(display_label, levels = display_label))

  title_lookup <- c(
    A = "Met\u02b0\u2071_Fib",
    B = "Met\u02b0\u2071_Fib",
    C = "pmTOR\u02b0\u2071CD98\u02b0\u2071_Fib",
    D = "Glycolysis\u2071\u207f\u1d57 PGC-1\u03b1\u02e1\u1d52\u02b7 FAO\u02e1\u1d52\u02b7_Fib",
    E = "Glycolysis\u2071\u207f\u1d57 PGC-1\u03b1\u02e1\u1d52\u02b7 FAO\u02e1\u1d52\u02b7_Fib"
  )
  subtitle_lookup <- c(
    A = "local mRSS | fibroblast neighborhoods",
    B = "local mRSS | endothelial neighborhoods",
    C = "local mRSS | fibroblast neighborhoods",
    D = "local mRSS | fibroblast neighborhoods",
    E = "local mRSS | macrophage neighborhoods"
  )

  p <- ggplot(plot_data, aes(x = rho, y = display_label)) +
    geom_vline(xintercept = 0, linewidth = 0.35, color = "grey75") +
    geom_segment(
      aes(x = 0, xend = rho, yend = display_label),
      linewidth = 0.55,
      color = "grey75"
    ) +
    geom_point(
      aes(size = abs(rho), color = p_spearman_raw),
      alpha = 1
    ) +
    scale_color_gradient(
      name = "p value",
      low = "red",
      high = "blue",
      limits = c(0.001, 0.05),
      breaks = c(0.01, 0.02, 0.03, 0.04, 0.05),
      labels = number_format(accuracy = 0.01),
      na.value = "grey52"
    ) +
    scale_size_continuous(
      name = "|cor coef|",
      range = c(2.4, 8.2),
      limits = c(0, 1),
      breaks = c(0.2, 0.4, 0.6, 0.8)
    ) +
    scale_x_continuous(
      breaks = pretty_breaks(n = 5),
      labels = number_format(accuracy = 0.01),
      expand = expansion(mult = c(0.08, 0.08))
    ) +
    labs(
      title = unname(title_lookup[panel_id]),
      subtitle = unname(subtitle_lookup[panel_id]),
      x = "Spearman correlation",
      y = NULL
    ) +
    theme_bw(base_size = 13) +
    theme(
      plot.title = element_text(size = 14, face = "plain", hjust = 0.5, margin = margin(b = 3)),
      plot.subtitle = element_text(size = 13, hjust = 0.5, margin = margin(b = 8)),
      axis.text.x = element_text(size = 11, color = "grey20"),
      axis.text.y = element_text(size = 12, color = "grey20", lineheight = 0.92),
      axis.title.x = element_text(size = 12),
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_line(color = "grey86", linewidth = 0.45),
      panel.grid.major.x = element_line(color = "grey92", linewidth = 0.4),
      legend.title = element_text(size = 11),
      legend.text = element_text(size = 10),
      legend.key.height = unit(0.48, "cm"),
      plot.margin = margin(8, 8, 8, 8)
    ) +
    guides(
      color = guide_colorbar(order = 1, barheight = unit(2.5, "cm"), barwidth = unit(0.38, "cm")),
      size = guide_legend(order = 2, override.aes = list(color = "black"))
    )

  if (!show_legends) {
    p <- p + theme(legend.position = "none")
  }
  p
}

panel_plots <- setNames(lapply(LETTERS[1:5], plot_local_mrss_panel), LETTERS[1:5])

panel_sizes <- list(
  A = c(width = 8.0, height = 5.6),
  B = c(width = 8.0, height = 4.8),
  C = c(width = 8.0, height = 5.6),
  D = c(width = 8.0, height = 5.6),
  E = c(width = 8.0, height = 4.8)
)

for (panel_id in names(panel_plots)) {
  p <- panel_plots[[panel_id]]
  dims <- panel_sizes[[panel_id]]
  stem <- file.path(output_dir, paste0("panel_", panel_id, "_local_mRSS_correlation"))
  ggsave(paste0(stem, ".svg"), p, width = dims[["width"]], height = dims[["height"]], units = "in", bg = "white")
  ggsave(paste0(stem, ".pdf"), p, width = dims[["width"]], height = dims[["height"]], units = "in", device = cairo_pdf, bg = "white")
  ggsave(paste0(stem, ".png"), p, width = dims[["width"]], height = dims[["height"]], units = "in", dpi = 400, bg = "white")
}
```

### Panel A

```{r panel-a, echo=FALSE, fig.width=8, fig.height=5.6}
panel_plots$A
```

### Panel B

```{r panel-b, echo=FALSE, fig.width=8, fig.height=4.8}
panel_plots$B
```

### Panel C

```{r panel-c, echo=FALSE, fig.width=8, fig.height=5.6}
panel_plots$C
```

### Panel D

```{r panel-d, echo=FALSE, fig.width=8, fig.height=5.6}
panel_plots$D
```

### Panel E

```{r panel-e, echo=FALSE, fig.width=8, fig.height=4.8}
panel_plots$E
```

## Combined figure

The combined layout follows the arrangement of the supplied first-cohort figure. Individual vector panels above are preferable for direct replacement in an assembled manuscript figure.

```{r combined-figure, echo=FALSE, fig.width=12.5, fig.height=16}
combined_plot <- (
  panel_plots$A + panel_plots$B + panel_plots$C +
  panel_plots$D + panel_plots$E
) +
  plot_layout(
    design = "
      AABB
      #CC#
      DDEE
    ",
    guides = "keep"
  ) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(size = 22, face = "bold"))

ggsave(
  file.path(output_dir, "combined_local_mRSS_correlations.svg"),
  combined_plot,
  width = 12.5,
  height = 16,
  units = "in",
  bg = "white"
)
ggsave(
  file.path(output_dir, "combined_local_mRSS_correlations.pdf"),
  combined_plot,
  width = 12.5,
  height = 16,
  units = "in",
  device = cairo_pdf,
  bg = "white"
)
ggsave(
  file.path(output_dir, "combined_local_mRSS_correlations.png"),
  combined_plot,
  width = 12.5,
  height = 16,
  units = "in",
  dpi = 400,
  bg = "white"
)

combined_plot
```

## Statistical results

```{r results-table, echo=FALSE}
results_for_display <- frequency_summary %>%
  transmute(
    Panel = panel,
    `Base cluster` = unname(cluster_labels[base_cluster]),
    `Neighbor cluster` = display_label,
    `n donors` = n_donors,
    `Spearman rho` = rho,
    `Raw Spearman p` = p_spearman_raw,
    `Exact permutation p` = p_exact_permutation,
    `BH q (33 tests)` = q_BH_33_tests
  )

kable(
  results_for_display,
  digits = 4,
  caption = "Local-mRSS correlations. The exact permutation p-value enumerates all 56 possible allocations of three score-2 labels among eight donors."
)
```

## Interpretation guardrails

- The analysis can be performed using local mRSS and is directly comparable in construction to the original current-mRSS analysis.
- The effective sample size is eight donors because `SSc8` has no local skin score.
- Local mRSS takes only two observed values (1 and 2), so it behaves as an ordered two-group variable rather than a finely graded severity measure.
- Raw Spearman p-values match the original display convention. The exact permutation p-values and BH q-values should be used when assessing robustness.
- These correlations are exploratory and should not be interpreted as independent cell-level tests.

```{r session-info}
sessionInfo()
```
