---
title: "First-cohort fibroblast-cluster frequencies and local mRSS"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    code_folding: hide
date: "`r format(Sys.Date(), '%d %B %Y')`"
---

```{r setup, include=FALSE}
source("data_preprocessing/common/config.R")
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dpi = 180
)

suppressPackageStartupMessages({
  library(SpatialExperiment)
  library(readxl)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(patchwork)
  library(scales)
  library(knitr)
})

spe_path <- project_path("IMC_metabolic_analysis_AEM/Datasets/spe20250127174306.rds")
clinical_path <- project_path("Revision/Clinical data_metabolic_AEM_therapy_localmRss.xlsx")
output_dir <- file.path(getwd(), "first_cohort_local_mRSS_frequency_outputs")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

## Analysis definition

This report adapts the original first-cohort plots generated in `Regr_clinicaldata_part2_selected.Rmd`. For each SSc donor, the frequency of each selected metabolic fibroblast cluster is calculated relative to **all stromal cells from that donor**, without conditioning on cellular neighborhoods or interactions. Donor-level frequencies are then correlated with the local skin score using Spearman's rank correlation.

Local mRSS is available for eight donors and takes only the values 1 and 2. The plots show the raw Spearman p-value used by the original analysis. Exact permutation p-values and Benjamini-Hochberg-adjusted q-values across the three requested correlations are supplied in the statistical table.

```{r helpers}
parse_local_score <- function(x) {
  x <- as.character(x)
  out <- suppressWarnings(as.numeric(x))
  needs_parse <- is.na(out) & grepl("[0-9]", x)
  out[needs_parse] <- suppressWarnings(as.numeric(
    sub(".*?([0-9]+(?:\\.[0-9]+)?).*", "\\1", x[needs_parse], perl = TRUE)
  ))
  out
}

exact_binary_permutation_p <- function(score, frequency, observed_rho) {
  n_high <- sum(score == max(score))
  allocations <- combn(seq_along(score), n_high)
  permuted_rho <- apply(allocations, 2, function(high_indices) {
    permuted_score <- rep(min(score), length(score))
    permuted_score[high_indices] <- max(score)
    suppressWarnings(cor(permuted_score, frequency, method = "spearman"))
  })
  mean(
    abs(permuted_rho) >= abs(observed_rho) - sqrt(.Machine$double.eps),
    na.rm = TRUE
  )
}

cluster_definitions <- tibble(
  panel = c("A", "B", "C"),
  metabolic_cluster = c("stromal_2", "stromal_4", "stromal_8"),
  display_label = c(
    "Met\u02b0\u2071_Fib",
    "pmTOR\u02b0\u2071CD98\u02b0\u2071_Fib",
    "Glycolysis\u2071\u207f\u1d57 PGC-1\u03b1\u02e1\u1d52\u02b7 FAO\u02e1\u1d52\u02b7_Fib"
  )
)
```

## Clinical data and donor frequencies

```{r calculate-data}
clinical_raw <- suppressWarnings(read_xlsx(clinical_path, sheet = "Summary"))
clinical <- clinical_raw %>%
  transmute(
    patient_id = `Label in metabolic panel`,
    subtype = `Subtype (dc/lc)`,
    current_mRSS = suppressWarnings(as.numeric(`current mRSS`)),
    local_mRSS = parse_local_score(`Local skin score`)
  ) %>%
  filter(subtype != "healthy") %>%
  mutate(included_local_mRSS = !is.na(local_mRSS))

spe <- readRDS(spe_path)
cell_data <- as.data.frame(colData(spe)) %>%
  transmute(
    patient_id = patient_id,
    indication = indication,
    celltype = new_celllabels,
    metabolic_cluster = pheno_clusters_new2
  ) %>%
  filter(indication == "SSc", celltype == "Stromal")

donor_denominators <- cell_data %>%
  count(patient_id, name = "all_stromal_cells")

frequency_long <- cell_data %>%
  count(patient_id, metabolic_cluster, name = "cluster_cells") %>%
  filter(metabolic_cluster %in% cluster_definitions$metabolic_cluster) %>%
  complete(
    patient_id = donor_denominators$patient_id,
    metabolic_cluster = cluster_definitions$metabolic_cluster,
    fill = list(cluster_cells = 0L)
  ) %>%
  left_join(donor_denominators, by = "patient_id") %>%
  mutate(frequency_percent = 100 * cluster_cells / all_stromal_cells) %>%
  left_join(cluster_definitions, by = "metabolic_cluster") %>%
  left_join(clinical, by = "patient_id") %>%
  arrange(panel, patient_id)

write.csv(
  frequency_long,
  file.path(output_dir, "donor_level_fibroblast_cluster_frequencies.csv"),
  row.names = FALSE
)

donor_display <- frequency_long %>%
  distinct(patient_id, current_mRSS, local_mRSS, included_local_mRSS, all_stromal_cells) %>%
  arrange(desc(included_local_mRSS), patient_id)

write.csv(
  donor_display,
  file.path(output_dir, "local_mRSS_donor_inclusion_and_denominators.csv"),
  row.names = FALSE
)

kable(
  donor_display,
  digits = 2,
  caption = "Donors, local-mRSS availability, and stromal-cell denominators"
)
```

## Statistical analysis

```{r statistics}
correlation_results <- frequency_long %>%
  filter(included_local_mRSS) %>%
  group_by(panel, metabolic_cluster, display_label) %>%
  group_modify(~{
    test <- suppressWarnings(cor.test(
      .x$local_mRSS,
      .x$frequency_percent,
      method = "spearman",
      exact = FALSE
    ))
    observed_rho <- unname(test$estimate)
    tibble(
      n_donors = nrow(.x),
      n_local_mRSS_1 = sum(.x$local_mRSS == 1),
      n_local_mRSS_2 = sum(.x$local_mRSS == 2),
      rho = observed_rho,
      p_spearman_raw = test$p.value,
      p_exact_permutation = exact_binary_permutation_p(
        .x$local_mRSS,
        .x$frequency_percent,
        observed_rho
      )
    )
  }) %>%
  ungroup() %>%
  mutate(q_BH_3_tests = p.adjust(p_spearman_raw, method = "BH")) %>%
  arrange(panel)

write.csv(
  correlation_results,
  file.path(output_dir, "local_mRSS_fibroblast_frequency_correlations.csv"),
  row.names = FALSE
)

# Recalculate the archived current-mRSS correlations as a pipeline check.
current_mrss_verification <- frequency_long %>%
  group_by(panel, metabolic_cluster, display_label) %>%
  group_modify(~{
    test <- suppressWarnings(cor.test(
      .x$current_mRSS,
      .x$frequency_percent,
      method = "spearman",
      exact = FALSE
    ))
    tibble(
      n_donors = nrow(.x),
      rho_current_mRSS = unname(test$estimate),
      p_current_mRSS = test$p.value
    )
  }) %>%
  ungroup()

write.csv(
  current_mrss_verification,
  file.path(output_dir, "current_mRSS_reproduction_check.csv"),
  row.names = FALSE
)

kable(
  correlation_results,
  digits = 4,
  caption = "Spearman correlations between fibroblast-cluster frequencies and local mRSS"
)
```

## Local-mRSS correlation plots

The slight horizontal jitter reveals donors with identical local-mRSS values. Correlations and fitted lines use the original, unjittered scores.

```{r plotting-functions}
format_p <- function(p) {
  ifelse(p < 0.001, format(p, scientific = TRUE, digits = 2), sprintf("%.2f", p))
}

make_frequency_plot <- function(panel_id) {
  plot_data <- frequency_long %>%
    filter(panel == panel_id, included_local_mRSS)
  result <- correlation_results %>%
    filter(panel == panel_id)

  annotation <- sprintf(
    "\u03c1 = %.2f, p = %s",
    result$rho,
    format_p(result$p_spearman_raw)
  )

  x_jitter <- c(A = -0.035, B = -0.035, C = -0.035)
  max_y <- max(plot_data$frequency_percent, na.rm = TRUE)

  ggplot(plot_data, aes(x = local_mRSS, y = frequency_percent)) +
    geom_smooth(
      method = "lm",
      formula = y ~ x,
      se = TRUE,
      color = "blue",
      fill = "grey82",
      linewidth = 0.9,
      alpha = 0.6,
      fullrange = TRUE
    ) +
    geom_jitter(
      width = 0.035,
      height = 0,
      shape = 21,
      size = 3.0,
      stroke = 0.65,
      color = "grey45",
      fill = "white"
    ) +
    annotate(
      "text",
      x = 0.94,
      y = max_y * 1.08,
      label = annotation,
      hjust = 0,
      vjust = 1,
      size = 4.2,
      fontface = "italic"
    ) +
    scale_x_continuous(
      breaks = c(1, 2),
      expand = expansion(mult = 0)
    ) +
    scale_y_continuous(
      expand = expansion(mult = c(0, 0.02))
    ) +
    coord_cartesian(
      xlim = c(0.9, 2.1),
      ylim = c(0, max_y * 1.14),
      expand = FALSE
    ) +
    labs(
      title = unique(plot_data$display_label),
      x = "Local mRSS",
      y = "% of all fibroblasts"
    ) +
    theme_classic(base_size = 14) +
    theme(
      plot.title = element_text(size = 15, hjust = 0.5, margin = margin(b = 8)),
      axis.title = element_text(size = 14),
      axis.text = element_text(size = 12, color = "black"),
      axis.line = element_line(linewidth = 0.55),
      axis.ticks = element_line(linewidth = 0.55),
      plot.margin = margin(10, 10, 8, 10)
    )
}

panel_plots <- setNames(lapply(c("A", "B", "C"), make_frequency_plot), c("A", "B", "C"))

for (panel_id in names(panel_plots)) {
  stem <- file.path(output_dir, paste0("panel_", panel_id, "_local_mRSS_frequency"))
  ggsave(paste0(stem, ".svg"), panel_plots[[panel_id]], width = 4.6, height = 3.6, units = "in", bg = "white")
  ggsave(paste0(stem, ".pdf"), panel_plots[[panel_id]], width = 4.6, height = 3.6, units = "in", device = cairo_pdf, bg = "white")
  ggsave(paste0(stem, ".png"), panel_plots[[panel_id]], width = 4.6, height = 3.6, units = "in", dpi = 400, bg = "white")
}
```

### Panel A

```{r panel-a, echo=FALSE, fig.width=4.6, fig.height=3.6}
panel_plots$A
```

### Panel B

```{r panel-b, echo=FALSE, fig.width=4.6, fig.height=3.6}
panel_plots$B
```

### Panel C

```{r panel-c, echo=FALSE, fig.width=4.6, fig.height=3.6}
panel_plots$C
```

## Combined figure

```{r combined, echo=FALSE, fig.width=9.2, fig.height=7.2}
combined_plot <- (
  panel_plots$A + panel_plots$B + panel_plots$C
) +
  plot_layout(
    design = "
      AABB
      #CC#
    "
  ) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(size = 20, face = "plain"))

ggsave(
  file.path(output_dir, "combined_local_mRSS_fibroblast_frequencies.svg"),
  combined_plot,
  width = 9.2,
  height = 7.2,
  units = "in",
  bg = "white"
)
ggsave(
  file.path(output_dir, "combined_local_mRSS_fibroblast_frequencies.pdf"),
  combined_plot,
  width = 9.2,
  height = 7.2,
  units = "in",
  device = cairo_pdf,
  bg = "white"
)
ggsave(
  file.path(output_dir, "combined_local_mRSS_fibroblast_frequencies.png"),
  combined_plot,
  width = 9.2,
  height = 7.2,
  units = "in",
  dpi = 400,
  bg = "white"
)

combined_plot
```

## Reproduction check

The table below recalculates the three correlations using current total mRSS. It is included solely to verify that the source object, fibroblast denominator, and cluster mapping reproduce the original figure.

```{r reproduction-check, echo=FALSE}
kable(
  current_mrss_verification,
  digits = 4,
  caption = "Reproduction of the archived current-mRSS correlations"
)
```

## Interpretation

- The analysis unit is one donor; individual cells are not treated as independent observations.
- `SSc8` is omitted because its local skin score is missing.
- With three donors at local mRSS 2 and five donors at local mRSS 1, the analysis is mathematically close to a rank-based two-group comparison.
- The fitted blue line is retained to match the original presentation, but it should not be interpreted as evidence of a continuous dose-response relationship.

```{r session-info}
sessionInfo()
```
