---
title: "Heatmap metabolic markers in healthy SSc"
author: "Alex Matei"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  html_document: 
    toc: true
    toc_float: true
    toc_depth: 4
    df_print: paged
    code_folding: hide
---

```{r setup, include=FALSE,results='hide'}
source("data_preprocessing/common/config.R")
knitr::opts_chunk$set(warning = FALSE, message = FALSE, eval=FALSE) 
```

```{r load data, eval=TRUE}

dir<- c(paste0(project_path(""), "/"))

dir_r <- "data_preprocessing/common"

source(file.path(dir_r, "get_ts_rightformat_fx.R"))

ts <- get_ts_rightformat()
ptm <- proc.time()

cat("This document was created with the time stamp",ts)

library(tidyverse)
library(imcRtools)
library(SpatialExperiment)
library(scuttle)
library(CATALYST)
library(dittoSeq)
library(ggpubr)
library(rstatix)
library(ggstatsplot)
library(viridis)

load(paste0(dir,"spe_list.RData"))
spe <- spe_list$spe
rm(spe_list)

source(file.path(dir_r, "output_dir_svg_plots_fx.R"))
output_dir <- output_dir_svg_plots_fx(ts)

```

# Prepare spe for analysis

```{r prep spe, eval=TRUE, cache=TRUE}
# Create a new data frame with all the colData of the spe object - to recode the patient_id column

metadata <- colData(spe)

# add a new column to metadata called Donor where you recode the patient_id column
# the elements starting with NH should be called Healthy1, Healthy2, etc.
# the elements starting with SSc should be called SSc1, SSc2, etc.
metadata$Donor <- dplyr::recode(metadata$patient_id,
                                     Healthy1="Healthy1",
                                     Healthy2="Healthy2", 
                                     SSc1="SSc1",
                                     SSc2="SSc2",
                                     SSc3="SSc3",
                                     SSc4="SSc4",
                                     SSc5="SSc5",
                                     SSc6="SSc6",
                                     SSc7="SSc7",
                                     SSc8="SSc8",
                                     Healthy3="Healthy3",
                                     Healthy4="Healthy4",
                                     SSc9="SSc9",
                                     Healthy5="Healthy5",
                                     Healthy6="Healthy6",
                                     Healthy7="Healthy7")

metadata$Disease <- dplyr::recode(metadata$indication,
                                  NH="Healthy",
                                  SSc="SSc")

metadata$ILD <- dplyr::recode(metadata$ILD,
                                     healthy="healthy",
                                     "1"="ILD", 
                                     "0"="no ILD")

metadata$Arthritis[is.na(metadata$Arthritis)] <- "no arthritis"

# Assign the new column in metadata to the spe object
spe$Donor <- metadata$Donor
spe$ILD <- metadata$ILD
spe$Arthritis <- metadata$Arthritis
spe$Disease <- metadata$Disease

names(colData(spe))[which(names(colData(spe))=="Skin.progression.at.biopsy")]="Progression_skin"
names(colData(spe))[which(names(colData(spe))=="Subtype.(dc/lc)")]="Subtype_dc_lc"

met_markers <- c("Glut1",
                 "HK1",
                 "PFKL1",
                 "PKM2",
                 "LDHA",
                 "GAPDH",
                 "CS",
                 "OGDH",
                 "ATP5A",
                 "SDHA",
                 "CD98",
                 "ACAC",
                 "CPT1a",
                 "G6PD",
                 "p_mTOR",
                 "pAMPK",
                 "NOX4",
                 "HIF1a",
                 "PGC-1a",
                 "TFAM",
                 "NRF2_p")

glycolysis_markers <- c("Glut1",
                 "HK1",
                 "PFKL1",
                 "PKM2",
                 "LDHA",
                 "GAPDH")

no_met_markers <- setdiff(rownames(spe),met_markers) 
rowData(spe)$use_channel_met_markers <- rownames(spe) %in% met_markers
rowData(spe)$use_channel_glyc_markers <- rownames(spe) %in% glycolysis_markers

# define clinical categories
clinical_cat <- c("Disease","Progression_skin", "Arthritis", "DU.history", "ILD", "Subtype_dc_lc", "Early.or.not")

spe[[clinical_cat[1]]] <- factor(spe[[clinical_cat[1]]], levels = c("Healthy", "SSc"))
spe[[clinical_cat[2]]] <- factor(spe[[clinical_cat[2]]], levels = c("healthy", "progressive", "stable"))
spe[[clinical_cat[3]]] <- factor(spe[[clinical_cat[3]]], levels = c("healthy","arthritis","no arthritis"))
spe[[clinical_cat[4]]] <- factor(spe[[clinical_cat[4]]], levels = c("healthy","DU history","no DU history"))
spe[[clinical_cat[5]]] <- factor(spe[[clinical_cat[5]]], levels = c("healthy","ILD","no ILD"))
spe[[clinical_cat[6]]] <- factor(spe[[clinical_cat[6]]], levels = c("healthy","dcSSc","lcSSc"))
spe[[clinical_cat[7]]] <- factor(spe[[clinical_cat[7]]], levels = c("healthy","early","late"))

# define color vectors
color_vectors <- list()
pheno_clusters_new <- setNames(c("#C4281C","#C2DAB8","#E8BAC8","#80BBDB","#C470A0","#0D69AC","#F5CD30","#287F47",
                                 "#A05F35","#D9856C","#6B327C","#008F9C","#C7D23C","#B7D7D5","#923978","#27462D",
                                 "#FF9494","#938767","#FFF67B","#E1A4C2","#FC7B1B","#CD6298","#102ADC"), 
                               as.character(unique(spe$pheno_clusters_new)))

indication <- setNames(c("#4DBBD533", "#E64B35AA"),
                       c(unique(colData(spe)$Disease)))

Donor <- setNames(c("#FFFF33", "#FFCC66",'#FF9933','#FF6633','#00FFFF','#FF0033','#79B5B5', '#C470A0',
                         '#996633','#999933','#99CC33','#FF9494', '#33FF33','#336633', '#330066','#923978'),
                       c(unique(colData(spe)$Donor)))

Neighbourhood_del_80_pheno_new <- setNames(c("#CC00CC", "#260EEC", "#FFCA28", "#004D40","#FF6633","#66FF66", 
                                             "#EF9EA7", "#990000", "#CC0033","#4FC3F7"),
                                           unique(spe$Neighbourhood_del_80_pheno_new))

color_vectors$clusters <- pheno_clusters_new
color_vectors$Disease <- indication
color_vectors$Donor <- Donor
color_vectors$neighbourhood_del_80_pheno_new <- Neighbourhood_del_80_pheno_new

color_vectors[[clinical_cat[2]]] <- setNames(c("#4DBBD533", "#C03830", "#F6BF93"), unique(spe[[clinical_cat[2]]]))
color_vectors[[clinical_cat[3]]] <- setNames(c("#4DBBD533","#377EB8","#4DAF4A"), unique(spe[[clinical_cat[3]]]))
color_vectors[[clinical_cat[4]]] <- setNames(c("#4DBBD533","#FC8D62","#8DA0CB"), unique(spe[[clinical_cat[4]]]))
color_vectors[[clinical_cat[5]]] <- setNames(c("#4DBBD533","#FFFFB3","#BEBADA"), unique(spe[[clinical_cat[5]]]))
color_vectors[[clinical_cat[6]]] <- setNames(c("#4DBBD533","#A6D854","#FFD92F"), unique(spe[[clinical_cat[6]]]))
color_vectors[[clinical_cat[7]]] <- setNames(c("#4DBBD533","#F781BF","#999999"), unique(spe[[clinical_cat[7]]]))

metadata(spe)$color_vectors <- color_vectors


# define an spe of all cells except epithelial cells
spe_not_Epith <- spe[,which(!spe$new_celllabels=="E-cad+")]

```

# Aggregate by healthy and SSc

```{r aggreg, eval=TRUE, cache=TRUE}

patient_mean_not_Epith <- aggregateAcrossCells(as(spe_not_Epith, "SingleCellExperiment"),  
                     ids = spe_not_Epith$patient_id, 
                     statistics = "mean",
                     use.assay.type = "z-rescaled", 
                     subset.row = rownames(spe_not_Epith)[rowData(spe_not_Epith)$use_channel_met_markers])


names(colData(patient_mean_not_Epith))[which(names(colData(patient_mean_not_Epith))=="ncells")]="Cells_number"

patient_mean_all_cells <- aggregateAcrossCells(as(spe, "SingleCellExperiment"),  
                     ids = spe$patient_id, 
                     statistics = "mean",
                     use.assay.type = "z-rescaled", 
                     subset.row = rownames(spe)[rowData(spe)$use_channel_met_markers])

names(colData(patient_mean_all_cells))[which(names(colData(patient_mean_all_cells))=="ncells")]="Cells_number"

```

# Heatmap z score rescaled - all cells except epithelial {.tabset .tabset-pill}

`r fig_dim <- c(10, 12)`

```{r heatmap1, eval=TRUE, fig.dim=fig_dim, results='asis'}

  while (!is.null(dev.list())) dev.off()
cat(sprintf("\n\n## %s {.tabset .tabset-pill}\n", "EMPTY"))
  counter <- 1
  plot_filename <- file.path(output_dir, paste0(counter,".svg"))
  counter <- counter + 1
  svg(filename = plot_filename, width = fig_dim[1], height = fig_dim[2])
  dittoHeatmap(patient_mean_not_Epith, 
             genes=met_markers,
             assay = "z-rescaled", 
             cluster_cols = TRUE,
             cluster_rows = FALSE,
             scale="row",
             annot_col=c(""),
             annot.by = c("Disease",
                          "Donor", 
                          "Cells_number"),
             annotation_colors = list(Disease = metadata(spe)$color_vectors$Disease,
                                      Donor = metadata(spe)$color_vectors$Donor,
                                      Cells_number = plasma(100)))
dev.off()

# Embed and provide download link
cat(sprintf('<object type="image/svg+xml" data="%s" width="600"></object>\n', plot_filename))
cat(sprintf('<a href="%s" download> Download Plot as SVG</a>\n\n', plot_filename))

for (cl_cat in clinical_cat[1:length(clinical_cat)]){
  
  cat(sprintf("\n\n## Heatmap annotated by %s {.tabset .tabset-pill}\n", cl_cat))
  
    annotation_colors <- list(
    Disease = metadata(spe)$color_vectors$Disease,
    Donor = metadata(spe)$color_vectors$Donor,
    Cells_number = plasma(100)
  )
   
  # Dynamically add the color for the current clinical category with proper naming
  annotation_colors[[cl_cat]] <- metadata(spe)$color_vectors[[cl_cat]]
  
  plot_filename <- file.path(output_dir, paste0(counter,".svg"))
  counter <- counter + 1
  svg(filename = plot_filename, width = fig_dim[1], height = fig_dim[2])
dittoHeatmap(patient_mean_not_Epith, 
             genes=met_markers,
             assay = "z-rescaled", 
             cluster_cols = TRUE,
             cluster_rows = FALSE,
             scale="row",
             annot_col=c(""),
             annot.by = c("Disease",
                          cl_cat,
                          "Donor", 
                          "Cells_number"),
             annotation_colors = annotation_colors)
  dev.off()
  
# Embed and provide download link
cat(sprintf('<object type="image/svg+xml" data="%s" width="600"></object>\n', plot_filename))
cat(sprintf('<a href="%s" download> Download Plot as SVG</a>\n\n', plot_filename))  
}

 annotation_colors <- list(
    Disease = metadata(spe)$color_vectors$Disease,
   # Donor = metadata(spe)$color_vectors$Donor,
    Cells_number = plasma(100)
  )
 
 for (i in 2:length(clinical_cat)){
   cl_cat <- clinical_cat[i]
   # Dynamically add the color for the current clinical category with proper naming
   annotation_colors[[cl_cat]] <- metadata(spe)$color_vectors[[cl_cat]]
 }
 
  
cat(sprintf("\n\n## Heatmap annotated by %s {.tabset .tabset-pill}\n", "all clinical groups")) 
  plot_filename <- file.path(output_dir, paste0(counter,".svg"))
  counter <- counter + 1
  
  svg(filename = plot_filename, width = fig_dim[1], height = fig_dim[2])
dittoHeatmap(patient_mean_not_Epith, 
             genes=met_markers,
             assay = "z-rescaled", 
             cluster_cols = TRUE,
             cluster_rows = FALSE,
             scale="row",
             annot_col=c(""),
             annot.by = c("Disease",
                          clinical_cat[2:length(clinical_cat)],
                          #"Donor", 
                          "Cells_number"),
             annotation_colors = annotation_colors)
dev.off()

cat(sprintf('<object type="image/svg+xml" data="%s" width="600"></object>\n', plot_filename))
cat(sprintf('<a href="%s" download> Download Plot as SVG</a>\n\n', plot_filename))
```

# Heatmap z score rescaled - all cells {.tabset .tabset-pill}

`r fig_dim <- c(10, 12)`

```{r heatmap2, eval=TRUE, eval=TRUE, fig.dim=fig_dim, results='asis'}

  while (!is.null(dev.list())) dev.off()

cat(sprintf("\n\n## %s {.tabset .tabset-pill}\n", "EMPTY"))
  plot_filename <- file.path(output_dir, paste0(counter,".svg"))
  counter <- counter + 1
  svg(filename = plot_filename, width = fig_dim[1], height = fig_dim[2])
dittoHeatmap(patient_mean_all_cells,
             genes=met_markers,
             assay = "z-rescaled", 
             cluster_cols = TRUE,
             cluster_rows = FALSE,
             scale="row",
             annot_col=c(""),
             annot.by = c("Disease",
                          "Progression_skin",
                          "Donor", 
                          "Cells_number"),
             annotation_colors = list(Disease = metadata(spe)$color_vectors$Disease,
                                      Donor = metadata(spe)$color_vectors$Donor,
                                      Cells_number = plasma(100)))
dev.off()

# Embed and provide download link
cat(sprintf('<object type="image/svg+xml" data="%s" width="600"></object>\n', plot_filename))
cat(sprintf('<a href="%s" download> Download Plot as SVG</a>\n\n', plot_filename))

for (cl_cat in clinical_cat[1:length(clinical_cat)]){
  
    cat(sprintf("\n\n## Heatmap annotated by %s {.tabset .tabset-pill}\n", cl_cat))
  
  annotation_colors <- list(
    Disease = metadata(spe)$color_vectors$Disease,
    Donor = metadata(spe)$color_vectors$Donor,
    Cells_number = plasma(100)
  )
   
  # Dynamically add the color for the current clinical category with proper naming
  annotation_colors[[cl_cat]] <- metadata(spe)$color_vectors[[cl_cat]]
  
  plot_filename <- file.path(output_dir, paste0(counter,".svg"))
  counter <- counter + 1
  svg(filename = plot_filename, width = fig_dim[1], height = fig_dim[2])
  dittoHeatmap(patient_mean_all_cells, 
             genes=met_markers,
             assay = "z-rescaled", 
             cluster_cols = TRUE,
             cluster_rows = FALSE,
             scale="row",
             annot_col=c(""),
             annot.by = c("Disease",
                          cl_cat,
                          "Donor", 
                          "Cells_number"),
             annotation_colors = annotation_colors)
  dev.off()
  
# Embed and provide download link
cat(sprintf('<object type="image/svg+xml" data="%s" width="600"></object>\n', plot_filename))
cat(sprintf('<a href="%s" download> Download Plot as SVG</a>\n\n', plot_filename))
}

 annotation_colors <- list(
    Disease = metadata(spe)$color_vectors$Disease,
   # Donor = metadata(spe)$color_vectors$Donor,
    Cells_number = plasma(100)
  )
 
 for (i in 2:length(clinical_cat)){
   cl_cat <- clinical_cat[i]
   # Dynamically add the color for the current clinical category with proper naming
   annotation_colors[[cl_cat]] <- metadata(spe)$color_vectors[[cl_cat]]
 }
 
 cat(sprintf("\n\n## Heatmap annotated by %s {.tabset .tabset-pill}\n", "all clinical groups")) 
 
  plot_filename <- file.path(output_dir, paste0(counter,".svg"))
  counter <- counter + 1
  
  svg(filename = plot_filename, width = fig_dim[1], height = fig_dim[2])
dittoHeatmap(patient_mean_all_cells, 
             genes=met_markers,
             assay = "z-rescaled", 
             cluster_cols = TRUE,
             cluster_rows = FALSE,
             scale="row",
             annot_col=c(""),
             annot.by = c("Disease",
                          clinical_cat[2:length(clinical_cat)],
                          #"Donor", 
                          "Cells_number"),
             annotation_colors = annotation_colors)
dev.off()

cat(sprintf('<object type="image/svg+xml" data="%s" width="600"></object>\n', plot_filename))
cat(sprintf('<a href="%s" download> Download Plot as SVG</a>\n\n', plot_filename))
```

# Create CATALYST-compatible spe

```{r CAT spe, eval=TRUE, cache=TRUE}

spe_cat <- spe

spe_cat_not_Epith <- spe_not_Epith

spe_cat$sample_id <- factor(spe$Donor)
spe_cat$condition <- factor(spe$Progression_skin)
spe_cat$cluster_id <- factor(spe$pheno_clusters_new)

spe_cat_not_Epith$sample_id <- factor(spe_not_Epith$Donor)
spe_cat_not_Epith$condition <- factor(spe_not_Epith$Progression_skin)
spe_cat_not_Epith$cluster_id <- factor(spe_not_Epith$pheno_clusters_new)

# Add celltype information to metadata
metadata(spe_cat)$cluster_codes <- data.frame(celltype = factor(spe_cat$pheno_clusters_new))
metadata(spe_cat_not_Epith)$cluster_codes <- data.frame(celltype = factor(spe_cat_not_Epith$pheno_clusters_new))

```

# MDS pseudobulk by patient - all cells except epithelial {.tabset .tabset-pill}

`r fig_dim <- c(10, 7)`

```{r MDS1, eval=TRUE, fig.dim=fig_dim, results='asis'}

metadata(spe_cat)$color_vectors$Progression_skin <- setNames(c("#6CC3F4", "#C03830", "#F6BF93"), unique(spe$Progression_skin))

pbMDS(spe_cat_not_Epith, 
      by = "sample_id", 
      assay = "z-rescaled",
      features = rownames(spe_cat_not_Epith)[rowData(spe_cat_not_Epith)$use_channel_met_markers])+
      #color_by="sample_id",
      #label_by = "sample_id") +
      #k = "celltype")
      #shape_by = "condition") 
     # size_by = TRUE) +
  scale_color_manual(values = metadata(spe_cat)$color_vectors$Progression_skin)

```

# MDS pseudobulk by patient - all cells {.tabset .tabset-pill}

`r fig_dim <- c(10, 7)`

```{r MDS2, eval=TRUE, fig.dim=fig_dim, results='asis'}

pbMDS(spe_cat, 
      by = "sample_id", 
      assay = "z-rescaled",
      features = rownames(spe_cat)[rowData(spe_cat)$use_channel_met_markers])+
      #color_by="sample_id",
      #label_by = "sample_id", 
      #k = "celltype")
      #shape_by = "condition") 
     # size_by = TRUE) #+
 scale_color_manual(values = metadata(spe_cat)$color_vectors$Progression_skin)

```

# Reduced dimension plot on CLR of proportions {.tabset .tabset-pill}

`r fig_dim <- c(10,10)`

```{r clr, eval=TRUE, fig.dim=fig_dim, results='asis'}

# CLR on sample proportions across clusters
clrDR(spe_cat, 
      dr = "PCA", 
      by = "sample_id", 
      label_by = "sample_id",
      point_col = "sample_id",
      arrow_col="cluster_id",
      #shape_by = "condition", 
      k = "celltype",
      point_pal = metadata(spe)$color_vectors$Donor) +
  scale_color_manual(values = metadata(spe)$color_vectors$clusters)

# CLR on cluster proportions across samples
clrDR(spe_cat, 
      dr = "PCA", 
      by = "cluster_id", 
      k = "celltype", 
      label_by = "cluster_id", 
      arrow_col = "sample_id", 
      point_pal = metadata(spe_cat)$color_vectors$clusters) 

```

# Boxplots glycolysis and TCA/ETC scores by disease - all cells except epithelial - ggplot {.tabset .tabset-pill}

`r fig_dim <- c(5, 5)`

```{r plot met scores1, eval=TRUE, fig.width=fig_dim[1], fig.height=fig_dim[2], results='asis'}

metabolic_scores <- c("Glycolysis_z_c","Glycolysis_z", "TCA_z_c", "TCA_z")
source(file.path(dir_r, "get_common_boxplot_components_fx.R"))

for_plot2 <- NULL
boxplot_plot_list <- list()
for(met_scores in metabolic_scores){
  for (cl_cat in clinical_cat){
for_plot2 <- NULL
#my_comparisons <- NULL
for_plot2 <- data.frame(score=colData(spe_not_Epith)[,met_scores],
                        Disease=colData(spe_not_Epith)[,cl_cat],
                        Patient_ID=spe_not_Epith$patient_id)%>%
  group_by(Disease,Patient_ID)%>%
  summarise(score=median(score,na.rm=T))


    disease_levels <- unique(for_plot2$Disease)
    if (length(disease_levels) > 2) {

# Statistics
kruskal_res <- compare_means (score ~ Disease, data=for_plot2,method="kruskal.test") %>% mutate (comparison=cl_cat)
dunn_res <- as.data.frame(for_plot2) %>% rstatix::dunn_test (score ~ Disease, 
                                                             p.adjust.method="none")%>%
  mutate (method="Dunn")

my_comparisons <- combn(unique(for_plot2$Disease), 2, simplify = FALSE)
my_comparisons <- lapply(my_comparisons,as.character)
    
    # Generate the boxplot
    boxplot_plot <- ggplot(data = for_plot2, aes(x = Disease, y = score)) +
      geom_boxplot(outlier.shape = NA, aes(fill=Disease)) +
      stat_pvalue_manual(dunn_res%>%
                           mutate(
                             p_label = case_when(
                               p < 0.001 ~ sprintf("p = %.2e", p),
                               p < 0.01 ~ sprintf("p = %.5f", p),
                               p < 0.1 ~ sprintf("p = %.4f", p),
                               TRUE ~ sprintf("p = %.3f", p)
                             ),
                             y.position=c(max(for_plot2$score+0.25),
                                          max(for_plot2$score+0.5),
                                          max(for_plot2$score+0.25))
                           ),
                         label = "p_label",
                         hide.ns = FALSE,
                         bracket.shorten = 0.05) +
      stat_compare_means(method="kruskal", label.y=max(for_plot2$score +0.75))+
      # stat_compare_means(
      #   comparisons = my_comparisons,
      #   method = "wilcox.test",
      #   aes(label = "p.format"),
      #   hide.ns = FALSE
      # ) +
      get_common_boxplot_components()+
      scale_fill_manual(values = metadata(spe)$color_vectors[[cl_cat]])+ 
      labs(x = "Disease", y = "Score") +
      ggtitle(paste0(met_scores, " score for ", cl_cat))
    
    # Store the plot in the list with a unique name
    plot_name <- paste0(met_scores, "_", cl_cat,"_plot")
    kruskal_name <- paste0(met_scores, "_", cl_cat,"_kruskal")
    dunn_name <- paste0(met_scores, "_", cl_cat,"_dunn")
    boxplot_plot_list[[plot_name]] <- boxplot_plot
    boxplot_plot_list[[kruskal_name]] <- kruskal_res
    boxplot_plot_list[[dunn_name]] <- dunn_res
    
  
    } else {
# Statistics
wilcox_res <- compare_means(score ~ Disease, data=for_plot2, method="wilcox.test")
my_comparisons <- combn(unique(for_plot2$Disease), 2, simplify = FALSE)
my_comparisons <- lapply(my_comparisons,as.character)
    
    # Generate the boxplot
    boxplot_plot <- ggplot(data = for_plot2, aes(x = Disease, y = score)) +
      geom_boxplot(outlier.shape = NA, aes(fill=Disease)) +
      stat_compare_means(comparisons=list(as.character(unique(for_plot2$Disease))),
                           method="wilcox", label.y=max(for_plot2$score+0.25))+
      # stat_compare_means(
      #   comparisons = my_comparisons,
      #   method = "wilcox.test",
      #   aes(label = "p.format"),
      #   hide.ns = FALSE
      # ) +
      get_common_boxplot_components()+
      scale_fill_manual(values = metadata(spe)$color_vectors$Disease)+ 
      labs(x = "Disease", y = "Score") +
      ggtitle(paste0(met_scores, " score for ", cl_cat))
    
     # Store the plot in the list with a unique name
    plot_name <- paste0(met_scores, "_", cl_cat,"_plot")
    wilcox_name <- paste0(met_scores, "_", cl_cat,"_wilcox")
    boxplot_plot_list[[plot_name]] <- boxplot_plot
    boxplot_plot_list[[wilcox_name]] <- wilcox_res
    }

  }
}

for(met_scores in metabolic_scores){
    # create a tab for each metabolic score
  cat(sprintf("\n\n## %s {.tabset .tabset-pill}\n", met_scores))
  # subset the boxplot_plot_list to names of the current metabolic score
  boxplot_plot_list_subset <- boxplot_plot_list[grep(met_scores, names(boxplot_plot_list), value = TRUE)]
  for (cl_cat in clinical_cat){
    # subset the boxplot_plot_list_subset to names of the current clinical category
    boxplot_plot_list_subset_subset <- boxplot_plot_list_subset[grep(paste0(met_scores,"_",cl_cat,"_plot"), names(boxplot_plot_list_subset), value = TRUE)]
    # save the plot as SVG
    svg_filename <- file.path(output_dir, paste0(met_scores, "_", cl_cat,".svg"))
    ggsave(filename = svg_filename, plot = boxplot_plot_list_subset_subset[[1]], device = "svg", width = fig_dim[1], height = fig_dim[2])
    # Embed and provide download link
    cat(sprintf("\n\n### %s\n", cl_cat))
    cat(sprintf('<object type="image/svg+xml" data="%s" width="600"></object>\n', svg_filename))
    cat(sprintf('<a href="%s" download> Download Plot as SVG</a>\n\n', svg_filename))
    }
  }

# Display the plots in R markdown (optional)
# for (plot_name in names(boxplot_plot_list)) {
#   print(boxplot_plot_list[[plot_name]])
# }
#boxplot_plot_list

```

# Boxplots glycolysis and TCA/ETC scores by disease - all cells except epithelial - ggstatsplot {.tabset .tabset-pill}

`r fig_dim <- c(5, 5)`
  
```{r plot met scores2, eval=TRUE, fig.dim=fig_dim, results='asis'}

metabolic_scores <- c("Glycolysis_z_c","Glycolysis_z", "TCA_z_c", "TCA_z")

for_plot3 <- NULL
boxplot_plot_list2 <- list()
for(met_scores in metabolic_scores){
  for (cl_cat in clinical_cat){
for_plot3 <- NULL
#my_comparisons <- NULL
for_plot3 <- data.frame(score=colData(spe_not_Epith)[,met_scores],
                        Disease=colData(spe_not_Epith)[,cl_cat],
                        Patient_ID=spe_not_Epith$patient_id)%>%
  group_by(Disease,Patient_ID)%>%
  summarise(score=median(score,na.rm=T))

    ggstats_boxplot <- ggbetweenstats(data = for_plot3, x = Disease, y = score, 
                              type = "nonparametric", 
                              p.adjust.method = "none",
                              ggplot.component = list(ggplot2::scale_color_manual(values=
                              metadata(spe)$color_vectors[[cl_cat]])), 
                              k = 4, 
                              pairwise.display = "all",
                              centrality.point.args = list(size = 0), 
                              violin.args = list(width = 0)) + 
      ylab('Score')+
      ggtitle(paste0(met_scores, " score for ", cl_cat))
    
     # Store the plot in the list with a unique name
    plot_name2 <- paste0(met_scores, "_", cl_cat,"_plot")
    boxplot_plot_list2[[plot_name2]] <- ggstats_boxplot

  }
}

for(met_scores in metabolic_scores){
    # create a tab for each metabolic score
  cat(sprintf("\n\n## %s {.tabset .tabset-pill}\n", met_scores))
  # subset the boxplot_plot_list to names of the current metabolic score
  boxplot_plot_list_subset2 <- boxplot_plot_list2[grep(met_scores, names(boxplot_plot_list2), value = TRUE)]
  for (cl_cat in clinical_cat){
    # subset the boxplot_plot_list_subset to names of the current clinical category
    boxplot_plot_list_subset_subset2 <- boxplot_plot_list_subset2[grep(paste0(met_scores,"_",cl_cat,"_plot"), names(boxplot_plot_list_subset2), value = TRUE)]
    # save the plot as SVG
    svg_filename <- file.path(output_dir, paste0(met_scores, "_", cl_cat,"_ggstatsplot.svg"))
    ggsave(filename = svg_filename, plot = boxplot_plot_list_subset_subset2[[1]], device = "svg", width = fig_dim[1], height = fig_dim[2])
    # Embed and provide download link
    cat(sprintf("\n\n### %s\n", cl_cat))
    cat(sprintf('<object type="image/svg+xml" data="%s" width="600"></object>\n', svg_filename))
    cat(sprintf('<a href="%s" download> Download Plot as SVG</a>\n\n', svg_filename))
    }
  }

# Display the plots in R markdown (optional)
# for (plot_name in names(boxplot_plot_list2)) {
#   print(boxplot_plot_list2[[plot_name]])
# }
#boxplot_plot_list

```

# Outputs

This analysis generates the following results: <br>

1. Plots of frequencies of clusters, split between healthy and SSc and between healthy, SSc stable and SSc progressive, for the metabolic skin IMC data, as boxplots, radar plots (with ggradar and base r) and ggstatsplot.
The output is in the format: html file, and a subfolder called `r output_dir`. 
Keep this subfolder in the same folder as the html file! <br>

# Session information

```{r session, eval=TRUE}
etime <- proc.time() - ptm
cat('Rendering this document took :', etime[3], ' s \n')

sessioninfo::session_info()

