---
title: "IMC-only figure panels: six matched IMC-Xenium donors"
author: "Alex Matei (with Codex assistance)"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    self_contained: true
    df_print: paged
---

```{r setup, include=FALSE}
source("data_preprocessing/common/config.R")
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 300,
  fig.retina = 2,
  fig.align = "center",
  fig.path = "six_donor_imc_only_outputs/report_figures/"
)
```

```{r libraries}
suppressPackageStartupMessages({
  library(SummarizedExperiment)
  library(SingleCellExperiment)
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(tibble)
  library(patchwork)
  library(knitr)
  library(grid)
})
```

```{r paths_and_constants}
report_dir <- normalizePath(
  dirname(knitr::current_input(dir = TRUE)),
  winslash = "/",
  mustWork = TRUE
)

imc_file <- paste0(
  paste0(project_path("Revision/"), "/"),
  "Nan annotation/Data/IMC spe objects/New IMC/spe2025-12-17 umap.RData"
)

source_scripts <- c(
  paste0(
    paste0(project_path("Revision/"), "/"),
    "Nan annotation/Data/RMD files/IMC heatmap sankey_new.Rmd"
  ),
  paste0(
    paste0(project_path("Revision/"), "/"),
    "Nan annotation/Data/RMD files/IMC protein and scMetabolism scores combined.Rmd"
  ),
  paste0(
    paste0(project_path("Revision/"), "/"),
    "Nan annotation/Data/R Scripts/Heatmap-IMC23.12.2025_Veda.R"
  )
)

output_dir <- file.path(report_dir, "six_donor_imc_only_outputs")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

excluded_sample <- "ExcludedValidation"

sample_to_donor <- c(
  "Validation1" = "Validation1",
  "Validation2" = "Validation2",
  "ExcludedValidation" = "ExcludedValidation",
  "Validation3" = "Validation3",
  "Validation4" = "Validation4",
  "Validation5" = "Validation5",
  "Validation6" = "Validation6"
)

matched_samples <- setdiff(names(sample_to_donor), excluded_sample)
matched_donor_levels <- c("Validation1", "Validation2", "Validation3", "Validation4", "Validation5", "Validation6")

fib_levels <- c("Met_hi_Fib", "Other_Fib")
fib_colors <- c("Met_hi_Fib" = "#F8766D", "Other_Fib" = "#00BFC4")

ec_levels <- c("Met_hi_EC", "Other_EC")
ec_colors <- c("Met_hi_EC" = "#E69F00", "Other_EC" = "#56B4E9")

score_palette <- c("#2C7BB6", "#ABD9E9", "#FFFFBF", "#FDAE61", "#D7191C")
score_limits <- c(-1, 1)
score_breaks <- seq(-1, 1, by = 0.5)
heatmap_palette <- viridisLite::viridis(100)

metabolic_markers <- c(
  "GLUT1", "HK1", "PFKL1_PFKM", "PKM2", "LDHA", "GAPDH",
  "CS", "OGDH", "ATP5A", "SDHA",
  "CD98", "ACAC", "CPT1a", "G6PD", "pAMPK", "p_mTOR",
  "NOX4", "HIF1a", "PGC1a", "mtTFA", "pNRF2"
)

glycolysis_markers <- c("GLUT1", "HK1", "PFKL1_PFKM", "PKM2", "LDHA", "GAPDH")
tca_oxphos_markers <- c("CS", "OGDH", "ATP5A", "SDHA")
protein_markers <- c("aSMA", "Cdh11", "FAP")

fib_umap_name <- "umap_n_neighbors_15_cell_type_Stromal_1"
ec_umap_name <- "umap_n_neighbors_15_cell_type_Endothelial_3"

save_plot_set <- function(plot, basename, width, height) {
  png_file <- file.path(output_dir, paste0(basename, ".png"))
  pdf_file <- file.path(output_dir, paste0(basename, ".pdf"))
  svg_file <- file.path(output_dir, paste0(basename, ".svg"))

  ggsave(
    png_file,
    plot = plot,
    width = width,
    height = height,
    dpi = 300,
    bg = "white"
  )

  grDevices::pdf(
    pdf_file,
    width = width,
    height = height,
    useDingbats = FALSE,
    bg = "white"
  )
  print(plot)
  grDevices::dev.off()

  grDevices::svg(svg_file, width = width, height = height, bg = "white")
  print(plot)
  grDevices::dev.off()

  invisible(c(png_file, pdf_file, svg_file))
}

draw_grob <- function(grob) {
  grid::grid.newpage()
  grid::grid.draw(grob)
}

save_grob_set <- function(grob, basename, width, height) {
  png_file <- file.path(output_dir, paste0(basename, ".png"))
  pdf_file <- file.path(output_dir, paste0(basename, ".pdf"))
  svg_file <- file.path(output_dir, paste0(basename, ".svg"))

  grDevices::png(
    png_file,
    width = width,
    height = height,
    units = "in",
    res = 300,
    bg = "white"
  )
  draw_grob(grob)
  grDevices::dev.off()

  grDevices::pdf(
    pdf_file,
    width = width,
    height = height,
    useDingbats = FALSE,
    bg = "white"
  )
  draw_grob(grob)
  grDevices::dev.off()

  grDevices::svg(svg_file, width = width, height = height, bg = "white")
  draw_grob(grob)
  grDevices::dev.off()

  invisible(c(png_file, pdf_file, svg_file))
}

format_p <- function(p_value) {
  dplyr::case_when(
    is.na(p_value) ~ "p = NA",
    p_value < 0.001 ~ "p < 0.001",
    TRUE ~ paste0("p = ", formatC(p_value, digits = 3, format = "f"))
  )
}

paired_test <- function(data, feature_col, value_col, group_col, group_1, group_2) {
  test_data <- data %>%
    dplyr::transmute(
      donor = as.character(donor),
      feature = as.character(.data[[feature_col]]),
      group = as.character(.data[[group_col]]),
      value = as.numeric(.data[[value_col]])
    )

  dplyr::bind_rows(lapply(unique(test_data$feature), function(current_feature) {
    wide <- test_data %>%
      dplyr::filter(feature == current_feature, group %in% c(group_1, group_2)) %>%
      dplyr::select(donor, group, value) %>%
      tidyr::pivot_wider(names_from = group, values_from = value) %>%
      dplyr::filter(
        is.finite(.data[[group_1]]),
        is.finite(.data[[group_2]])
      )

    test_result <- if (nrow(wide) >= 2) {
      suppressWarnings(stats::wilcox.test(
        wide[[group_1]],
        wide[[group_2]],
        paired = TRUE,
        exact = FALSE
      ))
    } else {
      NULL
    }

    delta <- wide[[group_1]] - wide[[group_2]]

    data.frame(
      feature = current_feature,
      group_1 = group_1,
      group_2 = group_2,
      n_paired_donors = nrow(wide),
      mean_delta_group_1_minus_group_2 = if (length(delta)) mean(delta) else NA_real_,
      median_delta_group_1_minus_group_2 = if (length(delta)) median(delta) else NA_real_,
      p_value = if (is.null(test_result)) NA_real_ else test_result$p.value,
      stringsAsFactors = FALSE
    )
  }))
}

base_umap_theme <- theme_classic(base_size = 8) +
  theme(
    axis.title = element_text(size = 8),
    axis.text = element_text(size = 6.8, color = "grey20"),
    axis.ticks = element_line(linewidth = 0.35),
    axis.line = element_line(linewidth = 0.4),
    legend.title = element_text(size = 6.6),
    legend.text = element_text(size = 6.8),
    legend.key.height = grid::unit(0.25, "cm"),
    legend.key.width = grid::unit(0.28, "cm"),
    legend.spacing.y = grid::unit(0.02, "cm"),
    legend.margin = margin(0, 0, 0, 1),
    plot.title = element_text(
      size = 9,
      face = "bold",
      hjust = 0.5,
      lineheight = 0.95,
      margin = margin(b = 2)
    ),
    plot.subtitle = element_blank(),
    plot.margin = margin(3, 4, 3, 3)
  )

figure_export_manifest <- data.frame(
  panel = c(
    "FibMET UMAP",
    "FibMET IMC score UMAPs",
    "FibMET protein expression",
    "FibMET metabolic heatmap",
    "FibMET donor score comparison",
    "EndMET UMAP",
    "EndMET metabolic heatmap"
  ),
  width_in = c(4.2, 6.8, 6.8, 4.8, 5.2, 4.2, 4.8),
  height_in = c(3.6, 4.0, 3.2, 3.35, 3.7, 3.6, 3.35),
  figure_ratio = c(4.2 / 3.6, 6.8 / 4.0, 6.8 / 3.2, 4.8 / 3.35, 5.2 / 3.7, 4.2 / 3.6, 4.8 / 3.35),
  color_scale = c(
    "Original FibMET salmon/cyan",
    "Original blue-yellow-red; fixed [-1, 1]",
    "Original FibMET salmon/cyan",
    "Viridis; fixed [-1, 0.5]",
    "Original FibMET salmon/cyan",
    "Original EndMET orange/blue",
    "Viridis; fixed [-1, 1]"
  ),
  stringsAsFactors = FALSE
)

write.csv(
  figure_export_manifest,
  file.path(output_dir, "figure_export_manifest.csv"),
  row.names = FALSE
)
```

```{r load_and_filter}
if (!file.exists(imc_file)) {
  stop("The annotated IMC SPE file was not found: ", imc_file)
}

imc_env <- new.env(parent = emptyenv())
load(imc_file, envir = imc_env)

if (!exists("spe", envir = imc_env, inherits = FALSE)) {
  stop("The IMC RData file does not contain an object named 'spe'.")
}

spe_all <- imc_env[["spe"]]
required_coldata <- c("sample_id", "gates", "Rphenograph_500", "Rphenograph_200EC")
missing_coldata <- setdiff(required_coldata, colnames(colData(spe_all)))
if (length(missing_coldata) > 0) {
  stop("Missing required colData fields: ", paste(missing_coldata, collapse = ", "))
}

if (!all(c(fib_umap_name, ec_umap_name) %in% reducedDimNames(spe_all))) {
  stop("The required finalized FibMET or EndMET UMAP embedding is missing.")
}

object_samples <- sort(unique(as.character(spe_all$sample_id)))
unexpected_samples <- setdiff(object_samples, names(sample_to_donor))
missing_samples <- setdiff(names(sample_to_donor), object_samples)

if (length(unexpected_samples) > 0 || length(missing_samples) > 0) {
  stop(
    "The annotated IMC object does not match the expected seven-sample manifest. ",
    "Unexpected: ", paste(unexpected_samples, collapse = ", "),
    "; missing: ", paste(missing_samples, collapse = ", ")
  )
}

keep_six <- as.character(spe_all$sample_id) %in% matched_samples
spe6 <- spe_all[, keep_six]

if (ncol(spe6) == 0 || length(unique(as.character(spe6$sample_id))) != 6) {
  stop("Six-donor filtering did not produce exactly six samples.")
}

spe6$donor <- unname(sample_to_donor[as.character(spe6$sample_id)])
spe6$donor <- factor(as.character(spe6$donor), levels = matched_donor_levels)

spe6$metfiblabel <- dplyr::case_when(
  as.character(spe6$Rphenograph_500) %in% c("1", "2") ~ "Met_hi_Fib",
  as.character(spe6$Rphenograph_500) == "3" ~ "Other_Fib",
  TRUE ~ NA_character_
)

spe6$metEClabel <- dplyr::case_when(
  as.character(spe6$Rphenograph_200EC) %in% c("1", "2") ~ "Met_hi_EC",
  as.character(spe6$Rphenograph_200EC) == "3" ~ "Other_EC",
  TRUE ~ NA_character_
)

asinh6 <- as.matrix(assay(spe6, "asinh"))
row_mu6 <- rowMeans(asinh6, na.rm = TRUE)
row_sd6 <- apply(asinh6, 1, stats::sd, na.rm = TRUE)
row_sd6[!is.finite(row_sd6) | row_sd6 == 0] <- 1

z6 <- sweep(sweep(asinh6, 1, row_mu6, "-"), 1, row_sd6, "/")
assay(spe6, "z_six_donor") <- z6

required_markers <- unique(c(metabolic_markers, protein_markers))
marker_manifest <- data.frame(
  marker = required_markers,
  available = required_markers %in% rownames(spe6),
  stringsAsFactors = FALSE
)

if (!all(marker_manifest$available)) {
  stop(
    "Required IMC markers are absent: ",
    paste(marker_manifest$marker[!marker_manifest$available], collapse = ", ")
  )
}

write.csv(marker_manifest, file.path(output_dir, "marker_availability.csv"), row.names = FALSE)
```

# Cohort audit

The six matched donors are retained from the finalized annotated post-Xenium
IMC object. `ExcludedValidation` is excluded before every plotted-cell selection,
six-donor z-score calculation, group summary, and donor-level test.

```{r cohort_audit}
sample_counts_all <- as.data.frame(table(as.character(spe_all$sample_id))) %>%
  dplyr::rename(sample_id = Var1, n_all_cells = Freq) %>%
  dplyr::mutate(
    donor = unname(sample_to_donor[as.character(sample_id)]),
    analysis_status = ifelse(sample_id == excluded_sample, "Excluded", "Included")
  ) %>%
  dplyr::arrange(factor(donor, levels = c(matched_donor_levels, "ExcludedValidation")))

gate_counts_six <- as.data.frame(table(
  sample_id = as.character(spe6$sample_id),
  gate = as.character(spe6$gates)
)) %>%
  dplyr::filter(Freq > 0) %>%
  dplyr::mutate(donor = unname(sample_to_donor[as.character(sample_id)])) %>%
  dplyr::rename(n_cells = Freq)

panel_counts <- dplyr::bind_rows(
  as.data.frame(table(
    group = factor(spe6$metfiblabel, levels = fib_levels)
  )) %>%
    dplyr::mutate(panel = "FibMET", .before = 1),
  as.data.frame(table(
    group = factor(spe6$metEClabel, levels = ec_levels)
  )) %>%
    dplyr::mutate(panel = "EndMET", .before = 1)
) %>%
  dplyr::filter(Freq > 0) %>%
  dplyr::rename(n_cells = Freq)

write.csv(sample_counts_all, file.path(output_dir, "donor_manifest.csv"), row.names = FALSE)
write.csv(gate_counts_six, file.path(output_dir, "six_donor_gate_counts.csv"), row.names = FALSE)
write.csv(panel_counts, file.path(output_dir, "panel_cell_counts.csv"), row.names = FALSE)

kable(
  sample_counts_all,
  caption = "Post-Xenium IMC donor manifest",
  align = c("l", "r", "l", "l")
)
```

```{r panel_counts_table}
kable(
  panel_counts,
  caption = "Cells contributing to the reproduced IMC panels",
  align = c("l", "l", "r")
)
```

**Scope note.** The five-state aSMA supplementary UMAP and heatmap in the
first supplied figure were generated from the separate reference-cohort
object (`spe_aSMA.rds`; 9 SSc and 7 healthy donors). Those six post-Xenium
donors do not exist in that object. The reference-cohort panels are therefore
not regenerated here.

# Fibroblasts {.tabset .tabset-pill}

## FibMET UMAP

```{r prepare_fib_umap}
fib_umap_matrix <- reducedDim(spe6, fib_umap_name)
fib_umap <- data.frame(
  cell = colnames(spe6),
  UMAP1 = fib_umap_matrix[, 1],
  UMAP2 = fib_umap_matrix[, 2],
  sample_id = as.character(spe6$sample_id),
  donor = as.character(spe6$donor),
  metfiblabel = as.character(spe6$metfiblabel),
  stringsAsFactors = FALSE
) %>%
  dplyr::filter(
    complete.cases(UMAP1, UMAP2),
    !is.na(metfiblabel),
    metfiblabel %in% fib_levels
  ) %>%
  dplyr::mutate(
    donor = factor(donor, levels = matched_donor_levels),
    metfiblabel = factor(metfiblabel, levels = fib_levels)
  )

if (nrow(fib_umap) != sum(!is.na(spe6$metfiblabel))) {
  stop("FibMET UMAP cell count does not match the finalized FibMET labels.")
}

write.csv(
  fib_umap,
  file.path(output_dir, "fibmet_umap_cells_six_donors.csv"),
  row.names = FALSE
)

p_fib_umap <- ggplot(
  fib_umap,
  aes(x = UMAP1, y = UMAP2, color = metfiblabel)
) +
  geom_point(size = 0.55, alpha = 0.9, stroke = 0) +
  scale_color_manual(values = fib_colors, drop = FALSE) +
  labs(
    title = "Metabolic phenotypes of fibroblasts (FibMET)",
    x = "UMAP1",
    y = "UMAP2",
    color = NULL
  ) +
  base_umap_theme +
  guides(
    color = guide_legend(
      override.aes = list(size = 2.2, alpha = 1),
      keyheight = grid::unit(0.28, "cm")
    )
  ) +
  theme(legend.position = "right")

save_plot_set(p_fib_umap, "fibmet_umap_six_donors", width = 4.2, height = 3.6)
```

```{r show_fib_umap, echo=FALSE, fig.width=4.2, fig.height=3.6}
p_fib_umap
```

## IMC protein metabolic scores

```{r prepare_fib_scores}
glycolysis_score <- colMeans(z6[glycolysis_markers, , drop = FALSE], na.rm = TRUE)
tca_oxphos_score <- colMeans(z6[tca_oxphos_markers, , drop = FALSE], na.rm = TRUE)
names(glycolysis_score) <- colnames(spe6)
names(tca_oxphos_score) <- colnames(spe6)

fib_umap_scores <- fib_umap %>%
  dplyr::mutate(
    glycolysis_score = unname(glycolysis_score[cell]),
    tca_oxphos_score = unname(tca_oxphos_score[cell])
  )

make_score_umap <- function(data, score_col, title) {
  ggplot(data, aes(x = UMAP1, y = UMAP2, color = .data[[score_col]])) +
    geom_point(size = 0.52, alpha = 0.92, stroke = 0) +
    scale_color_gradientn(
      colors = score_palette,
      limits = score_limits,
      breaks = score_breaks,
      oob = scales::squish,
      name = "Score",
      guide = guide_colorbar(
        barwidth = grid::unit(0.18, "cm"),
        barheight = grid::unit(0.95, "cm"),
        title.position = "top",
        title.hjust = 0.5
      )
    ) +
    labs(title = title, x = "UMAP1", y = "UMAP2") +
    base_umap_theme +
    theme(
      legend.position = "right",
      legend.title = element_text(size = 6.3, hjust = 0.5),
      legend.text = element_text(size = 5.9),
      legend.margin = margin(0, 0, 0, 0),
      plot.margin = margin(3, 2, 3, 2)
    )
}

p_fib_glycolysis <- make_score_umap(
  fib_umap_scores,
  "glycolysis_score",
  "Glycolysis score"
)

p_fib_tca <- make_score_umap(
  fib_umap_scores,
  "tca_oxphos_score",
  "TCA/OXPHOS score"
)

p_fib_scores <- (
  p_fib_glycolysis + p_fib_tca +
    patchwork::plot_layout(guides = "keep") +
    patchwork::plot_annotation(
      title = "Metabolic scores",
      theme = theme(
        plot.title = element_text(
          size = 9,
          face = "bold",
          hjust = 0.5,
          margin = margin(b = 1)
        )
      )
    )
)

write.csv(
  fib_umap_scores,
  file.path(output_dir, "fibmet_umap_imc_protein_scores_six_donors.csv"),
  row.names = FALSE
)

save_plot_set(
  p_fib_scores,
  "fibmet_imc_protein_score_umaps_six_donors",
  width = 6.8,
  height = 4.0
)
```

```{r show_fib_scores, echo=FALSE, fig.width=6.8, fig.height=4}
p_fib_scores
```

## Protein expression

```{r prepare_fib_protein_expression}
fib_index <- match(fib_umap$cell, colnames(spe6))

fib_protein_cell <- as.data.frame(
  t(z6[protein_markers, fib_index, drop = FALSE]),
  check.names = FALSE
) %>%
  tibble::rownames_to_column("cell") %>%
  dplyr::mutate(
    sample_id = fib_umap$sample_id[match(cell, fib_umap$cell)],
    donor = fib_umap$donor[match(cell, fib_umap$cell)],
    metfiblabel = as.character(fib_umap$metfiblabel[match(cell, fib_umap$cell)])
  ) %>%
  tidyr::pivot_longer(
    cols = dplyr::all_of(protein_markers),
    names_to = "marker",
    values_to = "expression_z"
  )

fib_protein_sample <- fib_protein_cell %>%
  dplyr::group_by(sample_id, donor, metfiblabel, marker) %>%
  dplyr::summarise(
    n_cells = dplyr::n(),
    mean_expression_z = mean(expression_z, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    donor = factor(donor, levels = matched_donor_levels),
    metfiblabel = factor(metfiblabel, levels = fib_levels),
    marker = factor(marker, levels = protein_markers)
  )

fib_protein_stats <- paired_test(
  fib_protein_sample,
  feature_col = "marker",
  value_col = "mean_expression_z",
  group_col = "metfiblabel",
  group_1 = "Met_hi_Fib",
  group_2 = "Other_Fib"
) %>%
  dplyr::mutate(marker = factor(feature, levels = protein_markers))

fib_protein_stat_positions <- fib_protein_sample %>%
  dplyr::group_by(marker) %>%
  dplyr::summarise(
    ymax = max(mean_expression_z, na.rm = TRUE),
    spread = diff(range(mean_expression_z, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    spread = ifelse(spread > 0, spread, 0.2),
    label_y = ymax + 0.17 * spread
  ) %>%
  dplyr::left_join(
    fib_protein_stats %>%
      dplyr::select(marker, p_value, n_paired_donors),
    by = "marker"
  ) %>%
  dplyr::mutate(label = format_p(p_value))

p_fib_protein <- ggplot(
  fib_protein_sample,
  aes(x = metfiblabel, y = mean_expression_z, fill = metfiblabel)
) +
  geom_violin(
    width = 0.78,
    alpha = 0.76,
    color = "grey25",
    linewidth = 0.28,
    trim = TRUE
  ) +
  geom_boxplot(
    width = 0.14,
    outlier.shape = NA,
    fill = "white",
    color = "grey20",
    linewidth = 0.28
  ) +
  geom_point(
    aes(group = donor),
    position = position_jitter(width = 0.045, height = 0),
    size = 1.05,
    color = "black",
    alpha = 0.9
  ) +
  geom_text(
    data = fib_protein_stat_positions,
    aes(x = 1.5, y = label_y, label = label),
    inherit.aes = FALSE,
    size = 2.05
  ) +
  scale_fill_manual(values = fib_colors, drop = FALSE) +
  facet_wrap(~ marker, nrow = 1, scales = "free_y") +
  labs(
    title = "Protein expression",
    x = NULL,
    y = "Mean expression z score"
  ) +
  theme_bw(base_size = 7.5) +
  theme(
    axis.text.x = element_text(size = 6.6, angle = 42, hjust = 1),
    axis.text.y = element_text(size = 6.2),
    axis.title.y = element_text(size = 7.4),
    axis.ticks = element_line(linewidth = 0.3),
    panel.border = element_rect(fill = NA, color = "grey25", linewidth = 0.35),
    panel.grid.major = element_line(color = "grey90", linewidth = 0.25),
    panel.grid.minor = element_line(color = "grey95", linewidth = 0.2),
    strip.background = element_blank(),
    strip.text = element_text(size = 8, face = "bold", margin = margin(b = 1)),
    legend.position = "none",
    plot.title = element_text(size = 9.5, face = "bold", hjust = 0.5, margin = margin(b = 2)),
    plot.margin = margin(3, 3, 4, 3)
  )

write.csv(
  fib_protein_sample,
  file.path(output_dir, "fibmet_protein_expression_sample_means_six_donors.csv"),
  row.names = FALSE
)
write.csv(
  fib_protein_stats,
  file.path(output_dir, "fibmet_protein_expression_paired_wilcoxon_six_donors.csv"),
  row.names = FALSE
)

save_plot_set(
  p_fib_protein,
  "fibmet_protein_expression_six_donors",
  width = 6.8,
  height = 3.2
)
```

```{r show_fib_protein, echo=FALSE, fig.width=6.8, fig.height=3.2}
p_fib_protein
```

## Metabolic-marker heatmap

```{r prepare_fib_heatmap}
fib_marker_cell <- as.data.frame(
  t(z6[metabolic_markers, fib_index, drop = FALSE]),
  check.names = FALSE
) %>%
  tibble::rownames_to_column("cell") %>%
  dplyr::mutate(
    metfiblabel = as.character(fib_umap$metfiblabel[match(cell, fib_umap$cell)])
  ) %>%
  tidyr::pivot_longer(
    cols = dplyr::all_of(metabolic_markers),
    names_to = "marker",
    values_to = "expression_z"
  )

fib_group_counts <- fib_umap %>%
  dplyr::count(metfiblabel, name = "n_cells") %>%
  dplyr::mutate(metfiblabel = as.character(metfiblabel))

fib_heatmap_data <- fib_marker_cell %>%
  dplyr::group_by(metfiblabel, marker) %>%
  dplyr::summarise(
    mean_expression_z = mean(expression_z, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::left_join(fib_group_counts, by = "metfiblabel") %>%
  dplyr::mutate(
    metfiblabel = factor(metfiblabel, levels = fib_levels),
    marker = factor(marker, levels = rev(metabolic_markers))
  )

fib_x_labels <- fib_group_counts %>%
  dplyr::mutate(label = paste0(metfiblabel, "\n", "n = ", n_cells)) %>%
  dplyr::select(metfiblabel, label) %>%
  tibble::deframe()

fib_heatmap_matrix_df <- fib_heatmap_data %>%
  dplyr::transmute(
    marker = as.character(marker),
    metfiblabel = as.character(metfiblabel),
    mean_expression_z
  ) %>%
  tidyr::pivot_wider(
    names_from = metfiblabel,
    values_from = mean_expression_z
  )

fib_heatmap_matrix_df <- fib_heatmap_matrix_df[
  match(metabolic_markers, fib_heatmap_matrix_df$marker),
  ,
  drop = FALSE
]

fib_heatmap_matrix <- as.matrix(
  fib_heatmap_matrix_df[, fib_levels, drop = FALSE]
)
rownames(fib_heatmap_matrix) <- fib_heatmap_matrix_df$marker

fib_ncells <- stats::setNames(
  fib_group_counts$n_cells,
  fib_group_counts$metfiblabel
)

fib_heatmap_annotation <- data.frame(
  ids = factor(fib_levels, levels = fib_levels),
  ncells = as.numeric(fib_ncells[fib_levels]),
  row.names = fib_levels,
  check.names = FALSE
)

fib_heatmap_annotation_colors <- list(
  ncells = grDevices::colorRampPalette(c("#F7F0F5", "#CC79A7"))(100),
  ids = fib_colors
)

fib_heatmap_object <- pheatmap::pheatmap(
  fib_heatmap_matrix,
  color = heatmap_palette,
  breaks = seq(-1, 0.5, length.out = length(heatmap_palette) + 1),
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = fib_heatmap_annotation,
  annotation_colors = fib_heatmap_annotation_colors,
  show_colnames = FALSE,
  show_rownames = TRUE,
  border_color = NA,
  legend = TRUE,
  legend_breaks = c(-1, -0.5, 0, 0.5),
  legend_labels = c("-1.0", "-0.5", "0.0", "0.5"),
  annotation_legend = TRUE,
  annotation_names_col = TRUE,
  fontsize = 7,
  fontsize_row = 6.2,
  fontsize_col = 6.2,
  treeheight_row = 0,
  treeheight_col = 0,
  silent = TRUE
)

p_fib_heatmap <- fib_heatmap_object$gtable

write.csv(
  fib_heatmap_data,
  file.path(output_dir, "fibmet_marker_group_means_six_donors.csv"),
  row.names = FALSE
)

save_grob_set(
  p_fib_heatmap,
  "fibmet_metabolic_marker_heatmap_six_donors",
  width = 4.8,
  height = 3.35
)
```

```{r show_fib_heatmap, echo=FALSE, fig.width=4.8, fig.height=3.35}
draw_grob(p_fib_heatmap)
```

## Six-donor metabolic score comparison

```{r prepare_fib_sample_scores}
fib_score_cell <- fib_umap_scores %>%
  dplyr::select(
    cell,
    sample_id,
    donor,
    metfiblabel,
    Glycolysis = glycolysis_score,
    `TCA/OXPHOS` = tca_oxphos_score
  ) %>%
  tidyr::pivot_longer(
    cols = c("Glycolysis", "TCA/OXPHOS"),
    names_to = "pathway",
    values_to = "score"
  ) %>%
  dplyr::mutate(
    pathway = factor(pathway, levels = c("Glycolysis", "TCA/OXPHOS")),
    metfiblabel = factor(metfiblabel, levels = fib_levels)
  )

fib_score_sample <- fib_score_cell %>%
  dplyr::group_by(sample_id, donor, metfiblabel, pathway) %>%
  dplyr::summarise(
    n_cells = dplyr::n(),
    mean_score = mean(score, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    donor = factor(donor, levels = matched_donor_levels),
    metfiblabel = factor(metfiblabel, levels = fib_levels),
    pathway = factor(pathway, levels = c("Glycolysis", "TCA/OXPHOS"))
  )

fib_score_stats <- paired_test(
  fib_score_sample,
  feature_col = "pathway",
  value_col = "mean_score",
  group_col = "metfiblabel",
  group_1 = "Met_hi_Fib",
  group_2 = "Other_Fib"
) %>%
  dplyr::mutate(pathway = factor(feature, levels = c("Glycolysis", "TCA/OXPHOS")))

fib_score_stat_positions <- fib_score_sample %>%
  dplyr::group_by(pathway) %>%
  dplyr::summarise(
    ymax = max(mean_score, na.rm = TRUE),
    spread = diff(range(mean_score, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    spread = ifelse(spread > 0, spread, 0.2),
    label_y = ymax + 0.16 * spread
  ) %>%
  dplyr::left_join(
    fib_score_stats %>% dplyr::select(pathway, p_value, n_paired_donors),
    by = "pathway"
  ) %>%
  dplyr::mutate(label = format_p(p_value))

make_fib_score_box <- function(current_pathway) {
  panel_data <- fib_score_sample %>%
    dplyr::filter(as.character(pathway) == current_pathway)

  panel_stat <- fib_score_stat_positions %>%
    dplyr::filter(as.character(pathway) == current_pathway)

  ggplot(
    panel_data,
    aes(x = metfiblabel, y = mean_score, fill = metfiblabel)
  ) +
    geom_boxplot(
      width = 0.52,
      outlier.shape = NA,
      alpha = 0.7,
      color = "grey30",
      linewidth = 0.3
    ) +
    geom_point(
      position = position_jitter(width = 0.045, height = 0),
      size = 1.15,
      color = "black",
      alpha = 0.9
    ) +
    geom_text(
      data = panel_stat,
      aes(x = 1.5, y = label_y, label = label),
      inherit.aes = FALSE,
      size = 2.15
    ) +
    scale_fill_manual(values = fib_colors, drop = FALSE) +
    labs(
      title = current_pathway,
      x = NULL,
      y = "Mean IMC protein score"
    ) +
    theme_bw(base_size = 7.5) +
    theme(
      axis.text.x = element_text(size = 6.6, angle = 20, hjust = 1),
      axis.text.y = element_text(size = 6.2),
      axis.title.y = element_text(size = 7),
      axis.ticks = element_line(linewidth = 0.3),
      panel.border = element_rect(fill = NA, color = "grey25", linewidth = 0.35),
      panel.grid.major = element_line(color = "grey90", linewidth = 0.25),
      panel.grid.minor = element_line(color = "grey95", linewidth = 0.2),
      legend.position = "none",
      plot.title = element_text(size = 8.5, face = "bold", hjust = 0.5, margin = margin(b = 1)),
      plot.margin = margin(2, 3, 3, 3)
    )
}

p_fib_sample_glycolysis <- make_fib_score_box("Glycolysis")
p_fib_sample_tca <- make_fib_score_box("TCA/OXPHOS")

p_fib_sample_scores <- (
  p_fib_sample_glycolysis /
    p_fib_sample_tca +
    patchwork::plot_layout(heights = c(1, 1))
)

write.csv(
  fib_score_cell,
  file.path(output_dir, "fibmet_imc_protein_score_cells_six_donors.csv"),
  row.names = FALSE
)
write.csv(
  fib_score_sample,
  file.path(output_dir, "fibmet_imc_protein_score_sample_means_six_donors.csv"),
  row.names = FALSE
)
write.csv(
  fib_score_stats,
  file.path(output_dir, "fibmet_imc_protein_score_paired_wilcoxon_six_donors.csv"),
  row.names = FALSE
)

save_plot_set(
  p_fib_sample_scores,
  "fibmet_imc_protein_score_sample_comparison_six_donors",
  width = 5.2,
  height = 3.7
)
```

```{r show_fib_sample_scores, echo=FALSE, fig.width=5.2, fig.height=3.7}
p_fib_sample_scores
```

# Endothelial cells {.tabset .tabset-pill}

## EndMET UMAP

```{r prepare_ec_umap}
ec_umap_matrix <- reducedDim(spe6, ec_umap_name)
ec_umap <- data.frame(
  cell = colnames(spe6),
  UMAP1 = ec_umap_matrix[, 1],
  UMAP2 = ec_umap_matrix[, 2],
  sample_id = as.character(spe6$sample_id),
  donor = as.character(spe6$donor),
  metEClabel = as.character(spe6$metEClabel),
  stringsAsFactors = FALSE
) %>%
  dplyr::filter(
    complete.cases(UMAP1, UMAP2),
    !is.na(metEClabel),
    metEClabel %in% ec_levels
  ) %>%
  dplyr::mutate(
    donor = factor(donor, levels = matched_donor_levels),
    metEClabel = factor(metEClabel, levels = ec_levels)
  )

if (nrow(ec_umap) != sum(!is.na(spe6$metEClabel))) {
  stop("EndMET UMAP cell count does not match the finalized EndMET labels.")
}

write.csv(
  ec_umap,
  file.path(output_dir, "endmet_umap_cells_six_donors.csv"),
  row.names = FALSE
)

p_ec_umap <- ggplot(
  ec_umap,
  aes(x = UMAP1, y = UMAP2, color = metEClabel)
) +
  geom_point(size = 0.68, alpha = 0.92, stroke = 0) +
  scale_color_manual(values = ec_colors, drop = FALSE) +
  labs(
    title = "Metabolic phenotypes of\nendothelial cells (End-Met)",
    x = "UMAP_EC_1",
    y = "UMAP_EC_2",
    color = NULL
  ) +
  base_umap_theme +
  guides(
    color = guide_legend(
      override.aes = list(size = 2.2, alpha = 1),
      keyheight = grid::unit(0.28, "cm")
    )
  ) +
  theme(legend.position = "right")

save_plot_set(p_ec_umap, "endmet_umap_six_donors", width = 4.2, height = 3.6)
```

```{r show_ec_umap, echo=FALSE, fig.width=4.2, fig.height=3.6}
p_ec_umap
```

## Metabolic-marker heatmap

```{r prepare_ec_heatmap}
ec_index <- match(ec_umap$cell, colnames(spe6))

ec_marker_cell <- as.data.frame(
  t(z6[metabolic_markers, ec_index, drop = FALSE]),
  check.names = FALSE
) %>%
  tibble::rownames_to_column("cell") %>%
  dplyr::mutate(
    metEClabel = as.character(ec_umap$metEClabel[match(cell, ec_umap$cell)])
  ) %>%
  tidyr::pivot_longer(
    cols = dplyr::all_of(metabolic_markers),
    names_to = "marker",
    values_to = "expression_z"
  )

ec_group_counts <- ec_umap %>%
  dplyr::count(metEClabel, name = "n_cells") %>%
  dplyr::mutate(metEClabel = as.character(metEClabel))

ec_heatmap_data <- ec_marker_cell %>%
  dplyr::group_by(metEClabel, marker) %>%
  dplyr::summarise(
    mean_expression_z = mean(expression_z, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::left_join(ec_group_counts, by = "metEClabel") %>%
  dplyr::mutate(
    metEClabel = factor(metEClabel, levels = ec_levels),
    marker = factor(marker, levels = rev(metabolic_markers))
  )

ec_x_labels <- ec_group_counts %>%
  dplyr::mutate(label = paste0(metEClabel, "\n", "n = ", n_cells)) %>%
  dplyr::select(metEClabel, label) %>%
  tibble::deframe()

ec_heatmap_matrix_df <- ec_heatmap_data %>%
  dplyr::transmute(
    marker = as.character(marker),
    metEClabel = as.character(metEClabel),
    mean_expression_z
  ) %>%
  tidyr::pivot_wider(
    names_from = metEClabel,
    values_from = mean_expression_z
  )

ec_heatmap_matrix_df <- ec_heatmap_matrix_df[
  match(metabolic_markers, ec_heatmap_matrix_df$marker),
  ,
  drop = FALSE
]

ec_heatmap_matrix <- as.matrix(
  ec_heatmap_matrix_df[, ec_levels, drop = FALSE]
)
rownames(ec_heatmap_matrix) <- ec_heatmap_matrix_df$marker

ec_ncells <- stats::setNames(
  ec_group_counts$n_cells,
  ec_group_counts$metEClabel
)

ec_heatmap_annotation <- data.frame(
  ids = factor(ec_levels, levels = ec_levels),
  ncells = as.numeric(ec_ncells[ec_levels]),
  row.names = ec_levels,
  check.names = FALSE
)

ec_heatmap_annotation_colors <- list(
  ncells = grDevices::colorRampPalette(c("#F7F0F5", "#CC79A7"))(100),
  ids = ec_colors
)

ec_heatmap_object <- pheatmap::pheatmap(
  ec_heatmap_matrix,
  color = heatmap_palette,
  breaks = seq(-1, 1, length.out = length(heatmap_palette) + 1),
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  annotation_col = ec_heatmap_annotation,
  annotation_colors = ec_heatmap_annotation_colors,
  show_colnames = FALSE,
  show_rownames = TRUE,
  border_color = NA,
  legend = TRUE,
  legend_breaks = c(-1, -0.5, 0, 0.5, 1),
  legend_labels = c("-1.0", "-0.5", "0.0", "0.5", "1.0"),
  annotation_legend = TRUE,
  annotation_names_col = TRUE,
  fontsize = 7,
  fontsize_row = 6.2,
  fontsize_col = 6.2,
  treeheight_row = 0,
  treeheight_col = 0,
  silent = TRUE
)

p_ec_heatmap <- ec_heatmap_object$gtable

write.csv(
  ec_heatmap_data,
  file.path(output_dir, "endmet_marker_group_means_six_donors.csv"),
  row.names = FALSE
)

save_grob_set(
  p_ec_heatmap,
  "endmet_metabolic_marker_heatmap_six_donors",
  width = 4.8,
  height = 3.35
)
```

```{r show_ec_heatmap, echo=FALSE, fig.width=4.8, fig.height=3.35}
draw_grob(p_ec_heatmap)
```

# Methods and provenance

- The annotated post-Xenium IMC object supplies the finalized
  `Rphenograph_500`, `Rphenograph_200EC`, FibMET UMAP, and EndMET UMAP.
- `ExcludedValidation` (`ExcludedValidation`) is excluded before all
  calculations shown in this report.
- Cluster assignments and stored UMAP coordinates are retained so this is a
  figure-only cohort restriction, not a new clustering analysis.
- Marker-wise z scores are recalculated from the IMC `asinh` assay using only
  the six retained donors.
- Glycolysis is the mean of GLUT1, HK1, PFKL1_PFKM, PKM2, LDHA, and GAPDH.
- TCA/OXPHOS is the mean of CS, OGDH, ATP5A, and SDHA.
- Donor-level comparisons use paired Wilcoxon tests on the six donor means.
- Figure-ready exports preserve the original categorical palettes and compact
  panel ratios. IMC score UMAPs are clipped to `[-1, 1]`; the FibMET heatmap
  is clipped to `[-1, 0.5]`, and the EndMET heatmap is clipped to `[-1, 1]`.

```{r provenance_table}
provenance <- data.frame(
  item = c("Annotated IMC object", paste0("Source script ", seq_along(source_scripts))),
  path = c(imc_file, source_scripts),
  exists = file.exists(c(imc_file, source_scripts)),
  stringsAsFactors = FALSE
)

write.csv(provenance, file.path(output_dir, "source_provenance.csv"), row.names = FALSE)
kable(provenance, caption = "Source provenance")
```
