---
title: "IMC cell-density control analysis"
subtitle: "Reviewer response for metabolic profiles around Met_hi_Fib"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    number_sections: false
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
report_dir <- normalizePath(
  dirname(knitr::current_input(dir = TRUE)),
  winslash = "/",
  mustWork = FALSE
)
output_dir <- file.path(report_dir, "outputs")
plot_dir <- file.path(output_dir, "plots")
table_dir <- file.path(output_dir, "tables")

donor_scores <- read.csv(
  file.path(table_dir, "donor_met_hi_fib_density_and_scores.csv"),
  check.names = FALSE
)
donor_counts <- read.csv(
  file.path(table_dir, "donor_analyzed_cell_counts_and_density.csv"),
  check.names = FALSE
)
primary_cor <- read.csv(
  file.path(table_dir, "primary_donor_level_spearman_correlations.csv"),
  check.names = FALSE
)
threshold_cor <- read.csv(
  file.path(table_dir, "minimum_met_hi_cell_threshold_sensitivity.csv"),
  check.names = FALSE
)
ssc_cor <- read.csv(
  file.path(table_dir, "ssc_only_spearman_correlations.csv"),
  check.names = FALSE
)
count_tests <- read.csv(
  file.path(table_dir, "cell_count_group_kruskal_tests.csv"),
  check.names = FALSE
)
local_tests <- read.csv(
  file.path(table_dir, "local_density_group_kruskal_tests.csv"),
  check.names = FALSE
)
graph_qc <- read.csv(
  file.path(table_dir, "delaunay_graph_qc.csv"),
  check.names = FALSE
)
source_manifest <- read.csv(
  file.path(table_dir, "source_manifest.csv"),
  check.names = FALSE
)

pretty_metric <- function(x) {
  x <- sub(
    "mean_delaunay_neighbors_80um",
    "Mean Delaunay neighbor count",
    x,
    fixed = TRUE
  )
  x <- sub(
    "mean_distance_5nn_um",
    "Mean five-nearest-cell distance",
    x,
    fixed = TRUE
  )
  x <- sub(
    "mean_glycolysis_score",
    "Glycolysis score",
    x,
    fixed = TRUE
  )
  sub(
    "mean_tca_oxphos_score",
    "TCA/OXPHOS score",
    x,
    fixed = TRUE
  )
}

figure_link <- function(stem, alt) {
  png <- file.path("outputs", "plots", paste0(stem, ".png"))
  pdf <- file.path("outputs", "plots", paste0(stem, ".pdf"))
  svg <- file.path("outputs", "plots", paste0(stem, ".svg"))
  cat(sprintf(
    "<p><a href='%s'>PDF</a> | <a href='%s'>SVG</a></p><img src='%s' alt='%s' style='width:100%%;height:auto;'>",
    pdf, svg, png, alt
  ))
}
```

# Analysis design

The finalized first-cohort IMC `SpatialExperiment` was used without
reclustering or recalculation of metabolic scores. `Met_hi_Fib` corresponds
to the established `stromal_2` state. The existing `Glycolysis_z_c` and
`TCA_z_c` columns were retained as the glycolysis and TCA/OXPHOS scores.

Two complementary local-density metrics were calculated for each
`Met_hi_Fib` cell:

1. **Number of neighboring cells:** out-degree in the stored
   `delaunay_80_new` graph, which is a Delaunay graph with a maximum edge
   distance of 80 µm.
2. **Five-nearest-cell distance:** mean Euclidean centroid distance to the
   five nearest cells in the same donor image.

Local-density metrics and metabolic scores were averaged within donor before
Spearman testing. Thus, the inferential unit is the donor, not the cell. This
avoids treating thousands of spatially dependent cells as independent
replicates.

```{r source-table}
knitr::kable(
  source_manifest,
  row.names = FALSE,
  caption = "Source object and analysis definitions"
)
```

```{r graph-table}
graph_show <- graph_qc
numeric_cols <- vapply(graph_show, is.numeric, logical(1))
graph_show[numeric_cols] <- lapply(graph_show[numeric_cols], round, 3)
knitr::kable(
  graph_show,
  row.names = FALSE,
  caption = "Stored Delaunay graph quality checks"
)
```

# Primary results

## Density-score correlations

```{r fig1, results='asis'}
figure_link(
  "figure_1_density_metabolic_score_correlations",
  "Donor-level correlations between local density and metabolic scores"
)
```

```{r primary-table}
primary_show <- primary_cor
primary_show$density_metric <- pretty_metric(primary_show$density_metric)
primary_show$metabolic_score <- pretty_metric(primary_show$metabolic_score)
numeric_cols <- vapply(primary_show, is.numeric, logical(1))
primary_show[numeric_cols] <- lapply(primary_show[numeric_cols], round, 3)
knitr::kable(
  primary_show,
  row.names = FALSE,
  caption = "Primary donor-level Spearman correlations across all 16 donors"
)
```

None of the four prespecified donor-level associations was statistically
significant (`r sprintf("raw p-values %.3f–%.3f", min(primary_cor$p_value), max(primary_cor$p_value))`).
The glycolysis estimates were moderate in magnitude but imprecise:
Delaunay-neighbor count \(\rho =
`r sprintf("%.2f", primary_cor$rho[primary_cor$density_metric == "mean_delaunay_neighbors_80um" & primary_cor$metabolic_score == "mean_glycolysis_score"])`\)
and five-nearest-cell distance \(\rho =
`r sprintf("%.2f", primary_cor$rho[primary_cor$density_metric == "mean_distance_5nn_um" & primary_cor$metabolic_score == "mean_glycolysis_score"])`\).
The appropriate conclusion is therefore **no statistically significant
donor-level association**, not proof that the true associations are exactly
zero.

## Analyzed cell numbers

```{r fig2, results='asis'}
figure_link(
  "figure_2_analyzed_cell_count_and_global_density",
  "Analyzed cell counts and area-normalized cellularity by clinical group"
)
```

The distributions of total analyzed cells and area-normalized analyzed-cell
density did not differ systematically among healthy, progressive SSc, and
stable SSc donors:

```{r count-table}
count_show <- count_tests
numeric_cols <- vapply(count_show, is.numeric, logical(1))
count_show[numeric_cols] <- lapply(count_show[numeric_cols], round, 3)
knitr::kable(
  count_show,
  row.names = FALSE,
  caption = "Kruskal-Wallis tests for analyzed cell counts and overall cellularity"
)
```

## Local cellularity by group

```{r fig3, results='asis'}
figure_link(
  "figure_3_local_density_metrics_by_group",
  "Local density metrics around Met_hi_Fib by clinical group"
)
```

```{r local-table}
local_show <- local_tests
numeric_cols <- vapply(local_show, is.numeric, logical(1))
local_show[numeric_cols] <- lapply(local_show[numeric_cols], round, 3)
knitr::kable(
  local_show,
  row.names = FALSE,
  caption = "Kruskal-Wallis tests for local-density metrics around Met_hi_Fib"
)
```

# Robustness checks

## Minimum Met_hi_Fib representation

Some donors contain few `Met_hi_Fib` cells. The primary figure retains all
donors and scales each point by the number of contributing cells. Repeating
the analysis after requiring at least 5, 10, or 20 `Met_hi_Fib` cells per
donor did not produce a statistically significant result.

```{r fig4, results='asis'}
figure_link(
  "figure_4_minimum_cell_threshold_sensitivity",
  "Sensitivity of correlations to minimum Met_hi_Fib cell counts"
)
```

## SSc-only analysis

```{r ssc-table}
ssc_show <- ssc_cor
ssc_show$density_metric <- pretty_metric(ssc_show$density_metric)
ssc_show$metabolic_score <- pretty_metric(ssc_show$metabolic_score)
numeric_cols <- vapply(ssc_show, is.numeric, logical(1))
ssc_show[numeric_cols] <- lapply(ssc_show[numeric_cols], round, 3)
knitr::kable(
  ssc_show,
  row.names = FALSE,
  caption = "Sensitivity analysis restricted to the nine SSc donors"
)
```

# Figure legend

The standalone vector exports intentionally omit panel letters so that they
can be added consistently during final figure assembly.

> **Figure S5. Cell-number and local-density controls for IMC metabolic
> profiles.** **(A)** Relationships between local cellularity and metabolic
> scores in Met-hi Fib cells. For each donor (`n = 16`), the mean number of
> adjacent cells in the stored Delaunay graph (maximum edge distance 80 µm)
> and the mean centroid distance to the five nearest cells within the same
> image were plotted against the donor-mean glycolysis and TCA/OXPHOS scores.
> Each point represents one donor; color indicates clinical group and point
> size indicates the number of Met-hi Fib cells contributing to the donor
> mean. Dashed lines are visual linear guides. Insets report two-sided
> donor-level Spearman correlation coefficients and unadjusted p-values.
> **(B)** Total analyzed cells and area-normalized analyzed-cell density for
> each donor, stratified by clinical group. Points represent donors; boxes
> show the median and interquartile range, with whiskers extending to the most
> extreme value within 1.5 times the interquartile range. Area-normalized
> density was calculated from the stored image dimensions assuming the 1-µm
> IMC pixel size. Insets report two-sided Kruskal-Wallis p-values.

# Recommended response wording

> We thank the reviewer for this important comment. We replaced the
> representative images in Fig. 2D with images selected from comparable
> dermal depths and anatomical localizations across groups, reducing potential
> bias from differences in skin layer. We additionally tested whether local
> cellularity could explain the metabolic profiles of Met_hi_Fib. For each
> Met_hi_Fib cell, we quantified the number of adjacent cells in the existing
> Delaunay graph (maximum edge distance 80 µm) and the mean distance to its
> five nearest cells. These quantities, together with glycolysis and
> TCA/OXPHOS scores, were averaged per donor, and donor-level associations were
> tested using Spearman correlation. Neither local-density metric showed a
> statistically significant donor-level association with either metabolic
> score. This conclusion was unchanged after restricting the analysis to SSc
> donors or to donors with greater numbers of Met_hi_Fib cells. Moreover, the
> total number of analyzed cells and the area-normalized analyzed-cell density
> did not differ significantly among healthy, progressive SSc, and stable SSc
> donors. These results do not support differences in cell number or local
> cellularity as the sole explanation for the observed metabolic profiles,
> although the modest donor number limits the precision with which small or
> moderate density effects can be excluded.

# Recommended Methods wording

> **Cell-density control analysis.** To assess whether increased metabolic
> scores in Met_hi_Fib could be attributable to local cellularity, two
> density-related parameters were calculated for every Met_hi_Fib cell: (i)
> the number of adjacent cells in the existing Delaunay graph with a maximum
> edge distance of 80 µm and (ii) the mean Euclidean centroid distance to the
> five nearest cells within the same donor image. Delaunay-neighbor count,
> five-nearest-cell distance, glycolysis score, and TCA/OXPHOS score were
> averaged per donor. Associations between donor-level density metrics and
> metabolic scores were assessed by two-sided Spearman correlation. Total
> analyzed-cell counts and area-normalized analyzed-cell density were compared
> among healthy, progressive SSc, and stable SSc donors using the
> Kruskal-Wallis test.

# Interpretation boundary

The data support **absence of a statistically significant donor-level
association**, not equivalence or proof of zero effect. The phrase “the
increased metabolic scores are not explained by local density” is too
absolute for 16 donors. “The results do not support local density as the sole
explanation” is accurate and better aligned with the confidence intervals.

# Reproducibility

All source tables are in `outputs/tables`. Every figure is exported as PDF,
SVG, and 450-dpi PNG. The complete calculation is in
`compute_cell_density_reviewer_qc.R`.
