---
title: "IMC final-panel signal-to-noise quality control"
subtitle: "First and second cohorts"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    number_sections: false
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
report_dir <- normalizePath(
  dirname(knitr::current_input(dir = TRUE)),
  winslash = "/",
  mustWork = FALSE
)
output_dir <- file.path(report_dir, "outputs")
plot_dir <- file.path(output_dir, "plots")
table_dir <- file.path(output_dir, "tables")

global_snr <- read.csv(file.path(table_dir, "global_marker_snr.csv"), check.names = FALSE)
sample_snr <- read.csv(file.path(table_dir, "per_sample_marker_snr.csv"), check.names = FALSE)
cohort_summary <- read.csv(file.path(table_dir, "cohort_snr_summary.csv"), check.names = FALSE)
source_manifest <- read.csv(file.path(table_dir, "source_manifest.csv"), check.names = FALSE)
shared_snr <- read.csv(file.path(table_dir, "cross_cohort_shared_marker_snr.csv"), check.names = FALSE)
diagnostics <- read.csv(file.path(table_dir, "mixture_fit_diagnostics.csv"), check.names = FALSE)

rho <- suppressWarnings(cor(
  shared_snr[["log2_snr__First cohort"]],
  shared_snr[["log2_snr__Second cohort"]],
  method = "spearman",
  use = "complete.obs"
))

figure_link <- function(stem, alt) {
  png <- file.path("outputs", "plots", paste0(stem, ".png"))
  pdf <- file.path("outputs", "plots", paste0(stem, ".pdf"))
  svg <- file.path("outputs", "plots", paste0(stem, ".svg"))
  cat(
    sprintf(
      "<p><a href='%s'>PDF</a> | <a href='%s'>SVG</a></p><img src='%s' alt='%s' style='width:100%%;height:auto;'>",
      pdf, svg, png, alt
    )
  )
}
```

# Purpose

This report quantitatively documents signal-to-noise characteristics of the
final IMC antibody panels in both cohorts. It follows the **cell-level**
quality-control strategy in the
[Bodenmiller-group IMCDataAnalysis workflow](https://bodenmillergroup.github.io/IMCDataAnalysis/image-and-cell-level-quality-control.html):

\[
\mathrm{SNR} =
\frac{\mathrm{mean\ intensity\ of\ the\ higher\ component}}
{\mathrm{mean\ intensity\ of\ the\ lower\ component}}
\]

For each antibody and cohort, a two-component Gaussian mixture was fitted to
the `asinh`-transformed segmented-cell mean intensities. The component with
the higher mean in the untransformed `counts` assay was designated **signal**;
the lower component was designated **empirical background/noise**. Component
means and SNR were calculated on the untransformed intensities.

This is a complementary quantitative summary of final-dilution staining. It
does not replace the serial-dilution experiments, expected-compartment review,
or direct inspection of raw images. Here, "background" means the lower
single-cell intensity component, not acellular optical background.

# Input data

```{r input-table}
knitr::kable(
  source_manifest,
  row.names = FALSE,
  caption = "Finalized SpatialExperiment objects used for signal-to-noise QC"
)
```

Technical channels (`Ar`, iodine/xenon/barium background channels, iridium
DNA intercalators, image-cell segmentation kit channels, lead, and the
unlabelled first-cohort bismuth channel) were excluded. Histone H3 was
retained. The second-cohort-only TCF21 channel remains in the mixture-fit
audit and source tables for provenance but is omitted from all displayed
figures.

# Results

```{r cohort-summary}
display_summary <- cohort_summary
numeric_cols <- vapply(display_summary, is.numeric, logical(1))
display_summary[numeric_cols] <- lapply(
  display_summary[numeric_cols],
  function(x) round(x, 3)
)
knitr::kable(
  display_summary,
  row.names = FALSE,
  caption = "Cohort-level summary of final antibody-channel mixture fits"
)
```

The cohort-level table includes TCF21 among the 37 second-cohort antibody
fits; the plotted second-cohort panel contains the remaining 36 channels.

All `r sum(cohort_summary$n_antibody_channels)` cohort-specific antibody
mixtures were evaluated. The cross-cohort rank correlation across the
`r nrow(shared_snr)` shared antibodies was Spearman
\(\rho = `r sprintf("%.2f", rho)`\). Because the panels were stained and
acquired separately, this correlation is a consistency check rather than an
assay-equivalence test.

**Recommended figure use.** The signal-intensity-versus-SNR plot below is the
stronger primary quantitative panel for the Methods and reviewer response. It
shows both quantities used in the QC argument for every antibody and cohort,
without imposing a pass/fail boundary. The per-sample interval plot is best
used as supporting evidence that the estimated SNR was reasonably consistent
across acquired samples.

## Manuscript-ready Figure S2

The combined file below was exported directly at 7.1 inches, corresponding to
the full A4 manuscript width. Its typography and spacing are therefore set for
the assembled figure rather than for a full-width report plot that is later
reduced. Separate half-width vector panels are also available.

```{r fig-s2, results='asis'}
figure_link(
  "figure_S2_manuscript_combined",
  "Manuscript-ready Figure S2 showing signal intensity, SNR, and empirical background"
)
```

Panel A: [PDF](outputs/plots/figure_S2A_manuscript_signal_vs_snr.pdf) |
[SVG](outputs/plots/figure_S2A_manuscript_signal_vs_snr.svg)

Panel B: [PDF](outputs/plots/figure_S2B_manuscript_signal_background.pdf) |
[SVG](outputs/plots/figure_S2B_manuscript_signal_background.svg)

### Figure legend

**Figure S2. Signal intensity and signal-to-noise characteristics of the final
IMC antibody panels.** **A,** Each filled point represents one antibody channel
in the first IMC cohort (blue) or second IMC cohort (orange). For each antibody
and cohort, a two-component Gaussian mixture model was fitted to
asinh-transformed segmented-cell mean intensities. The component with the
higher mean untransformed intensity was designated signal and the
lower-intensity component empirical background/noise. The x-axis shows the
log2-transformed mean signal intensity and the y-axis the log2-transformed
signal-to-noise ratio (SNR), calculated from the untransformed counts assay as
mean signal intensity divided by mean background intensity. Every antibody
channel is shown and labeled. **B,** Open and filled points show the lower- and
higher-intensity component means, respectively, on the log2 scale; connecting
lines show their separation for each antibody, with the complete antibody list
displayed on the y-axis. The second-cohort-only TCF21 channel is omitted from
the displayed panels but retained in the QC source tables. No fixed numerical
acceptance threshold was applied; these quantitative summaries were evaluated
together with expected compartmental localization and visual assessment of
staining quality in the raw images.

## Signal intensity versus SNR {.tabset}

### Figure

```{r fig1, results='asis'}
figure_link(
  "figure_1_reference_style_signal_vs_snr",
  "Signal intensity versus signal-to-noise ratio for both IMC cohorts"
)
```

### Interpretation

Each filled point represents one antibody; the cohorts are displayed in
separate panels. No numerical acceptance boundary is drawn because no
universal SNR or signal-intensity cutoff was applied. Low-abundance markers can
remain specific and useful, so final acceptance also depended on expected
spatial localization and visual review.

## Signal and empirical background

```{r fig3, results='asis'}
figure_link(
  "figure_3_signal_background_dumbbell",
  "Signal and empirical background component means for each antibody"
)
```

This view directly displays the separation between the lower and higher
single-cell intensity components. It is a useful mechanistic companion to the
primary plot because the numerator and denominator of SNR remain visible.

## Consistency across samples

```{r fig2, results='asis'}
figure_link(
  "figure_2_per_sample_snr_intervals",
  "Median and interquartile range of signal-to-noise ratios across samples"
)
```

For each marker, the cohort-wide mixture classification was applied to each
sample and SNR was recalculated from that sample's untransformed intensities.
Only sample-marker combinations with at least 20 cells in each component were
included in the interval summary.

## Per-sample heatmaps {.tabset}

### First cohort

```{r fig4a, results='asis'}
figure_link(
  "figure_4a_first_cohort_sample_snr_heatmap",
  "First cohort per-sample signal-to-noise heatmap"
)
```

### Second cohort

```{r fig4b, results='asis'}
figure_link(
  "figure_4b_second_cohort_sample_snr_heatmap",
  "Second cohort per-sample signal-to-noise heatmap"
)
```

Both heatmaps use the same full, unclipped color range. Gray entries indicate
that one fitted component contained fewer than 20 cells in that sample and
was therefore not treated as a stable sample-specific estimate.

## Cross-cohort concordance

```{r fig5, results='asis'}
figure_link(
  "figure_5_cross_cohort_snr_concordance",
  "Cross-cohort concordance of shared antibody signal-to-noise ratios"
)
```

# Suggested manuscript wording

## Response letter

> A good signal-to-background ratio was not defined by a single universal
> numerical cutoff. Antibodies were first evaluated by conventional
> immunofluorescence on FFPE skin sections across serial dilutions and were
> retained only when they showed the expected compartmental staining pattern
> and signal clearly distinguishable from background by visual inspection.
> Metal-conjugated antibodies were then retested by IMC on consecutive FFPE
> sections across serial dilutions to select the final concentration. As
> quantitative documentation of the final IMC panels, we additionally followed
> a published cell-level IMC quality-control approach. For each antibody and
> cohort, a two-component Gaussian mixture was fitted to asinh-transformed
> segmented-cell mean intensities. The higher-intensity component was treated
> as signal and the lower-intensity component as empirical background/noise;
> SNR was calculated from untransformed intensities as the ratio of their mean
> values. Signal-intensity-SNR plots and per-sample summaries were reviewed
> together with spatial staining patterns. This quantitative metric supported,
> but did not replace, visual assessment because low-abundance yet specific
> markers may not be appropriately judged by a fixed SNR threshold.

## Methods

> For IMC validation, antibody channels were reviewed for expected spatial
> distribution, consistency across serial sections, and acceptable
> signal-to-background characteristics. For quantitative quality control of
> the final panels, a two-component Gaussian mixture model was fitted
> separately for each antibody and cohort to asinh-transformed segmented-cell
> mean intensities. The component with the higher untransformed mean intensity
> was designated signal and the lower component empirical background/noise,
> and SNR was calculated as mean signal intensity divided by mean background
> intensity. Signal-intensity-SNR plots and per-sample summaries were inspected
> together with the raw images. No universal SNR cutoff was applied; final
> channel acceptance also required the expected compartmental localization
> and consistent staining quality.

# Important scope note

The first cohort retains multichannel TIFF stacks and masks, but equivalent
raw TIFF stacks for the second cohort were not found alongside the finalized
analysis files. A pixel-level Otsu SNR could therefore be added for the first
cohort, but it should not replace the matched cell-level figures above unless
the second-cohort raw stacks are recovered. Mixing first-cohort pixel SNR with
second-cohort cell-level SNR would compare different quantities.

# Reproducibility

The source tables are provided in `outputs/tables`; all plots are exported as
PDF, SVG, and high-resolution PNG. The calculation script is
`compute_imc_signal_to_noise_qc.R`.

```{r diagnostics-table}
flagged <- diagnostics[
  diagnostics$fit_status != "ok" |
    diagnostics$component_size_flag,
]
if (nrow(flagged)) {
  flagged_show <- flagged[, c(
    "cohort", "marker", "fit_status", "n_signal", "n_background",
    "fraction_signal", "signal_mean", "background_mean", "snr",
    "component_size_flag"
  )]
  numeric_cols <- vapply(flagged_show, is.numeric, logical(1))
  flagged_show[numeric_cols] <- lapply(
    flagged_show[numeric_cols],
    function(x) round(x, 3)
  )
  knitr::kable(
    flagged_show,
    row.names = FALSE,
    caption = "Mixture fits requiring cautious interpretation"
  )
} else {
  cat("No cohort-level mixture fits met the prespecified diagnostic flags.")
}
```
