---
title: "Interaction analysis"
author: "Alex Matei"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  html_document: 
    toc: true
    toc_float: true
    toc_depth: 4
    df_print: paged
    code_folding: hide
---

# Setup
```{r setup, include=FALSE}
source("data_preprocessing/common/config.R")

library(here)
source("data_preprocessing/common/original_setup.R")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, root.dir = here::here())
```

```{r timestamp}
ts <- format(Sys.time(), "%Y%m%d%H%M%S")
ptm <- proc.time()

cat("This document was created with the time stamp",ts)
```

# Load spe

```{r load spe}
spe <- readRDS(here("Datasets/spe_20241027221015.rds"))
```


# Generate: Plot expression of a marker gene in cells with and without selected neighbours - Boxplots

```{r plot expression of a marker gene in cells with and without selected neighbours 2,echo=TRUE, fig.width=10, fig.height=8}

# Define your features to plot and populations of interest
features_to_plot <- c(
  "aSMA",
  "FAP",
  "Cdh11",
  "Collagen",
  "Ki-67",
  "NOX4",
  "HIF1a",
  "Glycolysis_z_c",
  "Glycolysis_z",
  "TCA_z_c",
  "TCA_z"
)

has_neigh <- names(spe$hasNeigh_df)
pop_interest <- c("Stromal", "CD31+", "CD45+CD68+")

plot_list_pop_interest <- list()

for (k in pop_interest) {
  # Subset SingleCellExperiment object for cells of interest
  spe_subset <- spe[, spe$new_celllabels == k]
  
  # Turn "z-rescaled_new" assay data into a data frame (cells in rows)
  expression_data <- as.data.frame(t(assay(spe_subset, "z-rescaled_new")))
  
  # Add metadata columns to the data frame
  expression_data$Donor         <- colData(spe_subset)$Donor
  expression_data$Disease       <- colData(spe_subset)$Disease
  expression_data$Glycolysis_z_c <- colData(spe_subset)$Glycolysis_z_c
  expression_data$Glycolysis_z   <- colData(spe_subset)$Glycolysis_z
  expression_data$TCA_z_c        <- colData(spe_subset)$TCA_z_c
  expression_data$TCA_z          <- colData(spe_subset)$TCA_z
  
  # Initialize a list to store plots for each feature
  plot_list_features <- list()
  
  for (i in features_to_plot) {
    # Initialize a list to store plots for each neighbor type
    plot_list_has_neigh <- list()
    
    for (j in has_neigh) {
      # Convert neighbor column to a factor (ensure consistent ordering)
      expression_data[[j]] <- factor(as.character(spe_subset$hasNeigh_df[, j]))
      
      # Identify the first group/category on the x-axis
      first_group <- levels(expression_data[[j]])[1]
      
      # Calculate the median expression for the first group
      median_val <- median(expression_data[[i]][expression_data[[j]] == first_group], na.rm = TRUE)
      
      # Generate the violin plot with jittered points and a horizontal median line
      plot_exp <- ggplot(expression_data, aes(
        x   = !!sym(j),
        y   = .data[[i]],
        fill = !!sym(j)
      )) +
        geom_violin(trim = FALSE, alpha = 0.6) +
        geom_jitter(width = 0.05, size = 0.1, alpha = 0.1) +
        geom_hline(yintercept = median_val, linetype = "dashed", color = "blue") +
        labs(
          title = paste("Expression of", i, "in", k, "cells"),
          x = "Presence of target neighbour cell",
          y = "Marker expression",
          fill = "Presence of neighbour"
        ) +
        theme_bw(base_size = 12)+  # You can customize your theme here
      
      # Optionally color by j instead of fill, or add a manual fill scale:
      scale_fill_manual(values = c("#B2DF8A", "#FB9A99"))
      
      # Store the plot in the list
      plot_list_has_neigh[[j]] <- plot_exp
    }
    
    # Store all "neighbor" plots for the current feature
    plot_list_features[[i]] <- plot_list_has_neigh
  }
  
  # Store the plots for the current cell population
  plot_list_pop_interest[[k]] <- plot_list_features
}

```

# Display: Plot expression of a marker gene in cells with and without selected neighbours - Boxplots {.tabset .tabset-pills}

`r fig_dim <- c(6, 3)`

```{r display plot expr boxplot, fig.width=fig_dim[1], fig.height=fig_dim[2], results='asis'}
# Display the plots in tabs
for (k in pop_interest) {
  cat("## ", k, "{.tabset .tabset-pills}\n\n")
  for (i in features_to_plot) {
    cat("### ", i, "{.tabset .tabset-pills}\n\n")
    for (j in has_neigh) {
      cat("#### ", j, "\n\n")
      print(plot_list_pop_interest[[k]][[i]][[j]])
      cat("\n\n")
    }
  }
}
```

# Plot selected markers in cells with specific neighbours
```{r plot selected markers in cells with specific neighbours, fig.width=5, fig.height=3}
pop_interest <- c("CD45+CD68+")
features_to_plot <- c("Glycolysis_z_c")
has_neigh <- "has_stromal_2_neigh"
print(plot_list_pop_interest[[pop_interest]][[features_to_plot]][[has_neigh]])
```

# Outputs

This analysis generates the following results: <br>

1. Plots of expression of markers in cells with specific neighbours. The annotation of a cell type by its interaction with another cell type should be already present in the spe. <br>

# Session information

```{r session, eval=TRUE}
etime <- proc.time() - ptm
cat('Rendering this document took :', etime[3], ' s \n')

sessioninfo::session_info()
```