---
title: "Interaction analysis"
author: "Alex Matei"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  html_document: 
    toc: true
    toc_float: true
    toc_depth: 4
    df_print: paged
    code_folding: hide
---

# Setup
```{r setup}
source("data_preprocessing/common/config.R")

library(here)
source("data_preprocessing/common/original_setup.R")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, root.dir = here::here())
```

```{r timestamp}
ts <- format(Sys.time(), "%Y%m%d%H%M%S")
ptm <- proc.time()

cat("This document was created with the time stamp",ts)
```

# Load spe

```{r load spe}
spe <- readRDS(here("Datasets/spe_20241027221015.rds"))

spe$Disease <- as.character(spe$Disease)
spe <- spe[,spe$Disease=="Healthy"]
```


# Generate and display: Plot expression of a marker gene in cells with and without selected neighbours - Violin plot {.tabset .tabset-pills}

 `r fig_dim <- c(10, 8)`

```{r plot expression of a marker gene in cells with and without selected neighbours 1, fig.dim=fig_dim, results='asis'}
library(scater)

subset_variable <- c("new_celllabels", "Disease")
subset_values <- c("Stromal", "SSc")
features_to_plot <- c("aSMA","FAP","Cdh11","Collagen") 

# subset the spe to the cell type of interest for which to plot expression of markers if they have selected neighbours, and to a disease
for (i in 1:seq_along(subset_variable)) {
  spe_subset <- spe[, spe[[subset_variable[i]]] == subset_values[i]]
}

# Violin Plot - plotExpression
has_neigh <- names(spe$hasNeigh_df)
plot_list_violin <- list()
for (i in has_neigh) {
  spe_subset$has_x_Neigh <- spe_subset$hasNeigh_df[, i]
  #  colnames(spe_subset$has_x_Neigh) <- colnames(spe_df_hasNeigh_df[,i])
  plot <- plotExpression(
    spe_subset,
    features = c("aSMA", "FAP", "Cdh11", "Collagen"),
    x = "has_x_Neigh",
    exprs_values = "z-rescaled_new",
    colour_by = "has_x_Neigh"
  ) +
    theme(axis.text.x =  element_text(angle = 90))#+
  #scale_color_manual(values = metadata(spe)$color_vectors$celltype)
  
  plot_list_violin[[i]] <- plot
}

for(i in has_neigh){
  cat("## ", i, "\n\n")
  print(plot_list_violin[[i]])
  cat("\n\n")
}
```  

# Generate: Plot expression of a marker gene in cells with and without selected neighbours - Boxplots

```{r plot expression of a marker gene in cells with and without selected neighbours 2,echo=TRUE, fig.width=10, fig.height=8}

features_to_plot <- c(
  "aSMA",
  "FAP",
  "Cdh11",
  "Collagen",
  "Ki-67",
  "NOX4",
  "HIF1a",
  "Glycolysis_z_c",
  "Glycolysis_z",
  "TCA_z_c",
  "TCA_z"
) 
has_neigh <- names(spe$hasNeigh_df)
pop_interest <- c("Stromal","CD31+","CD45+CD68+")

plot_list_pop_interest <- list()
for (k in pop_interest){
spe_subset <- spe[,spe$new_celllabels == k]
plot_list_features <- list()

# Extract the relevant data for the plot
expression_data <- as.data.frame(t(assay(spe_subset, "z-rescaled_new"))) 
expression_data$Donor <- colData(spe_subset)$Donor
expression_data$Disease <- colData(spe_subset)$Disease
expression_data$Glycolysis_z_c <- colData(spe_subset)$Glycolysis_z_c
expression_data$Glycolysis_z <- colData(spe_subset)$Glycolysis_z
expression_data$TCA_z_c <- colData(spe_subset)$TCA_z_c
expression_data$TCA_z <- colData(spe_subset)$TCA_z

  for (i in features_to_plot){
    plot_list_has_neigh <- list()
    for (j in has_neigh){
      
col_name <- colnames(spe$hasNeigh_df[,j])
expression_data[[j]] <- as.character(spe_subset$hasNeigh_df[,j])
    
# Extract the marker of interest
expression_data <- expression_data %>%
  dplyr::select(all_of(features_to_plot), Donor, Disease, !!sym(j))
    # Calculate the average expression per donor and disease group
    avg_expression <- expression_data %>%
      group_by(Donor, Disease, !!sym(j)) %>%
      summarize(avg_marker_expression = median(.data[[i]], na.rm = TRUE))
    
    # Plot the data with one boxplot per neighbour presence 
    plot_exp <- ggplot(data=avg_expression, aes(x=!!sym(j),
                                             y=avg_marker_expression, 
                                             #fill=Disease
                                             )
    )+geom_boxplot(outlier.shape = NA,
                   aes(fill=!!sym(j))
                   )+
      stat_compare_means(comparisons=list(as.character(unique(avg_expression[j])[[1]])),
                         method="wilcox", 
                         hide.ns = TRUE,
                         label.y=max(avg_expression$avg_marker_expression)+0.2)+
      get_common_boxplot_components_fx()+
      scale_fill_manual(values = c("#B2DF8A", "#FB9A99"))+
      labs(title = paste("Average Expression of ", i, " per donor for ", k, " cells"),
        x="Presence of target neighbour cell",
        y="Average marker expression",
        fill="Presence of the neighbour x")
    
    plot_list_has_neigh[[j]] <- plot_exp
    }
    plot_list_features[[i]] <- plot_list_has_neigh
  }
  plot_list_pop_interest[[k]] <- plot_list_features
}

```

# Display: Plot expression of a marker gene in cells with and without selected neighbours - Boxplots {.tabset .tabset-pills}

`r fig_dim <- c(10, 8)`

```{r display plot expr boxplot, fig.dim=fig_dim, results='asis'}
# Display the plots in tabs
for (k in pop_interest) {
  cat("## ", k, "{.tabset .tabset-pills}\n\n")
  for (i in features_to_plot) {
    cat("### ", i, "{.tabset .tabset-pills}\n\n")
    for (j in has_neigh) {
      cat("#### ", j, "\n\n")
      print(plot_list_pop_interest[[k]][[i]][[j]])
      cat("\n\n")
    }
  }
}
```


# Outputs

This analysis generates the following results: <br>

1. Plots of expression of markers in cells with specific neighbours. The annotation of a cell type by its interaction with another cell type should be already present in the spe. <br>

# Session information

```{r session, eval=TRUE}
etime <- proc.time() - ptm
cat('Rendering this document took :', etime[3], ' s \n')

sessioninfo::session_info()
```