---
title: "Interaction frequency plots"
author: "Alex Matei"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  html_document: 
    toc: true
    toc_float: true
    toc_depth: 4
    df_print: paged
    code_folding: hide
---

# Setup
```{r setup, include=FALSE}
source("data_preprocessing/common/config.R")

library(here)
source("data_preprocessing/common/original_setup.R")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, root.dir = here::here())
```

```{r timestamp}
ts <- format(Sys.time(), "%Y%m%d%H%M%S")
ptm <- proc.time()

cat("This document was created with the time stamp",ts)
```

# Load spe

```{r load spe}
load(here("spe_list.RData"))
spe <- spe_list$spe
rm(spe_list)
```

# Prepare spe for analysis

```{r prep spe}
# Create a new data frame with all the colData of the spe object - to recode the patient_id column

metadata <- colData(spe)

# add a new column to metadata called Donor where you recode the patient_id column
# the elements starting with NH should be called Healthy1, Healthy2, etc.
# the elements starting with SSc should be called SSc1, SSc2, etc.
metadata$Donor <- dplyr::recode(metadata$patient_id,
                                     Healthy1="Healthy1",
                                     Healthy2="Healthy2", 
                                     SSc1="SSc1",
                                     SSc2="SSc2",
                                     SSc3="SSc3",
                                     SSc4="SSc4",
                                     SSc5="SSc5",
                                     SSc6="SSc6",
                                     SSc7="SSc7",
                                     SSc8="SSc8",
                                     Healthy3="Healthy3",
                                     Healthy4="Healthy4",
                                     SSc9="SSc9",
                                     Healthy5="Healthy5",
                                     Healthy6="Healthy6",
                                     Healthy7="Healthy7")


metadata$Disease <- dplyr::recode(metadata$indication,
                                  NH="Healthy",
                                  SSc="SSc")

metadata$ILD <- dplyr::recode(metadata$ILD,
                                     healthy="healthy",
                                     "1"="ILD", 
                                     "0"="no ILD")

metadata$Arthritis[is.na(metadata$Arthritis)] <- "no arthritis"

# Assign the new column in metadata to the spe object
spe$Donor <- metadata$Donor
spe$ILD <- metadata$ILD
spe$Arthritis <- metadata$Arthritis
spe$Disease <- metadata$Disease

names(colData(spe))[which(names(colData(spe))=="Skin.progression.at.biopsy")]="Progression_skin"
names(colData(spe))[which(names(colData(spe))=="Subtype.(dc/lc)")]="Subtype_dc_lc"

met_markers <- c("Glut1",
                 "HK1",
                 "PFKL1",
                 "PKM2",
                 "LDHA",
                 "GAPDH",
                 "CS",
                 "OGDH",
                 "ATP5A",
                 "SDHA",
                 "CD98",
                 "ACAC",
                 "CPT1a",
                 "G6PD",
                 "p_mTOR",
                 "pAMPK",
                 "NOX4",
                 "HIF1a",
                 "PGC-1a",
                 "TFAM",
                 "NRF2_p")

glycolysis_markers <- c("Glut1",
                 "HK1",
                 "PFKL1",
                 "PKM2",
                 "LDHA",
                 "GAPDH")

no_met_markers <- setdiff(rownames(spe),met_markers) 
rowData(spe)$use_channel_met_markers <- rownames(spe) %in% met_markers
rowData(spe)$use_channel_glyc_markers <- rownames(spe) %in% glycolysis_markers

# define clinical categories
clinical_cat <- c("Disease","Progression_skin", "Arthritis", "DU.history", "ILD", "Subtype_dc_lc", "Early.or.not")

spe[[clinical_cat[1]]] <- factor(spe[[clinical_cat[1]]], levels = c("Healthy", "SSc"))
spe[[clinical_cat[2]]] <- factor(spe[[clinical_cat[2]]], levels = c("healthy", "progressive", "stable"))
spe[[clinical_cat[3]]] <- factor(spe[[clinical_cat[3]]], levels = c("healthy","arthritis","no arthritis"))
spe[[clinical_cat[4]]] <- factor(spe[[clinical_cat[4]]], levels = c("healthy","DU history","no DU history"))
spe[[clinical_cat[5]]] <- factor(spe[[clinical_cat[5]]], levels = c("healthy","ILD","no ILD"))
spe[[clinical_cat[6]]] <- factor(spe[[clinical_cat[6]]], levels = c("healthy","dcSSc","lcSSc"))
spe[[clinical_cat[7]]] <- factor(spe[[clinical_cat[7]]], levels = c("healthy","early","late"))

# define color vectors
color_vectors <- list()
pheno_clusters_new <- setNames(c("#C4281C","#C2DAB8","#E8BAC8","#80BBDB","#C470A0","#0D69AC","#F5CD30","#287F47",
                                 "#A05F35","#D9856C","#6B327C","#008F9C","#C7D23C","#B7D7D5","#923978","#27462D",
                                 "#FF9494","#938767","#FFF67B","#E1A4C2","#FC7B1B","#CD6298","#102ADC"), 
                               as.character(unique(spe$pheno_clusters_new)))

indication <- setNames(c("#4DBBD533", "#E64B35AA"),
                       c(unique(colData(spe)$Disease)))

Donor <- setNames(c("#FFFF33", "#FFCC66",'#FF9933','#FF6633','#00FFFF','#FF0033','#79B5B5', '#C470A0',
                         '#996633','#999933','#99CC33','#FF9494', '#33FF33','#336633', '#330066','#923978'),
                       c(unique(colData(spe)$Donor)))

Neighbourhood_del_80_pheno_new <- setNames(c("#CC00CC", "#260EEC", "#FFCA28", "#004D40","#FF6633","#66FF66", 
                                             "#EF9EA7", "#990000", "#CC0033","#4FC3F7"),
                                           unique(spe$Neighbourhood_del_80_pheno_new))

color_vectors$clusters <- pheno_clusters_new
color_vectors$Disease <- indication
color_vectors$Donor <- Donor
color_vectors$neighbourhood_del_80_pheno_new <- Neighbourhood_del_80_pheno_new

color_vectors[[clinical_cat[2]]] <- setNames(c("#4DBBD533", "#C03830", "#F6BF93"), unique(spe[[clinical_cat[2]]]))
color_vectors[[clinical_cat[3]]] <- setNames(c("#4DBBD533","#377EB8","#4DAF4A"), unique(spe[[clinical_cat[3]]]))
color_vectors[[clinical_cat[4]]] <- setNames(c("#4DBBD533","#FC8D62","#8DA0CB"), unique(spe[[clinical_cat[4]]]))
color_vectors[[clinical_cat[5]]] <- setNames(c("#4DBBD533","#FFFFB3","#BEBADA"), unique(spe[[clinical_cat[5]]]))
color_vectors[[clinical_cat[6]]] <- setNames(c("#4DBBD533","#A6D854","#FFD92F"), unique(spe[[clinical_cat[6]]]))
color_vectors[[clinical_cat[7]]] <- setNames(c("#4DBBD533","#F781BF","#999999"), unique(spe[[clinical_cat[7]]]))

metadata(spe)$colors_vectors$Progression_skin <- setNames(c("#6CC3F4", "#C03830", "#F6BF93"),
                                                          unique(spe$Progression_skin))

metadata(spe)$color_vectors <- color_vectors

rm(metadata)
```

# Create plots: Interaction frequencies using CountNeighCells_fx and InteractFreq_fx

```{r create interaction frequencies data frame and plot, echo=TRUE, fig.width=10, fig.height=8}

# Add a new colData element with the counts of CD31_3 and CD31_5 neighbours for each cell
spe$pheno_clusters_new <- as.character(spe$pheno_clusters_new)

spe <- CountNeighCells_fx(
  spe,
  target = unique(spe$pheno_clusters_new),
  #("CD31_3","CD31_5", "CD45_CD3_3", "CD45_CD68_1", "stromal_2"),
  colPairs = 'delaunay_80_new',
  celltype = 'pheno_clusters_new',
  new_col_name = 'NeighCount'
)

# calculate interaction frequency for healthy - SSc

interact_freq <- InteractFreq_fx(
  spe,
  cell_of_interest = unique(spe$pheno_clusters_new),
  celltype = 'pheno_clusters_new',
  sample_id = 'Donor',
  grouping.var = 'Disease',
  neighbor_count = 'NeighCount',
  add_zeroes = TRUE
)

interact_freq$celltype <- as.character(interact_freq$celltype)

# plot interaction frequency for healthy - SSc
selected_target_cells <- unique(spe$pheno_clusters_new)
selected_cells_of_interest <- unique(spe$pheno_clusters_new)

cells_interest_1 <- list()
for (i in selected_cells_of_interest) {
  target_cells_1 <- list()
  for (j in selected_target_cells) {
    interact_freq_cell_interest <- interact_freq %>% filter(celltype == j)
    interact_freq_cell_interest_sub <- interact_freq_cell_interest %>% filter(Neighbor ==
                                                                                i)
    plot_freq_int <- ggplot(data = interact_freq_cell_interest_sub, aes(x =
                                                                          group, y = PercentHasNeigh)) +
      geom_boxplot(outlier.shape = NA, aes(fill = group)) +
      stat_compare_means(
        comparisons = list(as.character(
          unique(interact_freq_cell_interest_sub$group)
        )),
        method = "wilcox",
        hide.ns = TRUE,
        label.y = max(interact_freq_cell_interest_sub$PercentHasNeigh) +
          10
      ) +
      get_common_boxplot_components_fx() +
      scale_fill_manual(values = metadata(spe)$color_vectors$Disease) +
      labs(
        x = paste0("Frequency of ", i, " interacting with ", j),
        y = "Frequency (%)",
        fill = "Disease"
      )
    
    target_cells_1[[j]] <- plot_freq_int
  }
  cells_interest_1[[i]] <- target_cells_1
}

# calculate interaction frequency for healthy stable progressive

interact_freq2 <- InteractFreq_fx(
  spe,
  cell_of_interest = unique(spe$pheno_clusters_new),
  celltype = 'pheno_clusters_new',
  sample_id = 'patient_id',
  grouping.var = 'Progression_skin',
  neighbor_count = 'NeighCount',
  add_zeroes = TRUE
)

interact_freq2$celltype <- as.character(interact_freq2$celltype)

# plot interaction frequency for healthy stable progressive
selected_target_cells2 <- unique(spe$pheno_clusters_new)
selected_cells_of_interest2 <- unique(spe$pheno_clusters_new)


cells_interest_2 <- list()
for (i in selected_cells_of_interest2) {
  target_cells_2 <- list()
  for (j in selected_target_cells2) {
    interact_freq_cell_interest2 <- interact_freq2 %>% filter(celltype == j)
    interact_freq_cell_interest_sub2 <- interact_freq_cell_interest2 %>% filter(Neighbor ==
                                                                                  i)
    if (max(interact_freq_cell_interest_sub2$PercentHasNeigh) > 0) {
      dunn_res <- as.data.frame(interact_freq_cell_interest_sub2) %>% rstatix::dunn_test (PercentHasNeigh ~ group, p.adjust.method =
                                                                                            "none") %>%
        mutate (method = "Dunn")
      plot_freq_int2 <- ggplot(data = interact_freq_cell_interest_sub2, aes(x =
                                                                              group, y = PercentHasNeigh)) +
        geom_boxplot(outlier.shape = NA, aes(fill = group)) +
        stat_pvalue_manual(dunn_res %>%
                             mutate(y.position = c(
                               max(interact_freq_cell_interest_sub2$PercentHasNeigh + 10),
                               max(interact_freq_cell_interest_sub2$PercentHasNeigh +
                                     20),
                               max(interact_freq_cell_interest_sub2$PercentHasNeigh +
                                     10)
                             )), bracket.shorten = 0.05) +
        stat_compare_means(
          method = "kruskal",
          label.y = max(interact_freq_cell_interest_sub2$PercentHasNeigh + 30)
        ) +
        get_common_boxplot_components_fx() +
        scale_fill_manual(values = metadata(spe)$color_vectors$Progression_skin) +
        labs(
          x = paste0("Frequency of ", i, " interacting with ", j),
          y = "Frequency (%)",
          fill = "Disease"
        )
      target_cells_2[[j]] <- plot_freq_int2
    }
    else {
      print(paste0("No cells interacting with ", j))
    }
  }
  cells_interest_2[[i]] <- target_cells_2
}

```

# Display plots: Interaction frequency plots for healthy SSc {.tabset .tabset-pills}

`r fig_dim <- c(10, 8)`

```{r display interaction freq h ssc, fig.dim=fig_dim, results='asis'}
# Display the plots in tabs
for (i in names(cells_interest_1)) {
  cat("## Frequency of", i, "{.tabset .tabset-pills}\n\n")
  for (j in names(cells_interest_1[[i]])) {
    cat("### Interacting with", j, "\n\n")
    print(cells_interest_1[[i]][[j]])
    cat("\n\n")
  }
}
rm(cells_interest_1)
rm(target_cells_1)
```

# Display plots: Interaction frequency plots for healthy SSc stable SSc progressive {.tabset .tabset-pills}

`r fig_dim <- c(10, 8)`

```{r display interaction freq h ssc st ssc pr, fig.dim=fig_dim, results='asis'}
# Display the plots in tabs
for (i in names(cells_interest_2)) {
  cat("## Frequency of", i, "{.tabset .tabset-pills}\n\n")
  for (j in names(cells_interest_2[[i]])) {
    cat("### Interacting with", j, "\n\n")
    print(cells_interest_2[[i]][[j]])
    cat("\n\n")
  }
}
rm(cells_interest_2)
rm(target_cells_2)
```

# Create plots: Frequencies of a target cell among neighbours in groups from the aggregateNeighbours function

```{r plot frequencies of a target cell among neighbours, echo=TRUE, fig.width=10, fig.height=8}

neighbours_of_1 <- list()
for (i in unique(spe$pheno_clusters_new)) {
  spe_subset <- spe[, spe$pheno_clusters_new == i]
  freq_of_1 <- list()
  for (j in unique(spe$pheno_clusters_new)) {
    # Filter the frequencies of the neighbours of interest in the target cells as vectors
    nbh_freq <- colData(spe_subset)[, "metadata_del_80_pheno_new"][j][, 1]
    
    # Create data frame that summarizes the frequencies of the neighbours of interest in the target cells
    for_plot <- data.frame(
      "Freq" = nbh_freq,
      "Disease" = spe_subset$Disease,
      "Donor" = spe_subset$Donor
    ) %>%
      mutate(Freq = Freq * 100) %>%
      group_by(Donor, Disease) %>%
      summarize (Freq = mean(Freq, na.rm = T))
    
    #for_plot2 <- table(for_plot$Donor,for_plot$Disease)
    
    plot <- ggplot(data = for_plot, aes(x = Disease, y = Freq)) + geom_boxplot(outlier.shape = NA, aes(fill =
                                                                                                         Disease)) +
      stat_compare_means(
        comparisons = list(as.character(unique(
          for_plot$Disease
        ))),
        method = "wilcox",
        hide.ns = TRUE,
        label.y = max(for_plot$Freq) + 0.1
      ) +
      get_common_boxplot_components_fx() +
      scale_fill_manual(values = metadata(spe)$color_vectors$Disease) +
      labs(
        x = paste0("Frequency of ", j, " from the neighbours of ", i),
        y = "Frequency (%)",
        fill = "Disease"
      )
    freq_of_1[[j]] <- plot
  }
  neighbours_of_1[[i]] <- freq_of_1
}

neighbours_of_2 <- list()
for (i in unique(spe$pheno_clusters_new)) {
  spe_subset <- spe[, spe$pheno_clusters_new == i]
  freq_of_2 <- list()
  for (j in unique(spe$pheno_clusters_new)) {
    # Filter the frequencies of the neighbours of interest in the target cells as vectors
    nbh_freq2 <- colData(spe_subset)[, "metadata_del_80_pheno_new"][j][, 1]
    
    # Create data frame that summarizes the frequencies of the neighbours of interest in the target cells
    for_plot2 <- data.frame(
      "Freq" = nbh_freq2,
      "Disease" = spe_subset$Progression_skin,
      "Donor" = spe_subset$Donor
    ) %>%
      mutate(Freq = Freq * 100) %>%
      group_by(Donor, Disease) %>%
      summarize (Freq = mean(Freq, na.rm = T))
    
    if (max(for_plot2$Freq) > 0) {
      dunn_res <- as.data.frame(for_plot2) %>% rstatix::dunn_test (Freq ~ Disease, p.adjust.method =
                                                                     "none") %>%
        mutate (method = "Dunn")
      
      plot2 <- ggplot(data = for_plot2, aes(x = Disease, y = Freq)) + geom_boxplot(outlier.shape = NA, aes(fill =
                                                                                                             Disease)) +
        stat_pvalue_manual(dunn_res %>%
                             mutate(y.position = c(
                               max(for_plot2$Freq + 10),
                               max(for_plot2$Freq + 15),
                               max(for_plot2$Freq + 10)
                             )), bracket.shorten = 0.05) +
        stat_compare_means(method = "kruskal",
                           label.y = max(for_plot2$Freq +
                                           20)) +
        get_common_boxplot_components_fx() +
        scale_fill_manual(values = metadata(spe)$color_vectors$Progression) +
        labs(
          x = paste0("Frequency of ", j, " from the neighbours of ", i),
          y = "Frequency (%)",
          fill = "Disease"
        )
      freq_of_2[[j]] <- plot2
    }
    else {
      print(paste0("No cells interacting with ", i))
    }
  }
  neighbours_of_2[[i]] <- freq_of_2
}
```

# Display plots: Frequencies of a target cell among neighbours for healthy SSc {.tabset .tabset-pills}

`r fig_dim <- c(10, 8)`

```{r display freq among nbh h ssc, fig.dim=fig_dim, results='asis'}
# Display the plots in tabs
for (i in names(neighbours_of_1)) {
  cat("## From the neighbours of", i, "{.tabset .tabset-pills}\n\n")
  for (j in names(neighbours_of_1[[i]])) {
    cat("### Frequency of", j, "\n\n")
    print(neighbours_of_1[[i]][[j]])
    cat("\n\n")
  }
}

rm(neignbours_of_1)
rm(freq_of_1)
```

# Display plots: Frequencies of a target cell among neighbours for healthy SSc stable SSc progressive {.tabset .tabset-pills}

`r fig_dim <- c(10, 8)`

```{r display freq among nbh h ssc st ssc pr, fig.dim=fig_dim, results='asis'}
# Display the plots in tabs
for (i in names(neighbours_of_2)) {
  cat("## From the neighbours of", i, "{.tabset .tabset-pills}\n\n")
  for (j in names(neighbours_of_2[[i]])) {
    cat("### Frequency of", j, "\n\n")
    print(neighbours_of_2[[i]][[j]])
    cat("\n\n")
  }
}

rm(neighbours_of_2)
rm(freq_of_2)
```

# Create plots: Relative number of interacting cells

```{r plot number of interacting cells, echo=TRUE, fig.width=10, fig.height=8}

relative_counts_of_1 <- list()
for (i in unique(spe$pheno_clusters_new)) {
  spe_subset <- spe[, spe$pheno_clusters_new == i]
  interacting_with_1 <- list()
  for (j in unique(spe$pheno_clusters_new)) {
    # Filter the frequencies of the neighbours of interest in the target cells as vectors
    nbh_count <- as.data.frame(colData(spe_subset)[, "NeighCount"][paste0(j, ".count")][, 1]) %>%
      mutate(Count = !!sym(paste0(j, ".count")))
    
    # Create data frame that summarizes the frequencies of the neighbours of interest in the target cells
    for_plot3 <- data.frame(nbh_count[2],
                            "Disease" = spe_subset$Disease,
                            "Donor" = spe_subset$Donor) %>%
      group_by(Donor, Disease) %>%
      summarize (Count_rel = sum(Count, na.rm = T) / ncol(spe_subset))
    
    #for_plot2 <- table(for_plot$Donor,for_plot$Disease)
    
    plot3 <- ggplot(data = for_plot3, aes(x = Disease, y = Count_rel)) + geom_boxplot(outlier.shape = NA, aes(fill =
                                                                                                                Disease)) +
      stat_compare_means(
        comparisons = list(as.character(unique(
          for_plot3$Disease
        ))),
        method = "wilcox",
        hide.ns = TRUE,
        label.y = max(for_plot3$Count_rel) * 1.1
      ) +
      get_common_boxplot_components_fx() +
      scale_fill_manual(values = metadata(spe)$color_vectors$Disease) +
      labs(
        x = paste0("Relative count of ", i, " interacting with ", j),
        y = "Relative count",
        fill = "Disease"
      )
    interacting_with_1[[j]] <- plot3
  }
  relative_counts_of_1[[i]] <- interacting_with_1
}

relative_counts_of_2 <- list()
for (i in unique(spe$pheno_clusters_new)) {
  spe_subset <- spe[, spe$pheno_clusters_new == i]
  interacting_with_2 <- list()
  for (j in unique(spe$pheno_clusters_new)) {
    # Filter the frequencies of the neighbours of interest in the target cells as vectors
    nbh_count2 <- as.data.frame(colData(spe_subset)[, "NeighCount"][paste0(j, ".count")][, 1]) %>%
      mutate(Count = !!sym(paste0(j, ".count")))
    
    # Create data frame that summarizes the frequencies of the neighbours of interest in the target cells
    for_plot4 <- data.frame(nbh_count2[2],
                            "Disease" = spe_subset$Progression_skin,
                            "Donor" = spe_subset$Donor) %>%
      group_by(Donor, Disease) %>%
      summarize (Count_rel = sum(Count, na.rm = T) / ncol(spe_subset))
    
    if (max(for_plot4$Count_rel) > 0) {
      dunn_res <- as.data.frame(for_plot4) %>% rstatix::dunn_test (Count_rel ~ Disease, p.adjust.method =
                                                                     "none") %>%
        mutate (method = "Dunn")
      
      plot4 <- ggplot(data = for_plot4, aes(x = Disease, y = Count_rel)) + geom_boxplot(outlier.shape = NA, aes(fill =
                                                                                                                  Disease)) +
        stat_pvalue_manual(dunn_res %>%
                             mutate(y.position = c(
                               max(for_plot4$Count_rel) * 1.1,
                               max(for_plot4$Count_rel) *
                                 1.15,
                               max(for_plot4$Count_rel) *
                                 1.1
                             )), bracket.shorten = 0.05) +
        stat_compare_means(method = "kruskal",
                           label.y = max(for_plot4$Count_rel) * 1.2) +
        get_common_boxplot_components_fx() +
        scale_fill_manual(values = metadata(spe)$color_vectors$Progression) +
        labs(
          x = paste0("Relative count of ", i, " interacting with ", j),
          y = "Relative count",
          fill = "Disease"
        )
      interacting_with_2[[j]] <- plot4
    }
    else {
      print(paste0("No cells interacting with ", i))
    }
  }
  relative_counts_of_2[[i]] <- interacting_with_2
}
```

# Display plots: Relative number of interacting cells for healthy SSc {.tabset .tabset-pills}

`r fig_dim <- c(10, 8)`

```{r display number of interacting cells h ssc, fig.dim=fig_dim, results='asis'}
# Display the plots in tabs
for (i in names(relative_counts_of_1)) {
  cat("## Relative counts of", i, "{.tabset .tabset-pills}\n\n")
  for (j in names(relative_counts_of_1[[i]])) {
    cat("### Interacting with", j, "\n\n")
    print(relative_counts_of_1[[i]][[j]])
    cat("\n\n")
  }
}

rm(relative_counts_of_1)
rm(interacting_with_1)
```

# Display plots: Relative number of interacting cells for healthy SSc stable SSc progressive {.tabset .tabset-pills}

`r fig_dim <- c(10, 8)`

```{r display number of interacting cells h ssc st ssc pr, fig.dim=fig_dim, results='asis'}
# Display the plots in tabs
for (i in names(relative_counts_of_2)) {
  cat("## Relative counts of", i, "{.tabset .tabset-pills}\n\n")
  for (j in names(relative_counts_of_2[[i]])) {
    cat("### Interacting with", j, "\n\n")
    print(relative_counts_of_2[[i]][[j]])
    cat("\n\n")
  }
}

rm(relative_counts_of_2)
rm(interacting_with_2)
```


# Outputs

This analysis generates the following results: <br>

1. Plots of interaction frequencies of cells of interest with target cells, frequencies of a target cell among the neigbours of a cell of interest and relative number of cells of interest interacting with target cells. <br>

# Session information

```{r session, eval=TRUE}
etime <- proc.time() - ptm
cat('Rendering this document took :', etime[3], ' s \n')

sessioninfo::session_info()
