---
title: "SSc skin data from Aleix - data import"
author: "Yi-Nan Li"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
source("data_preprocessing/common/config.R")
knitr::opts_chunk$set(echo = TRUE)
```
> Text and R codes are adapted from Nils Eling at Bodenmiller Group, which is created for the Imaging Workshop 2023 More information can be found on [GitHub](https://github.com/BodenmillerGroup/ImagingWorkshop2023).

## Data import from steinbock

Run the following code chunk if the packages are **NOT** installed.

```{r install-packages, eval=FALSE}
# Install the dependencies listed in the repository environment documentation.
```

Now we load the 'imcRtools' and 'cytomapper' packages

```{r load-imcRtools, message=FALSE}
library(imcRtools)
library(cytomapper)
```

Now we load the data generated by steinbock. First, copy the scipt to
the steinbock folder to set up the working directory. Alternatively, it
can be set up by 'setwd()' function. The following chunk should return
the working directory and the information of the 'SpatialExperiment'
oject loaded.

```{r read-steinbock}
# Sets the working directory to where the script is located
getwd()
spe <- read_steinbock(getwd())
spe
```

Rename the sample ID and assign column names for each cell in the entire  dataset. 

*Hint: the excel file in the following example contains two columns, 'Sample ID' includes the name of the mcd file (e.g., Patient1) and 'indication' includes the disease or grouping condition (e.g., NH or SSc).*

```{r col-names}
library("stringr")
# rename sample ID
spe$patient_id[spe$sample_id == "Fibro_NH-Skin_193_196_20220131_Aleix_001"] <- "NH193"
spe$patient_id[spe$sample_id == "Fibro_NH-Skin_193_196_20220131_Aleix_002"] <- "NH196"
spe$patient_id[spe$sample_id == "Fibro_SSc-Skin_194733_20220214_Aleix_001"] <- "SSc194733"
spe$patient_id[spe$sample_id == "Fibro_SSc194994_20220207_Aleix_001"] <- "SSc194994"

# assign colnames
colnames(spe) <- paste0(spe$patient_id, "_", spe$ObjectNumber)

# assign indications
spe$indication[spe$patient_id %in% c("NH193", "NH196")] <- "NH"
spe$indication[spe$patient_id %in% c("SSc194733", "SSc194994")] <- "SSc"
```

## Data preprocessing

### Quality control by cell size

Here we plot the distribution of identified object sizes. 
```{r cell-size, message=FALSE, results='hold', fig.show='hold'}
library(dplyr)
library(ggplot2)
colData(spe) %>%
    as.data.frame() %>%
    group_by(patient_id) %>%
    ggplot() +
        geom_boxplot(aes(patient_id, area)) +
        theme_minimal(base_size = 15) + 
        theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8)) +
        ylab("Cell area") + xlab("")+
  ggtitle("before size cutoff")

# # size cut-off by percentile
# summary(spe$area)
# size.cutoff.lower <- quantile(spe$area, 0.01)
# cat(paste0("The 1st percentile of all object sizes is ", size.cutoff.lower, " µm^2. \n The objects smaller than this will be excluded.\n"))
# size.cutoff.upper <- quantile(spe$area, 0.99)
# cat(paste0("The 99th percentile of all object sizes is ", size.cutoff.upper, " µm^2. \n The objects larger than this will be excluded.\n"))

# manual size cut-off
size.cutoff.lower <- 10
size.cutoff.upper <- 300
cat(paste0("The lower and upper cut-off of object size are set to ",size.cutoff.lower, " and ", 
           size.cutoff.upper, " µm^2 \nThe object not within this range will be excluded!"))
n.exclude <- sum(sum(spe$area < size.cutoff.lower), sum(spe$area > size.cutoff.upper))

spe$size.filter[between(spe$area, size.cutoff.lower, size.cutoff.upper)] <- "include"
spe$size.filter[!between(spe$area,size.cutoff.lower, size.cutoff.upper)] <- "exclude"
spe.no.filter <- spe
spe <- spe[,spe$size.filter == "include"]

colData(spe) %>%
    as.data.frame() %>%
    group_by(patient_id) %>%
    ggplot() +
        geom_boxplot(aes(patient_id, area)) +
        theme_minimal(base_size = 15) + 
        theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8)) +
        ylab("Cell area") + xlab("")+
  ggtitle("after size cutoff")
```

<span style="font-size: 14pt;"> !!! There are **`r n.exclude` objects ** are excluded !!!

### Expression levels transformation

Asinh transformation with cofactor = 1 and min-max rescaling.

```{r asinh, message=FALSE, fig.show='hold'}
library(dittoSeq)
# Arcsinh transformation
assay(spe, "asinh") <- asinh(counts(spe)/1)


# Min-max rescaling
assay(spe, "rescaled") <- t(apply(assay(spe, "asinh"), 1, function(x){
  (x - min(x))/(max(x)-min(x))}))

# Ridge  plot
par(mfrow = c(3, 1)) # set plotting area into 3x1 array
dittoRidgePlot(spe, var = "PDGFRa", group.by = "patient_id", assay = "counts") +
    ggtitle("PDGFRa - before transformation") + 
  theme(plot.title = element_text(face = "bold"))
dittoRidgePlot(spe, var = "PDGFRa", group.by = "patient_id", assay = "asinh") +
    ggtitle("PDGFRa - after arcsinh transformation") + 
  theme(plot.title = element_text(face = "bold"))
dittoRidgePlot(spe, var = "PDGFRa", group.by = "patient_id", assay = "rescaled") +
    ggtitle("PDGFRa - after rescaling") + 
  theme(plot.title = element_text(face = "bold"))
```


*Hint: it is possible to create interactive plot by shiny. To do so, please directly run in RStudio*
```{r shiny-plot, echo=FALSE}
if (interactive()){
library(shiny)
# shiny app script is developed with ChatGPT Mar 23 Version.
library(ggplot2) # Add this line to load ggplot2
ui <- fluidPage(
  titlePanel("Expression ridge plot for data transformation"),
  sidebarLayout(position = "left",
                sidebarPanel("Select channels here", 
                             selectInput("ch", "Channel", choices = names(spe))),
                mainPanel("Expression ridge plots",
                          fluidRow(
                            column(width = 12, 
                                   plotOutput("p1", height = "300px")),
                            column(width = 12, 
                                   plotOutput("p2", height = "300px")),
                            column(width = 12, 
                                   plotOutput("p3", height = "300px"))
                          )
                )
  )
)
server <- shinyServer(function(input, output){
  
  # Reactive expression to get the selected channel
  channel <- reactive({
    input$ch
  })
  #  expression for plotting before transformation
  p1 <- reactive({
    dittoRidgePlot(spe, var = channel(), group.by = "patient_id", assay = "counts") +
      ggtitle(paste0(channel()," - before transformation")) + 
      theme(plot.title = element_text(face = "bold"))
  })
  #  expression for plotting after asinh transformation
  p2 <- reactive({
    dittoRidgePlot(spe, var = channel(), group.by = "patient_id", assay = "asinh") +
      ggtitle(paste0(channel()," - after transformation")) + 
      theme(plot.title = element_text(face = "bold"))
  })
  #  expression for plotting after rescaling
  p3 <- reactive({
    dittoRidgePlot(spe, var = channel(), group.by = "patient_id", assay = "rescaled") +
      ggtitle(paste0(channel()," - after rescaling")) + 
      theme(plot.title = element_text(face = "bold"))
  })
  
  # Render the plots
  output$p1 = renderPlot({p1()})
  output$p2 = renderPlot({p2()})
  output$p3 = renderPlot({p3()})
})

shinyApp(ui, server)
}
```

### Expression levels across samples

Here we plot the expression levels of every channels for each sample to detect sample-specific variation/ batch effects.

```{r expr-per-sample, message=FALSE, fig.width=7, fig.height=25}
library(dittoSeq)
multi_dittoPlot(spe, vars = rownames(spe),
               group.by = "patient_id", plots = "ridgeplot", 
               assay = "rescaled")
```




## Import images

### Construct cytomapper object

Here we import the image into Cytomapper object for visualization and further analysis. \

Set up channel names according to the `spe` object. The metadata for the images is also assigned here. \

The `on_disk` argument is turned on to store the images on the disk instead of memory. \


```{r import-images, message=FALSE}
library(cytomapper)
library(HDF5Array)
HDF5.Dump <- getHDF5DumpDir()
images <- loadImages(project_path("Steinbock/img"),
                     on_disk = TRUE, h5FilesPath = HDF5.Dump)
channelNames(images) <- names(spe)
images  ## printing the information of the images object
all.equal(names(images), names(masks)) ## confirm the names of images and masks are the identical

## assign metadata to the objects
# assign patient id
patient_id <- names(images)
patient_id[patient_id == "Fibro_NH-Skin_193_196_20220131_Aleix_001"] <- "NH193"
patient_id[patient_id == "Fibro_NH-Skin_193_196_20220131_Aleix_002"] <- "NH196"
patient_id[patient_id == "Fibro_SSc-Skin_194733_20220214_Aleix_001"] <- "SSc194733"
patient_id[patient_id == "Fibro_SSc194994_20220207_Aleix_001"] <- "SSc194994"

# extract indication from patient_id
library(stringr)
indication <- str_extract(patient_id, "^[A-z]{2,}")

# Store patient and image level information in elementMetadata
mcols(images) <- mcols(masks) <- DataFrame(sample_id = names(images),
                                           patient_id = patient_id,
                                           indication = indication)
mcols(images) ## printing the metadata of the images object
```
### Seqmentation quality control

Here we select 1 sample to plot the cell mask along with the marker expression with cytomapper.

```{r cytomapper-control, message=FALSE}
library(cytomapper)

# Normalize and clip images
cur_images <- images
cur_images <- normalize(cur_images, separateImages = TRUE)
cur_images <- normalize(cur_images, inputRange = c(0, 0.2))
plotPixels(cur_images,
           mask = masks,
           object = spe.no.filter,
           img_id = "sample_id",
           cell_id = "ObjectNumber",
           outline_by = "size.filter",
           missing_colour = "orange",
           colour_by = c("PDGFRa", "CD31", "E_cadherin", "Histone3"),
           colour = list(PDGFRa = c("black", "green"),
                         CD31 = c("black", "red"),
                         E_cadherin = c("black", "cyan"),
                         Histone3 = c("black", "blue"),
                         # indication = unlist(lapply(unique(mcols(cur_images)$indication), function(x){
                         #   col <- rep("white", length(x))
                         #   names(col) <- x
                         #   col
                         #   })),
                         size.filter = c(include = "white", exclude = "orange")),
           image_title = list(text = mcols(images)$patient_id),
           thick = TRUE, display ="single",
           legend = list(colour_by.title.cex = 0.7,
                         colour_by.labels.cex = 0.7))
```




*Hint: interactive cytomapper plot can be created to explore the data. To do so, please directly run the markdown file in RStudio*


```{r cytoviewer, eval=FALSE, message=FALSE}
library(cytoviewer)
# Use cytoviewer with images, masks and object
app <- cytoviewer(image = images, 
                  mask = masks, 
                  object = spe, 
                  img_id = "patient_id", 
                  cell_id = "ObjectNumber")

if (interactive()) {
  
  shiny::runApp(app, launch.browser = TRUE)

  }
```


## Save data

The R objects can be saved for future analysis.

```{r save-data, message=FALSE}
dir.create("data")
```


## Session information

```{r session-info}
sessionInfo()
```
