---
title: "First-cohort fibroblast frequencies: total and local mRSS"
output:
  html_document:
    self_contained: true
    toc: true
    toc_float: true
    code_folding: hide
date: "`r format(Sys.Date(), '%d %B %Y')`"
---

```{r setup, include=FALSE}
source("data_preprocessing/common/config.R")
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center", dpi = 180)
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(ggtext)
  library(patchwork)
  library(scales)
  library(knitr)
})
source("first_cohort_mrss_common.R")

output_dir <- file.path(getwd(), "first_cohort_total_local_mRSS_frequency_outputs")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

## Scope

This report evaluates fibroblast metabolic-cluster frequencies without neighborhood conditioning. It reproduces the three originally selected current-total-mRSS plots, applies the same selections to local mRSS, and ranks all eight fibroblast metabolic clusters by their local-mRSS association.

Each frequency is calculated per donor relative to all stromal cells. Donors, rather than cells, are the statistical units.

## Superscript audit

The metabolic-state labels use the same validated vector-safe superscripts throughout: `r superscript_state["hi"]` for hi, `r superscript_state["int"]` for int, and `r superscript_state["low"]` for low.

```{r label-audit, echo=FALSE}
kable(
  label_audit %>% filter(grepl("^stromal_", cluster_id)),
  caption = "Validated fibroblast metabolic-cluster labels"
)
```

## Donor-level frequencies

```{r calculate-frequencies}
inputs <- load_first_cohort_mrss_data()
clinical <- inputs$clinical

cluster_definitions <- tibble(
  metabolic_cluster = paste0("stromal_", 1:8),
  display_label = unname(cluster_labels[paste0("stromal_", 1:8)]),
  display_label_single = unname(cluster_labels_single_line[paste0("stromal_", 1:8)]),
  display_label_html = unname(cluster_labels_html[paste0("stromal_", 1:8)]),
  display_label_html_single = unname(cluster_labels_html_single_line[paste0("stromal_", 1:8)])
)

original_selection <- tibble(
  panel = c("A", "B", "C"),
  metabolic_cluster = c("stromal_2", "stromal_4", "stromal_8")
)

stromal_cells <- inputs$cell_data %>%
  filter(indication == "SSc", celltype == "Stromal")

donor_denominators <- stromal_cells %>%
  dplyr::count(patient_id, name = "all_stromal_cells")

frequency_long <- stromal_cells %>%
  dplyr::count(patient_id, metabolic_cluster, name = "cluster_cells") %>%
  filter(metabolic_cluster %in% cluster_definitions$metabolic_cluster) %>%
  complete(
    patient_id = donor_denominators$patient_id,
    metabolic_cluster = cluster_definitions$metabolic_cluster,
    fill = list(cluster_cells = 0L)
  ) %>%
  left_join(donor_denominators, by = "patient_id") %>%
  mutate(frequency_percent = 100 * cluster_cells / all_stromal_cells) %>%
  left_join(cluster_definitions, by = "metabolic_cluster") %>%
  left_join(clinical, by = "patient_id")

write.csv(
  frequency_long,
  file.path(output_dir, "all_8_donor_level_fibroblast_cluster_frequencies.csv"),
  row.names = FALSE
)

donor_audit <- frequency_long %>%
  distinct(patient_id, current_mRSS, local_mRSS, included_local_mRSS, all_stromal_cells) %>%
  arrange(desc(included_local_mRSS), patient_id)
write.csv(donor_audit, file.path(output_dir, "donor_inclusion_and_stromal_denominators.csv"), row.names = FALSE)
```

## Correlation audit

```{r calculate-correlations}
current_stats <- summarise_correlations(
  frequency_long,
  score_variable = "current_mRSS",
  group_variables = c(
    "metabolic_cluster", "display_label", "display_label_single",
    "display_label_html", "display_label_html_single"
  )
) %>%
  mutate(q_BH_8_tests = p.adjust(p_spearman_raw, method = "BH"))

local_stats <- summarise_correlations(
  frequency_long,
  score_variable = "local_mRSS",
  group_variables = c(
    "metabolic_cluster", "display_label", "display_label_single",
    "display_label_html", "display_label_html_single"
  ),
  exact_binary = TRUE
) %>%
  mutate(q_BH_8_tests = p.adjust(p_spearman_raw, method = "BH"))

selected_current_stats <- original_selection %>%
  left_join(current_stats, by = "metabolic_cluster")
selected_local_stats <- original_selection %>%
  left_join(local_stats, by = "metabolic_cluster")

write.csv(current_stats, file.path(output_dir, "all_8_current_total_mRSS_frequency_correlations.csv"), row.names = FALSE)
write.csv(local_stats, file.path(output_dir, "all_8_local_mRSS_frequency_correlations.csv"), row.names = FALSE)

# Values independently reproduced in the preceding denominator audit.
expected_current <- tibble(
  metabolic_cluster = c("stromal_2", "stromal_4", "stromal_8"),
  expected_rho = c(0.569042637953892, -0.00836827408755723, 0.510464719340991),
  expected_p = c(0.109804000209182, 0.982953286833085, 0.160266667914605)
)

reproduction_check <- selected_current_stats %>%
  left_join(expected_current, by = "metabolic_cluster") %>%
  mutate(
    rho_absolute_difference = abs(rho - expected_rho),
    p_absolute_difference = abs(p_spearman_raw - expected_p)
  )

write.csv(reproduction_check, file.path(output_dir, "original_total_mRSS_reproduction_check.csv"), row.names = FALSE)
stopifnot(
  max(reproduction_check$rho_absolute_difference) < 1e-12,
  max(reproduction_check$p_absolute_difference) < 1e-12
)
```

The original current-total-mRSS estimates are reproduced exactly: `rho = 0.5690`, `-0.00837`, and `0.5105` for stromal clusters 2, 4, and 8, respectively.

## Original selected plots: total mRSS

```{r correlation-plot-function}
make_frequency_scatter <- function(cluster_id, score = c("current", "local")) {
  score <- match.arg(score)
  score_variable <- if (score == "current") "current_mRSS" else "local_mRSS"
  result_table <- if (score == "current") current_stats else local_stats

  plot_data <- frequency_long %>%
    filter(metabolic_cluster == cluster_id, !is.na(.data[[score_variable]]))
  result <- result_table %>% filter(metabolic_cluster == cluster_id)
  max_y <- max(plot_data$frequency_percent)
  score_range <- range(plot_data[[score_variable]])
  annotation_x <- if (score == "local") {
    0.94
  } else {
    score_range[1] + 0.02 * diff(score_range)
  }

  annotation <- paste0(
    "\u03c1 = ", sprintf("%.2f", result$rho),
    ", p = ", format_p(result$p_spearman_raw)
  )

  p <- ggplot(plot_data, aes(x = .data[[score_variable]], y = frequency_percent)) +
    geom_smooth(
      method = "lm",
      formula = y ~ x,
      se = TRUE,
      color = "blue",
      fill = "grey82",
      linewidth = 0.9,
      alpha = 0.65,
      fullrange = TRUE
    )

  if (score == "local") {
    p <- p + geom_jitter(
      width = 0.035, height = 0, shape = 21, size = 3,
      stroke = 0.65, color = "grey45", fill = "white"
    ) +
      scale_x_continuous(breaks = c(1, 2), expand = expansion(mult = 0)) +
      coord_cartesian(xlim = c(0.9, 2.1), ylim = c(0, max_y * 1.14), expand = FALSE)
  } else {
    p <- p + geom_point(shape = 21, size = 3, stroke = 0.65, color = "grey45", fill = "white") +
      scale_x_continuous(breaks = pretty_breaks(n = 5), expand = expansion(mult = c(0.03, 0.03))) +
      coord_cartesian(ylim = c(0, max_y * 1.14), expand = FALSE)
  }

  p +
    scale_y_continuous(expand = expansion(mult = c(0, 0.02))) +
    annotate(
      "text",
      x = annotation_x,
      y = max_y * 1.08,
      label = annotation,
      hjust = 0,
      vjust = 1,
      size = 4.1,
      fontface = "italic"
    ) +
    labs(
      title = unique(plot_data$display_label_html_single),
      x = if (score == "local") "Local mRSS" else "mRSS",
      y = "% of all fibroblasts"
    ) +
    theme_classic(base_size = 14) +
    theme(
      plot.title = ggtext::element_markdown(size = 15, hjust = 0.5, margin = margin(b = 8)),
      axis.title = element_text(size = 14),
      axis.text = element_text(size = 12, color = "black"),
      axis.line = element_line(linewidth = 0.55),
      axis.ticks = element_line(linewidth = 0.55),
      plot.margin = margin(10, 10, 8, 10)
    )
}

current_selected_plots <- setNames(
  lapply(original_selection$metabolic_cluster, make_frequency_scatter, score = "current"),
  original_selection$panel
)
local_selected_plots <- setNames(
  lapply(original_selection$metabolic_cluster, make_frequency_scatter, score = "local"),
  original_selection$panel
)

for (panel_id in original_selection$panel) {
  save_vector_and_png(
    current_selected_plots[[panel_id]],
    file.path(output_dir, paste0("original_panel_", panel_id, "_current_total_mRSS")),
    4.6, 3.6
  )
  save_vector_and_png(
    local_selected_plots[[panel_id]],
    file.path(output_dir, paste0("original_panel_", panel_id, "_local_mRSS")),
    4.6, 3.6
  )
}

combine_three <- function(plot_list) {
  (plot_list$A + plot_list$B + plot_list$C) +
    plot_layout(design = "\nAABB\n#CC#\n") +
    plot_annotation(tag_levels = "A") &
    theme(plot.tag = element_text(size = 20, face = "plain"))
}

combined_current <- combine_three(current_selected_plots)
combined_local <- combine_three(local_selected_plots)
save_vector_and_png(combined_current, file.path(output_dir, "combined_original_frequencies_current_total_mRSS"), 9.2, 7.2)
save_vector_and_png(combined_local, file.path(output_dir, "combined_original_frequencies_local_mRSS"), 9.2, 7.2)
```

```{r show-current, echo=FALSE, fig.width=9.2, fig.height=7.2}
combined_current
```

## Same selected plots: local mRSS

```{r show-local, echo=FALSE, fig.width=9.2, fig.height=7.2}
combined_local
```

## All fibroblast clusters ranked by local mRSS

```{r ranked-local}
ranked_local <- local_stats %>%
  arrange(desc(abs(rho)), p_spearman_raw) %>%
  mutate(display_label_html = factor(display_label_html, levels = rev(display_label_html)))

ranked_local_plot <- ggplot(ranked_local, aes(x = rho, y = display_label_html)) +
  geom_vline(xintercept = 0, color = "grey70", linewidth = 0.4) +
  geom_segment(aes(x = 0, xend = rho, yend = display_label_html), color = "grey73", linewidth = 0.65) +
  geom_point(aes(size = abs(rho), color = p_spearman_raw)) +
  scale_color_gradient(
    name = "Raw p value",
    low = "red",
    high = "blue",
    limits = c(0.001, 0.05),
    na.value = "grey52"
  ) +
  scale_size_continuous(
    name = "|Spearman \u03c1|",
    range = c(3, 9),
    limits = c(0, 0.30),
    breaks = c(0.1, 0.2, 0.3)
  ) +
  scale_x_continuous(breaks = pretty_breaks(n = 5), expand = expansion(mult = c(0.08, 0.08))) +
  labs(
    title = "Fibroblast frequencies correlated with local mRSS",
    subtitle = "All eight metabolic fibroblast clusters",
    x = "Spearman correlation",
    y = NULL
  ) +
  theme_bw(base_size = 13) +
  theme(
    plot.title = element_text(size = 15, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    axis.text.y = ggtext::element_markdown(size = 12, lineheight = 0.95),
    axis.text.x = element_text(size = 11),
    panel.grid.minor = element_blank(),
    plot.margin = margin(8, 12, 8, 8)
  )

save_vector_and_png(
  ranked_local_plot,
  file.path(output_dir, "all_8_local_mRSS_fibroblast_frequency_correlations_lollipop"),
  8.5, 6.2
)
```

```{r show-ranked-local, echo=FALSE, fig.width=8.5, fig.height=6.2}
ranked_local_plot
```

### Donor-level views of the three strongest clusters

```{r top-three-local-details}
top_three_local <- local_stats %>%
  arrange(desc(abs(rho)), p_spearman_raw) %>%
  slice_head(n = 3) %>%
  mutate(rank = row_number())

make_top_cluster_detail <- function(rank_value) {
  definition <- top_three_local %>% filter(rank == rank_value)
  plot_data <- frequency_long %>%
    filter(metabolic_cluster == definition$metabolic_cluster, !is.na(local_mRSS)) %>%
    mutate(local_mRSS_group = factor(local_mRSS, levels = c(1, 2)))

  annotation <- paste0(
    "\u03c1 = ", sprintf("%.2f", definition$rho),
    ", raw p = ", format_p(definition$p_spearman_raw),
    "\nexact p = ", format_p(definition$p_exact_permutation),
    ", BH q = ", format_p(definition$q_BH_8_tests)
  )

  ggplot(plot_data, aes(x = local_mRSS_group, y = frequency_percent, fill = local_mRSS_group)) +
    geom_boxplot(width = 0.5, outlier.shape = NA, linewidth = 0.65, alpha = 0.65) +
    geom_jitter(width = 0.08, height = 0, shape = 21, size = 3.1, stroke = 0.65, color = "black") +
    annotate("text", x = 1.02, y = max(plot_data$frequency_percent) * 1.13,
             label = annotation, hjust = 0, vjust = 1, size = 3.7) +
    scale_fill_manual(values = c("1" = "#C9D8E8", "2" = "#EFA19A"), guide = "none") +
    scale_y_continuous(expand = expansion(mult = c(0.03, 0.20))) +
    labs(title = definition$display_label_html_single, x = "Local mRSS", y = "% of all fibroblasts") +
    theme_classic(base_size = 13) +
    theme(
      plot.title = ggtext::element_markdown(size = 13, hjust = 0.5, margin = margin(b = 8)),
      axis.title = element_text(size = 12),
      axis.text = element_text(size = 11, color = "black"),
      plot.margin = margin(10, 10, 8, 10)
    )
}

top_cluster_plots <- setNames(lapply(1:3, make_top_cluster_detail), paste0("rank", 1:3))
for (i in 1:3) {
  save_vector_and_png(
    top_cluster_plots[[i]],
    file.path(output_dir, paste0("top_local_frequency_rank_", i, "_donor_level")),
    5.0, 4.1
  )
}

combined_top_clusters <- top_cluster_plots$rank1 + top_cluster_plots$rank2 + top_cluster_plots$rank3 +
  plot_layout(ncol = 3) + plot_annotation(tag_levels = "A")
save_vector_and_png(
  combined_top_clusters,
  file.path(output_dir, "top_3_local_mRSS_fibroblast_frequencies_combined"),
  15, 4.3
)
```

```{r show-top-three, echo=FALSE, fig.width=15, fig.height=4.3}
combined_top_clusters
```

## Statistical table

```{r local-table, echo=FALSE}
kable(
  local_stats %>%
    arrange(desc(abs(rho))) %>%
    select(display_label_single, n_donors, rho, p_spearman_raw, p_exact_permutation, q_BH_8_tests),
  digits = 4,
  caption = "All eight fibroblast-cluster frequency correlations with local mRSS"
)
```

## Interpretation

- The three total-mRSS plots reproduce the original prespecified figure.
- The local-mRSS ranking evaluates all eight fibroblast metabolic clusters and does not introduce an unreported denominator change.
- Local mRSS is available for eight donors and has only two observed values. Exact permutation p-values and BH q-values are therefore reported with the Spearman estimates.
- The top-three donor-level plots are exploratory visualizations of the ranking, not independent confirmatory tests.

```{r session-info}
sessionInfo()
```
