---
title: "Frequencies of clusters - box plots and radar charts"
author: "Alex Matei"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  html_document: 
    toc: true
    toc_float: true
    toc_depth: 4
    df_print: paged
    code_folding: hide
---

```{r setup, include=FALSE,results='hide'}
source("data_preprocessing/common/config.R")
knitr::opts_chunk$set(warning = FALSE, message = FALSE, eval=FALSE) 
```

# Load spe

```{r load data, eval=TRUE}

ts <- format(Sys.time(), "%Y%m%d%H%M%S")
ptm <- proc.time()

cat("This document was created with the time stamp",ts)

library(tidyverse)
library(imcRtools)
library(SpatialExperiment)
library(scuttle)
library(CATALYST)
library(dittoSeq)
library(ggpubr)
library(rstatix)
library(ggstatsplot)
library(ggradar)
library(RColorBrewer)
library(fmsb)

source(file.path(here(),"config.R"))

dir<- here()

spe <- readRDS(file.path(dir,"Datasets/spe20250127122004.rds"))

dir_r <- "data_preprocessing/common"
source(file.path(dir_r, "output_dir_svg_plots_fx.R"))

output_dir <- output_dir_svg_plots_fx(ts)

```

# Prepare spe for analysis
```{r prep analysis, eval=TRUE}

# define color vectors
color_vectors <- list()
pheno_clusters_new <- setNames(c("#C4281C","#C2DAB8","#E8BAC8","#80BBDB","#C470A0","#0D69AC","#F5CD30","#287F47",
                                 "#A05F35","#D9856C","#6B327C","#008F9C","#C7D23C","#B7D7D5","#923978","#27462D",
                                 "#FF9494","#938767","#FFF67B","#E1A4C2","#FC7B1B","#CD6298","#102ADC"), 
                               as.character(unique(spe$pheno_clusters_new)))

indication <- setNames(c("#4DBBD533", "#E64B35AA"),
                       c(unique(colData(spe)$Disease)))

patient_id <- setNames(c("#FFFF33", "#FFCC66",'#FF9933','#FF6633','#00FFFF','#FF0033','#79B5B5', '#C470A0',
                         '#996633','#999933','#99CC33','#FF9494', '#33FF33','#336633', '#330066','#923978'),
                       c(unique(colData(spe)$patient_id)))

Neighbourhood_del_80_pheno_new <- setNames(c("#CC00CC", "#260EEC", "#FFCA28", "#004D40","#FF6633","#66FF66", 
                                             "#EF9EA7", "#990000", "#CC0033","#4FC3F7"),
                                           unique(spe$Neighbourhood_del_80_pheno_new))

spe$Progression_skin <- case_when(spe$Progression_skin=="progressive"~"Progressive SSc",
                                  spe$Progression_skin=="stable"~"Stable SSc",
                                  TRUE~"Healthy")

progressive_skin <- setNames(c("#6CC3F4", "#C03830", "#F6BF93"), unique(spe$Progression_skin))

color_vectors$clusters <- pheno_clusters_new
color_vectors$Disease <- indication
color_vectors$patient_id <- patient_id
color_vectors$neighbourhood_del_80_pheno_new <- Neighbourhood_del_80_pheno_new
color_vectors$progressive_skin <- progressive_skin
metadata(spe)$color_vectors <- color_vectors


# define an spe of all cells except epithelial cells
spe_not_Epith <- spe[,which(!spe$new_celllabels=="E-cad+")]

# Store reusable boxplot components in a variable
common_plot_components <- list(
  geom_point(position = position_dodge2(width = 0.5, padding = 1.5)),
  theme_minimal(),
  theme(
    axis.text.x = element_text(size = 14, face = "bold", angle = 45, hjust = 1),
    axis.text.y = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 14, face = "bold"),
    legend.title = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 14, face = "bold")
  )
)

```

# Boxplot frequencies selected plots - healthy and SSc 

```{r boxplot selected h_ssc, fig.width=5.5, fig.height=5}

frequency_var <- "CN_10_aggNbh_delaunay_80_new"
spe[[frequency_var]] <- as.character(spe[[frequency_var]])
spe_subset <- spe[,!spe[[frequency_var]]%in%c("1","2","3")]
data <- dittoSeq::dittoFreqPlot(spe_subset, var = frequency_var, 
                                  group.by = "Disease", 
                                  sample.by = "Donor",
                                  data.out = TRUE)$data

    data2 <- data %>% filter(label=="9")
    # Generate the boxplot frequencies
    boxplot_plot_freq <- ggplot(data = data2, aes(x = Disease, y = percent, fill = Disease)) +
      geom_boxplot(outlier.shape = NA) +
      stat_compare_means(comparisons=list(as.character(unique(data2$Disease))),
                         method="wilcox", 
                         hide.ns = TRUE,
                         size=5,
                         label.y=max(data2$percent)+0.1)+
      #facet_wrap(~label, nrow=2, scales = "free") +
      # stat_compare_means(
      #   comparisons = my_comparisons,
      #   method = "wilcox.test",
      #   aes(label = "p.format"),
      #   hide.ns = FALSE
      # ) +
      get_common_boxplot_components_fx() +
      scale_fill_manual(values = metadata(spe)$color_vectors$Disease) +
      labs(x = "Cluster", y = "Frequency") 
    boxplot_plot_freq
```


# Boxplot frequencies selected plots - healthy, SSc progressive, SSc stable

```{r boxplot selected h_prog_stable, fig.width=5.5, fig.height=5}

frequency_var <- "CN_10_aggNbh_delaunay_80_new"
spe[[frequency_var]] <- as.character(spe[[frequency_var]])
spe_subset <- spe[, !spe[[frequency_var]] %in% c("1", "2", "3")]
data <- dittoSeq::dittoFreqPlot(
  spe_subset,
  var = frequency_var,
  group.by = "Progression_skin",
  sample.by = "Donor",
  data.out = TRUE
)$data

data2 <- data %>% filter(label == "9")
dunn_res <- as.data.frame(data2) %>% rstatix::dunn_test (percent ~ Progression_skin, p.adjust.method =
                                                           "none") %>%
  mutate (method = "Dunn")

# Generate the boxplot
boxplot_plot_freq <- ggplot(data = data2, aes(x = Progression_skin, y = percent)) +
  geom_boxplot(outlier.shape = NA, aes(fill = Progression_skin)) +
  stat_pvalue_manual(dunn_res %>%
                       mutate(y.position = c(
                         max(data2$percent + 0.1),
                         max(data2$percent + 0.15),
                         max(data2$percent + 0.1)
                       )), bracket.shorten = 0.05) +
  stat_compare_means(method = "kruskal", label.y = max(data2$percent +
                                                         0.25)) +
  # stat_compare_means(
  #   comparisons = my_comparisons,
  #   method = "wilcox.test",
  #   aes(label = "p.format"),
  #   hide.ns = FALSE
  # ) +
  get_common_boxplot_components_fx() +
  scale_fill_manual(values = metadata(spe)$color_vectors$progressive_skin) +
  labs(x = "Cluster", y = "Frequency")
boxplot_plot_freq
```

# Outputs

This analysis generates the following results: <br>

1. Plots of frequencies of selected clusters or CNs, split between healthy and SSc and between healthy, SSc stable and SSc progressive, for the metabolic skin IMC data, as boxplots. <br>

# Session information

```{r session, eval=TRUE}
etime <- proc.time() - ptm
cat('Rendering this document took :', etime[3], ' s \n')

sessioninfo::session_info()
```
