---
title: "Frequencies of clusters - box plots and radar charts"
author: "Alex Matei"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  html_document: 
    toc: true
    toc_float: true
    toc_depth: 4
    df_print: paged
    code_folding: hide
---

```{r setup, include=FALSE,results='hide'}
source("data_preprocessing/common/config.R")
knitr::opts_chunk$set(warning = FALSE, message = FALSE, eval=FALSE) 
```

# Load spe

```{r load data, eval=TRUE}

ts <- format(Sys.time(), "%Y%m%d%H%M%S")
ptm <- proc.time()

cat("This document was created with the time stamp",ts)

library(tidyverse)
library(imcRtools)
library(SpatialExperiment)
library(scuttle)
library(CATALYST)
library(dittoSeq)
library(ggpubr)
library(rstatix)
library(ggstatsplot)
library(ggradar)
library(RColorBrewer)
library(fmsb)

dir<- c(paste0(project_path(""), "/"))

load(paste0(dir,"spe_list.RData"))
spe <- spe_list$spe
rm(spe_list)

dir_r <- "data_preprocessing/common"
source(file.path(dir_r, "output_dir_svg_plots_fx.R"))

output_dir <- output_dir_svg_plots_fx(ts)

```

# Prepare spe for analysis
```{r prep analysis, eval=TRUE}

# Create a new data frame with all the colData of the spe object - to recode the patient_id column

metadata <- colData(spe)

# add a new column to metadata called Donor where you recode the patient_id column
# the elements starting with NH should be called Healthy1, Healthy2, etc.
# the elements starting with SSc should be called SSc1, SSc2, etc.
metadata$Donor <- dplyr::recode(metadata$patient_id,
                                     Healthy1="Healthy1",
                                     Healthy2="Healthy2", 
                                     SSc1="SSc1",
                                     SSc2="SSc2",
                                     SSc3="SSc3",
                                     SSc4="SSc4",
                                     SSc5="SSc5",
                                     SSc6="SSc6",
                                     SSc7="SSc7",
                                     SSc8="SSc8",
                                     Healthy3="Healthy3",
                                     Healthy4="Healthy4",
                                     SSc9="SSc9",
                                     Healthy5="Healthy5",
                                     Healthy6="Healthy6",
                                     Healthy7="Healthy7")

metadata$Disease <- dplyr::recode(metadata$indication,
                                  NH="Healthy",
                                  SSc="SSc")

metadata$ILD <- dplyr::recode(metadata$ILD,
                                     healthy="healthy",
                                     "1"="ILD", 
                                     "0"="no ILD")

metadata$Arthritis[is.na(metadata$Arthritis)] <- "no arthritis"

# Assign the new column in metadata to the spe object
spe$Donor <- metadata$Donor
spe$ILD <- metadata$ILD
spe$Arthritis <- metadata$Arthritis

spe$Disease <- metadata$Disease

names(colData(spe))[which(names(colData(spe))=="Skin.progression.at.biopsy")]="Progression_skin"

met_markers <- c("Glut1",
                 "HK1",
                 "PFKL1",
                 "PKM2",
                 "LDHA",
                 "GAPDH",
                 "CS",
                 "OGDH",
                 "ATP5A",
                 "SDHA",
                 "CD98",
                 "ACAC",
                 "CPT1a",
                 "G6PD",
                 "p_mTOR",
                 "pAMPK",
                 "NOX4",
                 "HIF1a",
                 "PGC-1a",
                 "TFAM",
                 "NRF2_p")

glycolysis_markers <- c("Glut1",
                 "HK1",
                 "PFKL1",
                 "PKM2",
                 "LDHA",
                 "GAPDH")

no_met_markers <- setdiff(rownames(spe),met_markers) 
rowData(spe)$use_channel_met_markers <- rownames(spe) %in% met_markers
rowData(spe)$use_channel_glyc_markers <- rownames(spe) %in% glycolysis_markers

# define clinical categories
clinical_cat <- c("Disease","Progression_skin", "Arthritis", "DU.history", "ILD", "Subtype.(dc/lc)", "Early.or.not")

spe[[clinical_cat[1]]] <- factor(spe[[clinical_cat[1]]], levels = c("Healthy", "SSc"))
spe[[clinical_cat[2]]] <- factor(spe[[clinical_cat[2]]], levels = c("healthy", "progressive", "stable"))
spe[[clinical_cat[3]]] <- factor(spe[[clinical_cat[3]]], levels = c("healthy","arthritis","no arthritis"))
spe[[clinical_cat[4]]] <- factor(spe[[clinical_cat[4]]], levels = c("healthy","DU history","no DU history"))
spe[[clinical_cat[5]]] <- factor(spe[[clinical_cat[5]]], levels = c("healthy","ILD","no ILD"))
spe[[clinical_cat[6]]] <- factor(spe[[clinical_cat[6]]], levels = c("healthy","dcSSc","lcSSc"))
spe[[clinical_cat[7]]] <- factor(spe[[clinical_cat[7]]], levels = c("healthy","early","late"))

# define color vectors
color_vectors <- list()
pheno_clusters_new <- setNames(c("#C4281C","#C2DAB8","#E8BAC8","#80BBDB","#C470A0","#0D69AC","#F5CD30","#287F47",
                                 "#A05F35","#D9856C","#6B327C","#008F9C","#C7D23C","#B7D7D5","#923978","#27462D",
                                 "#FF9494","#938767","#FFF67B","#E1A4C2","#FC7B1B","#CD6298","#102ADC"), 
                               as.character(unique(spe$pheno_clusters_new)))

indication <- setNames(c("#4DBBD533", "#E64B35AA"),
                       c(unique(colData(spe)$Disease)))

patient_id <- setNames(c("#FFFF33", "#FFCC66",'#FF9933','#FF6633','#00FFFF','#FF0033','#79B5B5', '#C470A0',
                         '#996633','#999933','#99CC33','#FF9494', '#33FF33','#336633', '#330066','#923978'),
                       c(unique(colData(spe)$patient_id)))

Neighbourhood_del_80_pheno_new <- setNames(c("#CC00CC", "#260EEC", "#FFCA28", "#004D40","#FF6633","#66FF66", 
                                             "#EF9EA7", "#990000", "#CC0033","#4FC3F7"),
                                           unique(spe$Neighbourhood_del_80_pheno_new))

progressive_skin <- setNames(c("#6CC3F4", "#C03830", "#F6BF93"), unique(spe$Progression_skin))

color_vectors$clusters <- pheno_clusters_new
color_vectors$Disease <- indication
color_vectors$patient_id <- patient_id
color_vectors$neighbourhood_del_80_pheno_new <- Neighbourhood_del_80_pheno_new
color_vectors$progressive_skin <- progressive_skin
metadata(spe)$color_vectors <- color_vectors


# define an spe of all cells except epithelial cells
spe_not_Epith <- spe[,which(!spe$new_celllabels=="E-cad+")]

# Store reusable boxplot components in a variable
common_plot_components <- list(
  geom_point(position = position_dodge2(width = 0.5, padding = 1.5)),
  theme_minimal(),
  theme(
    axis.text.x = element_text(size = 14, face = "bold", angle = 45, hjust = 1),
    axis.text.y = element_text(size = 14, face = "bold"),
    legend.text = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 14, face = "bold"),
    legend.title = element_text(size = 14, face = "bold"),
    plot.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 14, face = "bold")
  )
)

```

# Boxplots frequencies and densities - healthy and SSc {.tabset .tabset-pill}

`r fig_dim <- c(6, 5)`

```{r boxplot h_ssc, eval=TRUE, fig.dim=fig_dim, results='asis'}

# Define the cell types to plot
celltypes <- c("Stromal", "CD31+", "CD45+CD68+", "CD45+CD3+")
spe_subset <- NULL
boxplot_plot_list <- list()

for (i in celltypes) {
  # Subset data for the current cell type
  spe_subset <- spe[, spe$new_celllabels == i]
  spe_subset$pheno_clusters_new <- as.character(spe_subset$pheno_clusters_new)
  
  # Obtain cell frequencies
  data <- dittoSeq::dittoFreqPlot(spe_subset, var = "pheno_clusters_new", 
                                  group.by = "Disease", 
                                  sample.by = "Donor",
                                  data.out = TRUE)$data

  for (j in unique(spe_subset$pheno_clusters_new)){
    data2 <- data %>% filter(label==j)
    # Generate the boxplot frequencies
    boxplot_plot_freq <- ggplot(data = data2, aes(x = Disease, y = percent, fill = Disease)) +
      geom_boxplot(outlier.shape = NA) +
      stat_compare_means(comparisons=list(as.character(unique(data2$Disease))),
                         method="wilcox", 
                         hide.ns = TRUE,
                         label.y=max(data2$percent)+0.1)+
      #facet_wrap(~label, nrow=2, scales = "free") +
      # stat_compare_means(
      #   comparisons = my_comparisons,
      #   method = "wilcox.test",
      #   aes(label = "p.format"),
      #   hide.ns = FALSE
      # ) +
      common_plot_components +
      scale_fill_manual(values = metadata(spe)$color_vectors$Disease) +
      labs(x = "Cluster", y = "Frequency") +
      ggtitle(paste0("Frequency of ",j, " in healthy and SSc"))
    

# calculate cell density 
    coldata.density <- data.frame(colData(spe_subset)) %>% 
      dplyr::select(Donor, img_area, pheno_clusters_new, Disease) %>% 
      group_by(Donor, Disease, pheno_clusters_new) %>%
      summarise(density = n()/mean(img_area))
coldata.density2 <- coldata.density %>% filter(pheno_clusters_new==j)
    
    # Generate the boxplot
    boxplot_plot_dens <- ggplot(data = coldata.density2, aes(x = Disease, y = density, fill = Disease)) +
      geom_boxplot(outlier.shape = NA) +
      stat_compare_means(comparisons=list(as.character(unique(coldata.density2$Disease))),
                           method="wilcox", label.y=max(coldata.density2$density)*1.1)+
#      facet_wrap(~pheno_clusters_new, nrow=2, scales = "free") +
      # stat_compare_means(
      #   comparisons = my_comparisons,
      #   method = "wilcox.test",
      #   aes(label = "p.format"),
      #   hide.ns = FALSE
      # ) +
      common_plot_components +
      scale_fill_manual(values = metadata(spe)$color_vectors$Disease) +
      labs(x = "Cluster", y = "Density per mm2") +
      ggtitle(paste0("Density of ",j, " in healthy and SSc"))
    
      # Store plots in respective lists
  plot_name_freq <- paste0("freq_of_",j,"_from_",i,"_plot")
  plot_name_dens <- paste0("dens_of_",j,"_from_",i,"_plot")
  boxplot_plot_list[[plot_name_freq]] <- boxplot_plot_freq
  boxplot_plot_list[[plot_name_dens]] <- boxplot_plot_dens
  }
}

for (i in celltypes) {
  cat(sprintf("\n\n## %s {.tabset .tabset-pill}\n", i))
  name_celltype_plot <- grep(paste0("_from_", i, "_plot"), names(boxplot_plot_list), value = TRUE, fixed=TRUE)
boxplot_plot_list_celltype <- boxplot_plot_list[name_celltype_plot]

name_clusters_for_celltype <- unique(gsub(".*of_(.*)_from_.*", "\\1", name_celltype_plot))

  for (j in name_clusters_for_celltype) {
 # Save the frequency plot as SVG
    name_freq_plot_clusters <- grep(paste0("freq_of_", j, "_from_"), names(boxplot_plot_list_celltype), value = TRUE, fixed=TRUE)
    freq_svg_filename <- file.path(output_dir, paste0(name_freq_plot_clusters, "_h_ssc.svg"))
    ggsave(filename = freq_svg_filename, plot = boxplot_plot_list_celltype[[name_freq_plot_clusters]], device = "svg", width = 6, height = 5)
    
    name_dens_plot_clusters <- grep(paste0("dens_of_", j, "_from_"), names(boxplot_plot_list_celltype), value = TRUE, fixed=TRUE)
    # Save the density plot as SVG
    dens_svg_filename <- file.path(output_dir, paste0(name_dens_plot_clusters, "_h_ssc.svg"))
    ggsave(filename = dens_svg_filename, plot = boxplot_plot_list_celltype[[name_dens_plot_clusters]], device = "svg", width = fig_dim[1], height = fig_dim[2])
    
# Embed and provide download link for frequency plot
    cat(sprintf("\n\n### %s_frequency\n", j))
    #print(boxplot_plot_freq)
    cat(sprintf('<object type="image/svg+xml" data="%s" width="600"></object>\n', freq_svg_filename))
    cat(sprintf('<a href="%s" download>Download Frequency Plot as SVG</a>\n\n', freq_svg_filename))
    
    # Embed and provide download link for density plot
    cat(sprintf("\n\n### %s_density\n", j))
   #print(boxplot_plot_dens)
    cat(sprintf('<object type="image/svg+xml" data="%s" width="600"></object>\n', dens_svg_filename))
    cat(sprintf('<a href="%s" download>Download Density Plot as SVG</a>\n\n', dens_svg_filename))
  }
}

# Return the lists of plots
# list(
#   box_plots = boxplot_plot_list
# )

```

# Radarchart ggradar - healthy and SSc {.tabset .tabset-pill}

`r fig_dim <- c(10, 8)`

```{r radarchart h_ssc, eval=TRUE, results='asis', fig.dim=fig_dim}

# Define the cell types for which to create the radar charts
celltypes <- c("Stromal", "CD31+", "CD45+CD68+", "CD45+CD3+")
spe_subset <- NULL
radar <- NULL
radar_zscore <- NULL
plot_radar_freq_list <- list()
plot_radar_zscore_freq_list <- list()

for (i in celltypes) {
  # Subset data for the current cell type
  spe_subset <- spe[, spe$new_celllabels == i]
  spe_subset$pheno_clusters_new <- as.character(spe_subset$pheno_clusters_new)
  
  # Obtain cell frequencies
  data <- dittoSeq::dittoFreqPlot(spe_subset, var = "pheno_clusters_new", 
                                  group.by = "Disease", 
                                  sample.by = "Donor",
                                  data.out = TRUE)$data

  # Prepare data for radar chart
  radar <- data %>% group_by(Disease, label) %>%
    summarize(p.label = median(percent, na.rm = TRUE)) %>% 
    pivot_wider(names_from = label, values_from = p.label) %>%
    column_to_rownames("Disease") %>%
    rownames_to_column("group")
  
  # Create radar chart for frequencies
  plot_radar_freq <- ggradar(
    radar, 
    #values.radar = seq(round(min(radar[, -1], na.rm = TRUE), 2),round(max(radar[, -1], na.rm = TRUE), 2),by = 0.1),
    grid.min = min(radar[, -1], na.rm = TRUE), 
    grid.mid = (max(radar[, -1], na.rm = TRUE)+min(radar[, -1], na.rm = TRUE))/2,
    grid.max = max(radar[, -1], na.rm = TRUE),
    label.gridline.min = FALSE,
    label.gridline.mid = FALSE,
    label.gridline.max = FALSE,
    group.line.width = 1, 
    group.point.size = 3,
    group.colours = metadata(spe)$color_vectors$Disease,
    fill = TRUE, fill.alpha = 0.5,
    background.circle.colour = "white",
    gridline.mid.colour = "grey",
    legend.position = "bottom"
  )
  
  # Z-score transformation for individual clusters
  zscores <- data %>% group_by(label) %>% 
    mutate(zscore = (percent - mean(percent, na.rm = TRUE)) / sd(percent, na.rm = TRUE))  
  
  radar_zscore <- zscores %>% group_by(Disease, label) %>%
    summarize(p.label = median(zscore, na.rm = TRUE)) %>% 
    pivot_wider(names_from = label, values_from = p.label) %>%
    column_to_rownames("Disease") %>%
    rownames_to_column("group")
  
  # Create radar chart for z-score frequencies
  plot_radar_zscore_freq <- ggradar(
    radar_zscore, 
    #values.radar = seq(round(min(radar_zscore[, -1], na.rm = TRUE), 2), round(max(radar_zscore[, -1], na.rm = TRUE), 2), by = 0.1),
    grid.min = min(radar_zscore[, -1], na.rm = TRUE), 
    grid.mid = (max(radar_zscore[, -1], na.rm = TRUE)+min(radar_zscore[, -1], na.rm = TRUE))/2,
    grid.max = max(radar_zscore[, -1], na.rm = TRUE),
    label.gridline.min = FALSE,
    label.gridline.mid = FALSE,
    label.gridline.max = FALSE,
    group.line.width = 1, 
    group.point.size = 3,
    group.colours = metadata(spe)$color_vectors$Disease,
    fill = TRUE, fill.alpha = 0.5,
    background.circle.colour = "white",
    gridline.mid.colour = "grey",
    legend.position = "bottom"
  )
  
  # Store plots in respective lists
  plot_radar_freq_list[[i]] <- plot_radar_freq
  plot_radar_zscore_freq_list[[i]] <- plot_radar_zscore_freq
}

for (i in celltypes) {
  cat(sprintf("\n\n## %s {.tabset .tabset-pill}\n", i))
  radar_svg_filename <- file.path(output_dir, paste0(i, "_radar_h_ssc.svg"))
  ggsave(filename = radar_svg_filename, plot = plot_radar_freq_list[[i]], device = "svg", width = fig_dim[1], height = fig_dim[2])
    
    radar_zscore_svg_filename <- file.path(output_dir, paste0(i, "_radar_zsc_h_ssc.svg"))
    ggsave(filename = radar_zscore_svg_filename, plot = plot_radar_zscore_freq_list[[i]], device = "svg", width = fig_dim[1], height = fig_dim[2])
    
# Embed and provide download link for radar plot
    cat(sprintf("\n\n### %s_radar\n", i))
    #print(boxplot_plot_freq)
    cat(sprintf('<object type="image/svg+xml" data="%s" width="600"></object>\n', radar_svg_filename))
    cat(sprintf('<a href="%s" download>Download Radar Plot as SVG</a>\n\n', radar_svg_filename))
    
    # Embed and provide download link for z-scored radar plot
    cat(sprintf("\n\n### %s_radar_zscore\n", i))
   #print(boxplot_plot_dens)
    cat(sprintf('<object type="image/svg+xml" data="%s" width="600"></object>\n', radar_zscore_svg_filename))
    cat(sprintf('<a href="%s" download>Download z-scored Radar Plot as SVG</a>\n\n', radar_zscore_svg_filename))
  }

# Return the lists of plots
# list(
#   freq_plots = plot_radar_freq_list,
#   zscore_plots = plot_radar_zscore_freq_list
# )

```

# Radarchart base r - healthy and SSc {.tabset .tabset-pill}

`r fig_dim <- c(10, 12)`

```{r radarchart2 h_ssc, eval=TRUE, fig.dim=fig_dim, results='asis'}

source(file.path(dir_r, "create_beautiful_radarchart_fx.R"))

# Define the cell types for which to create the radar charts
celltypes <- c("Stromal", "CD31+", "CD45+CD68+", "CD45+CD3+")
spe_subset <- NULL

for (i in celltypes){
  # Subset data for the current cell type
  spe_subset <- spe[,spe$new_celllabels == i]
  spe_subset$pheno_clusters_new <- as.character(spe_subset$pheno_clusters_new)
  
# obtain cell frequencies, optionally can also use table functions without dittoSeq package, e.g.:
# data <- as.data.frame(prop.table(table(spe.fib$Group, spe.fib$cell_clusters2)))
data <- dittoSeq::dittoFreqPlot(spe_subset, var = "pheno_clusters_new", 
                               group.by = "Disease", 
                               sample.by = "Donor",
                               data.out = TRUE)$data

radar <- data %>% group_by(Disease, label) %>%
  summarize(p.label=median(percent)) %>% pivot_wider(names_from = label, values_from = p.label) %>%
  column_to_rownames("Disease")
# reorder the element if needed
#radar <- radar[,order(-as.numeric(radar["NH",]))]


# the max and min values need to be set here.
max <- rep(max(radar)+0.1, ncol(radar)) %>% `names<-`(colnames(radar))
min <- rep(0, ncol(radar))
radar <- rbind(t(data.frame(Max = max, Min = min)), radar)

# Reduce plot margin using par()
opar <- par()
par(mar = rep(0.5,4))

 cat(sprintf("\n\n## %s {.tabset .tabset-pill}\n", i))
 
 # Base R radar chart saving as SVG
  baser_radar_svg_filename <- file.path(output_dir, paste0(i, "_baser_radar_h_ssc.svg"))
  svg(filename = baser_radar_svg_filename, width = fig_dim[1], height = fig_dim[2])

create_beautiful_radarchart(
  data = radar,
  color = metadata(spe)$color_vectors$Disease,
  title=paste0("Frequencies of ", i),
  caxislabels = rep("", 5)
); legend(
  x = "bottom", legend = rownames(radar[-c(1,2),]), horiz = TRUE,
  bty = "n", pch = 20 , col = metadata(spe)$color_vectors$Disease,
  text.col = "black", cex = 1, pt.cex = 1.5, x.intersp = 1
  )

dev.off()
# Restore the standard par() settings
par(opar)

# Embed and provide download link for the base R radar plot
  cat(sprintf("\n\n### %s_baser_radar\n", i))
  cat(sprintf('<object type="image/svg+xml" data="%s" width="600"></object>\n', baser_radar_svg_filename))
  cat(sprintf('<a href="%s" download>Download base r Radar Plot as SVG</a>\n\n', baser_radar_svg_filename))

## Z-score transformation for individual clusters
zscores <- data %>%  group_by(label) %>% mutate(zscore = (percent - mean(percent, na.rm = T)) / sd(percent, na.rm = T))  
radar <- zscores %>% group_by(Disease, label) %>%
  summarize(p.label=median(zscore)) %>% pivot_wider(names_from = label, values_from = p.label) %>%
  column_to_rownames("Disease")
#radar <- radar[,order(-as.numeric(radar["NH",]))]

max <- rep(max(radar)+0.2, ncol(radar)) %>% `names<-`(colnames(radar))
min <- rep(min(radar)-0.2, ncol(radar))
radar <- rbind(t(data.frame(Max = max, Min = min)), radar)

# Reduce plot margin using par()
opar <- par()
par(mar = rep(0.5,4))

# Base R radar chart for Z-scores
  baser_radar_zscore_svg_filename <- file.path(output_dir, paste0(i, "_baser_radar_zsc_h_ssc.svg"))
  svg(filename = baser_radar_zscore_svg_filename, width = fig_dim[1], height = fig_dim[2])

# Create the radar charts
 create_beautiful_radarchart(
  data = radar,
  color = metadata(spe)$color_vectors$Disease,
  title=paste0("Z-score transformed frequencies of ", i),
  caxislabels = rep("", 5)
); legend(
  x = "bottom", legend = rownames(radar[-c(1,2),]), horiz = TRUE,
  bty = "n", pch = 20 , col = metadata(spe)$color_vectors$Disease,
  text.col = "black", cex = 1.5, pt.cex = 1.5, x.intersp = 1
  )
 
dev.off()  # Close the SVG device to save the Z-score plot
# Restore the standard par() settings
par(opar)

  # Embed and provide download link for the Z-score base R radar plot
  cat(sprintf("\n\n### %s_baser_radar_zscore\n", i))
  cat(sprintf('<object type="image/svg+xml" data="%s" width="600"></object>\n', baser_radar_zscore_svg_filename))
  cat(sprintf('<a href="%s" download>Download Z-scored base r Radar Plot as SVG</a>\n\n', baser_radar_zscore_svg_filename))
}
  
```

# Boxplots ggstatsplot - healthy, SSc progressive, SSc stable {.tabset .tabset-pill}

`r fig_dim <- c(10, 15)`

```{r boxplot ggstatsplot h_pr_st, eval=TRUE, fig.dim=fig_dim, results='asis'}

# Define the cell types for which to create the frequency plots
celltypes <- c("Stromal", "CD31+", "CD45+CD68+", "CD45+CD3+")
spe_subset <- NULL
boxplot_list_ggstats <- list()

for (i in celltypes) {
  # Subset data for the current cell type
  spe_subset <- spe[, spe$new_celllabels == i]
  spe_subset$pheno_clusters_new <- as.character(spe_subset$pheno_clusters_new)
  
  # Obtain cell frequencies
  data <- dittoSeq::dittoFreqPlot(spe_subset, var = "pheno_clusters_new", 
                                  group.by = "Progression_skin", 
                                  sample.by = "Donor",
                                  data.out = TRUE)$data
  
# Generate the boxplots for frequencies with ggbetweenstats
 ggstats_boxplot_freq <- grouped_ggbetweenstats(data = data, x = Progression_skin, y = percent,
                              grouping.var=label,
                              ggplot.component = list(ggplot2::scale_color_manual(values=
                              metadata(spe)$color_vectors$progressive_skin)),
                              type = "nonparametric", 
                              p.adjust.method = "none", 
                              k = 4, 
                              pairwise.display = "all",
                              centrality.point.args = list(size = 0), 
                              violin.args = list(width = 0)) + 
      ylab('Frequency')+
      ggtitle(paste0("Frequency of clusters for ", i))
    

# calculate cell density 
    coldata.density <- data.frame(colData(spe_subset)) %>% 
      dplyr::select(Donor, img_area, pheno_clusters_new, Progression_skin) %>% 
      group_by(Donor, Progression_skin, pheno_clusters_new) %>%
      summarise(density = n()/mean(img_area))

    
    # Generate the boxplots for densities with ggbetweenstats
     ggstats_boxplot_dens <- grouped_ggbetweenstats(data = coldata.density, x = Progression_skin, y = density,
                              grouping.var=pheno_clusters_new,
                              ggplot.component = list(ggplot2::scale_color_manual(values=
                                metadata(spe)$color_vectors$progressive_skin)),
                              type = "nonparametric", 
                              p.adjust.method = "none", 
                              k = 4, 
                              pairwise.display = "all",
                              centrality.point.args = list(size = 0), 
                              violin.args = list(width = 0))+ 
      ylab('Density')+
      ggtitle(paste0("Density of clusters for ", i))

# Store plots in respective lists
  plot_name_freq_ggstats <- paste0("freq_of_", i,"_plot")
  plot_name_dens_ggstats <- paste0("dens_of_", i,"_plot")
  boxplot_list_ggstats[[plot_name_freq_ggstats]] <- ggstats_boxplot_freq
  boxplot_list_ggstats[[plot_name_dens_ggstats]] <- ggstats_boxplot_dens
}

for (i in celltypes) {
  cat(sprintf("\n\n## %s {.tabset .tabset-pill}\n", i))
  
  name_ggstplot_freq <- grep(paste0("freq_of_", i, "_plot"), names(boxplot_list_ggstats), value = TRUE, fixed=TRUE)
  ggstplot_freq_svg_filename <- file.path(output_dir, paste0(name_ggstplot_freq, "_ggst_h_st_pr.svg"))
  ggsave(filename = ggstplot_freq_svg_filename, plot = boxplot_list_ggstats[[name_ggstplot_freq]], device = "svg", width = fig_dim[1], height = fig_dim[2])
    
name_ggstplot_dens <- grep(paste0("dens_of_", i, "_plot"), names(boxplot_list_ggstats), value = TRUE, fixed=TRUE)
  ggstplot_dens_svg_filename <- file.path(output_dir, paste0(name_ggstplot_dens, "_ggst_h_st_pr.svg"))
  ggsave(filename = ggstplot_dens_svg_filename, plot = boxplot_list_ggstats[[name_ggstplot_dens]], device = "svg", width = fig_dim[1], height = fig_dim[2])
    
# Embed and provide download link for ggstatplot density plot
    cat(sprintf("\n\n### %s_ggstatsplot_frequency\n", i))
   #print(boxplot_plot_dens)
    cat(sprintf('<object type="image/svg+xml" data="%s" width="600"></object>\n', ggstplot_freq_svg_filename))
    cat(sprintf('<a href="%s" download>Download ggstatsplot for Frequency Radar Plot as SVG</a>\n\n', ggstplot_freq_svg_filename))
    
    # Embed and provide download link for ggstatplot density plot
    cat(sprintf("\n\n### %s_ggstatsplot_density\n", i))
   #print(boxplot_plot_dens)
    cat(sprintf('<object type="image/svg+xml" data="%s" width="600"></object>\n', ggstplot_dens_svg_filename))
    cat(sprintf('<a href="%s" download>Download ggstatsplot for Density as SVG</a>\n\n', ggstplot_dens_svg_filename))
  }

# Return the lists of plots
# list(
#   freq_box_plots_ggstats = boxplot_list_ggstats
# )

```

# Boxplots ggplot - healthy, SSc progressive, SSc stable {.tabset .tabset-pill}

`r fig_dim <- c(10, 8)`

```{r boxplot ggplot h_pr_st, eval=TRUE, fig.dim=fig_dim, results='asis'}

# Define the cell types for which to create the frequency and density plots
celltypes <- c("Stromal", "CD31+", "CD45+CD68+", "CD45+CD3+")
spe_subset <- NULL
boxplot_plot_list <- list()

for (i in celltypes) {
  # Subset data for the current cell type
  spe_subset <- spe[, spe$new_celllabels == i]
  spe_subset$pheno_clusters_new <- as.character(spe_subset$pheno_clusters_new)
  
  # Obtain cell frequencies
  data <- dittoSeq::dittoFreqPlot(spe_subset, var = "pheno_clusters_new", 
                                  group.by = "Progression_skin", 
                                  sample.by = "Donor",
                                  data.out = TRUE)$data

  for (j in unique(spe_subset$pheno_clusters_new)){
    data2 <- data %>% filter(label==j)
    dunn_res <- as.data.frame(data2) %>% rstatix::dunn_test (percent ~ Progression_skin, 
                                                             p.adjust.method="none")%>%
  mutate (method="Dunn")
    
    # Generate the boxplot
    boxplot_plot_freq <- ggplot(data = data2, aes(x = Progression_skin, y = percent)) +
      geom_boxplot(outlier.shape = NA, aes(fill = Progression_skin)) +
      stat_pvalue_manual(dunn_res%>%
                           mutate(y.position=c(max(data2$percent+0.1),
                                               max(data2$percent+0.15),
                                               max(data2$percent+0.1))),
                         bracket.shorten = 0.05) +
      stat_compare_means(method="kruskal", label.y=max(data2$percent +0.25))+
      # stat_compare_means(
      #   comparisons = my_comparisons,
      #   method = "wilcox.test",
      #   aes(label = "p.format"),
      #   hide.ns = FALSE
      # ) +
      common_plot_components +
      scale_fill_manual(values = metadata(spe)$color_vectors$progressive_skin) +
      labs(x = "Cluster", y = "Frequency") +
      ggtitle(paste0("Frequency of ",j, " in healthy and SSc"))
    
    # Calculate the densities
    coldata.density <- data.frame(colData(spe_subset)) %>% 
      dplyr::select(Donor, img_area, pheno_clusters_new, Progression_skin) %>% 
      group_by(Donor, Progression_skin, pheno_clusters_new) %>%
      summarise(density = n()/mean(img_area))
    
    coldata.density2 <- coldata.density %>% filter(pheno_clusters_new==j)
    dunn_res2 <- as.data.frame(coldata.density2) %>% rstatix::dunn_test (density ~ Progression_skin, 
                                                             p.adjust.method="none")%>%
  mutate (method="Dunn")
    
    # Generate the boxplot
    boxplot_plot_dens <- ggplot(data = coldata.density2, aes(x = Progression_skin, y = density)) +
      geom_boxplot(outlier.shape = NA, aes(fill = Progression_skin)) +
      stat_pvalue_manual(dunn_res2%>%
                           mutate(y.position=c(max(coldata.density2$density*1.1),
                                               max(coldata.density2$density*1.18),
                                               max(coldata.density2$density*1.1))),
                         bracket.shorten = 0.05) +
      stat_compare_means(method="kruskal", label.y=max(coldata.density2$density*1.28))+
      # stat_compare_means(
      #   comparisons = my_comparisons,
      #   method = "wilcox.test",
      #   aes(label = "p.format"),
      #   hide.ns = FALSE
      # ) +
      common_plot_components +
      scale_fill_manual(values = metadata(spe)$color_vectors$progressive_skin) +
      labs(x = "Cluster", y = "Density per mm2") +
      ggtitle(paste0("Density of ",j, " in healthy and SSc"))
    
   
  # Store plots in respective lists
          # Store plots in respective lists
  plot_name_freq <- paste0("freq_of_",j,"_from_",i,"_plot")
  plot_name_dens <- paste0("dens_of_",j,"_from_",i,"_plot")
  boxplot_plot_list[[plot_name_freq]] <- boxplot_plot_freq
  boxplot_plot_list[[plot_name_dens]] <- boxplot_plot_dens
  }
}

for (i in celltypes) {
  cat(sprintf("\n\n## %s {.tabset .tabset-pill}\n", i))
  name_celltype_plot <- grep(paste0("_from_", i, "_plot"), names(boxplot_plot_list), value = TRUE, fixed=TRUE)
boxplot_plot_list_celltype <- boxplot_plot_list[name_celltype_plot]

name_clusters_for_celltype <- unique(gsub(".*of_(.*)_from_.*", "\\1", name_celltype_plot))

  for (j in name_clusters_for_celltype) {
 # Save the frequency plot as SVG
    name_freq_plot_clusters <- grep(paste0("freq_of_", j, "_from_"), names(boxplot_plot_list_celltype), value = TRUE, fixed=TRUE)
    freq_svg_filename <- file.path(output_dir, paste0(name_freq_plot_clusters, "_h_sscst_sscpr.svg"))
    ggsave(filename = freq_svg_filename, plot = boxplot_plot_list_celltype[[name_freq_plot_clusters]], device = "svg", width = 6, height = 5)
    
    name_dens_plot_clusters <- grep(paste0("dens_of_", j, "_from_"), names(boxplot_plot_list_celltype), value = TRUE, fixed=TRUE)
    # Save the density plot as SVG
    dens_svg_filename <- file.path(output_dir, paste0(name_dens_plot_clusters, "_h_sscst_sscpr.svg"))
    ggsave(filename = dens_svg_filename, plot = boxplot_plot_list_celltype[[name_dens_plot_clusters]], device = "svg", width = fig_dim[1], height = fig_dim[2])
    
# Embed and provide download link for frequency plot
    cat(sprintf("\n\n### %s_frequency\n", j))
    #print(boxplot_plot_freq)
    cat(sprintf('<object type="image/svg+xml" data="%s" width="600"></object>\n', freq_svg_filename))
    cat(sprintf('<a href="%s" download>Download Frequency Plot as SVG</a>\n\n', freq_svg_filename))
    
    # Embed and provide download link for density plot
    cat(sprintf("\n\n### %s_density\n", j))
   #print(boxplot_plot_dens)
    cat(sprintf('<object type="image/svg+xml" data="%s" width="600"></object>\n', dens_svg_filename))
    cat(sprintf('<a href="%s" download>Download Density Plot as SVG</a>\n\n', dens_svg_filename))
  }
}

# Return the lists of plots
# list(
#   box_plots = boxplot_plot_list
# )
```

# Radarchart ggradar - healthy, SSc progressive, SSc stable {.tabset .tabset-pill}

`r fig_dim <- c(10, 8)`

```{r radarchart h_pr_st, eval=TRUE, fig.dim=fig_dim, results='asis'}

# Define the cell types for which to create the radar charts
celltypes <- c("Stromal", "CD31+", "CD45+CD68+", "CD45+CD3+")
spe_subset <- NULL
radar <- NULL
radar_zscore <- NULL
plot_radar_freq_list <- list()
plot_radar_zscore_freq_list <- list()

for (i in celltypes) {
  # Subset data for the current cell type
  spe_subset <- spe[, spe$new_celllabels == i]
  spe_subset$pheno_clusters_new <- as.character(spe_subset$pheno_clusters_new)
  
  # Obtain cell frequencies
  data <- dittoSeq::dittoFreqPlot(spe_subset, var = "pheno_clusters_new", 
                                  group.by = "Progression_skin", 
                                  sample.by = "Donor",
                                  data.out = TRUE)$data
  
   # Prepare data for radar chart
  radar <- data %>% group_by(Progression_skin, label) %>%
    summarize(p.label = median(percent, na.rm = TRUE)) %>% 
    pivot_wider(names_from = label, values_from = p.label) %>%
    column_to_rownames("Progression_skin") %>%
    rownames_to_column("group")
  
  # Create radar chart for frequencies
  plot_radar_freq <- ggradar(
    radar, 
    #values.radar = seq(round(min(radar[, -1], na.rm = TRUE), 2),round(max(radar[, -1], na.rm = TRUE), 2),by = 0.1),
    grid.min = min(radar[, -1], na.rm = TRUE), 
    grid.mid = (max(radar[, -1], na.rm = TRUE)+min(radar[, -1], na.rm = TRUE))/2,
    grid.max = max(radar[, -1], na.rm = TRUE),
    label.gridline.min = FALSE,
    label.gridline.mid = FALSE,
    label.gridline.max = FALSE,
    group.line.width = 1, 
    group.point.size = 3,
    group.colours = metadata(spe)$color_vectors$progressive_skin,
    fill = TRUE, fill.alpha = 0.5,
    background.circle.colour = "white",
    gridline.mid.colour = "grey",
    legend.position = "bottom"
  )
  
  # Z-score transformation for individual clusters
  zscores <- data %>% group_by(label) %>% 
    mutate(zscore = (percent - mean(percent, na.rm = TRUE)) / sd(percent, na.rm = TRUE))  
  
  radar_zscore <- zscores %>% group_by(Progression_skin, label) %>%
    summarize(p.label = median(zscore, na.rm = TRUE)) %>% 
    pivot_wider(names_from = label, values_from = p.label) %>%
    column_to_rownames("Progression_skin") %>%
    rownames_to_column("group")
  
  # Create radar chart for z-score frequencies
  plot_radar_zscore_freq <- ggradar(
    radar_zscore, 
    #values.radar = seq(round(min(radar_zscore[, -1], na.rm = TRUE), 2), round(max(radar_zscore[, -1], na.rm = TRUE), 2), by = 0.1),
    grid.min = min(radar_zscore[, -1], na.rm = TRUE), 
    grid.mid = (max(radar_zscore[, -1], na.rm = TRUE)+min(radar_zscore[, -1], na.rm = TRUE))/2,
    grid.max = max(radar_zscore[, -1], na.rm = TRUE),
    label.gridline.min = FALSE,
    label.gridline.mid = FALSE,
    label.gridline.max = FALSE,
    group.line.width = 1, 
    group.point.size = 3,
    group.colours = metadata(spe)$color_vectors$progressive_skin,
    fill = TRUE, fill.alpha = 0.5,
    background.circle.colour = "white",
    gridline.mid.colour = "grey",
    legend.position = "bottom"
  )
  
  plot_radar_freq_list[[i]] <- plot_radar_freq
  plot_radar_zscore_freq_list[[i]] <- plot_radar_zscore_freq
}

for (i in celltypes) {
  cat(sprintf("\n\n## %s {.tabset .tabset-pill}\n", i))
  radar_svg_filename <- file.path(output_dir, paste0(i, "_radar_h_sscst_sscpr.svg"))
  ggsave(filename = radar_svg_filename, plot = plot_radar_freq_list[[i]], device = "svg", width = fig_dim[1], height = fig_dim[2])
    
    radar_zscore_svg_filename <- file.path(output_dir, paste0(i, "_radar_zsc_h_sscst_sscpr.svg"))
    ggsave(filename = radar_zscore_svg_filename, plot = plot_radar_zscore_freq_list[[i]], device = "svg", width = fig_dim[1], height = fig_dim[2])
    
# Embed and provide download link for radar plot
    cat(sprintf("\n\n### %s_radar\n", i))
    #print(boxplot_plot_freq)
    cat(sprintf('<object type="image/svg+xml" data="%s" width="600"></object>\n', radar_svg_filename))
    cat(sprintf('<a href="%s" download>Download Radar Plot as SVG</a>\n\n', radar_svg_filename))
    
    # Embed and provide download link for z-scored radar plot
    cat(sprintf("\n\n### %s_radar_zscore\n", i))
   #print(boxplot_plot_dens)
    cat(sprintf('<object type="image/svg+xml" data="%s" width="600"></object>\n', radar_zscore_svg_filename))
    cat(sprintf('<a href="%s" download>Download z-scored Radar Plot as SVG</a>\n\n', radar_zscore_svg_filename))
  }

# Return the lists of plots
# list(
#   freq_plots = plot_radar_freq_list,
#   zscore_plots = plot_radar_zscore_freq_list
# )

```


# Radarchart base r - healthy, SSc progressive, SSc stable {.tabset .tabset-pill}

`r fig_dim <- c(10, 12)`

```{r radarchart2 h_pr_st, eval=TRUE, fig.dim=fig_dim, results='asis'}

# Define the cell types for which to create the radar charts
celltypes <- c("Stromal", "CD31+", "CD45+CD68+", "CD45+CD3+")
radarchart_freq_list <- list()
radarchart_zscaled_feq_list <- list()
spe_subset <- NULL


for (i in celltypes){
  radar <- NULL
  # Subset data for the current cell type
  spe_subset <- spe[,spe$new_celllabels == i]
  spe_subset$pheno_clusters_new <- as.character(spe_subset$pheno_clusters_new)
  
# obtain cell frequencies, optionally can also use table functions without dittoSeq package, e.g.:
# data <- as.data.frame(prop.table(table(spe.fib$Group, spe.fib$cell_clusters2)))
data <- dittoSeq::dittoFreqPlot(spe_subset, var = "pheno_clusters_new", 
                               group.by = "Progression_skin", 
                               sample.by = "Donor",
                               data.out = TRUE)$data

radar <- data %>% group_by(Progression_skin, label) %>%
  summarize(p.label=median(percent)) %>% pivot_wider(names_from = label, values_from = p.label) %>%
  column_to_rownames("Progression_skin")
# reorder the element if needed
#radar <- radar[,order(-as.numeric(radar["NH",]))]


# the max and min values need to be set here.
max <- rep(max(radar)+0.1, ncol(radar)) %>% `names<-`(colnames(radar))
min <- rep(0, ncol(radar))
radar <- rbind(t(data.frame(Max = max, Min = min)), radar)

# Reduce plot margin using par()
opar <- par()
par(mar = rep(0.5,4))

cat(sprintf("\n\n## %s {.tabset .tabset-pill}\n", i))
 
 # Base R radar chart saving as SVG
  baser_radar_svg_filename <- file.path(output_dir, paste0(i, "_baser_radar_h_sscst_sscpr.svg"))
  svg(filename = baser_radar_svg_filename, width = fig_dim[1], height = fig_dim[2])

create_beautiful_radarchart(
  data = radar,
  color = metadata(spe)$color_vectors$progressive_skin,
  title=paste0("Frequencies of ", i),
  caxislabels = rep("", 5)
); legend(
  x = "bottom", legend = rownames(radar[-c(1,2),]), horiz = TRUE,
  bty = "n", pch = 20 , col = metadata(spe)$color_vectors$progressive_skin,
  text.col = "black", cex = 1, pt.cex = 1.5, x.intersp = 1
  )

dev.off()
# Restore the standard par() settings
par(opar)

# Embed and provide download link for the base R radar plot
  cat(sprintf("\n\n### %s_baser_radar\n", i))
  cat(sprintf('<object type="image/svg+xml" data="%s" width="600"></object>\n', baser_radar_svg_filename))
  cat(sprintf('<a href="%s" download>Download base r Radar Plot as SVG</a>\n\n', baser_radar_svg_filename))

## Z-score transformation for individual clusters
zscores <- data %>%  group_by(label) %>% mutate(zscore = (percent - mean(percent, na.rm = T)) / sd(percent, na.rm = T))  
radar <- zscores %>% group_by(Progression_skin, label) %>%
  summarize(p.label=median(zscore)) %>% pivot_wider(names_from = label, values_from = p.label) %>%
  column_to_rownames("Progression_skin")
#radar <- radar[,order(-as.numeric(radar["NH",]))]

max <- rep(max(radar)+0.2, ncol(radar)) %>% `names<-`(colnames(radar))
min <- rep(min(radar)-0.2, ncol(radar))
radar <- rbind(t(data.frame(Max = max, Min = min)), radar)

# Reduce plot margin using par()
opar <- par()
par(mar = rep(0.5,4))

# Base R radar chart for Z-scores
  baser_radar_zscore_svg_filename <- file.path(output_dir, paste0(i, "_baser_radar_zsc_h_sscst_sscpr.svg"))
  svg(filename = baser_radar_zscore_svg_filename, width = fig_dim[1], height = fig_dim[2])

# Create the radar charts
create_beautiful_radarchart(
  data = radar,
  color = metadata(spe)$color_vectors$progressive_skin,
  title=paste0("Z-score transformed frequencies of ", i),
  caxislabels = rep("", 5)
); legend(
  x = "bottom", legend = rownames(radar[-c(1,2),]), horiz = TRUE,
  bty = "n", pch = 20 , col = metadata(spe)$color_vectors$progressive_skin,
  text.col = "black", cex = 1.5, pt.cex = 1.5, x.intersp = 1
  )

dev.off()  # Close the SVG device to save the Z-score plot
# Restore the standard par() settings
par(opar)

  # Embed and provide download link for the Z-score base R radar plot
  cat(sprintf("\n\n### %s_baser_radar_zscore\n", i))
  cat(sprintf('<object type="image/svg+xml" data="%s" width="600"></object>\n', baser_radar_zscore_svg_filename))
  cat(sprintf('<a href="%s" download>Download Z-scored base r Radar Plot as SVG</a>\n\n', baser_radar_zscore_svg_filename))
}

```

# Outputs

This analysis generates the following results: <br>

1. Plots of frequencies of clusters, split between healthy and SSc and between healthy, SSc stable and SSc progressive, for the metabolic skin IMC data, as boxplots, radar plots (with ggradar and base r) and ggstatsplot.
The output is in the format: html file, and a subfolder called `r output_dir`. 
Keep this subfolder in the same folder as the html file! <br>

# Session information

```{r session, eval=TRUE}
etime <- proc.time() - ptm
cat('Rendering this document took :', etime[3], ' s \n')

sessioninfo::session_info()
```
