---
title: "Ridge plot for distance to "
author: "Alex Matei"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  html_document: 
    toc: true
    toc_float: true
    toc_depth: 4
    df_print: paged
    code_folding: hide
---

# Setup
```{r setup, include=FALSE}
source("data_preprocessing/common/config.R")

library(here)
source("data_preprocessing/common/original_setup.R")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, root.dir = here::here())
```

```{r timestamp}
ts <- format(Sys.time(), "%Y%m%d%H%M%S")
ptm <- proc.time()

cat("This document was created with the time stamp",ts)
```

# Load spe

```{r load spe}
load(here("spe_list.RData"))
spe <- spe_list$spe

rm(spe_list)

color_vectors <- list()
color_vectors$clusters <- setNames(c("#C4281C","#C2DAB8","#E8BAC8","#80BBDB","#C470A0","#0D69AC",
                                     "#F5CD30","#287F47","#A05F35","#D9856C","#6B327C","#008F9C","#C7D23C","#B7D7D5","#923978",
                                     "#27462D","#FF9494","#938767","#FFF67B","#E1A4C2","#FC7B1B","#CD6298","#102ADC"), 
                               as.character(unique(spe$pheno_clusters_new)))
metadata(spe)$color_vectors <- color_vectors

spe$pheno_clusters_new2 <- case_when(
  grepl("CD45_CD3",spe$pheno_clusters_new) ~ "CD45",
  TRUE ~ spe$pheno_clusters_new
)

unique(spe$pheno_clusters_new2)
spe$disease <- case_when(spe$indication=="NH" ~ "Healthy", 
                         spe$indication=="SSc" ~ "SSc")
```

# Calculate minimum distance to endothelium (and others)

```{r mindist, echo=TRUE, cache=TRUE}

spe <- minDistToCells(spe, 
                      x_cells = (spe$new_celllabels=="CD31+"), 
                      img_id = "sample_id",
                      name="distToEndothCells")

spe <- minDistToCells(spe, 
                      x_cells = (spe$new_celllabels %in% c("CD45+CD68+","CD45+CD68-CD3-","CD45+CD3+","CD45+CD68+CD3+")), 
                      img_id = "sample_id",
                      name="distToLeukocytes")

spe <- minDistToCells(spe, 
                      x_cells = (spe$new_celllabels == "Stromal"), 
                      img_id = "sample_id",
                      name="distToStromal")

spe <- minDistToCells(spe, 
                      x_cells = (spe$new_celllabels == "E-cad+"), 
                      img_id = "sample_id",
                      name="distToEpid")

spe <- minDistToCells(spe, 
                      x_cells = (spe$pheno_clusters_new=="stromal_2"), 
                      img_id = "sample_id",
                      name="distToStromal2")

spe <- minDistToCells(spe, 
                      x_cells = (spe$pheno_clusters_new %in% c("CD31_3")), 
                      img_id = "sample_id",
                      name="distToCD31_3")

spe <- minDistToCells(spe, 
                      x_cells = (spe$pheno_clusters_new == "CD45_CD68_1"), 
                      img_id = "sample_id",
                      name="distToCD68_1")

spe$pheno_clusters_new2_fib2vsrestfib <- case_when(
  grepl("^stromal", spe$pheno_clusters_new2) ~ "stromal_others",
  TRUE ~ spe$pheno_clusters_new2
)
spe$pheno_clusters_new2_fib2vsrestfib <- case_when(
  grepl("stromal_2", spe$pheno_clusters_new2) ~ "stromal_2",
  TRUE ~ spe$pheno_clusters_new2_fib2vsrestfib
)

spe <- minDistToCells(
  spe,
  x_cells = (spe$pheno_clusters_new2_fib2vsrestfib ==
               "stromal_others"),
  img_id = "sample_id",
  name = "distToStromalothers"
)

spe$pheno_clusters_new2_cd31_3vsrestEC <- case_when(
  grepl("^CD31", spe$pheno_clusters_new2) ~ "EC_others",
  TRUE ~ spe$pheno_clusters_new2
)
spe$pheno_clusters_new2_cd31_3vsrestEC <- case_when(
  grepl("CD31_3", spe$pheno_clusters_new2) ~ "CD31_3",
  TRUE ~ spe$pheno_clusters_new2_cd31_3vsrestEC
)

spe <- minDistToCells(
  spe,
  x_cells = (spe$pheno_clusters_new2_cd31_3vsrestEC ==
               "EC_others"),
  img_id = "sample_id",
  name = "distToECothers"
)

spe$pheno_clusters_new2_CD68_1vsrestMf <- case_when(
  grepl("^stromal", spe$pheno_clusters_new2) ~ "Mf_others",
  TRUE ~ spe$pheno_clusters_new2
)
spe$pheno_clusters_new2_CD68_1vsrestMf <- case_when(
  grepl("CD45_CD68_1", spe$pheno_clusters_new2) ~ "CD45_CD68_1",
  TRUE ~ spe$pheno_clusters_new2_CD68_1vsrestMf
)

spe <- minDistToCells(
  spe,
  x_cells = (spe$pheno_clusters_new2_CD68_1vsrestMf ==
               "Mf_others"),
  img_id = "sample_id",
  name = "distToMfothers"
)

spe$pheno_clusters_new2_fib2vsrest <- case_when(
  !spe$pheno_clusters_new2%in%c("stromal_2","Ecad") ~ "otherdermalcells",
  TRUE ~ spe$pheno_clusters_new2
)

spe$pheno_clusters_new2_cd31_3vsrest <- case_when(
  !spe$pheno_clusters_new2%in%c("CD31_3","Ecad") ~ "otherdermalcells",
  TRUE ~ spe$pheno_clusters_new2
)

spe$pheno_clusters_new2_cd68_1vsrest <- case_when(
  !spe$pheno_clusters_new2%in%c("CD45_CD68_1","Ecad") ~ "otherdermalcells",
  TRUE ~ spe$pheno_clusters_new2
)

spe <- minDistToCells(
  spe,
  x_cells = (spe$pheno_clusters_new2_fib2vsrest ==
               "otherdermalcells"),
  img_id = "sample_id",
  name = "distToOthercells_not_str2"
)

spe <- minDistToCells(
  spe,
  x_cells = (spe$pheno_clusters_new2_cd31_3vsrest ==
               "otherdermalcells"),
  img_id = "sample_id",
  name = "distToOthercells_not_CD31_3"
)

spe <- minDistToCells(
  spe,
  x_cells = (spe$pheno_clusters_new2_cd68_1vsrest ==
               "otherdermalcells"),
  img_id = "sample_id",
  name = "distToOthercells_not_CD68_1"
)

```


# Density plots - EXAMPLE, ADJUST TO SELECT CELL TYPE AND DISTANCE TO CELL TYPE

```{r density plots, echo=TRUE, fig.width=15, fig.height=7}
library(ggsci)
library(ggtext)
library(grid)

spe$Progressive_SSc <- recode(spe$Skin.progression.at.biopsy,
                              "stable" = "Stable SSc",
                              "progressive" = "Progressive SSc",
                              "Healthy" = "Healthy")

spe$pheno_clusters_new3 <- recode(
  spe$pheno_clusters_new2,
  "CD31_1" = "Met<sup>low</sup>_EC",
  "CD31_2" = "TCA/OXPHOS<sup>hi</sup><br>PGC-1α<sup>hi</sup>_EC",
  "CD31_3" = "Met<sup>hi</sup>_EC",
  "CD31_4" = "pmTOR<sup>hi</sup>CD98<sup>hi</sup>_EC",
  "CD31_5" = "Glycolysis<sup>hi</sup><br>TCA/OXPHOS<sup>int</sup>PPP<sup>hi</sup>_EC",
  "CD45_CD68_1" = "Met<sup>hi</sup>_Mf",
  "CD45_CD68_2" = "Met<sup>low</sup>_Mf",
  "CD45_CD68_3" = "Glycolysis<sup>low</sup><br>TCA/OXPHOS<sup>hi</sup>_Mf",
  "CD45_CD68_4" = "Glycolysis<sup>hi</sup><br>TCA/OXPHOS<sup>hi</sup>_Mf",
  "stromal_1" = "TCA/OXPHOS<sup>hi</sup><br>PGC-1α<sup>hi</sup>_Fib",
  "stromal_2" = "Met<sup>hi</sup>_Fib",
  "stromal_3" = "TCA/OXPHOS<sup>int</sup><br>PGC-1α<sup>hi</sup>_Fib",
  "stromal_4" = "pmTOR<sup>hi</sup>CD98<sup>hi</sup>_Fib",
  "stromal_5" = "Met<sup>low</sup>_Fib",
  "stromal_6" = "Glycolysis<sup>int</sup><br>CS<sup>hi</sup>OGDH<sup>low</sup>_Fib",
  "stromal_7" = "Glut<sup>hi</sup>_Fib",
  "stromal_8" = "Glycolysis<sup>int</sup><br>PGC-1α<sup>low</sup>FAO<sup>low</sup>_Fib"
)

ggplot(
  as.data.frame(colData(spe)) %>%
    filter(pheno_clusters_new %in% grep("^CD31", # ADJUST TO SELECT CELL TYPE
                                        unique(spe$pheno_clusters_new2), 
                                        value = TRUE)) %>%
    pivot_longer(
      cols = starts_with("DistToStromal2"), # ADJUST TO SELECT DISTANCE TO CELL TYPE
      names_to = "DistanceType",
      values_to = "Distance"
    )
) +
  geom_density(
    aes(Distance, ..density.., color = disease, fill = disease),
    alpha = 0.05,
    size = 1,
    bounds = c(0, 2000)
  ) +
  xlab("Distance (µm)") +
  ylab("Density") +
  scale_x_continuous(breaks = seq(0, 160, 40)) +
  scale_color_manual(values = rev(pal_npg("nrc")(2))) +
  scale_fill_manual(values = rev(pal_npg("nrc")(2))) +
  # scale_color_npg() + scale_fill_npg() +
  coord_cartesian(expand = FALSE,
                  xlim = c(0, 120),
                  ylim = c(0, 0.04)) +
  facet_wrap(~ pheno_clusters_new3, 
             ncol = 4,
             labeller = label_wrap_gen(width = 40)) +  # Wrap labels to multiple lines
  geom_vline(xintercept = 40,
             linetype = "solid",
             size = 1) +
  theme_bw() + ggtitle("Distance profile of FibMET subsets to Met<sup>hi</sup>_Mf") +
  theme(
    strip.text = element_markdown(
      size = 16,
      face = "bold",
      lineheight = 0.8
    ),
    # Use element_markdown
    plot.title = element_markdown(size = 16, face = "bold.italic"),
    legend.key.size = grid::unit(0.8, "cm"),
    legend.title.align = 0.5,
    panel.spacing = grid::unit(0.5, "cm"),
    plot.margin = margin(10, 10, 10, 10),
    legend.title = element_blank(),
    legend.text = element_text(size = 16),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18)
  )

```


# Heatmap distances - EXAMPLE, ADJUST TO SELECT CELL TYPE AND DISTANCE TO CELL TYPE

```{r heatmap, echo=TRUE, fig.width=3, fig.height=4}

distance_df <-  as.data.frame(colData(spe)) %>% select(pheno_clusters_new2, starts_with("DistTo"), disease)
for (i in grep("^distTo", colnames(distance_df), value = TRUE)) {
  # Dynamically create new column names
  new_col <- paste0("within_", gsub("^distTo", "", i), "_innerniche")
  
  # Add the new column using mutate and dynamic column name
  distance_df <- distance_df %>%
    mutate(!!sym(new_col) := ifelse(.data[[i]] < 40, "Within Niche", "Distant"))
}

distance_df_sub <- distance_df %>% filter(
  pheno_clusters_new2 %in% grep("^str", unique(pheno_clusters_new2), value = TRUE), # ADJUST TO SELECT CELL TYPE
  disease == "SSc" # ADJUST TO SELECT DISEASE GROUP
)
selected_niche <- "within_CD68_1_innerniche" # ADJUST TO SELECT WHAT CELL TYPE IS PRESENT IN THE NICHE
distance_df_sub[[selected_niche]] <- factor(distance_df_sub[[selected_niche]], levels = c("Distant", "Within Niche"))
distance_df_sub_closeby_cells <- distance_df_sub #%>% filter(!!sym(selected_niche) %in% c("inner", "outer"))

# prop_table <- prop.table(
#   table(
#     distance_df_sub_closeby_cells$pheno_clusters_new2,
#     distance_df_sub_closeby_cells[[selected_niche]]
#   ),
#   1
# )

#### Create a prop table and filter the rows with at least n cells ####

# Step 1: Create the contingency table
contingency_table <- table(
  distance_df_sub_closeby_cells$pheno_clusters_new2,
  distance_df_sub_closeby_cells[[selected_niche]]
)

# Step 2: Initialize the proportion table with NA
prop_table <- matrix(
  NA,
  nrow = nrow(contingency_table),
  ncol = ncol(contingency_table),
  dimnames = dimnames(contingency_table)
)

# Step 3: Identify rows with at least 25 cells
valid_rows <- rowSums(contingency_table) >= 25

# Step 4: Calculate proportions for valid rows
prop_table[valid_rows, ] <- prop.table(contingency_table[valid_rows, ], 1)

#### Create the heatmap ####

# Define the breaks for the color scale
scale_max <- 1
breaks <- seq(0,scale_max, length.out = 100)  # Three breaks for three colors)
#breaks <- seq(0.1, scale_max, length.out = 100)  # Three breaks for three colors

# Define the corresponding colors
#colors <- colorRampPalette(c("blue", "lightyellow", "red"))(length(breaks))
colors <- colorRampPalette(rev(brewer.pal(n = 11, name = "RdYlBu")))(100)

# Create the color mapping function
color_function <- colorRamp2(breaks, colors)

heatmap_fib <- Heatmap(prop_table, 
                       name = "matrix", 
                       col = color_function,
                       na_col = "lightgrey",
                       cluster_rows = FALSE, 
                       cluster_columns = FALSE, 
                       rect_gp = gpar(col = "grey", lwd = 0,1),
                       #row_title = "BANKSY clustering", 
                       #column_title = "non-spatial clustering",
                       row_title_gp = gpar(fontsize = 10, fontface = "bold"),
                       column_title_gp = gpar(fontsize = 10, fontface = "bold")) 
draw(heatmap_fib, heatmap_legend_side = "right")


# # Convert the proportion table to a data frame
# prop_df <- as.data.frame(prop_table)
# 
# # Rename columns to something more intuitive
# prop_df_long <- prop_df %>% 
#   rename(
#     cluster = Var1,
#     DistanceCategory = Var2,
#     Proportion = Freq
#   )
# 
# ggplot(prop_df_long, aes(x = cluster, y = Proportion, fill = DistanceCategory)) +
#   geom_col(position = "stack") +
#   theme_bw() +
#   labs(x = "Cluster", y = "Proportion", fill = "Category") +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))

``` 

# Score distance - TEST, IGNORE
```{r scoredistance, echo=TRUE, fig.width=2, fig.height=2, eval=FALSE}

distance_df <-  as.data.frame(colData(spe)) %>% select(pheno_clusters_new2, starts_with("DistTo"),disease)

distance_df$distance_stromal_score <- distance_df$distToOthercells_not_CD31_3/distance_df$distToCD31_3

distance_df_sub <- distance_df %>% filter(pheno_clusters_new2 %in% grep("^str", unique(pheno_clusters_new2), value = TRUE))

ggplot(distance_df_sub, aes(x = pheno_clusters_new2, y = distance_stromal_score)) +
  geom_boxplot(outlier.shape = NA, fill = "lightgrey", color = "black") +
  geom_jitter(aes(color = disease), width = 0.2, alpha = 0.6) +
  scale_color_manual(values = rev(pal_npg("nrc")(2)))+
  theme_bw() +
  theme(legend.position = "right",
        legend.title = element_blank(),
        legend.text = element_text(size=14)) +  
  ggtitle("Distance Score to Stromal 2") #+
  #ylim(c(0, 10))

```

# Outputs

This analysis generates the following results: <br>

1. Density plots of distance of a specific cell type to other cell types, and heatmaps that illustrate the proportion of a particular cell type within a niche of another cell type. <br>

# Session information

```{r session, eval=TRUE}
etime <- proc.time() - ptm
cat('Rendering this document took :', etime[3], ' s \n')

sessioninfo::session_info()
```
