---
title: "Ridge plot for distance to "
author: "Alex Matei"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  html_document: 
    toc: true
    toc_float: true
    toc_depth: 4
    df_print: paged
    code_folding: hide
---

# Setup
```{r setup, include=FALSE}
source("data_preprocessing/common/config.R")

library(here)
source("data_preprocessing/common/original_setup.R")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, root.dir = here::here())
```

```{r timestamp}
ts <- format(Sys.time(), "%Y%m%d%H%M%S")
ptm <- proc.time()

cat("This document was created with the time stamp",ts)
```

# Load spe

```{r load spe}
load(here("spe_list.RData"))
spe <- spe_list$spe

rm(spe_list)

color_vectors <- list()
color_vectors$clusters <- setNames(c("#C4281C","#C2DAB8","#E8BAC8","#80BBDB","#C470A0","#0D69AC",
                                     "#F5CD30","#287F47","#A05F35","#D9856C","#6B327C","#008F9C","#C7D23C","#B7D7D5","#923978",
                                     "#27462D","#FF9494","#938767","#FFF67B","#E1A4C2","#FC7B1B","#CD6298","#102ADC"), 
                               as.character(unique(spe$pheno_clusters_new)))
metadata(spe)$color_vectors <- color_vectors

spe$pheno_clusters_new2 <- case_when(
  grepl("CD45_CD3",spe$pheno_clusters_new) ~ "CD45",
  TRUE ~ spe$pheno_clusters_new
)

unique(spe$pheno_clusters_new2)
```

# Calculate minimum distance to endothelium (and others)

```{r mindist, echo=TRUE, cache=TRUE}

spe <- minDistToCells(spe, 
                      x_cells = (spe$new_celllabels=="CD31+"), 
                      img_id = "sample_id",
                      name="distToEndothCells")

spe <- minDistToCells(spe, 
                      x_cells = (spe$new_celllabels %in% c("CD45+CD68+","CD45+CD68-CD3-","CD45+CD3+","CD45+CD68+CD3+")), 
                      img_id = "sample_id",
                      name="distToLeukocytes")

spe <- minDistToCells(spe, 
                      x_cells = (spe$new_celllabels == "Stromal"), 
                      img_id = "sample_id",
                      name="distToStromal")

spe <- minDistToCells(spe, 
                      x_cells = (spe$new_celllabels == "E-cad+"), 
                      img_id = "sample_id",
                      name="distToEpid")

spe <- minDistToCells(spe, 
                      x_cells = (spe$pheno_clusters_new=="stromal_2"), 
                      img_id = "sample_id",
                      name="distToStromal2")

spe <- minDistToCells(spe, 
                      x_cells = (spe$pheno_clusters_new %in% c("CD31_3")), 
                      img_id = "sample_id",
                      name="distToCD31_3")

spe <- minDistToCells(spe, 
                      x_cells = (spe$pheno_clusters_new == "CD45_CD68_1"), 
                      img_id = "sample_id",
                      name="distToCD68_1")


spe_not_EC <- spe[,which(!spe$new_celllabels == "CD31+")]
spe_stromal <- spe[,which(spe$new_celllabels == "Stromal")]
spe_leukocytes <- spe[,which(spe$new_celllabels %in% c("CD45+CD68+","CD45+CD68-CD3-","CD45+CD3+","CD45+CD68+CD3+"))]
```

# Ridge plot - distance to endothelium

`r fig_dim=c(6,5)`

```{r ridgeplot 1, echo=TRUE, fig.dim=fig_dim}

ggplot(as.data.frame(colData(spe_stromal))) + 
    geom_density_ridges(aes(distToEndothCells, pheno_clusters_new, fill = pheno_clusters_new)) +
    geom_vline(xintercept = 0, color = "dark red", linewidth = 2) +
    scale_x_continuous(limits = c(-10,300))+
    theme_bw()+
    scale_fill_manual(values = metadata(spe)$color_vectors$clusters)+
  ggtitle("Distance to endothelium")

```

# Ridge plot - distance to leukocytes

```{r ridgeplot 2, echo=TRUE, fig.dim=fig_dim}

ggplot(as.data.frame(colData(spe_stromal))) + 
    geom_density_ridges(aes(distToLeukocytes, pheno_clusters_new, fill = pheno_clusters_new)) +
    geom_vline(xintercept = 0, color = "dark red", linewidth = 2) +
    scale_x_continuous(limits = c(-10,300))+
    theme_bw()+
    scale_fill_manual(values = metadata(spe)$color_vectors$clusters)+
  ggtitle("Distance to leukocytes")

```

# Ridge plot - distance to epidermis

```{r ridgeplot 4, echo=TRUE, fig.dim=fig_dim}

ggplot(as.data.frame(colData(spe_stromal))) + 
    geom_density_ridges(aes(distToEpid, pheno_clusters_new, fill = pheno_clusters_new)) +
    geom_vline(xintercept = 0, color = "dark red", linewidth = 2) +
    scale_x_continuous(limits = c(-50,500))+
    theme_bw()+
    scale_fill_manual(values = metadata(spe)$color_vectors$clusters)+
  ggtitle("Distance to keratinocytes")

```

# Ridge plot - distance to stromal cells

```{r ridgeplot 3, echo=TRUE, fig.dim=fig_dim}

ggplot(as.data.frame(colData(spe))) + 
    geom_density_ridges(aes(distToStromal, pheno_clusters_new, fill = pheno_clusters_new)) +
    geom_vline(xintercept = 0, color = "dark red", linewidth = 2) +
    scale_x_continuous(limits = c(-50,150))+
    theme_bw()+
    scale_fill_manual(values = metadata(spe)$color_vectors$clusters)+
  ggtitle("Distance to stromal cells")

```


# Ridge plot - distance to stromal 2

```{r ridgeplot 5, echo=TRUE, fig.dim=fig_dim}

spe_EC <- spe[,which(spe$new_celllabels == "CD31+")]

ggplot(as.data.frame(colData(spe_EC))) + 
    geom_density_ridges(aes(distToStromal2, pheno_clusters_new, fill = pheno_clusters_new)) +
    geom_vline(xintercept = 0, color = "dark red", linewidth = 2) +
    scale_x_continuous(limits = c(-80,700))+
    theme_bw()+
    scale_fill_manual(values = metadata(spe)$color_vectors$clusters)+
  ggtitle("Distance to stromal 2")

spe_Mf <- spe[,which(spe$new_celllabels == "CD45+CD68+")]

ggplot(as.data.frame(colData(spe_Mf))) + 
    geom_density_ridges(aes(distToStromal2, pheno_clusters_new, fill = pheno_clusters_new)) +
    geom_vline(xintercept = 0, color = "dark red", linewidth = 2) +
    scale_x_continuous(limits = c(-80,700))+
    theme_bw()+
    scale_fill_manual(values = metadata(spe)$color_vectors$clusters)+
  ggtitle("Distance to stromal 2")

```

# Ridge plot - distance to CD31 3

```{r ridgeplot 6, echo=TRUE, fig.dim=fig_dim}

ggplot(as.data.frame(colData(spe_stromal))) + 
    geom_density_ridges(aes(distToCD31_3, pheno_clusters_new, fill = pheno_clusters_new)) +
    geom_vline(xintercept = 0, color = "dark red", linewidth = 2) +
    scale_x_continuous(limits = c(-70,1000))+
    theme_bw()+
    scale_fill_manual(values = metadata(spe)$color_vectors$clusters)+
  ggtitle("Distance to CD31 3")

```


# Ridge plot - distance to CD68 1

```{r ridgeplot 8, echo=TRUE, fig.dim=fig_dim}

ggplot(as.data.frame(colData(spe_stromal))) + 
    geom_density_ridges(aes(distToCD68_1, pheno_clusters_new, fill = pheno_clusters_new)) +
    geom_vline(xintercept = 0, color = "dark red", linewidth = 2) +
    scale_x_continuous(limits = c(-70,1500))+
    theme_bw()+
    scale_fill_manual(values = metadata(spe)$color_vectors$clusters)+
  ggtitle("Distance to CD68 1")

```

# Outputs

This analysis generates the following results: <br>

1. Ridge plots of distance of a specific cell type to other cell types. <br>

# Session information

```{r session, eval=TRUE}
etime <- proc.time() - ptm
cat('Rendering this document took :', etime[3], ' s \n')

sessioninfo::session_info()
```
