---
title: "Neighbourhood analysis for CosMx data - healthy and SSc"
author: "Alex Matei"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  html_document: 
    toc: true
    toc_float: true
    toc_depth: 4
    df_print: paged
    code_folding: hide
---

# Setup

```{r setup, include=FALSE}
source("data_preprocessing/common/config.R")

library(here)
source("data_preprocessing/common/original_setup.R")

knitr::opts_chunk$set(echo = TRUE, eval=TRUE, message = FALSE, warning = FALSE, root.dir = here::here())
```

```{r timestamp}
ts <- format(Sys.time(), "%Y%m%d%H%M%S")
ptm <- proc.time()

cat("This document was created with the time stamp",ts)
```

# User-specified variables for loading the spe - **USER INPUTS REQUIRED** 

```{r user_var_load_spe}
# Path to the directory containing the input 'spe' files
data_dir <- here('Datasets/')  # Specify the data directory

# Name of the input 'spe' file to load
input_spe_filename <- "spe20241110033414.rds"  # Specify the 'spe' file name

# Explanations:
# - **data_dir**: The directory where your input 'spe' files are located.
# - **input_spe_filename**: The name of the 'spe' file to load.
```
# Load the spe

```{r load spe}

spe2 <- readRDS(file.path(data_dir, input_spe_filename))

spe2$pheno_clusters_new3 <- case_when(
  spe2$new_celllabels == "Stromal" ~ spe$pheno_clusters_new,
  spe$new_celllabels == "CD31+" ~ "Endothelial cells",
  spe$new_celllabels == "E-cad+" ~ "Keratinocytes",
  spe$new_celllabels == "CD45+CD68-CD3-" ~ "Immune cells",
  spe$new_celllabels == "CD45+CD68+CD3+" ~ "Immune cells",
  spe$new_celllabels == "CD45+CD68+" ~ "Macrophages",
  spe$new_celllabels == "CD45+CD3+" ~ "T cells"
)

for(i in grep(colnames(colData(spe2)), pattern = "^CN|aggNbh", value = TRUE)){
  spe2[[i]] <- NULL
}

spe <- spe2

```

# Other user-specified variables - **USER INPUTS REQUIRED** 

```{r user_var_other}
# Path to the directory where the spe will be saved
output_dir <- here('Datasets/')  # Specify the output directory

# Name of the cell type variable in 'spe' (e.g., "celltype")
cat(" Choose the cell type variable from:\n")
cat(colnames(colData(spe)), sep = "\n")
celltype_variable <- "pheno_clusters_new3"  # Specify the cell type variable in 'spe'

# Define the donor variable from the variables printed above for celltype
donor_variable <- "patient_id"

# Define the donor variable from the variables printed above for celltype
disease_variable <- "Disease"

spe[[disease_variable]] <- as.character(spe[[disease_variable]])

# Define colors for disease that match the number of unique values in the disease variable
cat("\n The number of unique values in the disease variable are:", length(unique(spe[[disease_variable]])))
cat("\n The unique values in the disease variable are:", unique(spe[[disease_variable]]), sep = "\n")
color_vector_disease <- c("#4DBBD533", "#E64B35AA") # Specify colors for disease; the order will match to the unique values in the disease variable


# Variable to subset the spe by, if needed
subset_variable <- "lv1.anno_bank" # Specify the column in 'spe' to subset by
cat(" Choose the cell type variable from:\n")
cat(unique(spe[[subset_variable]]), sep = "\n")
value_to_exclude <- "Undefined"

# Should the spe be subset? Do not subset the spe if creating spatial graphs!
  subset_flag <- FALSE  # Set to TRUE to subset the spe or FALSE to not subset it

# Parameters for building spatial graphs, if not already present in the spe, or if new ones should be added
# A list of spatial graphs to build, each with its parameters
cat("\n The spatial graphs already present are:\n")
cat(colPairNames(spe), sep = "\n")
spatial_graph_params <- list(
  list(type = "delaunay", max_dist = 200, name = "delaunay_200"),
  list(type = "delaunay", max_dist = 300, name = "delaunay_300"),
  list(type = "knn", k = 7, max_dist = 200, name = "knn_7_dist_200")
  # Add more spatial graphs as needed
)

# Should spatial graphs be created?
create_spatial_graphs_flag <- FALSE  # Set to TRUE to create spatial graphs or FALSE to not create them

# Names of the spatial graphs (colPairNames) to use for plotting spatial graphs
colPair_selected <- c("delaunay_80_new")  # Specify colPairs to plot

# Parameters for creating neighbourhoods
neighbourhood_params <- list(
  aggregate_by = "expression",
  # Method for calculating neighbourhoods, choose between "metadata" or "expression" (expression of specified genes or proteins)
  count_by = NULL,
  # If method is "metadata", specify the column in 'spe' to aggregate by (an annotation from the ones listed above for celltype), otherwise specify "NULL"
  assay_type = "z-rescaled",
  # If method is "expression", specify the assay type to use for calculating neighbourhoods, otherwise specify "NULL"
  subset_row = rownames(spe)[rowData(spe)$use_channel_met_markers],
  # If method is "expression", specify the genes or proteins to use for calculating neighbourhoods, otherwise specify "NULL"
  k_values = c(10)          # Values of 'k' for k-means clustering (number of neighbourhoods)
)

# Should spatial graphs be created?
create_neighbourhoods_flag <- TRUE  # Set to TRUE to create spatial graphs or FALSE to not create them

# Any other variables that require manual input can be added here

# Explanations:
# - **output_dir**: The directory where the modified 'spe' files will be saved.
# - **celltype_variable**: The column name in 'spe' that contains cell type annotations.
# - **donor_variable**: The column name in 'spe' that contains donor IDs.
# - **disease_variable**: The column name in 'spe' that contains disease annotations.
# - **color_vector_disease**: A vector of colors for each unique value in the disease variable.
# - **subset_variable**: The column name in 'spe' to subset by.
# - **value_to_exclude**: The value to exclude when subsetting the 'spe'.
# - **subset_flag**: A flag to determine whether to subset the 'spe'.
# - **spatial_graph_params**: A list defining the spatial graphs to build, each with type, parameters, and name.
# - **colPair_selected**: A vector of colPairNames (names of spatial graphs) to use when plotting spatial graphs.
# - **create_spatial_graphs_flag**: A flag to determine whether to create spatial graphs.
# - **neighbourhood_params**: Parameters for creating neighbourhoods, including how to aggregate and cluster.
# - **create_neighbourhoods_flag**: A flag to determine whether to create neighbourhoods.
```



# Prepare spe for analysis

```{r prepare spe for analysis, eval=TRUE, echo=TRUE}
# Subset the spe
 if (create_spatial_graphs_flag) {
  subset_flag <- FALSE
  cat("The spe will not be subset as spatial graphs are being created. Set create_spatial_graphs_flag to FALSE if you want to subset the spe.")
} 
if (subset_flag) {
  spe <- spe2[, spe[[subset_variable]] != value_to_exclude]
} else {
  cat("The spe was not subset.")
  spe <- spe2
}

# Assign the cell type variable
spe$celltype <- spe[[celltype_variable]]

# Set up color vectors for plotting
color_vectors <- list()

disease_color <- setNames(color_vector_disease, unique(spe[[disease_variable]]))

color_vectors$disease <- disease_color
metadata(spe)$color_vectors <- color_vectors
```
# Create new spatial graphs

```{r create spatial graphs}
# Function to build spatial graphs
build_spatial_graphs <- function(spe_object, params_list) {
  for (params in params_list) {
    if (params$type == "delaunay") {
      spe_object <- buildSpatialGraph(
        spe_object,
        img_id = "donor",
        type = params$type,
        max_dist = params$max_dist,
        name = params$name
      )
    } else if (params$type == "knn") {
      spe_object <- buildSpatialGraph(
        spe_object,
        img_id = "donor",
        type = params$type,
        k = params$k,
        max_dist = params$max_dist,
        name = params$name
      )
    }
  }
  return(spe_object)
}

if (create_spatial_graphs_flag) {
  # Build spatial graphs
  spe <- build_spatial_graphs(spe, spatial_graph_params)

# Build spatial graphs
spe <- build_spatial_graphs(spe, spatial_graph_params)
} else {
  cat("Spatial graphs were not created.")
}
```

# Create spatial graph plots

```{r create spatial graph plots, cache=TRUE}

  spatial_plot_list <- list()

# Create color vector for cell types
color_vector_celltype <- setNames(
  distinct_colors(length(unique(spe$celltype)))$name,
  unique(spe$celltype)
)
  
  # Loop over each unique donor
  for (sample_id in unique(spe[[donor_variable]])) {
    # Initialize a list to store plots for each colPairName for the current sample ID
    sample_plots <- list()
    
    # Loop over each colPairName
    for (colPair in colPair_selected) {
      # Generate the spatial plot for the current sample ID and colPairName
      plot_spatial_graph <- plotSpatial(
        spe[, spe[[donor_variable]] == sample_id],
        node_color_by = "celltype",
        img_id = "sample_id",
        draw_edges = TRUE,
        colPairName = colPair,
        nodes_first = FALSE,
        edge_color_fix = "grey"
      ) + 
        ggtitle(paste0(sample_id, " ", colPair)) +
        scale_color_manual(values = color_vector_celltype)
      
      # Store the plot in the nested list
      sample_plots[[colPair]] <- plot_spatial_graph
    }
    # Store the nested list of plots in the main list
    spatial_plot_list[[sample_id]] <- sample_plots
  }

```

# Display the spatial graph plots {.tabset .tabset-pills}

`r fig_dim <- c(10, 10)`

```{r display spatial graph plots, fig.dim=fig_dim, results='asis'}

for (sample_id in names(spatial_plot_list)) {
  cat("## ", sample_id, "{.tabset .tabset-pills}\n\n")
  for (colPair in names(spatial_plot_list[[sample_id]])) {
    cat("### ", colPair, "\n\n")
    print(spatial_plot_list[[sample_id]][[colPair]])
    cat("\n\n")
  }
}
```

# Create neighbourhoods

```{r create neighbourhoods}

# Function to create neighbourhoods
create_neighbourhoods <- function(spe_object, neighbourhood_params) {
  # Aggregate neighbors for each spatial graph
  for (colPair in colPairNames(spe_object)) {
    spe_object <- aggregateNeighbors(
      spe_object,
      colPairName = colPair,
      aggregate_by = neighbourhood_params$aggregate_by,
      count_by = neighbourhood_params$count_by,
      assay_type = neighbourhood_params$assay_type,
      subset_row = neighbourhood_params$subset_row,
      name = paste0("aggNbh_", colPair)
    )
  }
  
  # Perform k-means clustering to create neighbourhoods
  aggNbh_elem <- grep("^aggNbh", colnames(colData(spe_object)), value = TRUE)
  for (aggNbh in aggNbh_elem) {
    for (k in neighbourhood_params$k_values) {
      set.seed(123)
      cn <- kmeans(spe_object[[aggNbh]], centers = k)
      spe_object[[paste0("CN_", k, "_", aggNbh)]] <- as.factor(cn$cluster)
    }
  }
  return(spe_object)
}

if (create_neighbourhoods_flag) {
  # Create neighbourhoods
  spe <- create_neighbourhoods(spe, neighbourhood_params)
} else {
  cat("Neighbourhoods were not created.")
}

```

# User-specified variable for plotting neighbourhoods - **USER INPUTS REQUIRED** 

```{r user_var_nbh, include=FALSE}

# Names of the neighbourhood variables to use for plotting
all_nbh <- grep("^CN_", colnames(colData(spe)), value = TRUE)
cat(" Choose the neighbourhood variable from:\n")
cat(all_nbh, sep = "\n")
nbh_selected <- c("CN_10_aggNbh_delaunay_80_new")  # Specify neighbourhood variables to plot

# Explanations:
# - **nbh_selected**: A vector of neighbourhood variable names to use when plotting neighbourhoods.
```

# Plot neighbourhoods

```{r neighbourhood analysis - plot neighbourhoods, cache=TRUE} 

# Initialize list to store the spatial plots
neighbourhood_plot_list <- list()

# Ensure neighbourhood variables are factors
for (nbh in nbh_selected) {
  spe[[nbh]] <- as.factor(spe[[nbh]])
}

# Loop over each unique donor
for (sample_id in unique(spe[[donor_variable]])) {
  # Initialize a list to store plots for each neighbourhood
  sample_plots <- list()
  for (nbh in nbh_selected) {
    # Create color vector for neighbourhoods
    nbh_levels <- levels(spe[[nbh]])
    color_vector_nbh <- setNames(distinct_colors(length(nbh_levels))$name, nbh_levels)
    
    # Generate the spatial plot for the current sample ID and neighbourhood
    plot_spatial_graph <- plotSpatial(
      spe[, spe[[donor_variable]] == sample_id],
      node_color_by = nbh,
      img_id = "sample_id",
      node_size_fix = 3,
      scales = "free",
      aspect_ratio = NULL
    ) +
      ggtitle(paste0(sample_id, " ", nbh)) +
      scale_color_manual(values = color_vector_nbh)
    
    # Store the plot
    sample_plots[[nbh]] <- plot_spatial_graph
  }
  neighbourhood_plot_list[[sample_id]] <- sample_plots
}

```

# Display the neighbourhood plots {.tabset .tabset-pills}

`r fig_dim <- c(10, 10)`

```{r display neighbourhood plots, fig.dim=fig_dim, results='asis'}
# Display the neighbourhood plots in tabs
for (sample_id in names(neighbourhood_plot_list)) {
  cat("## ", sample_id, "{.tabset .tabset-pills}\n\n")
  for (nbh in names(neighbourhood_plot_list[[sample_id]])) {
    cat("### ", nbh, "\n\n")
    print(neighbourhood_plot_list[[sample_id]][[nbh]])
    cat("\n\n")
  }
}
```

# Heatmaps rescaled for neighbourhood analysis

```{r neighbourhood analysis - heatmap rescaled}
# Initialize the list to store heatmap objects
heatmap_plot_list <- list()

# Loop over each selected neighbourhood
for (nbh in nbh_selected) {
  #message(paste("Processing neighbourhood:", nbh))
  
  # Prepare the data for plotting
  for_plot <- prop.table(
    table(spe[[nbh]], spe$celltype),
    margin = 1
  )
  
  #message("  Prepared for_plot data.")
  
  # Check the structure of for_plot
  #print(head(for_plot))
  
  # Create the heatmap plot and store the object
  heatmap_plot <- tryCatch({
    pheatmap(
      for_plot,
      color = colorRampPalette(c("dark blue", "white", "dark red"))(100),
      scale = "column",
      main = paste0("Heatmap for ", nbh)
    )
  }, error = function(e){
    message(paste("Error generating heatmap for neighbourhood:", nbh))
    message(e$message)
    return(NULL)
  })
  
  # Check if pheatmap succeeded
  if (is.null(heatmap_plot)) {
    next  # Skip to the next iteration if there was an error
  }
  
  #message("  Generated pheatmap.")
  
  # Store the pheatmap object in the list
  heatmap_plot_list[[nbh]] <- heatmap_plot
}

```

# Display the neighbourhood heatmaps rescaled {.tabset .tabset-pills}

`r fig_dim <- c(7, 7)`

```{r display neighbourhood heatmaps rescaled, fig.dim=fig_dim, results='asis'}
# Loop over each stored heatmap object and plot it
for (nbh in names(heatmap_plot_list)) {
  cat("## ", nbh, "\n\n")
  print(heatmap_plot_list[[nbh]])
  cat("\n\n")
}
```

```{r display neighbourhood heatmaps nice, fig.width=4, fig.height=4, results='asis'}
# Loop over each stored heatmap object and plot it
for (nbh in names(heatmap_plot_list)) {
  cat("## ", nbh, "\n\n")
  print(heatmap_plot_list[[nbh]])
  cat("\n\n")
}
```

# Heatmaps counts for neighbourhood analysis

```{r neighbourhood analysis - heatmap counts}
# Initialize the list to store heatmap objects
heatmap_plot_list <- list()

# Loop over each selected neighbourhood
for (nbh in nbh_selected) {
  #message(paste("Processing neighbourhood:", nbh))
  
  # Prepare the data for plotting
  for_plot <- table(spe[[nbh]], spe$celltype)
  
  #message("  Prepared for_plot data.")
  
  # Check the structure of for_plot
  #print(head(for_plot))
  
  # Create the heatmap plot and store the object
  heatmap_plot <- tryCatch({
    pheatmap(
      for_plot,
      color = viridis(100),
      display_numbers = TRUE,
      number_color = "white",
      number_format = "%.0f",
      main = paste0("Heatmap for ", nbh)
    )
  }, error = function(e){
    message(paste("Error generating heatmap for neighbourhood:", nbh))
    message(e$message)
    return(NULL)
  })
  
  # Check if pheatmap succeeded
  if (is.null(heatmap_plot)) {
    next  # Skip to the next iteration if there was an error
  }
  
  #message("  Generated pheatmap.")
  
  # Store the pheatmap object in the list
  heatmap_plot_list[[nbh]] <- heatmap_plot
}

```

# Display the neighbourhood heatmaps counts {.tabset .tabset-pills}

`r fig_dim <- c(7, 7)`

```{r display neighbourhood heatmaps counts, fig.dim=fig_dim, results='asis'}
# Loop over each stored heatmap object and plot it
for (nbh in names(heatmap_plot_list)) {
  cat("## ", nbh, "\n\n")
  print(heatmap_plot_list[[nbh]])
  cat("\n\n")
}
```

# Frequencies of neighbourhoods

```{r neighbourhood analysis - plot frequencies}

boxplot_plot_list <- list()

# Extract colData as a dataframe
coldata_df <- as.data.frame(colData(spe))
coldata_df_to_merge <- coldata_df %>% select(disease_variable, donor_variable)
# Extract unique donor-disease pairs
donor_disease <- unique(coldata_df_to_merge[, c(disease_variable, donor_variable)])

for (nbh in nbh_selected) {
  # Initialize a list to store plots for each neighbourhood
  cluster_plots <- list()
  for_plot2 <- as.data.frame(prop.table(table(spe[[nbh]], spe[[donor_variable]]), margin = 2))
  
  for_plot2 <- for_plot2 %>%
    mutate(Donor = Var2)
  # Merge for_plot2 with donor_disease on the Donor and donor columns
  for_plot2 <- merge(
    for_plot2,
    donor_disease,
    by.x = 'Donor',
    by.y = donor_variable,
    all.x = TRUE
  )
  for_plot2 <- for_plot2 %>% rename_with( ~ "Disease", all_of(disease_variable))
  for_plot2$Disease <- as.character(for_plot2$Disease)
  
  for (cluster in unique(for_plot2$Var1)) {
    for_plot_sub <- for_plot2[for_plot2$Var1 == cluster, ]
    if (length(unique(for_plot2$Disease)) == 2) {
      boxplot_plot <- ggplot(data = for_plot_sub, aes(x = Disease, y = Freq, fill = Disease)) +
        geom_boxplot(outlier.shape = NA) +
        stat_compare_means(
          comparisons = list(unique(for_plot_sub$Disease)),
          method = "wilcox.test",
          hide.ns = TRUE,
          label.y = max(for_plot_sub$Freq) + 0.05
        ) +
        get_common_boxplot_components_fx() +
        scale_fill_manual(values = metadata(spe)$color_vectors$disease) +
        labs(
          x = "Disease",
          y = "Frequency",
          fill = "Disease",
          title = paste0("Frequency of Cluster ", cluster, " for ", nbh)
        )
      
    } else {
      dunn_res <- as.data.frame(for_plot_sub) %>% rstatix::dunn_test (Freq ~ Disease, p.adjust.method =
                                                                        "none") %>% mutate (method = "Dunn")
      # Get unique disease levels from your data
      disease_levels <- unique(for_plot_sub$Disease)
      
      # Convert Disease to factor with specified levels
      for_plot_sub <- for_plot_sub %>%
        mutate(Disease = factor(Disease, levels = disease_levels))
      
      # Convert group1 and group2 to factors with the same levels
      dunn_res <- dunn_res %>%
        mutate(
          group1 = factor(group1, levels = disease_levels),
          group2 = factor(group2, levels = disease_levels)
        )
      
      # Calculate the number of comparisons
      num_comparisons <- nrow(dunn_res)
      
      # Define the base y-position and the increment
      base_y <- max(for_plot_sub$Freq) + 0.1
      increment <- 0.05
      
      # Add y.position to dunn_res
      dunn_res <- dunn_res %>%
        mutate(y.position = base_y + (seq_len(num_comparisons) - 1) * increment)
      
      boxplot_plot <- ggplot(data = for_plot_sub, aes(x = Disease, y = Freq, fill = Disease)) +
        geom_boxplot(outlier.shape = NA) +
        stat_pvalue_manual(
          data = dunn_res,
          inherit.aes = FALSE,
          mapping = aes(
            x = group1,
            xend = group2,
            y.position = y.position,
            label = p.adj
          ),
          bracket.shorten = 0.05
        ) +
        stat_compare_means(method = "kruskal",
                           label.y = base_y + num_comparisons * increment + 0.05) +
        get_common_boxplot_components_fx() +
        #     scale_fill_manual(values = metadata(spe)$color_vectors$disease) +
        labs(
          x = "Disease",
          y = "Frequency",
          fill = "Disease",
          title = paste0("Frequency of Cluster ", cluster, " for ", nbh)
        )
    }
    # Store the plot
    cluster_plots[[cluster]] <- boxplot_plot
  }
    boxplot_plot_list[[nbh]] <- cluster_plots
}

```

# Display the neighbourhood frequencies {.tabset .tabset-pills}

`r fig_dim <- c(5, 4)`

```{r display neighbourhood frequencies, fig.dim=fig_dim, results='asis'}
# Display the frequency plots
for (nbh in names(boxplot_plot_list)) {
  cat("## ", nbh, "{.tabset .tabset-pills}\n\n")
  for (cluster in names(boxplot_plot_list[[nbh]])) {
    cat("### ", cluster, "\n\n")
    print(boxplot_plot_list[[nbh]][[cluster]])
    cat("\n\n")
  }
}

```

# Save the spe if spatial graphs or neighbourhoods were created
```{r save spe}

all_nbh <- grep("^CN_", colnames(colData(spe)), value = TRUE)

if (create_neighbourhoods_flag) {
  if (subset_flag) {
    for (nbh in all_nbh) {
      spe2[[nbh]] <- rep(NA, ncol(spe2))
      match_var <- match(colnames(spe), colnames(spe2))
      spe2[[nbh]][match_var] <- spe[[nbh]]
    }
  } else {
    spe2 <- spe
  }
}

  
  if (create_spatial_graphs_flag | create_neighbourhoods_flag) {
    # Save the modified 'spe' object with a timestamp
    output_spe_filename <- paste0(gsub("20.*", "", input_spe_filename),ts, ".rds")
    
    saveRDS(spe2, file.path(output_dir, output_spe_filename))
    cat("The modified 'spe' object was saved as:", output_spe_filename)
  } else {
    cat("No spatial graphs or neighbourhoods were created. No 'spe' object was saved.")
  }
  
```

# Outputs

This analysis generates the following results: <br>

1. It creates spatial graphs for the spe if required.
2. It creates neighbourhoods for the spe if required.
3. It plots the spatial graphs, colored by selected cell types or neighbourhoods.
4. It creates heatmaps for the neighbourhood analysis.
5. It plots the frequencies of neighbourhoods.
6. It saves the spe with the newly created spatial graphs and neighbourhoods at '`r paste0(output_dir, output_spe_filename, sep="")`'. If neither spatial graphs nor neighbourhoods were created, it does not create an spe object. <br>

# Session information

```{r session}
etime <- proc.time() - ptm
cat('Rendering this document took :', etime[3], ' s \n')

sessioninfo::session_info()
```
