---
title: "Neighbourhood analysis for CosMx data - healthy and SSc"
author: "Alex Matei"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  html_document: 
    toc: true
    toc_float: true
    toc_depth: 4
    df_print: paged
    code_folding: hide
---

# Setup

```{r setup, include=FALSE}
source("data_preprocessing/common/config.R")

library(here)
source("data_preprocessing/common/original_setup.R")

knitr::opts_chunk$set(echo = TRUE, eval=TRUE, message = FALSE, warning = FALSE, root.dir = here::here())
```

```{r timestamp}
ts <- format(Sys.time(), "%Y%m%d%H%M%S")
ptm <- proc.time()

cat("This document was created with the time stamp",ts)
```

# User-specified variables for loading the spe - **USER INPUTS REQUIRED** 

```{r user_var_load_spe}
# Path to the directory containing the input 'spe' files
data_dir <- here('Datasets/')  # Specify the data directory

# Name of the input 'spe' file to load
input_spe_filename <- "spe20250127174306.rds"  # Specify the 'spe' file name

# Explanations:
# - **data_dir**: The directory where your input 'spe' files are located.
# - **input_spe_filename**: The name of the 'spe' file to load.
```
# Load the spe

```{r load spe}

# Load the 'spe' object
spe <- readRDS(file.path(data_dir, input_spe_filename))
```

# Other user-specified variables - **USER INPUTS REQUIRED** 

```{r user_var_other}
# Path to the directory where the spe will be saved
output_dir <- here('Datasets/')  # Specify the output directory

# Name of the cell type variable in 'spe' (e.g., "celltype")
cat(" Choose the cell type variable from:\n")
cat(colnames(colData(spe)), sep = "\n")
celltype_variable <- "pheno_clusters_new2"  # Specify the cell type variable in 'spe'

# Define the donor variable from the variables printed above for celltype
donor_variable <- "patient_id"

# Define the donor variable from the variables printed above for celltype
disease_variable <- "Disease"

spe[[disease_variable]] <- as.character(spe[[disease_variable]])

# Define colors for disease that match the number of unique values in the disease variable
cat("\n The number of unique values in the disease variable are:", length(unique(spe[[disease_variable]])))
cat("\n The unique values in the disease variable are:", unique(spe[[disease_variable]]), sep = "\n")
color_vector_disease <- c("#4DBBD533", "#E64B35AA") # Specify colors for disease; the order will match to the unique values in the disease variable

nbh_selected <- "CN_10_aggNbh_delaunay_80_new"

```


# Heatmaps rescaled for neighbourhood analysis

```{r neighbourhood analysis - heatmap rescaled}
# Initialize the list to store heatmap objects
heatmap_plot_list <- list()
spe2 <- spe
spe <- spe[,spe$Disease=="Healthy"] 
#spe <- spe[,spe$Progression_skin=="progressive"]
spe <- spe[,grepl("^str",spe[[celltype_variable]])]
spe[[nbh_selected]] <- as.numeric(spe[[nbh_selected]])
spe <- spe[,!spe[[nbh_selected]]%in%c(3,1,2)]
spe[[nbh_selected]] <- as.factor(spe[[nbh_selected]])
spe <- spe[,!spe[[celltype_variable]]%in%c("Ecad", "CD45")]

# Loop over each selected neighbourhood
for (nbh in nbh_selected) {
  #message(paste("Processing neighbourhood:", nbh))
  
  # Prepare the data for plotting
  for_plot <- prop.table(
    table(spe[[nbh]], spe$celltype),
    margin = 1
  ) %>% pheatmap:::scale_mat(scale = "column")
  
  #message("  Prepared for_plot data.")
  
  # Check the structure of for_plot
  #print(head(for_plot))
  
  # Create the heatmap plot and store the object
  heatmap_plot <- tryCatch({
    #scaled_data <- scale(for_plot, center = TRUE, scale = TRUE)
    #scaled_data <- pmax(pmin(scaled_data, 3), -3)
    pheatmap(
      for_plot,
      #cluster_cols = FALSE,
      cluster_rows = FALSE,
      color = colorRampPalette(c("dark blue", "white", "dark red"))(100),
      #breaks = seq(-3, 3, length.out = 101),
      scale = "column",
      main = paste0("Heatmap for ", nbh)
    )
  }, error = function(e) {
    message(paste("Error generating heatmap for neighbourhood:", nbh))
    message(e$message)
    return(NULL)
  })
  
  # Check if pheatmap succeeded
  if (is.null(heatmap_plot)) {
    next  # Skip to the next iteration if there was an error
  }
  
  #message("  Generated pheatmap.")
  
  # Store the pheatmap object in the list
  heatmap_plot_list[[nbh]] <- heatmap_plot
}

spe <- spe2
rm(spe2)
```

# Display the neighbourhood heatmaps rescaled

```{r display neighbourhood heatmaps rescaled, fig.width=7, fig.height=7}
# Loop over each stored heatmap object and plot it
for (nbh in names(heatmap_plot_list)) {
  cat("## ", nbh, "\n\n")
  print(heatmap_plot_list[[nbh]])
  cat("\n\n")
}
```

```{r display neighbourhood heatmaps nice, fig.width=4, fig.height=4, results='asis'}
# Loop over each stored heatmap object and plot it
for (nbh in names(heatmap_plot_list)) {
  cat("## ", nbh, "\n\n")
  print(heatmap_plot_list[[nbh]])
  cat("\n\n")
}
```



# Outputs

This analysis generates the following results: <br>
1. It creates heatmaps for the neighbourhood analysis. <br>

# Session information

```{r session}
etime <- proc.time() - ptm
cat('Rendering this document took :', etime[3], ' s \n')

sessioninfo::session_info()
```
