---
title: "Xenium fibroblast ECM scores by IMC metfiblabel"
author: "Codex"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    code_folding: hide
---

```{r setup, include=FALSE}
source("data_preprocessing/common/config.R")
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dev = "png",
  dpi = 300,
  fig.retina = 2,
  dev.args = list(bg = "white")
)
```

```{r libraries}
library(SpatialFeatureExperiment)
library(SummarizedExperiment)
library(Matrix)
library(decoupleR)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(knitr)
```

```{r configuration}
sfe_path <- project_path("Revision/Nan annotation/Data/Xenium sfe objects/Xenium sfe object/sfe_09032026.rds")
spe_path <- project_path("Revision/Nan annotation/Data/IMC spe objects/New IMC/spe2025-12-17 umap.RData")
matrisome_path <- project_path("Revision/Nan annotation/Data/Hs_Matrisome_Masterlist_Naba.csv")
output_dir <- file.path(getwd(), "xenium_fib_ecm_by_metfiblabel_outputs")
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

lv3_preferred_order <- c(
  "papillary_Fib",
  "PI16_Fib",
  "CXCL12_Fib",
  "COL8A1_Fib",
  "CCL19_Fib",
  "COMP_Fib",
  "COCH_Fib",
  "NGFR_Fib"
)

metfib_levels <- c("Other_Fib", "Met_hi_Fib")
metfib_colors <- c("Other_Fib" = "#4C78A8", "Met_hi_Fib" = "#E45756")

score_levels <- c(
  "CoreMatrisome score",
  "Collagens score",
  "ECM Glycoproteins score",
  "Proteoglycans score"
)

score_labels <- c(
  "CoreMatrisome score" = "Core\nmatrisome",
  "Collagens score" = "Collagens",
  "ECM Glycoproteins score" = "ECM\nglycoproteins",
  "Proteoglycans score" = "Proteoglycans"
)

to_dgC <- function(x) {
  if (inherits(x, "dgCMatrix")) return(x)
  x <- as(x, "CsparseMatrix")
  as(x, "dgCMatrix")
}

rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  if (!all(is.finite(rng)) || rng[1] == rng[2]) {
    return(rep(0.5, length(x)))
  }
  (x - rng[1]) / (rng[2] - rng[1])
}

format_p <- function(p_value) {
  dplyr::case_when(
    is.na(p_value) ~ "p = NA",
    p_value < 0.001 ~ "p < 0.001",
    TRUE ~ paste0("p = ", formatC(p_value, format = "f", digits = 3))
  )
}

save_plot_set <- function(plot, basename, width, height) {
  png_file <- file.path(output_dir, paste0(basename, ".png"))
  pdf_file <- file.path(output_dir, paste0(basename, ".pdf"))
  svg_file <- file.path(output_dir, paste0(basename, ".svg"))

  ggplot2::ggsave(png_file, plot, width = width, height = height, dpi = 300, bg = "white")

  grDevices::pdf(pdf_file, width = width, height = height, useDingbats = FALSE, bg = "white")
  print(plot)
  grDevices::dev.off()

  grDevices::svg(svg_file, width = width, height = height, bg = "white")
  print(plot)
  grDevices::dev.off()

  invisible(c(png_file, pdf_file, svg_file))
}
```

# Data And ECM Score Setup

```{r prepare_fib_data}
if (!exists("spe")) {
  load(spe_path)
}

if (!exists("sfe")) {
  sfe <- readRDS(sfe_path)
}

if (!"Rphenograph_500" %in% colnames(SummarizedExperiment::colData(spe))) {
  stop("The IMC spe object must contain Rphenograph_500 for metfiblabel.")
}

if (!"KEY" %in% colnames(SummarizedExperiment::colData(sfe))) {
  stop("The Xenium sfe object must contain KEY to transfer IMC metfiblabel.")
}

spe$metfiblabel <- dplyr::case_when(
  as.character(spe$Rphenograph_500) %in% c("1", "2") ~ "Met_hi_Fib",
  as.character(spe$Rphenograph_500) == "3" ~ "Other_Fib",
  TRUE ~ NA_character_
)

fib_map <- data.frame(
  KEY = colnames(spe),
  metfiblabel = as.character(spe$metfiblabel),
  stringsAsFactors = FALSE
)

fib_idx <- match(SummarizedExperiment::colData(sfe)$KEY, fib_map$KEY)
sfe$metfiblabel <- fib_map$metfiblabel[fib_idx]

if (!"lv3_anno" %in% colnames(SummarizedExperiment::colData(sfe))) {
  if ("lv2_anno_new" %in% colnames(SummarizedExperiment::colData(sfe))) {
    sfe$lv3_anno <- sfe$lv2_anno_new
  } else {
    stop("The Xenium sfe object must contain lv3_anno or lv2_anno_new.")
  }
}

sfe_coldata <- as.data.frame(SummarizedExperiment::colData(sfe))
sample_col <- dplyr::case_when(
  "sample_id" %in% colnames(sfe_coldata) ~ "sample_id",
  "SampleId" %in% colnames(sfe_coldata) ~ "SampleId",
  "SampleID" %in% colnames(sfe_coldata) ~ "SampleID",
  TRUE ~ NA_character_
)

keep_fib <- !is.na(sfe_coldata[["lv3_anno"]]) &
  sfe_coldata[["lv3_anno"]] %in% lv3_preferred_order &
  !is.na(sfe_coldata[["metfiblabel"]]) &
  sfe_coldata[["metfiblabel"]] %in% metfib_levels

sfe_fib <- sfe[, keep_fib]

lv3_levels <- c(
  intersect(lv3_preferred_order, unique(as.character(sfe_fib$lv3_anno))),
  setdiff(sort(unique(as.character(sfe_fib$lv3_anno))), lv3_preferred_order)
)

sfe_fib$lv3_anno <- factor(as.character(sfe_fib$lv3_anno), levels = lv3_levels)
sfe_fib$metfiblabel <- factor(as.character(sfe_fib$metfiblabel), levels = metfib_levels)
sfe_fib$lv3_by_met <- paste(sfe_fib$lv3_anno, sfe_fib$metfiblabel, sep = "_")

lv3_by_met_levels <- as.vector(t(outer(lv3_levels, metfib_levels, paste, sep = "_")))
lv3_by_met_levels <- intersect(lv3_by_met_levels, unique(as.character(sfe_fib$lv3_by_met)))
sfe_fib$lv3_by_met <- factor(as.character(sfe_fib$lv3_by_met), levels = lv3_by_met_levels)

fib_counts <- as.data.frame(table(
  lv3_anno = as.character(sfe_fib$lv3_anno),
  metfiblabel = as.character(sfe_fib$metfiblabel)
)) %>%
  dplyr::filter(Freq > 0) %>%
  dplyr::mutate(
    lv3_anno = factor(lv3_anno, levels = lv3_levels),
    metfiblabel = factor(metfiblabel, levels = metfib_levels)
  ) %>%
  dplyr::arrange(lv3_anno, metfiblabel)

write.csv(fib_counts, file.path(output_dir, "xenium_fib_ecm_by_metfiblabel_counts.csv"), row.names = FALSE)

kable(fib_counts)
```

```{r matrisome_network}
matrisome <- read.csv(matrisome_path, check.names = FALSE, stringsAsFactors = FALSE)

ecm_core <- matrisome %>%
  dplyr::filter(`Matrisome Division` == "Core matrisome") %>%
  dplyr::transmute(
    source = `Matrisome Category`,
    target = `Gene Symbol`
  )

ecm_network <- dplyr::bind_rows(
  ecm_core,
  ecm_core %>% dplyr::transmute(source = "CoreMatrisome", target = target)
) %>%
  dplyr::distinct(source, target)

gene_universe <- rownames(sfe_fib)

gene_availability <- ecm_network %>%
  dplyr::group_by(source) %>%
  dplyr::summarise(
    n_genes = dplyr::n_distinct(target),
    n_detected = sum(unique(target) %in% gene_universe),
    detected_fraction = n_detected / n_genes,
    .groups = "drop"
  ) %>%
  dplyr::mutate(score = paste0(source, " score")) %>%
  dplyr::filter(score %in% score_levels) %>%
  dplyr::mutate(score = factor(score, levels = score_levels)) %>%
  dplyr::arrange(score)

write.csv(gene_availability, file.path(output_dir, "xenium_fib_ecm_gene_availability.csv"), row.names = FALSE)

kable(gene_availability, digits = 3)
```

```{r run_ecm_aucell}
score_cache <- file.path(output_dir, "xenium_fib_ecm_by_metfiblabel_aucell_scores.rds")

if (file.exists(score_cache)) {
  ecm_scores <- readRDS(score_cache)
} else {
  expr_mat <- to_dgC(SummarizedExperiment::assay(sfe_fib, "logcounts"))

  ecm_scores <- decoupleR::run_aucell(
    mat = expr_mat,
    network = ecm_network,
    minsize = 0,
    .source = source,
    .target = target,
    nproc = 1
  )

  saveRDS(ecm_scores, score_cache)
}
```

```{r score_tables}
fib_meta <- as.data.frame(SummarizedExperiment::colData(sfe_fib))
fib_meta$cell <- colnames(sfe_fib)
fib_meta$sample_id <- if (!is.na(sample_col) && sample_col %in% colnames(fib_meta)) {
  as.character(fib_meta[[sample_col]])
} else {
  NA_character_
}

score_long <- ecm_scores %>%
  dplyr::transmute(
    score_name = paste0(source, " score"),
    cell = as.character(condition),
    score = as.numeric(score)
  ) %>%
  dplyr::filter(score_name %in% score_levels) %>%
  dplyr::left_join(
    fib_meta %>%
      dplyr::transmute(
        cell,
        sample_id,
        lv3_anno = as.character(lv3_anno),
        metfiblabel = as.character(metfiblabel),
        lv3_by_met = as.character(lv3_by_met)
      ),
    by = "cell"
  ) %>%
  dplyr::filter(
    !is.na(lv3_anno),
    !is.na(metfiblabel),
    lv3_anno %in% lv3_levels,
    metfiblabel %in% metfib_levels
  ) %>%
  dplyr::mutate(
    score_name = factor(score_name, levels = score_levels),
    lv3_anno = factor(lv3_anno, levels = lv3_levels),
    metfiblabel = factor(metfiblabel, levels = metfib_levels),
    lv3_by_met = factor(lv3_by_met, levels = lv3_by_met_levels)
  )

score_summary <- score_long %>%
  dplyr::group_by(score_name, lv3_anno, metfiblabel, lv3_by_met) %>%
  dplyr::summarise(
    n_cells = dplyr::n(),
    n_samples = dplyr::n_distinct(sample_id[!is.na(sample_id) & sample_id != ""]),
    mean_score = mean(score, na.rm = TRUE),
    median_score = median(score, na.rm = TRUE),
    q25 = stats::quantile(score, 0.25, na.rm = TRUE),
    q75 = stats::quantile(score, 0.75, na.rm = TRUE),
    sd_score = stats::sd(score, na.rm = TRUE),
    se_score = sd_score / sqrt(n_cells),
    ci95 = 1.96 * se_score,
    .groups = "drop"
  )

sample_summary <- score_long %>%
  dplyr::filter(!is.na(sample_id), sample_id != "") %>%
  dplyr::group_by(sample_id, score_name, lv3_anno, metfiblabel, lv3_by_met) %>%
  dplyr::summarise(
    n_cells = dplyr::n(),
    mean_score = mean(score, na.rm = TRUE),
    median_score = median(score, na.rm = TRUE),
    .groups = "drop"
  )

delta_summary <- score_summary %>%
  dplyr::select(score_name, lv3_anno, metfiblabel, mean_score, median_score, n_cells) %>%
  tidyr::pivot_wider(
    names_from = metfiblabel,
    values_from = c(mean_score, median_score, n_cells)
  ) %>%
  dplyr::filter(!is.na(mean_score_Met_hi_Fib), !is.na(mean_score_Other_Fib)) %>%
  dplyr::mutate(
    mean_delta_met_hi_vs_other = mean_score_Met_hi_Fib - mean_score_Other_Fib,
    median_delta_met_hi_vs_other = median_score_Met_hi_Fib - median_score_Other_Fib,
    lv3_anno = factor(as.character(lv3_anno), levels = lv3_levels),
    score_name = factor(as.character(score_name), levels = score_levels)
  )

within_lv3_wilcox <- dplyr::bind_rows(lapply(levels(score_long$score_name), function(cur_score) {
  dplyr::bind_rows(lapply(lv3_levels, function(cur_lv3) {
    cur_df <- score_long %>%
      dplyr::filter(score_name == cur_score, lv3_anno == cur_lv3) %>%
      dplyr::mutate(metfiblabel = factor(as.character(metfiblabel), levels = metfib_levels))

    if (dplyr::n_distinct(cur_df$metfiblabel) < 2) {
      return(data.frame())
    }

    data.frame(
      score_name = cur_score,
      lv3_anno = cur_lv3,
      n_other_fib = sum(cur_df$metfiblabel == "Other_Fib"),
      n_met_hi_fib = sum(cur_df$metfiblabel == "Met_hi_Fib"),
      median_other_fib = median(cur_df$score[cur_df$metfiblabel == "Other_Fib"], na.rm = TRUE),
      median_met_hi_fib = median(cur_df$score[cur_df$metfiblabel == "Met_hi_Fib"], na.rm = TRUE),
      difference_median_met_hi_minus_other =
        median(cur_df$score[cur_df$metfiblabel == "Met_hi_Fib"], na.rm = TRUE) -
        median(cur_df$score[cur_df$metfiblabel == "Other_Fib"], na.rm = TRUE),
      p_value = stats::wilcox.test(score ~ metfiblabel, data = cur_df, exact = FALSE)$p.value,
      stringsAsFactors = FALSE
    )
  }))
})) %>%
  dplyr::group_by(score_name) %>%
  dplyr::mutate(
    p_adjust_bh = stats::p.adjust(p_value, method = "BH"),
    p_label = format_p(p_adjust_bh)
  ) %>%
  dplyr::ungroup()

write.csv(score_long, file.path(output_dir, "xenium_fib_ecm_by_metfiblabel_cell_scores.csv"), row.names = FALSE)
write.csv(score_summary, file.path(output_dir, "xenium_fib_ecm_by_metfiblabel_summary.csv"), row.names = FALSE)
write.csv(sample_summary, file.path(output_dir, "xenium_fib_ecm_by_metfiblabel_sample_summary.csv"), row.names = FALSE)
write.csv(delta_summary, file.path(output_dir, "xenium_fib_ecm_by_metfiblabel_delta.csv"), row.names = FALSE)
write.csv(within_lv3_wilcox, file.path(output_dir, "xenium_fib_ecm_by_metfiblabel_within_lv3_wilcox_tests.csv"), row.names = FALSE)
```

# Summary Tables

```{r summary_table}
kable(score_summary, digits = 4)
```

```{r delta_table}
kable(delta_summary, digits = 4)
```

```{r single_cell_test_table}
kable(within_lv3_wilcox, digits = 4)
```

The Wilcoxon tests are exploratory single-cell comparisons of `Met_hi_Fib`
versus `Other_Fib` inside each Xenium fibroblast annotation and ECM score. They
do not model sample nesting.

# Mean ECM Scores

```{r split_dotplots, fig.width=10.5, fig.height=5.2}
make_split_dotplot <- function(summary_df, scale_by_score = FALSE) {
  plot_df <- summary_df %>%
    dplyr::mutate(
      score_name = factor(as.character(score_name), levels = rev(score_levels)),
      lv3_anno = factor(as.character(lv3_anno), levels = lv3_levels),
      metfiblabel = factor(as.character(metfiblabel), levels = metfib_levels)
    )

  if (isTRUE(scale_by_score)) {
    plot_df <- plot_df %>%
      dplyr::group_by(score_name) %>%
      dplyr::mutate(
        fill_score = rescale01(mean_score),
        size_score = fill_score
      ) %>%
      dplyr::ungroup()
    legend_title <- "Score-scaled mean"
    plot_title <- "Fibroblast ECM scores by Xenium label and IMC metfiblabel, scaled"
  } else {
    plot_df <- plot_df %>%
      dplyr::mutate(
        fill_score = mean_score,
        size_score = rescale01(mean_score)
      )
    legend_title <- "Mean AUCell score"
    plot_title <- "Fibroblast ECM scores by Xenium label and IMC metfiblabel"
  }

  ggplot(plot_df, aes(x = lv3_anno, y = score_name)) +
    geom_point(
      aes(fill = fill_score, size = size_score, color = metfiblabel),
      shape = 21,
      stroke = 0.85,
      position = position_dodge(width = 0.68)
    ) +
    scale_y_discrete(labels = score_labels) +
    scale_fill_gradient(low = "white", high = "#D7301F", name = legend_title) +
    scale_color_manual(values = metfib_colors, name = NULL) +
    scale_size(range = c(3.8, 10.5), guide = "none") +
    labs(title = plot_title, x = NULL, y = NULL) +
    theme_bw(base_size = 14) +
    theme(
      axis.text.x = element_text(size = 11.5, face = "bold", angle = 35, hjust = 1),
      axis.text.y = element_text(size = 12.5, lineheight = 0.95),
      legend.position = "right",
      legend.title = element_text(size = 12),
      legend.text = element_text(size = 10.5),
      panel.grid.major = element_line(color = "grey88"),
      panel.grid.minor = element_blank(),
      plot.title = element_text(size = 15.5, hjust = 0.5, margin = margin(b = 6)),
      plot.margin = margin(5.5, 6, 5.5, 5.5)
    )
}

p_dot_raw <- make_split_dotplot(score_summary, scale_by_score = FALSE)
p_dot_scaled <- make_split_dotplot(score_summary, scale_by_score = TRUE)

save_plot_set(p_dot_raw, "xenium_fib_ecm_by_metfiblabel_dotplot_raw", width = 10.5, height = 5.2)
save_plot_set(p_dot_scaled, "xenium_fib_ecm_by_metfiblabel_dotplot_score_scaled", width = 10.5, height = 5.2)

p_dot_raw
p_dot_scaled
```

# Presentation Dot Matrices {.tabset .tabset-pill}

## Mean Scores - Raw

```{r presentation_two_color_dot_matrix_raw, fig.width=8.8, fig.height=7.4}
make_presentation_mean_dot_matrix <- function(summary_df, scale_by_score = FALSE) {
  plot_df <- summary_df %>%
    dplyr::mutate(
      score_name = factor(as.character(score_name), levels = score_levels),
      lv3_anno = factor(as.character(lv3_anno), levels = rev(lv3_levels)),
      metfiblabel = factor(as.character(metfiblabel), levels = metfib_levels)
    )

  if (isTRUE(scale_by_score)) {
    plot_df <- plot_df %>%
      dplyr::group_by(score_name) %>%
      dplyr::mutate(
        fill_score = rescale01(mean_score),
        size_score = fill_score
      ) %>%
      dplyr::ungroup()
    legend_title <- "Scaled mean\nwithin score"
    plot_title <- "ECM score presentation matrix, scaled"
  } else {
    plot_df <- plot_df %>%
      dplyr::mutate(
        fill_score = mean_score,
        size_score = mean_score
      )
    legend_title <- "Mean\nAUCell score"
    plot_title <- "ECM score presentation matrix"
  }

  ggplot(plot_df, aes(x = metfiblabel, y = lv3_anno)) +
    geom_point(
      aes(fill = fill_score, size = size_score),
      shape = 21,
      color = "grey25",
      stroke = 0.35
    ) +
    facet_wrap(~ score_name, ncol = 2, labeller = labeller(score_name = score_labels)) +
    scale_x_discrete(labels = c("Other_Fib" = "Other", "Met_hi_Fib" = "Met_hi")) +
    scale_fill_gradient(low = "white", high = "#B30000", name = legend_title) +
    scale_size(range = c(5, 13), guide = "none") +
    labs(
      title = plot_title,
      x = NULL,
      y = NULL
    ) +
    theme_bw(base_size = 16) +
    theme(
      panel.grid = element_blank(),
      axis.text.x = element_text(size = 14, face = "bold"),
      axis.text.y = element_text(size = 13.5, face = "bold"),
      strip.background = element_rect(fill = "grey95", color = "grey35"),
      strip.text = element_text(size = 13.5, face = "bold"),
      legend.title = element_text(size = 12.5),
      legend.text = element_text(size = 11.5),
      legend.position = "right",
      plot.title = element_text(size = 17, hjust = 0.5),
      plot.margin = margin(7, 7, 7, 7)
    )
}

p_presentation_two_color_raw <- make_presentation_mean_dot_matrix(
  score_summary,
  scale_by_score = FALSE
)

save_plot_set(
  p_presentation_two_color_raw,
  "xenium_fib_ecm_by_metfiblabel_presentation_two_color_mean_dot_matrix",
  width = 8.8,
  height = 7.4
)

p_presentation_two_color_raw
```

## Mean Scores - Scaled

```{r presentation_two_color_dot_matrix_scaled, fig.width=8.8, fig.height=7.4}
p_presentation_two_color_scaled <- make_presentation_mean_dot_matrix(
  score_summary,
  scale_by_score = TRUE
)

save_plot_set(
  p_presentation_two_color_scaled,
  "xenium_fib_ecm_by_metfiblabel_presentation_two_color_mean_dot_matrix_scaled_by_score",
  width = 8.8,
  height = 7.4
)

p_presentation_two_color_scaled
```

## Delta Matrix - Raw

```{r presentation_divergent_delta_dot_matrix_raw, fig.width=8.8, fig.height=5.8}
make_presentation_delta_dot_matrix <- function(delta_df, scale_by_score = FALSE) {
  plot_df <- delta_df %>%
    dplyr::mutate(
      score_name = factor(as.character(score_name), levels = score_levels),
      lv3_anno = factor(as.character(lv3_anno), levels = rev(lv3_levels))
    )

  if (isTRUE(scale_by_score)) {
    plot_df <- plot_df %>%
      dplyr::group_by(score_name) %>%
      dplyr::mutate(
        max_abs_delta = max(abs(mean_delta_met_hi_vs_other), na.rm = TRUE),
        delta_fill = dplyr::if_else(
          is.finite(max_abs_delta) & max_abs_delta > 0,
          mean_delta_met_hi_vs_other / max_abs_delta,
          0
        ),
        size_score = abs(delta_fill)
      ) %>%
      dplyr::ungroup()
    legend_title <- "Scaled delta\nwithin score"
    plot_title <- "ECM score delta presentation matrix, scaled"
    fill_limits <- c(-1, 1)
  } else {
    plot_df <- plot_df %>%
      dplyr::mutate(
        delta_fill = mean_delta_met_hi_vs_other,
        size_score = abs(mean_delta_met_hi_vs_other)
      )
    legend_title <- "Mean delta\nMet_hi - Other"
    plot_title <- "ECM score delta presentation matrix"
    max_abs <- max(abs(plot_df$delta_fill), na.rm = TRUE)
    fill_limits <- c(-max_abs, max_abs)
  }

  ggplot(plot_df, aes(x = score_name, y = lv3_anno)) +
    geom_point(
      aes(fill = delta_fill, size = size_score),
      shape = 21,
      color = "grey25",
      stroke = 0.35
    ) +
    scale_x_discrete(labels = score_labels) +
    scale_fill_gradient2(
      low = "#3B6FB6",
      mid = "white",
      high = "#B30000",
      midpoint = 0,
      limits = fill_limits,
      name = legend_title
    ) +
    scale_size(range = c(4.5, 13), guide = "none") +
    labs(
      title = plot_title,
      x = NULL,
      y = NULL
    ) +
    theme_bw(base_size = 16) +
    theme(
      panel.grid = element_blank(),
      axis.text.x = element_text(size = 12.5, face = "bold", lineheight = 0.95),
      axis.text.y = element_text(size = 13.5, face = "bold"),
      legend.title = element_text(size = 12.5),
      legend.text = element_text(size = 11.5),
      legend.position = "right",
      plot.title = element_text(size = 17, hjust = 0.5),
      plot.margin = margin(7, 7, 7, 7)
    )
}

p_presentation_divergent_raw <- make_presentation_delta_dot_matrix(
  delta_summary,
  scale_by_score = FALSE
)

save_plot_set(
  p_presentation_divergent_raw,
  "xenium_fib_ecm_by_metfiblabel_presentation_divergent_delta_dot_matrix",
  width = 8.8,
  height = 5.8
)

p_presentation_divergent_raw
```

## Delta Matrix - Scaled

```{r presentation_divergent_delta_dot_matrix_scaled, fig.width=8.8, fig.height=5.8}
p_presentation_divergent_scaled <- make_presentation_delta_dot_matrix(
  delta_summary,
  scale_by_score = TRUE
)

save_plot_set(
  p_presentation_divergent_scaled,
  "xenium_fib_ecm_by_metfiblabel_presentation_divergent_delta_dot_matrix_scaled_by_score",
  width = 8.8,
  height = 5.8
)

p_presentation_divergent_scaled
```

```{r delta_plots, fig.width=9.6, fig.height=5.2}
p_delta_heatmap <- delta_summary %>%
  dplyr::mutate(
    score_name = factor(as.character(score_name), levels = rev(score_levels)),
    lv3_anno = factor(as.character(lv3_anno), levels = lv3_levels)
  ) %>%
  ggplot(aes(x = lv3_anno, y = score_name, fill = mean_delta_met_hi_vs_other)) +
  geom_tile(color = "white", linewidth = 0.7) +
  scale_y_discrete(labels = score_labels) +
  scale_fill_gradient2(
    low = "#3B6FB6",
    mid = "white",
    high = "#D7301F",
    midpoint = 0,
    name = "Mean delta\nMet_hi - Other"
  ) +
  labs(title = "ECM score delta by Xenium fibroblast label", x = NULL, y = NULL) +
  theme_bw(base_size = 14) +
  theme(
    axis.text.x = element_text(size = 11.5, face = "bold", angle = 35, hjust = 1),
    axis.text.y = element_text(size = 12.5, lineheight = 0.95),
    panel.grid = element_blank(),
    plot.title = element_text(size = 15.5, hjust = 0.5)
  )

p_delta_bar <- delta_summary %>%
  dplyr::mutate(
    score_name = factor(as.character(score_name), levels = score_levels),
    lv3_anno = factor(as.character(lv3_anno), levels = lv3_levels)
  ) %>%
  ggplot(aes(x = lv3_anno, y = mean_delta_met_hi_vs_other, fill = mean_delta_met_hi_vs_other > 0)) +
  geom_hline(yintercept = 0, color = "grey35", linewidth = 0.4) +
  geom_col(width = 0.72, color = "grey30", linewidth = 0.25) +
  facet_wrap(~ score_name, ncol = 2, scales = "free_y", labeller = labeller(score_name = score_labels)) +
  scale_fill_manual(values = c("TRUE" = "#D7301F", "FALSE" = "#3B6FB6"), guide = "none") +
  labs(title = "Mean ECM score delta, Met_hi_Fib minus Other_Fib", x = NULL, y = "Mean AUCell delta") +
  theme_bw(base_size = 14) +
  theme(
    axis.text.x = element_text(size = 10.5, face = "bold", angle = 35, hjust = 1),
    strip.text = element_text(size = 12.5, face = "bold"),
    plot.title = element_text(size = 15.5, hjust = 0.5)
  )

save_plot_set(p_delta_heatmap, "xenium_fib_ecm_by_metfiblabel_delta_heatmap", width = 9.6, height = 5.2)
save_plot_set(p_delta_bar, "xenium_fib_ecm_by_metfiblabel_delta_bar", width = 9.6, height = 5.2)

p_delta_heatmap
p_delta_bar
```

# Cell-Level Distributions

```{r distribution_plots, fig.width=12, fig.height=7.2}
p_violin <- score_long %>%
  dplyr::mutate(
    score_name = factor(as.character(score_name), levels = score_levels),
    lv3_anno = factor(as.character(lv3_anno), levels = lv3_levels),
    metfiblabel = factor(as.character(metfiblabel), levels = metfib_levels)
  ) %>%
  ggplot(aes(x = lv3_anno, y = score, fill = metfiblabel)) +
  geom_violin(
    position = position_dodge(width = 0.82),
    width = 0.78,
    alpha = 0.72,
    color = "grey30",
    trim = TRUE
  ) +
  geom_boxplot(
    position = position_dodge(width = 0.82),
    width = 0.14,
    outlier.shape = NA,
    fill = "white",
    color = "grey15",
    linewidth = 0.35
  ) +
  facet_wrap(~ score_name, ncol = 2, scales = "free_y", labeller = labeller(score_name = score_labels)) +
  scale_fill_manual(values = metfib_colors) +
  labs(title = "Single-cell fibroblast ECM scores split by IMC metfiblabel", x = NULL, y = "AUCell score", fill = NULL) +
  theme_bw(base_size = 14) +
  theme(
    axis.text.x = element_text(size = 10.5, face = "bold", angle = 35, hjust = 1),
    strip.text = element_text(size = 12.5, face = "bold"),
    legend.position = "top",
    plot.title = element_text(size = 16, hjust = 0.5)
  )

p_density <- score_long %>%
  dplyr::mutate(score_name = factor(as.character(score_name), levels = score_levels)) %>%
  ggplot(aes(x = score, color = metfiblabel, fill = metfiblabel)) +
  geom_density(alpha = 0.18, linewidth = 0.8, adjust = 1.05) +
  facet_grid(score_name ~ lv3_anno, scales = "free", labeller = labeller(score_name = score_labels)) +
  scale_color_manual(values = metfib_colors) +
  scale_fill_manual(values = metfib_colors) +
  labs(title = "ECM score density within each Xenium fibroblast label", x = "AUCell score", y = "Density", color = NULL, fill = NULL) +
  theme_bw(base_size = 12) +
  theme(
    axis.text.x = element_text(size = 8, angle = 35, hjust = 1),
    axis.text.y = element_text(size = 8),
    strip.text = element_text(size = 9.5, face = "bold"),
    legend.position = "top",
    plot.title = element_text(size = 15.5, hjust = 0.5)
  )

save_plot_set(p_violin, "xenium_fib_ecm_by_metfiblabel_violin_box", width = 12, height = 7.2)
save_plot_set(p_density, "xenium_fib_ecm_by_metfiblabel_density_grid", width = 12, height = 8.4)

p_violin
p_density
```

# Sample Means

```{r sample_mean_plot, fig.width=12, fig.height=7}
if (nrow(sample_summary) > 0) {
  p_sample <- sample_summary %>%
    dplyr::mutate(
      score_name = factor(as.character(score_name), levels = score_levels),
      lv3_anno = factor(as.character(lv3_anno), levels = lv3_levels),
      metfiblabel = factor(as.character(metfiblabel), levels = metfib_levels)
    ) %>%
    ggplot(aes(x = lv3_anno, y = mean_score, color = metfiblabel)) +
    geom_point(
      aes(size = n_cells),
      alpha = 0.82,
      position = position_jitterdodge(jitter.width = 0.08, dodge.width = 0.58)
    ) +
    facet_wrap(~ score_name, ncol = 2, scales = "free_y", labeller = labeller(score_name = score_labels)) +
    scale_color_manual(values = metfib_colors) +
    scale_size_continuous(range = c(2, 6), name = "Cells per sample") +
    labs(title = "Sample-level fibroblast ECM score means", x = NULL, y = "Mean AUCell score", color = NULL) +
    theme_bw(base_size = 14) +
    theme(
      axis.text.x = element_text(size = 10.5, face = "bold", angle = 35, hjust = 1),
      strip.text = element_text(size = 12.5, face = "bold"),
      legend.position = "right",
      plot.title = element_text(size = 16, hjust = 0.5)
    )

  save_plot_set(p_sample, "xenium_fib_ecm_by_metfiblabel_sample_mean_dots", width = 12, height = 7)
  p_sample
} else {
  cat("No sample identifier was available for the fibroblast subset.")
}
```

# Output Files

```{r output_files}
output_files <- data.frame(
  file = sort(list.files(output_dir, full.names = FALSE)),
  stringsAsFactors = FALSE
)
kable(output_files)
```
