---
title: "MintFlow Met_hi_Fib perturbation: EC response"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    df_print: paged
---

```{r setup, message=FALSE, warning=FALSE}
source("data_preprocessing/common/config.R")
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(scales)
library(sf)
library(SpatialFeatureExperiment)
library(SummarizedExperiment)
library(knitr)
library(patchwork)
library(viridis)

base_dir <- project_path("external/perturb_met_hi_fib_ec")
sfe_path <- project_path("Revision/Nan annotation/Data/Xenium sfe objects/Xenium sfe object/sfe_09032026.rds")
obs_path <- project_path("external/obs.csv")

dir.create(file.path(base_dir, "r_plots"), showWarnings = FALSE, recursive = TRUE)

theme_set(theme_bw(base_size = 12))
theme_update(
  panel.grid.minor = element_blank(),
  strip.background = element_rect(fill = "grey92", colour = NA),
  legend.position = "right"
)

scenario_levels <- c("baseline", "remove_Met_hi_Fib", "replace_Met_hi_Fib_with_Other_Fib")
scenario_labels <- c(
  baseline = "Baseline",
  remove_Met_hi_Fib = "Remove Met_hi_Fib",
  replace_Met_hi_Fib_with_Other_Fib = "Replace Met_hi_Fib with Other_Fib"
)
state_levels <- c("Met_hi_EC", "Met_int_EC")
score_cols <- c("glycolysis_score", "tca_score", "hif1a_score", "nox4_score")

cell_df <- read_csv(file.path(base_dir, "ec_cell_level_perturbation_results.csv"), show_col_types = FALSE) %>%
  mutate(
    scenario = factor(scenario, levels = scenario_levels, labels = scenario_labels),
    predicted_ec_state = factor(predicted_ec_state, levels = state_levels),
    metEClabel = factor(metEClabel, levels = state_levels)
  )

obs_df <- read_csv(obs_path, show_col_types = FALSE) %>%
  mutate(
    cell_id = as.character(cell_id),
    combined_met_label = as.character(combined_met_label)
  )

write_plot <- function(plot, filename, width, height) {
  pdf_path <- file.path(base_dir, "r_plots", paste0(filename, ".pdf"))
  png_path <- file.path(base_dir, "r_plots", paste0(filename, ".png"))
  ggsave(pdf_path, plot, width = width, height = height, limitsize = FALSE)
  ggsave(png_path, plot, width = width, height = height, dpi = 250, limitsize = FALSE)
  invisible(list(pdf = pdf_path, png = png_path))
}

fmt_p <- function(p) {
  case_when(
    is.na(p) ~ "NA",
    p < 0.001 ~ "Mann-Whitney p < 0.001",
    TRUE ~ paste0("Mann-Whitney p = ", sprintf("%.3f", p))
  )
}
```

## Inputs and Definitions

This report is generated in R from the MintFlow perturbation outputs in:

```{r input-paths}
tibble(
  input = c("cell-level perturbation CSV", "training-cell metadata CSV", "SFE with cell boundaries"),
  path = c(
    file.path(base_dir, "ec_cell_level_perturbation_results.csv"),
    obs_path,
    sfe_path
  )
) %>% kable()
```

The perturbations are:

- `Remove Met_hi_Fib`: remove all cells labelled `Met_hi_Fib` from the in silico tissue.
- `Replace Met_hi_Fib with Other_Fib`: keep the cells but relabel `Met_hi_Fib` as `Other_Fib`.

EC-state frequencies are computed per sample as the fraction of ECs classified as `Met_hi_EC` or `Met_int_EC` by the baseline-generated-expression EC-state classifier. The statistical tests below use an unpaired Mann-Whitney test, as requested.

## EC-State Frequencies per Sample

```{r frequency-data}
sample_totals <- cell_df %>%
  dplyr::count(scenario, sample_id, name = "n_ec_total")

sample_freq <- cell_df %>%
  dplyr::count(scenario, sample_id, predicted_ec_state, name = "n_ec_state") %>%
  complete(
    nesting(scenario, sample_id),
    predicted_ec_state = factor(state_levels, levels = state_levels),
    fill = list(n_ec_state = 0)
  ) %>%
  left_join(sample_totals, by = c("scenario", "sample_id")) %>%
  mutate(freq = n_ec_state / n_ec_total)

make_pair <- function(target_scenario, comparison_label) {
  sample_freq %>%
    filter(as.character(scenario) %in% c("Baseline", target_scenario)) %>%
    mutate(
      comparison = comparison_label,
      scenario_pair = factor(as.character(scenario), levels = c("Baseline", target_scenario))
    )
}

freq_pair <- bind_rows(
  make_pair("Remove Met_hi_Fib", "Baseline vs remove"),
  make_pair("Replace Met_hi_Fib with Other_Fib", "Baseline vs replace")
)

freq_tests <- freq_pair %>%
  group_by(comparison, predicted_ec_state) %>%
  summarise(
    p_value = tryCatch(wilcox.test(freq ~ scenario_pair, exact = FALSE)$p.value, error = function(e) NA_real_),
    y = min(1.06, max(freq, na.rm = TRUE) + 0.12),
    .groups = "drop"
  ) %>%
  mutate(label = fmt_p(p_value), x = 1.5)

freq_summary <- sample_freq %>%
  group_by(scenario, predicted_ec_state) %>%
  summarise(
    n_samples = n_distinct(sample_id),
    mean_freq = mean(freq),
    median_freq = median(freq),
    min_freq = min(freq),
    max_freq = max(freq),
    .groups = "drop"
  )

freq_summary %>% kable(digits = 3)
```

```{r frequency-plot, fig.width=12, fig.height=6.8}
p_freq <- ggplot(freq_pair, aes(x = scenario_pair, y = freq)) +
  geom_point(
    aes(colour = sample_id),
    position = position_jitter(width = 0.08, height = 0),
    size = 2.5,
    alpha = 0.9
  ) +
  stat_summary(fun = median, geom = "crossbar", width = 0.45, fatten = 0, colour = "black") +
  geom_text(
    data = freq_tests,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    size = 3.2
  ) +
  facet_grid(predicted_ec_state ~ comparison) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 1.1)) +
  labs(
    x = NULL,
    y = "Frequency among ECs",
    colour = "Sample",
    title = "EC-state frequencies per sample after Met_hi_Fib perturbation"
  ) +
  theme(
    axis.text.x = element_text(angle = 25, hjust = 1),
    legend.position = "bottom",
    legend.text = element_text(size = 8)
  )
p_freq
write_plot(p_freq, "r_sample_level_ec_state_frequencies_mann_whitney", 12, 6.8)
```

## Metabolic Scores among ECs

```{r score-data}
sample_scores <- cell_df %>%
  group_by(scenario, sample_id) %>%
  summarise(across(all_of(score_cols), mean, na.rm = TRUE), .groups = "drop") %>%
  pivot_longer(all_of(score_cols), names_to = "score", values_to = "value")

make_score_pair <- function(target_scenario, comparison_label) {
  sample_scores %>%
    filter(as.character(scenario) %in% c("Baseline", target_scenario)) %>%
    mutate(
      comparison = comparison_label,
      scenario_pair = factor(as.character(scenario), levels = c("Baseline", target_scenario))
    )
}

score_pair <- bind_rows(
  make_score_pair("Remove Met_hi_Fib", "Baseline vs remove"),
  make_score_pair("Replace Met_hi_Fib with Other_Fib", "Baseline vs replace")
)

score_tests <- score_pair %>%
  group_by(comparison, score) %>%
  summarise(
    p_value = tryCatch(wilcox.test(value ~ scenario_pair, exact = FALSE)$p.value, error = function(e) NA_real_),
    y = max(value, na.rm = TRUE) + 0.12,
    .groups = "drop"
  ) %>%
  mutate(label = fmt_p(p_value), x = 1.5)

sample_scores %>%
  group_by(scenario, score) %>%
  summarise(
    n_samples = n_distinct(sample_id),
    mean = mean(value),
    median = median(value),
    .groups = "drop"
  ) %>%
  kable(digits = 3)
```

```{r score-plot, fig.width=13, fig.height=8}
p_scores <- ggplot(score_pair, aes(x = scenario_pair, y = value)) +
  geom_point(
    aes(colour = sample_id),
    position = position_jitter(width = 0.08, height = 0),
    size = 2.3,
    alpha = 0.9
  ) +
  stat_summary(fun = median, geom = "crossbar", width = 0.45, fatten = 0, colour = "black") +
  geom_text(
    data = score_tests,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    size = 3
  ) +
  facet_grid(score ~ comparison, scales = "free_y") +
  labs(
    x = NULL,
    y = "Mean log1p generated Xmic score per sample",
    colour = "Sample",
    title = "EC metabolic scores per sample"
  ) +
  theme(
    axis.text.x = element_text(angle = 25, hjust = 1),
    legend.position = "bottom",
    legend.text = element_text(size = 8)
  )
p_scores
write_plot(p_scores, "r_sample_level_ec_metabolic_scores_mann_whitney", 13, 8)
```

## Tissue Maps using Cell Boundaries

```{r load-boundaries, message=FALSE, warning=FALSE}
sfe <- readRDS(sfe_path)
geometry_names <- colGeometryNames(sfe)
use_boundaries <- "cellSeg" %in% geometry_names

if (use_boundaries) {
  boundary_geom <- st_as_sf(colGeometry(sfe, "cellSeg"))
  boundary_geom$cell_id <- rownames(boundary_geom)
  geometry_source <- "cellSeg polygons where available; centroids only for cells without cellSeg polygons"
} else {
  boundary_geom <- st_sf(cell_id = character(), geometry = st_sfc(crs = NA))
  geometry_source <- "centroid points only; no cellSeg polygons found"
}

centroid_geom <- st_as_sf(colGeometry(sfe, "centroids"))
centroid_geom$cell_id <- rownames(centroid_geom)
if (all(is.na(centroid_geom$cell_id)) || any(centroid_geom$cell_id == "")) {
  centroid_geom$cell_id <- colnames(sfe)[seq_len(nrow(centroid_geom))]
}

map_base_boundary <- boundary_geom %>%
  inner_join(obs_df %>% select(cell_id, combined_met_label), by = "cell_id")

missing_boundary_ids <- setdiff(obs_df$cell_id, map_base_boundary$cell_id)
map_base_centroid <- centroid_geom %>%
  semi_join(tibble(cell_id = missing_boundary_ids), by = "cell_id") %>%
  inner_join(obs_df %>% select(cell_id, combined_met_label), by = "cell_id")

make_map_labels <- function(map_base) {
  bind_rows(
    map_base %>% mutate(scenario = "Baseline", map_label = combined_met_label),
    map_base %>%
      filter(combined_met_label != "Met_hi_Fib") %>%
      mutate(scenario = "Remove Met_hi_Fib", map_label = combined_met_label),
    map_base %>%
      mutate(
        scenario = "Replace Met_hi_Fib with Other_Fib",
        map_label = if_else(combined_met_label == "Met_hi_Fib", "Other_Fib", combined_met_label)
      )
  ) %>%
    mutate(
      scenario = factor(scenario, levels = unname(scenario_labels)),
      map_label = factor(map_label, levels = c("Met_hi_Fib", "Other_Fib", "Met_hi_EC", "Met_int_EC"))
    )
}

map_labels_boundary <- make_map_labels(map_base_boundary)
map_labels_centroid <- make_map_labels(map_base_centroid)

ec_meta <- cell_df %>%
  mutate(scenario = as.character(scenario)) %>%
  select(cell_id, scenario, metEClabel, predicted_ec_state, prob_Met_hi_EC)

ec_prob_sf_boundary <- boundary_geom %>%
  inner_join(ec_meta, by = "cell_id") %>%
  mutate(
    scenario = factor(scenario, levels = unname(scenario_labels)),
    metEClabel = factor(metEClabel, levels = state_levels)
  )

ec_missing_boundary_ids <- setdiff(ec_meta$cell_id, ec_prob_sf_boundary$cell_id)
ec_prob_sf_centroid <- centroid_geom %>%
  semi_join(tibble(cell_id = ec_missing_boundary_ids), by = "cell_id") %>%
  inner_join(ec_meta, by = "cell_id") %>%
  mutate(
    scenario = factor(scenario, levels = unname(scenario_labels)),
    metEClabel = factor(metEClabel, levels = state_levels)
  )

tibble(
  geometry_used = geometry_source,
  training_cells_with_boundaries = n_distinct(map_base_boundary$cell_id),
  training_cells_with_centroid_fallback = n_distinct(map_base_centroid$cell_id),
  ec_cells_with_boundaries = n_distinct(ec_prob_sf_boundary$cell_id),
  ec_cells_with_centroid_fallback = n_distinct(ec_prob_sf_centroid$cell_id)
) %>%
  kable()
```

```{r celltype-map, fig.width=16, fig.height=6}
label_cols <- c(
  Met_hi_Fib = "#c0392b",
  Other_Fib = "#7f8c8d",
  Met_hi_EC = "#2166ac",
  Met_int_EC = "#1b9e77"
)

p_map_labels <- ggplot() +
  geom_sf(data = map_labels_boundary, aes(fill = map_label), colour = NA, alpha = 0.9) +
  geom_sf(data = map_labels_centroid, aes(colour = map_label), size = 0.22, alpha = 0.7) +
  facet_wrap(~scenario, nrow = 1) +
  scale_fill_manual(values = label_cols, na.value = "grey85", drop = FALSE) +
  scale_colour_manual(values = label_cols, guide = "none", na.value = "grey70", drop = FALSE) +
  coord_sf(datum = NA) +
  scale_y_reverse() +
  labs(
    title = "In silico tissue layouts after perturbation",
    subtitle = geometry_source,
    fill = "Cell label"
  ) +
  theme_void(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "bottom",
    plot.title = element_text(face = "bold")
  )
p_map_labels
write_plot(p_map_labels, "r_tissue_map_cell_boundaries_celltype", 16, 6)
```

```{r ec-probability-map, fig.width=16, fig.height=6}
p_map_prob <- ggplot() +
  geom_sf(data = map_base_boundary, fill = "grey92", colour = NA, alpha = 0.35) +
  geom_sf(data = map_base_centroid, colour = "grey78", size = 0.12, alpha = 0.3) +
  geom_sf(data = ec_prob_sf_boundary, aes(fill = prob_Met_hi_EC), colour = NA, alpha = 0.95) +
  geom_sf(data = ec_prob_sf_centroid, aes(colour = prob_Met_hi_EC), size = 0.55, alpha = 0.95) +
  facet_wrap(~scenario, nrow = 1) +
  scale_fill_viridis_c(option = "magma", limits = c(0, 1), name = "P(Met_hi_EC)") +
  scale_colour_viridis_c(option = "magma", limits = c(0, 1), guide = "none") +
  coord_sf(datum = NA) +
  scale_y_reverse() +
  labs(
    title = "Generated EC Met_hi_EC probability after perturbation",
    subtitle = geometry_source
  ) +
  theme_void(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "bottom",
    plot.title = element_text(face = "bold")
  )
p_map_prob
write_plot(p_map_prob, "r_tissue_map_cell_boundaries_ec_probability", 16, 6)
```

```{r ec-probability-map-by-population, fig.width=16, fig.height=8}
p_map_prob_by_pop <- ggplot() +
  geom_sf(data = map_base_boundary, fill = "grey94", colour = NA, alpha = 0.25) +
  geom_sf(data = map_base_centroid, colour = "grey82", size = 0.1, alpha = 0.25) +
  geom_sf(data = ec_prob_sf_boundary, aes(fill = prob_Met_hi_EC), colour = NA, alpha = 0.95) +
  geom_sf(data = ec_prob_sf_centroid, aes(colour = prob_Met_hi_EC), size = 0.55, alpha = 0.95) +
  facet_grid(metEClabel ~ scenario) +
  scale_fill_viridis_c(option = "magma", limits = c(0, 1), name = "P(Met_hi_EC)") +
  scale_colour_viridis_c(option = "magma", limits = c(0, 1), guide = "none") +
  coord_sf(datum = NA) +
  scale_y_reverse() +
  labs(
    title = "Generated EC Met_hi_EC probability by original EC population",
    subtitle = geometry_source
  ) +
  theme_void(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "bottom",
    plot.title = element_text(face = "bold")
  )
p_map_prob_by_pop
write_plot(p_map_prob_by_pop, "r_tissue_map_cell_boundaries_ec_probability_by_population", 16, 8)
```

## Exported Files

```{r exported-files}
tibble(
  file = list.files(file.path(base_dir, "r_plots"), pattern = "\\.(pdf|png)$", full.names = TRUE)
) %>%
  mutate(file = basename(file)) %>%
  arrange(file) %>%
  kable()
```
