---
title: "Clincal correlations as dotplots"
author: "Alex Matei"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  html_document: 
    toc: true
    toc_float: true
    toc_depth: 4
    df_print: paged
    code_folding: hide
---

# Setup

```{r setup, include=FALSE}
source("data_preprocessing/common/config.R")

library(here)
source("data_preprocessing/common/original_setup.R")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, eval = TRUE, root.dir = here::here())

```

```{r timestamp, eval=TRUE}
ts <- format(Sys.time(), "%Y%m%d%H%M%S")
ptm <- proc.time()

cat("This document was created with the time stamp",ts)
```

Import the datasets created with the RMarkdown `Clinicaldata_tables.Rmd` (export firstly the Datatable outputs as .csv files and save them in the format "cor\_... .csv" in `r here("Datasets")`).

# Load datasets

```{r load datasets, eval=TRUE}
file_paths <- list.files(here("Datasets"), pattern = "cor_", full.names = TRUE)
# Example: read in your CSV (modify path and column names as appropriate)
list_df <- setNames(lapply(file_paths, read.csv), 
                    gsub(".*cor_(.*?)_hasneigh\\.csv", "\\1", file_paths))
```

# Prepare datasets for analysis

```{r prep datasets, eval=TRUE}

list_df <- lapply(list_df, function(x) {
  x$label2 <- recode(
  x$label,
  "with_CD31_1_neigh" = "Met<sup>low</sup>_EC",
  "with_CD31_2_neigh" = "TCA/OXPHOS<sup>hi</sup>PGC-1α<sup>hi</sup>_EC",
  "with_CD31_3_neigh" = "Met<sup>hi</sup>_EC",
  "with_CD31_4_neigh" = "pmTOR<sup>hi</sup>CD98<sup>hi</sup>_EC",
  "with_CD31_5_neigh" = "Glycolysis<sup>hi</sup>TCA/OXPHOS<sup>int</sup>PPP<sup>hi</sup>_EC",
  "with_CD45_CD68_1_neigh" = "Met<sup>hi</sup>_Mf",
  "with_CD45_CD68_2_neigh" = "Met<sup>low</sup>_Mf",
  "with_CD45_CD68_3_neigh" = "Glycolysis<sup>low</sup>TCA/OXPHOS<sup>hi</sup>_Mf",
  "with_CD45_CD68_4_neigh" = "Glycolysis<sup>hi</sup>TCA/OXPHOS<sup>hi</sup>_Mf",
  "with_stromal_1_neigh" = "TCA/OXPHOS<sup>hi</sup>PGC-1α<sup>hi</sup>_Fib",
  "with_stromal_2_neigh" = "Met<sup>hi</sup>_Fib",
  "with_stromal_3_neigh" = "TCA/OXPHOS<sup>int</sup>PGC-1α<sup>hi</sup>_Fib",
  "with_stromal_4_neigh" = "pmTOR<sup>hi</sup>CD98<sup>hi</sup>_Fib",
  "with_stromal_5_neigh" = "Met<sup>low</sup>_Fib",
  "with_stromal_6_neigh" = "Glycolysis<sup>int</sup>CS<sup>hi</sup>OGDH<sup>low</sup>_Fib",
  "with_stromal_7_neigh" = "Glut<sup>hi</sup>_Fib",
  "with_stromal_8_neigh" = "Glycolysis<sup>int</sup>PGC-1α<sup>low</sup>FAO<sup>low</sup>_Fib"
)
  return(x)
})

list_df <- lapply(list_df, function(x) {
  x$cluster_var2 <- recode(
  x$cluster_var,
  "CD31_1" = "Met<sup>low</sup>_EC",
  "CD31_2" = "TCA/OXPHOS<sup>hi</sup>PGC-1α<sup>hi</sup>_EC",
  "CD31_3" = "Met<sup>hi</sup>_EC",
  "CD31_4" = "pmTOR<sup>hi</sup>CD98<sup>hi</sup>_EC",
  "CD31_5" = "Glycolysis<sup>hi</sup>TCA/OXPHOS<sup>int</sup>PPP<sup>hi</sup>_EC",
  "CD45_CD68_1" = "Met<sup>hi</sup>_Mf",
  "CD45_CD68_2" = "Met<sup>low</sup>_Mf",
  "CD45_CD68_3" = "Glycolysis<sup>low</sup>TCA/OXPHOS<sup>hi</sup>_Mf",
  "CD45_CD68_4" = "Glycolysis<sup>hi</sup>TCA/OXPHOS<sup>hi</sup>_Mf",
  "stromal_1" = "TCA/OXPHOS<sup>hi</sup>PGC-1α<sup>hi</sup>_Fib",
  "stromal_2" = "Met<sup>hi</sup>_Fib",
  "stromal_3" = "TCA/OXPHOS<sup>int</sup>PGC-1α<sup>hi</sup>_Fib",
  "stromal_4" = "pmTOR<sup>hi</sup>CD98<sup>hi</sup>_Fib",
  "stromal_5" = "Met<sup>low</sup>_Fib",
  "stromal_6" = "Glycolysis<sup>int</sup>CS<sup>hi</sup>OGDH<sup>low</sup>_Fib",
  "stromal_7" = "Glut<sup>hi</sup>_Fib",
  "stromal_8" = "Glycolysis<sup>int</sup>PGC-1α<sup>low</sup>FAO<sup>low</sup>_Fib"
)
  return(x)
})

list_df  <- lapply(list_df, function(x) {
  x %>% mutate(celltype_with_nbh = paste(cluster_var,label, sep = "_"))
})

celltype_neigh_plot <- c("stromal_2_with_stromal_4_neigh",
                         "stromal_2_with_stromal_5_neigh",
                         "stromal_2_with_CD31_4_neigh",
                         "stromal_2_with_CD31_5_neigh",
                         "stromal_4_with_stromal_2_neigh",
                         "stromal_8_with_stromal_2_neigh",
                         "stromal_8_with_CD45_CD68_1_neigh")

df_comb <- bind_rows(list_df) 

df_comb$clinical_var <- case_when(
  df_comb$clinical_var == "Age_at_the_time_of_biopsy_years" ~ "Age",
  df_comb$clinical_var == "CRP" ~ "CRP",
  df_comb$clinical_var == "non_RP_disease_duration" ~ "Disease duration",
  df_comb$clinical_var == "EUSTAR_score_2017" ~ "Disease activity",
  df_comb$clinical_var == "current_mRSS" ~ "mRss")

df_comb2 <- df_comb %>% filter(celltype_with_nbh %in% celltype_neigh_plot)

```

# Plot overall differences

```{r plot overall, eval=TRUE, fig.width=9, fig.height=5}
df_comb <- df_comb %>% 
  mutate(celltype_with_nbh = paste(cluster_var2, label2, sep = " &rarr; ")) %>% 
  filter(!grepl("Ecad_neigh$|CD45_neigh$|CD45_CD3_[1-4]_neigh$", celltype_with_nbh))

# Filter for significant differences
df_filtered <- df_comb %>%
  filter(p.value < 0.05 & abs(estimate) > 0.3)

df_filtered <- df_filtered %>% filter (clinical_var=="mRss")

# Order the factor levels for plotting
df_filtered$celltype_with_nbh <- factor(
  df_filtered$celltype_with_nbh,
  levels = df_filtered$celltype_with_nbh[order(df_filtered$estimate)]
)

p <- ggplot(df_filtered, aes(x = estimate, y = celltype_with_nbh)) +
  # Draw a segment from correlation=0 to the correlation value, just for a "lollipop" style
  geom_segment(aes(x = 0, xend = estimate, y = celltype_with_nbh, yend = celltype_with_nbh),
               color = "grey80") +
  # Plot the points; color by p-value and size by absolute correlation
  geom_point(aes(color = p.value, size = abs(estimate)), shape = 16) +
  # Color scale: red for larger p-values, blue for smaller (adjust as you see fit)
    scale_color_gradient(low = "red", high = "blue", limits = c(0.001, 0.05)) +
  # Adjust point size range as desired
  scale_size(range = c(2, 8)) +
  # Nice minimal theme
  theme_minimal(base_size = 14) +
  # Axis labels, legend labels, etc.
  labs(x = "Correlation",
       y = NULL,
       color = "p value",
       size = "|cor coef|")  + 
  theme(axis.text.y = ggtext::element_markdown(size = 14, color="black"),
        title = element_text(size = 8),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10),
        axis.title.x = element_text(size = 10),color = "black")

    plot(p)

```

# Plot one celltype - neighbour combination across several clincal outcomes {.tabset .tabset-pills}

```{r plot across clinical, eval=TRUE, fig.width=5, fig.height=4, results='asis'}
for(i in unique(df_comb2$celltype_with_nbh)){
  df_comb_filtered <- df_comb2 %>% filter(celltype_with_nbh == i)
df_comb_filtered$clinical_var <- factor(df_comb_filtered$clinical_var, levels = df_comb_filtered$clinical_var[order(df_comb_filtered$estimate)])

p <- ggplot(df_comb_filtered, aes(x = estimate, y = clinical_var)) +
  # Draw a segment from correlation=0 to the correlation value, just for a "lollipop" style
  geom_segment(aes(x = 0, xend = estimate, y = clinical_var, yend = clinical_var),
               color = "grey80") +
  # Plot the points; color by p-value and size by absolute correlation
  geom_point(aes(color = p.value, size = abs(estimate)), shape = 16) +
  # Color scale: red for larger p-values, blue for smaller (adjust as you see fit)
    scale_color_gradient(low = "red", high = "blue", limits = c(0.001, 0.05)) +
  # Adjust point size range as desired
  scale_size(range = c(2, 8)) +
  # Nice minimal theme
  theme_minimal(base_size = 14) +
  # Axis labels, legend labels, etc.
  labs(x = "Correlation",
       y = NULL,
       color = "p value",
       size = "|cor coef|",
       title = i)  + 
  theme(axis.text.y = element_text(size = 12, color="black"),
        title = element_text(size = 8),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10),
        axis.title.x = element_text(size = 10),color = "black")

cat("## ", i, "\n\n")
    plot(p)
    cat("\n\n")

}
```

# Plot one clinical outcome across several celltype - neighbour combinations {.tabset .tabset-pills}

```{r plot across celltypes, eval=TRUE, fig.width=5, fig.height=4, results='asis'}

list_df$mrss$label <- recode(
  list_df$mrss$label,
  "with_CD31_1_neigh" = "Met<sup>low</sup>_EC",
  "with_CD31_2_neigh" = "TCA/OXPHOS<sup>hi</sup><br>PGC-1α<sup>hi</sup>_EC",
  "with_CD31_3_neigh" = "Met<sup>hi</sup>_EC",
  "with_CD31_4_neigh" = "pmTOR<sup>hi</sup>CD98<sup>hi</sup>_EC",
  "with_CD31_5_neigh" = "Glycolysis<sup>hi</sup><br>TCA/OXPHOS<sup>int</sup>PPP<sup>hi</sup>_EC",
  "with_CD45_CD68_1_neigh" = "Met<sup>hi</sup>_Mf",
  "with_CD45_CD68_2_neigh" = "Met<sup>low</sup>_Mf",
  "with_CD45_CD68_3_neigh" = "Glycolysis<sup>low</sup><br>TCA/OXPHOS<sup>hi</sup>_Mf",
  "with_CD45_CD68_4_neigh" = "Glycolysis<sup>hi</sup><br>TCA/OXPHOS<sup>hi</sup>_Mf",
  "with_stromal_1_neigh" = "TCA/OXPHOS<sup>hi</sup><br>PGC-1α<sup>hi</sup>_Fib",
  "with_stromal_2_neigh" = "Met<sup>hi</sup>_Fib",
  "with_stromal_3_neigh" = "TCA/OXPHOS<sup>int</sup><br>PGC-1α<sup>hi</sup>_Fib",
  "with_stromal_4_neigh" = "pmTOR<sup>hi</sup>CD98<sup>hi</sup>_Fib",
  "with_stromal_5_neigh" = "Met<sup>low</sup>_Fib",
  "with_stromal_6_neigh" = "Glycolysis<sup>int</sup><br>CS<sup>hi</sup>OGDH<sup>low</sup>_Fib",
  "with_stromal_7_neigh" = "Glut<sup>hi</sup>_Fib",
  "with_stromal_8_neigh" = "Glycolysis<sup>int</sup><br>PGC-1α<sup>low</sup>FAO<sup>low</sup>_Fib"
)

# Define a function to generate a lollipop plot for a given set of filter parameters
plot_lollipop <- function(data, clinvar, clustervar, include_pattern, exclude_pattern, title_suffix = "") {
  
  # Combine the list of data frames into one and filter using the given parameters
  df <- bind_rows(data) %>%
    filter(clinical_var == clinvar,
           cluster_var == clustervar,
           !grepl(exclude_pattern, neigh_var),
           grepl(include_pattern, neigh_var))
  
  # Order the factor levels for the y-axis based on the estimate values
  df$label <- factor(df$label, 
                                 levels = df$label[order(df$estimate)])
  
  # Create the ggplot object
  p <- ggplot(df, aes(x = estimate, y = label)) +
    geom_segment(aes(x = 0, xend = estimate, y = label, yend = label),
                 color = "grey80") +
    geom_point(aes(color = p.value, size = abs(estimate)), shape = 16) +
    scale_color_gradient(low = "red", high = "blue", limits = c(0.001, 0.05)) +
    scale_size(range = c(2, 8)) +
    theme_minimal(base_size = 14) +
    labs(x = "Correlation",
         y = NULL,
         color = "p value",
         size = "|cor coef|",
         title = paste("mRss", title_suffix)) +
    theme(axis.text.y = ggtext::element_markdown(size = 10, color="black"),
          title = element_text(size = 8),
          legend.text = element_text(size = 10),
          legend.title = element_text(size = 10),
          axis.title.x = element_text(size = 10),color = "black")  
  
  return(p)
}

# Define the parameters for each plot in a list
plot_params <- list(
  list(clustervar = "stromal_2", include_pattern = "^has_str", 
       exclude_pattern = "^has_CD45_CD3|^has_Ecad|^has_CD45_n", 
       title_suffix = "- stromal_2, has_str"),
  list(clustervar = "stromal_2", include_pattern = "^has_CD31", 
       exclude_pattern = "^has_CD45_CD3|^has_Ecad|^has_CD45_n", 
       title_suffix = "- stromal_2, has_CD31"),
  list(clustervar = "stromal_4", include_pattern = "^has_str", 
       exclude_pattern = "^has_CD45_CD3|^has_Ecad|^has_CD45_n", 
       title_suffix = "- stromal_4, has_str"),
  list(clustervar = "stromal_8", include_pattern = "^has_str", 
       exclude_pattern = "^has_CD45_CD3|^has_Ecad|^has_CD45_n", 
       title_suffix = "- stromal_8, has_str"),
  list(clustervar = "stromal_8", include_pattern = "^has_CD45_CD68", 
       exclude_pattern = "^has_CD45_CD3|^has_Ecad|^has_CD45_n", 
       title_suffix = "- stromal_8, has_CD45_CD68")
)

# Generate the plots by applying the function to each set of parameters.
# (Assuming your data is in the list `list_df`.)
plots <- lapply(plot_params, function(params) {
  plot_lollipop(data = list_df, 
                clinvar = "current_mRSS", 
                clustervar = params$clustervar,
                include_pattern = params$include_pattern, 
                exclude_pattern = params$exclude_pattern,
                title_suffix = params$title_suffix)
})

counter <- 1
# Display the plots (one after the other)
for (p in plots) {
  cat("## Plot ",counter, "\n\n")
    plot(p)
    counter <- counter + 1
    cat("\n\n")
}
```

# Outputs

This analysis generates the following results: <br>

1.  Dot plots of correlations with clinical data. <br>

# Session information

```{r session, eval=TRUE}
etime <- proc.time() - ptm
cat('Rendering this document took :', etime[3], ' s \n')

sessioninfo::session_info()
```
