---
title: "Differences in a binary clinical outcome as dotplots"
author: "Alex Matei"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  html_document: 
    toc: true
    toc_float: true
    toc_depth: 4
    df_print: paged
    code_folding: hide
---

# Setup
```{r setup, include=FALSE}
source("data_preprocessing/common/config.R")

library(here)
source("data_preprocessing/common/original_setup.R")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, eval = TRUE, root.dir = here::here())

```

```{r timestamp, eval=TRUE}
ts <- format(Sys.time(), "%Y%m%d%H%M%S")
ptm <- proc.time()

cat("This document was created with the time stamp",ts)
```
Import the datasets created with the RMarkdown `Clinicaldata_tables.Rmd` (export firstly the Datatable outputs as .csv files and save them in `r here("Datasets")`).

# Load datasets

```{r load datasets, eval=TRUE}
df <- read.csv(here("Datasets/dif_prog_hasneigh.csv"), stringsAsFactors = FALSE)
```

# Prepare datasets for analysis

```{r prep datasets, eval=TRUE}

df$label2 <- recode(
  df$label,
  "with_CD31_1_neigh" = "Met<sup>low</sup>_EC",
  "with_CD31_2_neigh" = "TCA/OXPHOS<sup>hi</sup>PGC-1α<sup>hi</sup>_EC",
  "with_CD31_3_neigh" = "Met<sup>hi</sup>_EC",
  "with_CD31_4_neigh" = "pmTOR<sup>hi</sup>CD98<sup>hi</sup>_EC",
  "with_CD31_5_neigh" = "Glycolysis<sup>hi</sup>TCA/OXPHOS<sup>int</sup>PPP<sup>hi</sup>_EC",
  "with_CD45_CD68_1_neigh" = "Met<sup>hi</sup>_Mf",
  "with_CD45_CD68_2_neigh" = "Met<sup>low</sup>_Mf",
  "with_CD45_CD68_3_neigh" = "Glycolysis<sup>low</sup>TCA/OXPHOS<sup>hi</sup>_Mf",
  "with_CD45_CD68_4_neigh" = "Glycolysis<sup>hi</sup>TCA/OXPHOS<sup>hi</sup>_Mf",
  "with_stromal_1_neigh" = "TCA/OXPHOS<sup>hi</sup>PGC-1α<sup>hi</sup>_Fib",
  "with_stromal_2_neigh" = "Met<sup>hi</sup>_Fib",
  "with_stromal_3_neigh" = "TCA/OXPHOS<sup>int</sup>PGC-1α<sup>hi</sup>_Fib",
  "with_stromal_4_neigh" = "pmTOR<sup>hi</sup>CD98<sup>hi</sup>_Fib",
  "with_stromal_5_neigh" = "Met<sup>low</sup>_Fib",
  "with_stromal_6_neigh" = "Glycolysis<sup>int</sup>CS<sup>hi</sup>OGDH<sup>low</sup>_Fib",
  "with_stromal_7_neigh" = "Glut<sup>hi</sup>_Fib",
  "with_stromal_8_neigh" = "Glycolysis<sup>int</sup>PGC-1α<sup>low</sup>FAO<sup>low</sup>_Fib"
)


df$cluster_var2 <- recode(
  df$cluster_var,
  "CD31_1" = "Met<sup>low</sup>_EC",
  "CD31_2" = "TCA/OXPHOS<sup>hi</sup>PGC-1α<sup>hi</sup>_EC",
  "CD31_3" = "Met<sup>hi</sup>_EC",
  "CD31_4" = "pmTOR<sup>hi</sup>CD98<sup>hi</sup>_EC",
  "CD31_5" = "Glycolysis<sup>hi</sup>TCA/OXPHOS<sup>int</sup>PPP<sup>hi</sup>_EC",
  "CD45_CD68_1" = "Met<sup>hi</sup>_Mf",
  "CD45_CD68_2" = "Met<sup>low</sup>_Mf",
  "CD45_CD68_3" = "Glycolysis<sup>low</sup>TCA/OXPHOS<sup>hi</sup>_Mf",
  "CD45_CD68_4" = "Glycolysis<sup>hi</sup>TCA/OXPHOS<sup>hi</sup>_Mf",
  "stromal_1" = "TCA/OXPHOS<sup>hi</sup>PGC-1α<sup>hi</sup>_Fib",
  "stromal_2" = "Met<sup>hi</sup>_Fib",
  "stromal_3" = "TCA/OXPHOS<sup>int</sup>PGC-1α<sup>hi</sup>_Fib",
  "stromal_4" = "pmTOR<sup>hi</sup>CD98<sup>hi</sup>_Fib",
  "stromal_5" = "Met<sup>low</sup>_Fib",
  "stromal_6" = "Glycolysis<sup>int</sup>CS<sup>hi</sup>OGDH<sup>low</sup>_Fib",
  "stromal_7" = "Glut<sup>hi</sup>_Fib",
  "stromal_8" = "Glycolysis<sup>int</sup>PGC-1α<sup>low</sup>FAO<sup>low</sup>_Fib"
)

# Create a combined cell type + neighbor variable and remove unwanted patterns
df <- df %>% 
  mutate(celltype_with_nbh = paste(cluster_var2, label2, sep = " &rarr; ")) %>% 
  filter(!grepl("Ecad_neigh$|CD45_neigh$|CD45_CD3_[1-4]_neigh$", celltype_with_nbh))

```

# Plot 1: overall differences

```{r plot across clinical, eval=TRUE, fig.width=9, fig.height=5}

# Filter for significant differences
df_filtered <- df %>%
  mutate(difference = median_group2 - median_group1) %>%
  filter(p < 0.05 & abs(difference) > 1) 

# Order the factor levels for plotting
df_filtered$celltype_with_nbh <- factor(
  df_filtered$celltype_with_nbh,
  levels = df_filtered$celltype_with_nbh[order(df_filtered$difference)]
)

# Create the lollipop plot
p <- ggplot(df_filtered, aes(x = difference, y = celltype_with_nbh)) +
  # Draw segments from 0 to the difference for a “lollipop” effect
  geom_segment(aes(x = 0, xend = difference, 
                   y = celltype_with_nbh, yend = celltype_with_nbh), 
               color = "grey80") +
  # Plot points, color by p-value, size by absolute difference
  geom_point(aes(color = p, size = abs(difference)), shape = 16) +
  # Color and size scales
  scale_color_gradient(low = "red", high = "blue", limits = c(0, 0.05)) +
  scale_size(range = c(2, 8)) +
  theme_minimal(base_size = 14) +
  labs(
    x = "Difference",
    y = NULL,
    color = "P value",
    size = "|Difference|"
  ) +
 theme(axis.text.y = ggtext::element_markdown(size = 14, color="black"),
        title = element_text(size = 8),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10),
        axis.title.x = element_text(size = 10),color = "black")

print(p)

```

# Plot 2: stromal_2 with "has_str" Neighbors {.tabset .tabset-pills}

```{r plot across celltypes, eval=TRUE, fig.width=6, fig.height=5, results='asis'}

df$label2 <- recode(
  df$label,
  "with_CD31_1_neigh" = "Met<sup>low</sup>_EC",
  "with_CD31_2_neigh" = "TCA/OXPHOS<sup>hi</sup><br>PGC-1α<sup>hi</sup>_EC",
  "with_CD31_3_neigh" = "Met<sup>hi</sup>_EC",
  "with_CD31_4_neigh" = "pmTOR<sup>hi</sup>CD98<sup>hi</sup>_EC",
  "with_CD31_5_neigh" = "Glycolysis<sup>hi</sup><br>TCA/OXPHOS<sup>int</sup>PPP<sup>hi</sup>_EC",
  "with_CD45_CD68_1_neigh" = "Met<sup>hi</sup>_Mf",
  "with_CD45_CD68_2_neigh" = "Met<sup>low</sup>_Mf",
  "with_CD45_CD68_3_neigh" = "Glycolysis<sup>low</sup><br>TCA/OXPHOS<sup>hi</sup>_Mf",
  "with_CD45_CD68_4_neigh" = "Glycolysis<sup>hi</sup><br>TCA/OXPHOS<sup>hi</sup>_Mf",
  "with_stromal_1_neigh" = "TCA/OXPHOS<sup>hi</sup><br>PGC-1α<sup>hi</sup>_Fib",
  "with_stromal_2_neigh" = "Met<sup>hi</sup>_Fib",
  "with_stromal_3_neigh" = "TCA/OXPHOS<sup>int</sup><br>PGC-1α<sup>hi</sup>_Fib",
  "with_stromal_4_neigh" = "pmTOR<sup>hi</sup>CD98<sup>hi</sup>_Fib",
  "with_stromal_5_neigh" = "Met<sup>low</sup>_Fib",
  "with_stromal_6_neigh" = "Glycolysis<sup>int</sup><br>CS<sup>hi</sup>OGDH<sup>low</sup>_Fib",
  "with_stromal_7_neigh" = "Glut<sup>hi</sup>_Fib",
  "with_stromal_8_neigh" = "Glycolysis<sup>int</sup><br>PGC-1α<sup>low</sup>FAO<sup>low</sup>_Fib"
)

counter <- 1
# Define the patterns and a descriptive name for each plot
patterns <- list(
  "has_str"      = "^has_str",
  "has_CD31"     = "^has_CD31",
  "has_CD45_CD68" = "^has_CD45_CD68"
)

# Loop over each pattern to generate and print the corresponding plot
for (pattern_name in names(patterns)) {
  
  # Filter the dataset for cluster "stromal_2" and the current neighbor pattern
  df_filtered <- df %>%
    mutate(difference = median_group2 - median_group1) %>%
    filter(cluster_var == "stromal_2" & grepl(patterns[[pattern_name]], neigh_var))
  
  # Order the factor levels based on the difference for a neat y-axis ordering
  df_filtered$label2 <- factor(
    df_filtered$label2,
    levels = df_filtered$label2[order(df_filtered$difference)]
  )
  
  # Create the lollipop plot
  p <- ggplot(df_filtered, aes(x = difference, y = label2)) +
    geom_segment(aes(x = 0, xend = difference, 
                     y = label2, yend = label2), 
                 color = "grey80") +
    geom_point(aes(color = p, size = abs(difference)), shape = 16) +
    scale_color_gradient(low = "red", high = "blue", limits = c(0, 0.05)) +
    scale_size(range = c(2, 8)) +
    theme_minimal(base_size = 14) +
    labs(
      x = "Difference",
      y = NULL,
      color = "P value",
      size = "|Difference|",
      title = paste("Lollipop Plot for stromal_2 ", pattern_name, "neighbors")
    ) +
 theme(axis.text.y = ggtext::element_markdown(size = 10, color="black"),
        title = element_text(size = 8),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10),
        axis.title.x = element_text(size = 10),color = "black")
  
  # Print the plot
  cat("## Plot ",counter, "\n\n")
  print(p)
  counter <- counter + 1
  cat("\n\n")
}

```


# Outputs

This analysis generates the following results: <br>

1. Dot plots of differences in median frequency between the groups of a binary clinical outcome. <br>

# Session information

```{r session, eval=TRUE}
etime <- proc.time() - ptm
cat('Rendering this document took :', etime[3], ' s \n')

sessioninfo::session_info()
```