---
title: "Heatmap regression models clinical data"
author: "Alex Matei"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  html_document: 
    toc: true
    toc_float: true
    toc_depth: 4
    df_print: paged
    code_folding: hide
---

# Setup
```{r setup, include=FALSE}
source("data_preprocessing/common/config.R")

library(here)
source("data_preprocessing/common/original_setup.R")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, eval = FALSE, root.dir = here::here())

plan_future_fx(memory_fraction = 0.5, workers = 6)
options(future.seed = TRUE)
```

```{r timestamp, eval=TRUE}
ts <- format(Sys.time(), "%Y%m%d%H%M%S")
ptm <- proc.time()

cat("This document was created with the time stamp",ts)
```

# Load spe

```{r load spe, eval=TRUE}
spe <- readRDS(here("Datasets/spe20250127174306.rds"))
```

# Prepare spe for analysis

```{r prep spe, eval=TRUE}

# Create a new data frame with all the colData of the spe object - to recode the patient_id column

metadata <- colData(spe)

# add a new column to metadata called Donor where you recode the patient_id column
# the elements starting with NH should be called Healthy1, Healthy2, etc.
# the elements starting with SSc should be called SSc1, SSc2, etc.
metadata$Donor <- dplyr::recode(
  metadata$patient_id,
  Healthy1 = "Healthy1",
  Healthy2 = "Healthy2",
  SSc1 = "SSc1",
  SSc2 = "SSc2",
  SSc3 = "SSc3",
  SSc4 = "SSc4",
  SSc5 = "SSc5",
  SSc6 = "SSc6",
  SSc7 = "SSc7",
  SSc8 = "SSc8",
  Healthy3 = "Healthy3",
  Healthy4 = "Healthy4",
  SSc9 = "SSc9",
  Healthy5 = "Healthy5",
  Healthy6 = "Healthy6",
  Healthy7 = "Healthy7"
)


metadata$Disease <- dplyr::recode(metadata$indication, NH = "Healthy", SSc =
                                    "SSc")

# Assign the new columns in metadata to the spe object
spe$Donor <- metadata$Donor
spe$Disease <- metadata$Disease

# Verify the assignment
unique(spe$Donor)
unique(spe$Disease)

```

# Preparing data for heatmap

```{r preparing data, eval=TRUE}

clinical <- readxl::read_xlsx(here("Clinical data_AEM.xlsx")) %>% filter(`Subtype (dc/lc)` !=
                                                                           "healthy")
clinical$Donor <- dplyr::recode(
  clinical$`Label in metabolic panel`,
  Healthy1 = "Healthy1",
  Healthy2 = "Healthy2",
  SSc1 = "SSc1",
  SSc2 = "SSc2",
  SSc3 = "SSc3",
  SSc4 = "SSc4",
  SSc5 = "SSc5",
  SSc6 = "SSc6",
  SSc7 = "SSc7",
  SSc8 = "SSc8",
  Healthy3 = "Healthy3",
  Healthy4 = "Healthy4",
  SSc9 = "SSc9",
  Healthy5 = "Healthy5",
  Healthy6 = "Healthy6",
  Healthy7 = "Healthy7"
)


clinical$`Subtype (dc/lc)` <- dplyr::recode(clinical$`Subtype (dc/lc)`,
                                            dcSSc = 0,
                                            lcSSc = 1)

clinical$`Skin progression at biopsy` <- dplyr::recode(clinical$`Skin progression at biopsy`,
                                                       stable = 0,
                                                       progressive = 1)

clinical$`Early or not` <- dplyr::recode(clinical$`Early or not`, late =
                                           0, early = 1)

clinical$`current DU` <- dplyr::recode(clinical$`current DU`,
                                       "no DU" = 0,
                                       "current DU" = 1)

clinical$`DU history` <- dplyr::recode(clinical$`DU history`,
                                       "no DU history" = 0,
                                       "DU history" = 1)

clinical$`Myositis` <- dplyr::recode(clinical$`Myositis`,
                                     "no myositis" = 0,
                                     "myositis" = 1)

clinical$`Arthritis` <- dplyr::recode(clinical$`Arthritis`,
                                      "no arthritis" = 0,
                                      "arthritis" = 1)

clinical$`CRP(mg/l)(5.2mg/l)` <- dplyr::recode(
  clinical$`CRP(mg/l)(5.2mg/l)`,
  "6,1" = "6.1",
  "1,7" = "1.7",
  "<5.2" = "5.2"
)
clinical$`CRP(mg/l)(5.2mg/l)` <- as.numeric(clinical$`CRP(mg/l)(5.2mg/l)`)
clinical$CRP <- clinical$`CRP(mg/l)(5.2mg/l)`

binary <- c(
  "Subtype (dc/lc)",
  "Skin progression at biopsy",
  "Early or not",
  "current DU",
  "DU history",
  "Myositis",
  "Arthritis",
  "ILD"
)

for (i in binary) {
  clinical[i] <- as.logical(pull(clinical, i))
}
colSums(clinical[binary], na.rm = TRUE)
# exclude parameters with less than 3 samples
binary <- names(which(colSums(clinical[binary], na.rm = TRUE) >= 2))
binary <- c("Gender", binary)


# define cell types for which to calculate frequencies
unique(spe$new_celllabels)
celltypes_interest <- c("Stromal", "CD31+", "CD45+CD68+", "CD45+CD3+")

coldata_df_hasNeigh <- spe$hasNeigh_df
for (i in colnames(coldata_df_hasNeigh)) {
  spe[[i]] <- coldata_df_hasNeigh[[i]]
}

clinical_vars <- c(
  binary,
  "current mRSS",
  "Age at the time of biopsy (years)",
  "EUSTAR score (2017)",
  "non-RP disease duration",
  "CRP"
)

```

# Calculating linear regression models and preparing heatmaps

```{r calculate lm and prep heatmap, cache=FALSE}
# Iterate over the unique labels and apply the function
freq_clusters_list_lm <- lm_model_clinical_data_fx(
  spe = spe,
  var_analysis = "pheno_clusters_new2",
  disease_var = "Disease",
  disease_var_value = "SSc",
  donor_var = "Donor",
  clinical_data = clinical,
  clinical_vars = clinical_vars,
  subset_spe = TRUE,
  var_subset = "new_celllabels",
  value_var_subset = celltypes_interest
)

freq_neighbourhood_list_lm <- lm_model_clinical_data_fx (
  spe = spe,
  var_analysis = "CN_10_aggNbh_delaunay_80_new",
  disease_var = "Disease",
  disease_var_value = "SSc",
  donor_var = "Donor",
  clinical_data = clinical,
  clinical_vars = clinical_vars,
  subset_spe = FALSE,
  var_subset = NULL,
  value_var_subset = NULL
)

# calculate lm for cells that have CD68_1 as neighbours splitted per cell type (stromal, CD31 ...)


freq_has_neigh_list_lm <- setNames(future_lapply(names(coldata_df_hasNeigh), function(neigh_var) {
  has_x_neigh <- sub("^has_", "with_", neigh_var)
  
  spe[["celltype_has_neigh"]] <- as.character(spe[["pheno_clusters_new2"]])
  
  # For those fibroblasts that have the specific neighbour:
spe[["celltype_has_neigh"]][spe[[neigh_var]] == has_x_neigh] <- paste0(
  spe[["pheno_clusters_new2"]][spe[[neigh_var]] == has_x_neigh],
  "_",
  has_x_neigh
)

# For those that do not have this neighbour, optionally mark them differently:
spe[["celltype_has_neigh"]][spe[[neigh_var]] != has_x_neigh] <- paste0(
  spe[["pheno_clusters_new2"]][spe[[neigh_var]] != has_x_neigh],
  "_no_",
  has_x_neigh
)
  
  plot <- tryCatch({
    lm_model_clinical_data_fx(
      spe = spe,
      var_analysis = "celltype_has_neigh",
      disease_var = "Disease",
      disease_var_value = "SSc",
      donor_var = "Donor",
      clinical_data = clinical,
      clinical_vars = clinical_vars,
      subset_spe = TRUE,
      var_subset = "new_celllabels",
      value_var_subset = celltypes_interest
    )
  }, error = function(e) {
    message(paste("Skipping plot for:", neigh_var, "-", e$message))
    return(NULL)  # Return NULL to indicate no plot generated
  })
  return(plot)
}),
names(coldata_df_hasNeigh)
)

```

# Logistic regression models for binary outcomes - **SKIPPED**

```{r lm}
# ## logistic regression models for binary clinical outcome
### Less statistical power -> use linear models ###

# glm.res <- data.frame()
# for (i in binary) {
#   if (i != "gender") {
#     for (j in unique(spe.fib$cell_clusters2)) {
#       formula <- as.formula(paste(i, " ~ PercentOfStromal" ))
#       temp <- data[data$label == j, c("label", "PercentOfStromal", i), ]
#       
#       # Check if there are enough observations for the test
#       if (sum(temp[["label"]] == j) > 1) {
#         res <- tryCatch({
#           glm.summary <- summary(glm(formula, family = 'binomial', data = temp))
#           glm.summary$coefficients
#         }, error = function(e) {
#           message("Error in glm() for combination: ", i, " and cluster: ", j)
#           return(NULL)
#         })
#         
#         if (!is.null(res)) {
#           res_temp <- data.frame(
#             cluster = j,
#             clinical_par = i,
#             p_value = res[2,4],
#             glm.estimate = res[2,1]
#           )
#           glm.res <- rbind(glm.res, res_temp)
#         } else {
#           message("No successful results for glm() for combination: ", i, " and cluster: ", j)
#         }
#       }
#     }
#   }
# }
# glm.res[glm.res$p_value < 0.2,]

# linear regression models for all clinical outcome
# lm.res <- data.frame()
# for (i in c(binary, "current mRSS", "Age at the time of biopsy (years)", 
#             "EUSTAR score (2017)", "non-RP disease duration", "CRP")) {
#   if (i != "Gender") {
#     for (j in unique(spe.fib$pheno_clusters_new)) {
#       temp <- data[data$label == j, c("label", "PercentOfStromal", i), ]
#       # Check if there are enough observations for the test
#       if (sum(temp[["label"]] == j) > 1) {
#         res <- tryCatch({
#           formula <- as.formula(paste0("PercentOfStromal ~ `", i, "`"))
#           summary(lm(formula = formula, data = temp))
#         }, error = function(e) {
#           message("Error in lm() for combination: ", i, " and cluster: ", j)
#           return(NULL)
#         })
#         
#         if (!is.null(res)) {
#           res_temp <- data.frame(
#             cluster = j,
#             clinical_par = i,
#             p_value = res$coefficients[2,'Pr(>|t|)'],
#             lm.estimate = res$coefficients[2,'Estimate']
#           )
#           lm.res <- rbind(lm.res, res_temp)
#         } else {
#           message("No successful results for lm() for combination: ", i, " and cluster: ", j)
#         }
#       }
#     }
#   }
# }
# lm.res[lm.res$p_value < 0.2,]
# lm.res$log.p_value <- -log10(lm.res$p_value)
# mt <- lm.res %>% tidyr::pivot_wider(id_cols = c('cluster'), names_from = 'clinical_par', values_from = 'log.p_value') %>%
#   #filter(cluster %in% c('MyoFib', 'FAPhigh', 'Thy1+ADAM12highPU1high', 'S1PR+', 'ADAM12+Gli1+')) %>%
#   column_to_rownames('cluster') %>% 
#   #select(-DU_history) %>% 
#   as.matrix()

```

# Plot heatmaps freq clusters {.tabset .tabset-pills}

```{r plot heatmap and regression plots freq clusters, results='asis'}

for (i in names(freq_clusters_list_lm)) {
  cat("## ", i, "\n\n")
  print(freq_clusters_list_lm[[i]][["heatmap"]])
  cat("\n\n")
}

```

# Plot heatmaps freq neighbourhoods {.tabset .tabset-pills}

```{r plot heatmap and regression plots nbh, results='asis'}

cat("## Neighbourhoods\n\n")
print(freq_neighbourhood_list_lm[["heatmap"]])

```

# Plot heatmaps freq has neighbour {.tabset .tabset-pills}

```{r plot heatmap and regression plots has neighbour, results='asis'}

for (i in names(freq_has_neigh_list_lm)) {
  cat("## ", i, "{.tabset .tabset-pills}\n\n")
  for (j in names(freq_has_neigh_list_lm[[i]])) {
    cat("### ", j, "\n\n")
    print(freq_has_neigh_list_lm[[i]][[j]][["heatmap"]])
    cat("\n\n")
  }
  cat("\n\n")
}

```


# Correlation plots with ggscater

```{r corr plots, cache=FALSE, eval=TRUE}

coldata_df_hasNeigh <- coldata_df_hasNeigh %>% select(!starts_with("has_CD45_CD3"))

for (neigh_var in names(coldata_df_hasNeigh)) {
  has_x_neigh <- sub("^has_", "with_", neigh_var)
  spe[[paste0("pheno_clusters_new2_",neigh_var)]] <- NA
  
  # For those fibroblasts that have the specific neighbour:
  spe[[paste0("pheno_clusters_new2_",neigh_var)]][spe[[neigh_var]] == has_x_neigh] <- paste0(spe[["pheno_clusters_new2"]][spe[[neigh_var]] == has_x_neigh], "_", has_x_neigh)
  
  # For those that do not have this neighbour, optionally mark them differently:
  # spe[["celltype_has_neigh"]][spe[[neigh_var]] != has_x_neigh] <- paste0(spe[["lv3.anno_bank"]][spe[[neigh_var]] != has_x_neigh], "_no_", has_x_neigh)
}

names_coldata_df_celltype_hasNeigh <- grep("^pheno_clusters_new2_has", colnames(colData(spe)),value=TRUE)

names(clinical) <- sanitize_filename_fx(names(clinical))

names(clinical)

clinical_values_plot <- c(
  "current_mRSS"
)

clusters_var <- c("stromal_2",
                  "stromal_4",
                  "stromal_8")

plot_list_stromal <- setNames(future_lapply(clinical_values_plot, function(clinical_var_name) {
  # Assign names to the inner list based on var_analysis
  setNames(lapply(clusters_var, function(cluster_var) {
    plot_spearman_clinical_data_fx(
      spe = spe,
      subset_vars = list(
        "new_celllabels" = c("Stromal"),
        Disease = c("SSc")
      ),
      var_analysis = "pheno_clusters_new2",
      value_analysis =  cluster_var,
      disease_var = "Disease",
      donor_var = "Donor",
      clinical_data = clinical,
      clinical_var = clinical_var_name
    )
  }), clusters_var  # Inner list names
  )
}),
  clinical_values_plot  # Outer list names
)

```

```{r corr plots nbh, cache=FALSE, eval=FALSE}
plot_list_nbh <- setNames(lapply(clinical_values_plot, function(clinical_var_name) {
  # Assign names to the inner list based on var_analysis
  setNames(lapply(unique(spe$CN_10_aggNbh_delaunay_80_new), function(cluster_var) {
    plot_spearman_clinical_data_fx(
      spe = spe,
      subset_vars = list(
        Disease = c("SSc")
      ),
      var_analysis = "CN_10_aggNbh_delaunay_80_new",
      value_analysis =  cluster_var,
      disease_var = "Disease",
      donor_var = "Donor",
      clinical_data = clinical,
      clinical_var = clinical_var_name
    )
  }), unique(spe$CN_10_aggNbh_delaunay_80_new)  # Inner list names
  )
}),
  clinical_values_plot  # Outer list names
)
```

```{r corr plots hasnbh, cache=FALSE, eval=TRUE}

celltype_neigh_plot <- c("has_stromal_4_neigh",
                         "has_stromal_5_neigh",
                         "has_CD31_4_neigh",
                         "has_CD31_5_neigh",
                         "has_stromal_2_neigh",
                         "has_CD45_CD68_1_neigh")

clinical_values_plot <- c(
  "current_mRSS"
)

clusters_var <- c("stromal_2",
                  "stromal_4",
                  "stromal_8")

plot_list_stromal_hasneigh <- setNames(lapply(clinical_values_plot, function(clinical_var_name) {
  # Outer list: clinical_var_name
  setNames(lapply(clusters_var, function(cluster_var) {
    # Middle list: cluster_var
    setNames(lapply(celltype_neigh_plot, function(neigh_var) {
      # Inner list: neigh_var
      
      # Create the subset_vars list
      subset_vars <- list("new_celllabels" = c("Stromal"),
                          "Disease" = c("SSc"))
      
      match_index <- match(colnames(spe[, spe[["pheno_clusters_new2"]] == cluster_var]), colnames(spe))
      spe[["celltype_hasNeigh"]] <- rep(NA, ncol(spe))
      spe[["celltype_hasNeigh"]][match_index] <- spe[, spe[["pheno_clusters_new2"]] ==
                                                       cluster_var][[neigh_var]]
      
      plot <- tryCatch({
        plot_spearman_clinical_data_fx(
          spe = spe,
          subset_vars = subset_vars,
          var_analysis = "celltype_hasNeigh",
          # Remains constant
          value_analysis =  grep("^with",na.omit(unique(spe[["celltype_hasNeigh"]])),value=TRUE),
          # Remains as cluster_var
          disease_var = "Disease",
          donor_var = "Donor",
          clinical_data = clinical,
          clinical_var = clinical_var_name
        )
      }, error = function(e) {
        message(
          paste(
            "Skipping plot for:",
            clinical_var_name,
            cluster_var,
            neigh_var,
            "-",
            e$message
          )
        )
        return(NULL)  # Return NULL to indicate no plot generated
      })
      
      return(plot)
      
    }), celltype_neigh_plot  # Assign names to the inner list based on has_neigh
    )
  }), clusters_var  # Assign names to the middle list based on clusters_var
  )
}
),
clinical_values_plot  # Assign names to the outer list based on clinical_values_plot
)

```

# Plot correlation plots frequency stromal cells {.tabset .tabset-pills}

```{r plot corr plot freq str, results='asis', eval=TRUE, fig.width=4, fig.height=3}

for (i in names(plot_list_stromal)) {
  cat("## ", i, "{.tabset .tabset-pills}\n\n")
  for (j in names(plot_list_stromal[[i]])) {
    cat("### ", j, "\n\n")
    print(plot_list_stromal[[i]][[j]])
    cat("\n\n")
  }
}

```

# Plot correlation plots frequency CN {.tabset .tabset-pills}

```{r plot corr plot freq CN, results='asis', eval=FALSE}

for (i in names(plot_list_nbh)) {
  cat("## ", i, "{.tabset .tabset-pills}\n\n")
  for (j in names(plot_list_nbh[[i]])) {
    cat("### ", j, "\n\n")
    print(plot_list_nbh[[i]][[j]])
    cat("\n\n")
  }
}
```

# Plot correlation plots frequency stromal cells with neighbours {.tabset .tabset-pills}

```{r plot corr plot freq str has neigh, results='asis', eval=TRUE, fig.width=4, fig.height=3}

for (i in names(plot_list_stromal_hasneigh)) {
  cat("## ", i, "{.tabset .tabset-pills}\n\n")
  for (j in names(plot_list_stromal_hasneigh[[i]])) {
    cat("### ", j, "{.tabset .tabset-pills}\n\n")
    for (k in names(plot_list_stromal_hasneigh[[i]][[j]])) {
      cat("#### ", k, "\n\n")
      print(plot_list_stromal_hasneigh[[i]][[j]][[k]])
      cat("\n\n")
    }
  }
}

```
   
# Boxplots for categorical clinical data

```{r boxplots, cache=FALSE}

clinical_values_boxplot <- c(
  "Subtype_dclc",
  "Skin_progression_at_biopsy",
  "Early_or_not",
  "current_DU",
  "DU_history",
  "ILD",
  "TFR_0w0_1with",
  "SSc_antibody_profile"
)

# for (i in binary) {
#   clinical[i] <- as.numeric(pull(clinical, i))
# }

for (i in clinical_values_boxplot) {
  # clinical[i] <- factor(pull(clinical, i), levels =
  #                         unique(pull(clinical, i)))
    clinical[i] <- as.character(pull(clinical, i))
}

clusters_var <- grep("FB$", unique(spe$lv3.anno_bank), value = TRUE)

#options(future.seed=TRUE)

plot_list_stromal <- setNames(future_lapply(clinical_values_boxplot, function(clinical_var_name) {
  # Assign names to the inner list based on var_analysis
  setNames(lapply(clusters_var, function(cluster_var) {
    plot_categorical_clinical_data_fx(
      spe = spe,
      subset_vars = list(
        "new_celllabels" = c("Stromal"),
        "Disease" = c("SSc")
      ),
      var_analysis = "pheno_clusters_new2",
      value_analysis =  cluster_var,
      disease_var = "Disease",
      donor_var = "Donor",
      clinical_data = clinical,
      clinical_var = clinical_var_name
    )
  }), clusters_var  # Inner list names
  )
}),
  clinical_values_boxplot  # Outer list names
)
```

```{r boxplots nbh, cache=FALSE}

plot_list_nbh <- setNames(future_lapply(clinical_values_boxplot, function(clinical_var_name) {
  # Assign names to the inner list based on var_analysis
  setNames(lapply(unique(spe$CN_10_aggNbh_delaunay_80_new), function(cluster_var) {
    plot_categorical_clinical_data_fx(
      spe = spe,
      subset_vars = list(
        "Disease" = c("SSc")
      ),
      var_analysis = "CN_10_aggNbh_delaunay_80_new",
      value_analysis =  cluster_var,
      disease_var = "Disease",
      donor_var = "Donor",
      clinical_data = clinical,
      clinical_var = clinical_var_name
    )
  }), clusters_var  # Inner list names
  )
}),
  clinical_values_boxplot  # Outer list names
)

```

```{r boxplots str has neigh, cache=FALSE}

plot_list_stromal_hasneigh <- setNames(future_lapply(clinical_values_boxplot, function(clinical_var_name) {
  # Outer list: clinical_var_name
  setNames(lapply(clusters_var, function(cluster_var) {
    # Middle list: cluster_var
    setNames(lapply(names(coldata_df_hasNeigh), function(neigh_var) {
      # Inner list: neigh_var
      
      # Create the subset_vars list
      subset_vars <- list("new_celllabels" = c("Stromal"),
                          "Disease" = c("SSc"))
      
      match_index <- match(colnames(spe[, spe[["pheno_clusters_new2"]] == cluster_var]), colnames(spe))
      spe[["celltype_hasNeigh"]] <- rep(NA, ncol(spe))
      spe[["celltype_hasNeigh"]][match_index] <- spe[, spe[["pheno_clusters_new2"]] ==
                                                       cluster_var][[neigh_var]]
      
      plot <- tryCatch({
        plot_categorical_clinical_data_fx(
          spe = spe,
          subset_vars = subset_vars,
          var_analysis = "celltype_hasNeigh",
          # Remains constant
          value_analysis = na.omit(unique(spe[["celltype_hasNeigh"]]))[1],
          # Remains as cluster_var
          disease_var = "Disease",
          donor_var = "Donor",
          clinical_data = clinical,
          clinical_var = clinical_var_name
        )
      }, error = function(e) {
        message(
          paste(
            "Skipping plot for:",
            clinical_var_name,
            cluster_var,
            neigh_var,
            "-",
            e$message
          )
        )
        return(NULL)  # Return NULL to indicate no plot generated
      })
      
      return(plot)
      
    }), names(coldata_df_hasNeigh)  # Assign names to the inner list based on has_neigh
    )
  }), clusters_var  # Assign names to the middle list based on clusters_var
  )
},
future.globals = list(
  spe = spe,
  clinical = clinical,
  clusters_var = clusters_var,
  coldata_df_hasNeigh = coldata_df_hasNeigh,
  get_common_boxplot_components_fx = get_common_boxplot_components_fx,
  plot_categorical_clinical_data_fx = plot_categorical_clinical_data_fx
)
),
clinical_values_boxplot  # Assign names to the outer list based on clinical_values_plot
)

```

```{r boxplots str has neigh healthy ssc, cache=FALSE}

clinical <- readxl::read_xlsx(here("Clinical data_AEM.xlsx")) %>% mutate(Disease=ifelse(`Subtype (dc/lc)`=="healthy","Healthy","SSc"))

clinical$Donor <- dplyr::recode(
  clinical$`Label in metabolic panel`,
  Healthy1 = "Healthy1",
  Healthy2 = "Healthy2",
  SSc1 = "SSc1",
  SSc2 = "SSc2",
  SSc3 = "SSc3",
  SSc4 = "SSc4",
  SSc5 = "SSc5",
  SSc6 = "SSc6",
  SSc7 = "SSc7",
  SSc8 = "SSc8",
  Healthy3 = "Healthy3",
  Healthy4 = "Healthy4",
  SSc9 = "SSc9",
  Healthy5 = "Healthy5",
  Healthy6 = "Healthy6",
  Healthy7 = "Healthy7"
)

plot_list_stromal_hasneigh_h_ssc <- setNames(future_lapply("Disease", function(clinical_var_name) {
  # Outer list: clinical_var_name
  setNames(lapply(clusters_var, function(cluster_var) {
    # Middle list: cluster_var
    setNames(lapply(names(coldata_df_hasNeigh), function(neigh_var) {
      # Inner list: neigh_var
      
      # Create the subset_vars list
      subset_vars <- list("new_celllabels" = c("Stromal"))
      
      match_index <- match(colnames(spe[, spe[["pheno_clusters_new2"]] == cluster_var]), colnames(spe))
      spe[["celltype_hasNeigh"]] <- rep(NA, ncol(spe))
      spe[["celltype_hasNeigh"]][match_index] <- spe[, spe[["pheno_clusters_new2"]] ==
                                                       cluster_var][[neigh_var]]
      
      plot <- tryCatch({
        plot_categorical_clinical_data_fx(
          spe = spe,
          subset_vars = subset_vars,
          var_analysis = "celltype_hasNeigh",
          # Remains constant
          value_analysis = na.omit(unique(spe[["celltype_hasNeigh"]]))[1],
          # Remains as cluster_var
          disease_var = "Disease",
          donor_var = "Donor",
          clinical_data = clinical,
          clinical_var = clinical_var_name
        )
      }, error = function(e) {
        message(
          paste(
            "Skipping plot for:",
            clinical_var_name,
            cluster_var,
            neigh_var,
            "-",
            e$message
          )
        )
        return(NULL)  # Return NULL to indicate no plot generated
      })
      
      return(plot)
      
    }), names(coldata_df_hasNeigh)  # Assign names to the inner list based on has_neigh
    )
  }), clusters_var  # Assign names to the middle list based on clusters_var
  )
},
future.globals = list(
  spe = spe,
  clinical = clinical,
  clusters_var = clusters_var,
  coldata_df_hasNeigh = coldata_df_hasNeigh,
  get_common_boxplot_components_fx = get_common_boxplot_components_fx,
  plot_categorical_clinical_data_fx = plot_categorical_clinical_data_fx
)
),
clinical_values_boxplot  # Assign names to the outer list based on clinical_values_plot
)

```

# Plot boxplots frequency stromal cells {.tabset .tabset-pills}

```{r plot boxplot freq str, results='asis'}

for (i in names(plot_list_stromal)) {
  cat("## ", i, "{.tabset .tabset-pills}\n\n")
  for (j in names(plot_list_stromal[[i]])) {
    cat("### ", j, "\n\n")
    print(plot_list_stromal[[i]][[j]])
    cat("\n\n")
  }
}

```


```{r plot boxplot freq nbh, results='asis'}

for (i in names(plot_list_nbh)) {
  cat("## ", i, "{.tabset .tabset-pills}\n\n")
  for (j in names(plot_list_nbh[[i]])) {
    cat("### ", j, "\n\n")
    print(plot_list_nbh[[i]][[j]])
    cat("\n\n")
  }
}

```

# Plot boxplots frequency stromal cells with neighbours {.tabset .tabset-pills}

```{r plot boxplot freq str has neigh, results='asis'}

for (i in names(plot_list_stromal_hasneigh)) {
  cat("## ", i, "{.tabset .tabset-pills}\n\n")
  for (j in names(plot_list_stromal_hasneigh[[i]])) {
    cat("### ", j, "{.tabset .tabset-pills}\n\n")
    for (k in names(plot_list_stromal_hasneigh[[i]][[j]])) {
      cat("#### ", k, "\n\n")
      print(plot_list_stromal_hasneigh[[i]][[j]][[k]])
      cat("\n\n")
    }
  }
}

```

# Plot boxplots frequency stromal cells with neighbours healthy SSc {.tabset .tabset-pills}

```{r plot boxplot freq str has neigh h ssc, results='asis'}

for (i in names(plot_list_stromal_hasneigh_h_ssc)) {
  cat("## ", i, "{.tabset .tabset-pills}\n\n")
  for (j in names(plot_list_stromal_hasneigh_h_ssc[[i]])) {
    cat("### ", j, "{.tabset .tabset-pills}\n\n")
    for (k in names(plot_list_stromal_hasneigh_h_ssc[[i]][[j]])) {
      cat("#### ", k, "\n\n")
      print(plot_list_stromal_hasneigh_h_ssc[[i]][[j]][[k]])
      cat("\n\n")
    }
  }
}

```

# Outputs

This analysis generates the following results: <br>

1. Heatmaps, scatter plots and box plots of the relationship of cell/neighbourhood frequencies with clinical data. <br>

# Session information

```{r session, eval=TRUE}
etime <- proc.time() - ptm
cat('Rendering this document took :', etime[3], ' s \n')

sessioninfo::session_info()
```