---
title: "Bulk RNA-seq differential expression: 2DG + OX and LDHA + OX versus control"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: false
    theme: cosmo
    df_print: paged
params:
  out_dir: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
out_dir <- params$out_dir
summary <- readRDS(file.path(out_dir, 'objects', 'analysis_summary.rds'))
fmt_int <- function(x) format(x, big.mark = ',', scientific = FALSE, trim = TRUE)
fmt_num <- function(x, digits = 3) format(round(x, digits), nsmall = digits, trim = TRUE)
fig <- function(name) knitr::include_graphics(file.path(out_dir, 'figures', name))
deg_summary <- summary$deg_summary
top_genes <- summary$top_genes_summary
top_gsea <- summary$top_gsea_summary
filtering <- summary$filtering_summary
sample_meta <- summary$sample_meta
deseq_summary <- summary$deseq_summary
```

## Technical Summary

This analysis compares two treatment conditions against paired controls: `2DG` is interpreted as `2DG + OX`, and `LDHA` is interpreted as `LDHA + OX`. The input contained `r fmt_int(filtering$value[filtering$metric == 'input_genes'])` gene rows across `r fmt_int(filtering$value[filtering$metric == 'sequencing_lanes'])` sequencing-lane columns. Technical lanes belonging to the same PCS-condition sample were summed, yielding `r fmt_int(nrow(sample_meta))` biological profiles: five PCS blocks measured under control, 2DG + OX, and LDHA + OX.

The primary differential-expression model uses TMM normalization and `limma-voom` with sample quality weights in a paired/block design: `~ PCS + condition`. This is the most defensible primary model for this matrix because counts are fractional rather than strict raw integers. High-confidence DEG calls are defined as FDR < 0.05 and absolute log2 fold-change >= 1; genes passing FDR alone are retained in the all-gene tables.

```{r summary-table}
show_summary <- deg_summary[, c('contrast_label', 'genes_tested', 'fdr_0_05', 'up_fdr_0_05_abslog2fc_1', 'down_fdr_0_05_abslog2fc_1', 'deg_fdr_0_05_abslog2fc_1')]
colnames(show_summary) <- c('Contrast', 'Genes tested', 'FDR < 0.05', 'Up DEG', 'Down DEG', 'Total DEG')
knitr::kable(show_summary, digits = 3)
```

## The paired samples support treatment-level comparisons after lane consolidation

The sample set is balanced at the biological level after technical lane consolidation: each PCS block has one control, one 2DG + OX, and one LDHA + OX profile. The library-size plot is therefore a QC check for depth and technical-lane consolidation rather than evidence of biological replication.

```{r qc-library, out.width='100%'}
fig('qc_library_sizes.png')
```

The PCA uses the top 5,000 most variable retained genes after TMM/voom transformation. The plot should be read as a sample-level QC view: treatment separation is interpretable only after accounting for the paired PCS structure in the model.

```{r qc-pca, out.width='100%'}
fig('qc_pca_voom_top_variable_genes.png')
```

The correlation heatmap checks whether any sample behaves as a strong outlier after normalization. Correlation structure is expected to reflect both paired PCS identity and treatment response.

```{r qc-cor, out.width='100%'}
fig('qc_sample_correlation_heatmap.png')
```

## Both treatment contrasts were tested with the same PCS-blocked model

The DEG summary below separates high-confidence up/down genes from genes that meet FDR but have smaller absolute effect sizes. This avoids mixing statistical detectability with a publication-scale effect threshold.

```{r deg-summary-plot, out.width='100%'}
fig('deg_summary_bar.png')
```

### 2DG + OX versus control

The volcano and MA plots show the same contrast from complementary angles: the volcano emphasizes significance and effect size, while the MA plot checks whether fold-change estimates depend strongly on expression level.

```{r volcano-2dg, out.width='100%'}
fig('deg_volcano_2DG_vs_C.png')
```

```{r ma-2dg, out.width='100%'}
fig('deg_ma_2DG_vs_C.png')
```

### LDHA + OX versus control

The LDHA + OX contrast was modeled with the identical normalization, filter, and paired design, so effect sizes and FDR values are directly comparable to the 2DG + OX contrast.

```{r volcano-ldha, out.width='100%'}
fig('deg_volcano_LDHA_vs_C.png')
```

```{r ma-ldha, out.width='100%'}
fig('deg_ma_LDHA_vs_C.png')
```

## Top differential genes provide a manuscript-ready shortlist

The heatmap shows the strongest differential genes across both contrasts, using row-scaled voom logCPM values. It is intended as a compact figure-panel candidate; exact statistics should be taken from the exported all-gene tables.

```{r heatmap, out.width='100%'}
fig('deg_top_genes_heatmap.png')
```

```{r top-gene-table}
tg <- top_genes[, c('contrast_label', 'gene_label', 'GENENAME', 'logFC', 'AveExpr', 'adj.P.Val', 'direction')]
colnames(tg) <- c('Contrast', 'Gene', 'Gene name', 'log2FC', 'Ave logCPM', 'FDR', 'Class')
knitr::kable(tg, digits = 4)
```

## GO Biological Process analysis summarizes pathway-level signal

GO Biological Process GSEA was run on the full ranked gene list for each contrast, using moderated t-statistics as the ranking metric. Positive NES values indicate enrichment toward the treatment-up end of the ranked list; negative NES values indicate enrichment toward control.

```{r gsea-2dg, out.width='100%'}
fig('pathway_go_bp_gsea_2DG_vs_C.png')
```

```{r gsea-ldha, out.width='100%'}
fig('pathway_go_bp_gsea_LDHA_vs_C.png')
```

```{r gsea-table}
if (nrow(top_gsea) > 0) {
  gsea_show <- top_gsea[, c('contrast_label', 'Description', 'setSize', 'NES', 'p.adjust', 'qvalue')]
  colnames(gsea_show) <- c('Contrast', 'GO Biological Process', 'Set size', 'NES', 'FDR', 'q value')
  knitr::kable(gsea_show, digits = 4)
} else {
  cat('No GO GSEA results were returned.')
}
```

## Scope, data, and definitions

- Biological comparison: 2DG + OX vs control and LDHA + OX vs control.
- Biological blocking factor: PCS sample identity, modeled as a fixed paired effect.
- Technical lanes: sequencing-lane columns with the same PCS and condition were summed before modeling.
- Primary expression unit: TMM-normalized voom logCPM for QC, heatmaps, and linear modeling.
- DEG threshold: FDR < 0.05 and |log2FC| >= 1 for the publication shortlist; all tested genes are exported without effect-size filtering.
- Gene identifiers: Ensembl IDs were version-stripped for annotation through `org.Hs.eg.db`; unmapped genes remain in the analysis with their Ensembl IDs.

```{r sample-table}
sm <- sample_meta[, c('sample_id', 'pcs', 'condition_label', 'n_technical_lanes', 'library_size')]
colnames(sm) <- c('Sample', 'PCS block', 'Condition', 'Technical lanes', 'Summed library size')
knitr::kable(sm)
```

## Methodology

1. Imported the Galaxy gene-level matrix and parsed sample names into PCS block, condition, and technical-lane metadata.
2. Summed technical lanes belonging to the same PCS-condition biological sample.
3. Removed genes with zero total signal, then applied `edgeR::filterByExpr` using the paired design matrix.
4. Normalized retained genes with TMM and fitted `limma::voomWithQualityWeights` using `~ PCS + condition`.
5. Tested `condition2DG` and `conditionLDHA` coefficients against control with robust empirical Bayes moderation.
6. Annotated genes with `org.Hs.eg.db`, exported all-gene and high-confidence DEG tables, and generated publication-style QC and DEG figures.
7. Ran a DESeq2 sensitivity analysis on rounded lane-summed counts. This is secondary because the source matrix contains fractional counts.
8. Ran GO Biological Process GSEA with `clusterProfiler::gseGO` on moderated t-statistics, plus GO over-representation tables for up/down DEG sets when enough genes were available.

```{r deseq-sensitivity}
if (nrow(deseq_summary) > 0) {
  ds <- deseq_summary[, c('contrast_label', 'deseq2_fdr_0_05', 'deseq2_deg_fdr_0_05_abslog2fc_1', 'logfc_spearman_vs_limma', 'logfc_pearson_vs_limma')]
  colnames(ds) <- c('Contrast', 'DESeq2 FDR < 0.05', 'DESeq2 DEG', 'Spearman log2FC vs limma', 'Pearson log2FC vs limma')
  knitr::kable(ds, digits = 4)
} else {
  cat(summary$source_notes$sensitivity_model)
}
```

## Limitations, uncertainty, and robustness checks

- The supplied matrix contains fractional count values. The primary limma-voom model is appropriate for non-integer estimated counts, while DESeq2 is provided only as a rounded-count sensitivity analysis.
- No raw sequencing QC, alignment metrics, strandedness checks, or gene-length offsets were provided. The report therefore validates the count matrix and modeled contrasts, not upstream read processing.
- PCS is modeled as a fixed paired effect. With five biological blocks per condition, large effects are more reliable than borderline FDR-only calls.
- GO terms can be redundant and broad. The exported GSEA and ORA tables should be used to choose a concise pathway narrative for the manuscript.

## Recommended next steps

1. Review the DEG shortlists against expected biology and remove genes that are unannotated or low-confidence for the planned figure panel.
2. If FASTQ/BAM QC is available, add upstream metrics such as mapping rate, duplication, strandedness, and gene-body coverage before submission.
3. For manuscript pathway claims, collapse redundant GO terms into a small set of biological themes and validate those against the leading-edge genes.
4. Confirm whether `2DG + OX` and `LDHA + OX` should be interpreted as paired perturbations in the same PCS system for all downstream text.

## Further questions

- Are there additional covariates such as sequencing batch, library prep batch, or RNA integrity that should be included in the design?
- Should the final manuscript emphasize high-confidence |log2FC| >= 1 genes, or all FDR-significant genes regardless of effect size?
- Are there curated gene sets specific to the biological system that should supplement the GO analysis?
